---
title: "Short-Read Vs Long-Read: Gene Expression Correlation"
author: "Leshem, Yul"
date: "2026-02-12"
output:
  html_document:
    toc: true
    toc_float: true
    df_print: paged
    code_folding: hide
    number_sections: true
---

Code location: `/udd/reyle/nmd_lungcells_2026/code/NMD_correlation_analysis.Rmd`

# Overview

This report measures concordance between short-read (Illumina/Salmon) and
long-read (PacBio Kinnex/SQANTI3) RNA-seq at the gene level, restricted to
the four cell types kept in the paper (AT2, LAE, FB, MV). Long-read isoforms
are aggregated to genes, both pipelines are filtered to a common
gene-and-sample set, and TMM-normalized log2 CPM values are compared via
sample-wise Pearson and Spearman correlation. A gene-length-vs-expression
diagnostic is run on both pipelines using two length measures (Salmon
effective length and expression-weighted exonic length).

Note on naming: sample names and the internal cell-type codes still use the
original `AT` (AT2) and `DD` (LAE) labels, because those codes are what the
Salmon column names and the project DGEList are keyed on. We keep them for all
sample matching and relabel to the display names (AT2, LAE) only at the
plotting / summary boundary.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 10, fig.height = 6)
options(bitmapType = "cairo")

```

# Libraries

```{r load-libs}
suppressPackageStartupMessages({
  library(data.table)
  library(edgeR)
  library(SummarizedExperiment)
  library(dplyr)
  library(tidyr)
  library(tibble)
  library(ggplot2)
  library(patchwork)
})

# Bioconductor namespace masking guard
select <- dplyr::select
filter <- dplyr::filter
rename <- dplyr::rename
count  <- dplyr::count
slice  <- dplyr::slice
first  <- dplyr::first
```

# Configuration

```{r config}
BASE       <- "/udd/reyle/nmd_lungcells_2026"
OUTPUT_DIR <- file.path(BASE, "final_figures/comparison")
dir.create(OUTPUT_DIR, recursive = TRUE, showWarnings = FALSE)

# Inputs
COUNTS_PATH    <- "/proj/regeps/regep00/studies/ExternalCellLines/data/rnaseq/Randell_Lung_Cells_2025/results/nfcore-rnaseq/20250811_plus_20260129/star_salmon/salmon.merged.gene_counts_length_scaled.rds"
SALMON_DIR     <- "/proj/regeps/regep00/studies/ExternalCellLines/data/rnaseq/Randell_Lung_Cells_2025/results/nfcore-rnaseq/20250811_plus_20260129/star_salmon"
DGELIST_GENE   <- file.path(BASE, "dge_gene_unfiltered_2026.1.2.rds")
DGELIST_ISO    <- file.path(BASE, "long_read/dge_isoform_longread_2026.3.3.rds")
ISO_COUNT_MAT  <- "/proj/regeps/regep00/studies/ExternalCellLines/analyses/repjc/Randell_Lung_Cells_2025/code/sqanti3_by_chromosome/projects/nmd_lungcells/results/nmd_lungcells_filtered.count_matrix.txt"

# Four cell types — DD_ALI and DO_ALI are cut from the paper.
# These are the internal codes used for sample matching (AT = AT2, DD = LAE).
CT_KEEP <- c("AT", "DD", "FB", "MV")

# Display relabel applied at the plotting / summary boundary only.
CT_DISPLAY <- c("AT" = "AT2", "DD" = "LAE", "FB" = "FB", "MV" = "MV")

# Helper: extract cell type from a "ID_TREATMENT_CT[_ALI]" sample name.
# The CT can be one or two underscore-delimited tokens (e.g., "AT" vs "DD_ALI"),
# so we glue everything from position 3 onward.
extract_ct <- function(sample_names) {
  sapply(strsplit(sample_names, "_"), function(x) {
    paste(x[3:length(x)], collapse = "_")
  })
}

```

# Short-Read

Load the gene-level Salmon counts, harmonize sample names, strip Ensembl
version suffixes, and restrict to the four kept cell types.


```{r short-read}
dge_gene  <- readRDS(DGELIST_GENE)
se_counts <- readRDS(COUNTS_PATH)

counts_matrix <- assay(se_counts, "counts")
cat("Counts matrix:", dim(counts_matrix), "\n")

# ---- Parse Salmon-style "AT2003C_DMSO_..." into "003C_DMSO_AT" ----
parse_colname <- function(colname) {
  parts      <- strsplit(colname, "_")[[1]]
  first_part <- parts[1]

  if (startsWith(first_part, "AT2")) {
    ct        <- "AT"
    sample_id <- substring(first_part, 4)
  } else {
    ct        <- substring(first_part, 1, 2)
    sample_id <- substring(first_part, 3)
  }
  treatment <- parts[2]
  paste(sample_id, treatment, ct, sep = "_")
}

colnames(counts_matrix) <- sapply(colnames(counts_matrix), parse_colname)

# ---- Identify DD samples that are canonically DD_ALI ----
# The Salmon-derived names label some samples as "DD" that are actually
# DD_ALI in the project DGEList (donors grown under air-liquid interface).
# Use the DGEList's canonical sample IDs to detect them, then drop them.
modified_ct <- gsub("DO_ALI", "DO", dge_gene$samples$ct)
modified_ct <- gsub("AT2",    "AT", modified_ct)
canonical_ids <- paste(dge_gene$samples$sample_id,
                       dge_gene$samples$treatment,
                       modified_ct, sep = "_")

dd_ali_in_sr <- sapply(colnames(counts_matrix), function(nm) {
  if (!grepl("_DD$", nm)) return(FALSE)
  potential_ali <- gsub("_DD$", "_DD_ALI", nm)
  potential_ali %in% canonical_ids
})
n_dd_ali <- sum(dd_ali_in_sr)
if (n_dd_ali > 0) {
  cat("FILTER: dropping", n_dd_ali, "SR samples labeled 'DD' that are canonically DD_ALI (cut from paper):\n")
  print(colnames(counts_matrix)[dd_ali_in_sr])
  counts_matrix <- counts_matrix[, !dd_ali_in_sr]
}

# ---- Restrict to four kept cell types (drops DO_ALI; DD_ALI already gone) ----
ct_per_sample <- extract_ct(colnames(counts_matrix))
keep_idx      <- ct_per_sample %in% CT_KEEP
n_dropped     <- sum(!keep_idx)
if (n_dropped > 0) {
  cat("FILTER: dropping", n_dropped, "samples outside CT_KEEP:",
      paste(unique(ct_per_sample[!keep_idx]), collapse = ", "), "\n")
  counts_matrix <- counts_matrix[, keep_idx]
}

# ---- Strip Ensembl version suffixes ----
gene_ids_clean <- sub("\\..*", "", rownames(counts_matrix))
n_dup_total    <- sum(duplicated(gene_ids_clean))
rownames(counts_matrix) <- gene_ids_clean

cat("Final SR matrix:", dim(counts_matrix), "\n")
cat("Duplicate gene IDs after version strip:", n_dup_total, "\n")
cat("Cell types (internal codes):",
    paste(sort(unique(extract_ct(colnames(counts_matrix)))), collapse = ", "), "\n")

counts_combined <- counts_matrix
```

# Long-Read

Load the SQANTI3 isoform count matrix, reformat sample names to match
short-read, restrict to CT_KEEP, then aggregate isoforms to gene level by
summing.

```{r long-read}
counts_iso <- fread(ISO_COUNT_MAT, data.table = FALSE)
cat("Isoform count matrix:", dim(counts_iso), "\n")

# Transcript-to-gene mapping from the long-read DGEList
dge_iso <- readRDS(DGELIST_ISO)
cat("Long-read DGEList:", dim(dge_iso), "\n")

tx2gene <- unique(dge_iso$genes[!is.na(dge_iso$genes$txid), c("txid", "gene_id")])
cat("Transcript-to-gene mappings:", nrow(tx2gene), "\n")

# ---- Format isoform counts ----
rownames(counts_iso) <- counts_iso$id
counts_iso$id        <- NULL
count_mat            <- round(as.matrix(counts_iso))

# ---- Reformat sample names: "Sample##_CT_ID_Treatment" -> "ID_Treatment_CT" ----
nms     <- gsub("^Sample\\d+_", "", colnames(count_mat))
new_nms <- sapply(nms, function(x) {
  parts     <- strsplit(x, "_")[[1]]
  treatment <- parts[length(parts)]
  sample_id <- parts[length(parts) - 1]
  ct        <- paste(parts[1:(length(parts) - 2)], collapse = "_")
  paste(sample_id, treatment, ct, sep = "_")
})
colnames(count_mat) <- new_nms

# ---- Restrict to four kept cell types (drops DD_ALI and DO_ALI) ----
ct_per_sample_lr <- extract_ct(colnames(count_mat))
keep_idx_lr      <- ct_per_sample_lr %in% CT_KEEP
n_dropped_lr     <- sum(!keep_idx_lr)
if (n_dropped_lr > 0) {
  cat("FILTER: dropping", n_dropped_lr, "long-read samples outside CT_KEEP:",
      paste(unique(ct_per_sample_lr[!keep_idx_lr]), collapse = ", "), "\n")
  count_mat <- count_mat[, keep_idx_lr]
}

# ---- Aggregate isoforms -> genes ----
idx               <- match(rownames(count_mat), tx2gene$txid)
gene_ids_for_iso  <- tx2gene$gene_id[idx]
has_gene          <- !is.na(gene_ids_for_iso)
counts_with_gene  <- count_mat[has_gene, ]
gene_ids_for_iso  <- sub("\\..*", "", gene_ids_for_iso[has_gene])

cat("Isoforms matched to gene:", sum(has_gene), "of", nrow(count_mat), "\n")
cat("Isoforms dropped (no gene ID):", sum(!has_gene), "\n")

counts_genelevel <- rowsum(counts_with_gene, group = gene_ids_for_iso)
cat("After aggregation:", nrow(counts_with_gene), "isoforms ->",
    nrow(counts_genelevel), "genes\n")

# ---- Sample overlap report ----
short_names <- colnames(counts_combined)
long_names  <- colnames(counts_genelevel)

cat("\nSample overlap:\n")
cat("  Common:           ", length(intersect(short_names, long_names)), "\n")
cat("  Short-read only:  ", length(setdiff(short_names, long_names)),   "\n")
cat("  Long-read only:   ", length(setdiff(long_names,  short_names)),  "\n")

if (length(setdiff(short_names, long_names)) > 0) {
  cat("\nShort-read only:\n"); print(setdiff(short_names, long_names))
}
if (length(setdiff(long_names, short_names)) > 0) {
  cat("\nLong-read only:\n");  print(setdiff(long_names,  short_names))
}

```

# Combine Short And Long Read

Subset both matrices to the common gene set and common sample set, drop
genes with all-zero counts in both pipelines, TMM-normalize each pipeline
separately, and compute log2 CPM.

```{r combine-short-long}
common_genes   <- intersect(rownames(counts_combined), rownames(counts_genelevel))
common_samples <- intersect(colnames(counts_combined), colnames(counts_genelevel))

cat("Common genes:  ", length(common_genes),   "\n")
cat("Common samples:", length(common_samples), "\n")

stopifnot("No common genes — check gene IDs"     = length(common_genes)   > 0)
stopifnot("No common samples — check name format" = length(common_samples) > 0)

counts_short_matched <- counts_combined[common_genes, common_samples]
counts_long_matched  <- counts_genelevel[common_genes, common_samples]

# ---- Drop genes with all zeros in BOTH pipelines (uninformative) ----
expressed_short <- rowSums(counts_short_matched) > 0
expressed_long  <- rowSums(counts_long_matched)  > 0
keep_genes      <- expressed_short | expressed_long

cat("\nFILTER: drop genes all-zero in BOTH pipelines.\n")
cat("  All-zero in SR only:    ", sum(!expressed_short),  "\n")
cat("  All-zero in LR only:    ", sum(!expressed_long),    "\n")
cat("  All-zero in BOTH (drop):", sum(!keep_genes),        "\n")
cat("  Genes before:", length(keep_genes), "/ after:", sum(keep_genes), "\n")

counts_short_matched   <- counts_short_matched[keep_genes, ]
counts_long_matched    <- counts_long_matched[keep_genes, ]
common_genes_filtered  <- rownames(counts_short_matched)

# ---- TMM-normalize and compute log2 CPM ----
dge_short_matched <- calcNormFactors(DGEList(counts = counts_short_matched), method = "TMM")
dge_long_matched  <- calcNormFactors(DGEList(counts = counts_long_matched),  method = "TMM")

cpm_short_matched <- cpm(dge_short_matched, normalized.lib.sizes = TRUE, log = TRUE)
cpm_long_matched  <- cpm(dge_long_matched,  normalized.lib.sizes = TRUE, log = TRUE)

stopifnot(identical(rownames(cpm_short_matched), rownames(cpm_long_matched)))
stopifnot(identical(colnames(cpm_short_matched), colnames(cpm_long_matched)))

cat("\nFinal CPM matrices:", dim(cpm_short_matched), "\n")
```

## Sample-Wise Correlation

```{r corr-overall}
sample_cors_pearson  <- sapply(seq_len(ncol(cpm_short_matched)), function(i) {
  cor(cpm_short_matched[, i], cpm_long_matched[, i], method = "pearson")
})
sample_cors_spearman <- sapply(seq_len(ncol(cpm_short_matched)), function(i) {
  cor(cpm_short_matched[, i], cpm_long_matched[, i], method = "spearman")
})

cor_pearson_samplewise  <- mean(sample_cors_pearson)
cor_spearman_samplewise <- mean(sample_cors_spearman)

cat("Sample-wise Pearson  (mean):", sprintf("%.3f", cor_pearson_samplewise),
    "| range:", sprintf("%.3f", min(sample_cors_pearson)),
    "-",        sprintf("%.3f", max(sample_cors_pearson)), "\n")
cat("Sample-wise Spearman (mean):", sprintf("%.3f", cor_spearman_samplewise),
    "| range:", sprintf("%.3f", min(sample_cors_spearman)),
    "-",        sprintf("%.3f", max(sample_cors_spearman)), "\n")

cor_table <- data.frame(
  sample   = common_samples,
  pearson  = sprintf("%.3f", sample_cors_pearson),
  spearman = sprintf("%.3f", sample_cors_spearman)
)
print(cor_table)

```

```{r corr-overall-plot, fig.width=10, fig.height=6}
plot_data <- data.frame(
  short_read = as.vector(cpm_short_matched),
  long_read  = as.vector(cpm_long_matched)
)

p_overall <- ggplot(plot_data, aes(x = short_read, y = long_read)) +
  geom_hex(bins = 50) +
  scale_fill_gradientn(colors = c("white", "lightblue", "blue", "darkblue", "red"),
                       trans = "log10", name = "Count") +
  annotate("text", x = min(plot_data$short_read), y = max(plot_data$long_read),
           hjust = 0, vjust = 1,
           label = paste0("Sample-wise: Spearman = ",
                          sprintf("%.3f", cor_spearman_samplewise),
                          ", Pearson = ",
                          sprintf("%.3f", cor_pearson_samplewise)),
           size = 3.5) +
  labs(title = "Short-Read Vs Long-Read Gene Expression",
       x = "Short-Read Gene log2(CPM)",
       y = "Long-Read Gene log2(CPM)") +
  theme_minimal() +
  theme(panel.grid.minor = element_blank(),
        plot.title       = element_text(face = "bold"))

print(p_overall)

ggsave(file.path(OUTPUT_DIR, "sr_vs_lr_gene_expression_correlation.png"),
       p_overall, width = 10, height = 6, dpi = 300, bg = "white")
```

# Correlation By Cell Type

```{r corr-by-celltype}
sample_names <- colnames(cpm_short_matched)
cell_types   <- extract_ct(sample_names)
unique_cts   <- intersect(CT_KEEP, unique(cell_types))  # preserves CT_KEEP order

cat("Cell type counts (internal codes):\n"); print(table(cell_types))

ct_results <- list()
for (ct in unique_cts) {
  ct_samples   <- sample_names[cell_types == ct]
  cpm_short_ct <- cpm_short_matched[, ct_samples]
  cpm_long_ct  <- cpm_long_matched[, ct_samples]

  spear <- sapply(seq_len(ncol(cpm_short_ct)), function(i) {
    cor(cpm_short_ct[, i], cpm_long_ct[, i], method = "spearman")
  })
  pears <- sapply(seq_len(ncol(cpm_short_ct)), function(i) {
    cor(cpm_short_ct[, i], cpm_long_ct[, i], method = "pearson")
  })

  ct_results[[ct]] <- list(
    samples              = ct_samples,
    spearman             = mean(spear),
    pearson              = mean(pears),
    sample_cors_spearman = spear,
    sample_cors_pearson  = pears
  )
}

# Relabel internal codes to display names (AT->AT2, DD->LAE) for the summary.
summary_df <- data.frame(
  CellType      = unname(CT_DISPLAY[unique_cts]),
  N_Samples     = sapply(unique_cts, function(ct) length(ct_results[[ct]]$samples)),
  Spearman_Mean = sapply(unique_cts, function(ct) sprintf("%.3f", ct_results[[ct]]$spearman)),
  Pearson_Mean  = sapply(unique_cts, function(ct) sprintf("%.3f", ct_results[[ct]]$pearson))
)
cat("\nPer-cell-type sample-wise correlation:\n"); print(summary_df)

```

## Multi-Panel: Per-Cell-Type Hex Scatter

With four cell types this is a clean 2x2 grid.

```{r corr-by-celltype-panel, fig.width=12, fig.height=10}
ct_plots <- list()

for (ct in unique_cts) {
  ct_samples   <- ct_results[[ct]]$samples
  cpm_short_ct <- cpm_short_matched[, ct_samples]
  cpm_long_ct  <- cpm_long_matched[, ct_samples]

  plot_data <- data.frame(short_read = as.vector(cpm_short_ct),
                          long_read  = as.vector(cpm_long_ct))

  ct_plots[[ct]] <- ggplot(plot_data, aes(x = short_read, y = long_read)) +
    geom_hex(bins = 50) +
    scale_fill_gradientn(
      colors = c("white", "lightblue", "blue", "darkblue", "red"),
      trans  = "log10", name = "Count"
    ) +
    annotate("text",
             x = min(plot_data$short_read), y = max(plot_data$long_read),
             hjust = 0, vjust = 1,
             label = paste0("r = ",       sprintf("%.3f", ct_results[[ct]]$pearson),
                            "\nρ = ", sprintf("%.3f", ct_results[[ct]]$spearman)),
             size = 4, fontface = "bold") +
    labs(title = CT_DISPLAY[ct],   # display name (AT2 / LAE / FB / MV)
         x = "Short-Read log2(CPM)",
         y = "Long-Read log2(CPM)") +
    theme_minimal() +
    theme(panel.grid.minor = element_blank(),
          plot.title       = element_text(size = 14, face = "bold"),
          legend.position  = "none")
}

p_cor_panel <- wrap_plots(ct_plots, ncol = 2) +
  plot_annotation(
    title    = "Short-Read Vs Long-Read Gene Expression By Cell Type",
    subtitle = "log2(CPM) with TMM normalization; Pearson r and Spearman ρ per cell type",
    theme    = theme(plot.title    = element_text(size = 16, face = "bold"),
                     plot.subtitle = element_text(size = 12, color = "grey40"))
  )

print(p_cor_panel)

ggsave(file.path(OUTPUT_DIR, "sr_vs_lr_expression_by_celltype_panel.png"),
       p_cor_panel, width = 12, height = 10, dpi = 300, bg = "white")
```

# Save Results

```{r save-results}
fwrite(cor_table,  file.path(OUTPUT_DIR, "samplewise_correlation.csv"))
fwrite(summary_df, file.path(OUTPUT_DIR, "celltype_correlation_summary.csv"))

cat("All outputs saved to:", OUTPUT_DIR, "\n")
print(list.files(OUTPUT_DIR, pattern = "\\.(csv|png)$"))

```

# Session Info

```{r session-info}
sessionInfo()
```
