---
title: "DE by Cell Type Under DMSO"
subtitle: "Identifying cell-type-specific isoforms and genes at baseline (no Smg1i)"
author: "Leshem, Yul"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: hide
    df_print: paged
    theme: flatly
---
Code location: `/udd/reyle/nmd_lungcells_2026/code/final/ct_de.Rmd`

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 8, fig.height = 5)
options(bitmapType = "cairo")
```

# Overview

This analysis identifies isoforms and genes whose **baseline (DMSO) expression is cell-type-specific** — i.e., that distinguish AT2, LAE, FB, MV from each other independent of NMD inhibition.

**Contrast structure:** one-vs-rest per cell type (CT_X vs mean of the other three) → defines CT-marker features.

**Primary level: long-read isoform (LR DIE).** Secondary level: short-read gene (SR DGE), which is also used for topGO overrepresentation.

**Adaptive shrinkage:** after the per-CT limma contrasts, we run `mashr` jointly across all four one-vs-rest effects to (a) borrow strength across CTs, (b) get `lfsr`-based significance that respects the joint structure, and (c) enable sharing analysis (how many markers are CT-specific vs broadly distinctive).

**Prerequisite packages:** `mashr` and `ashr` must be installed *before* knitting
(`install.packages("ashr"); BiocManager::install("mashr")`). This document does not
install packages at render time.

DMSO samples only.

# Configuration

```{r config}
suppressPackageStartupMessages({
  library(data.table)
  library(dplyr)
  library(tidyr)
  library(ggplot2)
  library(patchwork)
  library(edgeR)
  library(limma)
  library(pheatmap)
  library(RColorBrewer)
  library(reshape2)
})

# Paths
BASE         <- "/udd/reyle/nmd_lungcells_2026"
LR_ISO_PATH  <- file.path(BASE, "long_read",  "dge_isoform_longread_filtered_2026.3.3.rds")
SR_GENE_PATH <- file.path(BASE, "short_read", "dge_shortread_gene_filtered_2026.3.2.rds")
OUT_DIR      <- file.path(BASE, "long_read", "de_by_celltype_dmso")
MASH_DIR     <- file.path(OUT_DIR, "mashr")
FIG_DIR      <- file.path(OUT_DIR, "figures")   # keep figures with this analysis (was BASE/figure)
dir.create(OUT_DIR,  recursive = TRUE, showWarnings = FALSE)
dir.create(MASH_DIR, recursive = TRUE, showWarnings = FALSE)
dir.create(FIG_DIR,  recursive = TRUE, showWarnings = FALSE)

stopifnot(file.exists(LR_ISO_PATH))
stopifnot(file.exists(SR_GENE_PATH))

# Cell types kept, in display naming (AT2 = old AT, LAE = old DD)
CT_KEEP <- c("AT2", "LAE", "FB", "MV")

# Relabel helper: map the DGEList's internal ct codes to display names.
relabel_ct <- function(samples_df) {
  samples_df$ct <- as.character(samples_df$ct)
  samples_df$ct[samples_df$ct == "AT"] <- "AT2"
  samples_df$ct[samples_df$ct == "DD"] <- "LAE"
  samples_df
}

# Significance thresholds
LFSR_CUT      <- 0.05    # for mashr-shrunk results
PADJ_CUT      <- 0.05    # for limma BH-adjusted results
LOGFC_MARKER  <- 1       # |logFC| > 1 (~2-fold) for CT-marker classification

DATE_STAMP <- format(Sys.Date(), "%Y.%m.%d")
```

# Load Data

We load both the long-read isoform DGEList and the short-read gene DGEList that were used in the main DGE/DIE pipelines, subset to DMSO samples and the four kept cell types.

```{r donor_helper}
# DGELists in this project use different donor column names:
#   LR isoform DGEList: $samples$id
#   SR gene DGEList:    $samples$sample_id
# Pick whichever exists.
get_donor <- function(samples_df) {
  candidates <- c("id", "sample_id", "donor", "donor_id")
  hit <- candidates[candidates %in% colnames(samples_df)][1]
  if (is.na(hit)) stop("No donor column found in samples. Available: ",
                        paste(colnames(samples_df), collapse = ", "))
  samples_df[[hit]]
}
```

```{r load_lr_isoform}
# Long-read isoform DGEList (filtered, same object as DIE pipeline input)
y_iso <- readRDS(LR_ISO_PATH)

# Relabel ct to display names AFTER loading (AT -> AT2, DD -> LAE)
y_iso$samples <- relabel_ct(y_iso$samples)

# Subset to DMSO samples and four kept CTs
keep_iso <- y_iso$samples$treatment == "DMSO" &
            y_iso$samples$ct %in% CT_KEEP
cat(sprintf("FILTER: DMSO + CT_KEEP {%s}. Samples before: %d / after: %d.\n",
            paste(CT_KEEP, collapse = ", "), ncol(y_iso), sum(keep_iso)))
y_iso_dmso <- y_iso[, keep_iso]

# Drop unused factor levels and rebuild as plain character → factor
y_iso_dmso$samples$ct        <- factor(as.character(y_iso_dmso$samples$ct), levels = CT_KEEP)
y_iso_dmso$samples$treatment <- factor(as.character(y_iso_dmso$samples$treatment))
# Donor: copy whichever column exists into a uniform `donor_factor`
y_iso_dmso$samples$donor_factor <- factor(as.character(get_donor(y_iso_dmso$samples)))

stopifnot(nrow(y_iso_dmso$samples) == ncol(y_iso_dmso))
stopifnot(!any(is.na(y_iso_dmso$samples$ct)))
stopifnot(!any(is.na(y_iso_dmso$samples$donor_factor)))

cat("LR isoform DMSO DGEList:",
    nrow(y_iso_dmso), "isoforms x",
    ncol(y_iso_dmso), "samples\n")
print(table(y_iso_dmso$samples$ct, y_iso_dmso$samples$donor_factor))
```

```{r load_sr_gene}
# Short-read gene DGEList (filtered, same object as SR DGE pipeline input)
y_gene <- readRDS(SR_GENE_PATH)

# Relabel ct to display names AFTER loading (AT -> AT2, DD -> LAE)
y_gene$samples <- relabel_ct(y_gene$samples)

keep_gene <- y_gene$samples$treatment == "DMSO" &
             y_gene$samples$ct %in% CT_KEEP
cat(sprintf("FILTER: DMSO + CT_KEEP {%s}. Samples before: %d / after: %d.\n",
            paste(CT_KEEP, collapse = ", "), ncol(y_gene), sum(keep_gene)))
y_gene_dmso <- y_gene[, keep_gene]

y_gene_dmso$samples$ct        <- factor(as.character(y_gene_dmso$samples$ct), levels = CT_KEEP)
y_gene_dmso$samples$treatment <- factor(as.character(y_gene_dmso$samples$treatment))
y_gene_dmso$samples$donor_factor <- factor(as.character(get_donor(y_gene_dmso$samples)))

stopifnot(nrow(y_gene_dmso$samples) == ncol(y_gene_dmso))
stopifnot(!any(is.na(y_gene_dmso$samples$ct)))
stopifnot(!any(is.na(y_gene_dmso$samples$donor_factor)))

cat("SR gene DMSO DGEList:",
    nrow(y_gene_dmso), "genes x",
    ncol(y_gene_dmso), "samples\n")
print(table(y_gene_dmso$samples$ct, y_gene_dmso$samples$donor_factor))
```

# Helper: Fit Limma-Voom and Extract Contrasts

Wraps the standard pipeline: `voom` → `duplicateCorrelation` → `lmFit` → `contrasts.fit` → `eBayes` → `topTable`. Donor structure: donors are shared across three of the four cell types (partially crossed, unbalanced). Donor therefore enters as a random effect via duplicateCorrelation(block = donor).

The helper also returns the post-contrast `fit2` object so that downstream `mashr` can pull effect sizes (`coefficients`) and moderated SEs (`coefficients / t`).

```{r fit_helper}
fit_voom_contrasts <- function(y, contrasts_mat, design, block, level = "feature") {
  # Refilter for DMSO-only subset
  keep <- filterByExpr(y, design = design, min.count = 5, min.total.count = 10)
  y2 <- y[keep, , keep.lib.sizes = FALSE]
  y2 <- calcNormFactors(y2, method = "TMM")

  cat(sprintf("FILTER: filterByExpr on DMSO subset — %d %ss retained (was %d)\n",
              nrow(y2), level, nrow(y)))

  v <- voom(y2, design, plot = FALSE)

  corfit <- duplicateCorrelation(v, design, block = block)
  cat(sprintf("Within-donor correlation: %.3f\n", corfit$consensus.correlation))

  fit <- lmFit(v, design, block = block, correlation = corfit$consensus.correlation)

  fit2 <- contrasts.fit(fit, contrasts_mat)
  fit2 <- eBayes(fit2)

  out <- lapply(colnames(contrasts_mat), function(cn) {
    tt <- topTable(fit2, coef = cn, number = Inf, sort.by = "none")
    tt$feature_id <- rownames(tt)
    tt$contrast <- cn
    as.data.table(tt)
  })
  names(out) <- colnames(contrasts_mat)
  list(toptables = out, fit2 = fit2, dge_filtered = y2)
}
```

# One-vs-Rest Contrasts

For each CT, define a one-vs-rest contrast: that CT minus the mean of the other three. This identifies CT-marker features — features whose expression in CT_X differs from the rest of the panel.

```{r ovr_design}
build_ovr_contrasts <- function(y) {
  ct  <- droplevels(factor(y$samples$ct, levels = CT_KEEP))
  don <- droplevels(factor(y$samples$donor_factor))

  cat(sprintf("ncol(y)=%d  length(ct)=%d  length(don)=%d\n",
              ncol(y), length(ct), length(don)))
  stopifnot(length(ct)  == ncol(y))
  stopifnot(length(don) == ncol(y))

  # Match SR pipeline: ct only in design, donor enters via duplicateCorrelation block
  design <- model.matrix(~ 0 + ct)
  colnames(design) <- gsub("^ct", "", colnames(design))

  # One-vs-rest: each CT minus mean of other three (display names AT2/LAE)
  cm <- makeContrasts(
    AT2_vs_rest = AT2 - (LAE + FB + MV)/3,
    LAE_vs_rest = LAE - (AT2 + FB + MV)/3,
    FB_vs_rest  = FB  - (AT2 + LAE + MV)/3,
    MV_vs_rest  = MV  - (AT2 + LAE + FB)/3,
    levels = design
  )
  list(design = design, contrasts = cm, donor = don)
}
```

## LR Isoform Level (Primary)

```{r ovr_iso}
ovr_iso <- build_ovr_contrasts(y_iso_dmso)
res_iso <- fit_voom_contrasts(
  y_iso_dmso, ovr_iso$contrasts, ovr_iso$design,
  block = ovr_iso$donor, level = "isoform"
)
res_ovr_iso     <- res_iso$toptables
fit2_iso        <- res_iso$fit2
y_iso_dmso_filt <- res_iso$dge_filtered

ovr_iso_dt <- rbindlist(res_ovr_iso, idcol = "ct_contrast")
invisible(ovr_iso_dt[, `:=`(
  ct          = sub("_vs_rest$", "", ct_contrast),
  is_marker   = adj.P.Val < PADJ_CUT & logFC >  LOGFC_MARKER,
  is_depleted = adj.P.Val < PADJ_CUT & logFC < -LOGFC_MARKER
)])

cat("\n=== LR Isoform CT-marker counts (one-vs-rest, unshrunken limma) ===\n")
print(ovr_iso_dt[, .(
  n_marker_up   = sum(is_marker),
  n_marker_down = sum(is_depleted),
  n_tested      = .N
), by = ct])
```

## SR Gene Level (Secondary)

```{r ovr_gene}
ovr_gene <- build_ovr_contrasts(y_gene_dmso)
res_gene <- fit_voom_contrasts(
  y_gene_dmso, ovr_gene$contrasts, ovr_gene$design,
  block = ovr_gene$donor, level = "gene"
)
res_ovr_gene     <- res_gene$toptables
fit2_gene        <- res_gene$fit2
y_gene_dmso_filt <- res_gene$dge_filtered

ovr_gene_dt <- rbindlist(res_ovr_gene, idcol = "ct_contrast")
invisible(ovr_gene_dt[, `:=`(
  ct          = sub("_vs_rest$", "", ct_contrast),
  is_marker   = adj.P.Val < PADJ_CUT & logFC >  LOGFC_MARKER,
  is_depleted = adj.P.Val < PADJ_CUT & logFC < -LOGFC_MARKER
)])

cat("\n=== SR Gene CT-marker counts (one-vs-rest, unshrunken limma) ===\n")
print(ovr_gene_dt[, .(
  n_marker_up   = sum(is_marker),
  n_marker_down = sum(is_depleted),
  n_tested      = .N
), by = ct])
```

# Adaptive Shrinkage Across Cell Types (mashr)

We jointly shrink the four one-vs-rest contrast effects with `mashr`. The four contrasts are non-independent by construction (each CT minus mean of the other three), so the **null correlation matrix `Vhat`** is essential — it is what lets `mashr` not over-call shared signal that is just the residual structure of the contrast design.

The procedure mirrors the SR DGE pipeline exactly:

1. Pull `bhat` (effect sizes) and `shat` (moderated SEs = `coef / t`) from `fit2`.
2. Filter to features with finite, positive SEs in all four contrasts.
3. Estimate `Vhat` from a random subset (the `mashr`-recommended workflow).
4. Build a covariance library (data-driven PCA + canonical patterns).
5. Fit, extract posterior means (`pm`) and `lfsr`.

```{r mashr-libs}
# mashr + ashr are prerequisites (see Overview). Fail loudly if absent rather
# than installing at render time on someone else's machine.
.need <- c("ashr", "mashr")
.miss <- .need[!vapply(.need, requireNamespace, logical(1), quietly = TRUE)]
if (length(.miss))
  stop("Missing required package(s): ", paste(.miss, collapse = ", "),
       ". Install before knitting:  install.packages('ashr'); BiocManager::install('mashr')")
suppressPackageStartupMessages({
  library(mashr)
  library(ashr)
})
```

```{r mashr-helper}
# Generic mashr runner for any limma fit2 object with multiple contrasts.
# Returns a list with the model, posterior means, lfsr, sharing matrices,
# and Vhat — keyed by the contrast names from fit2.
# seed = 42 matches the short-read pipeline's RNG for the random Vhat subset.
run_mashr_on_fit <- function(fit2, level_label = "feature", n_random = 20000, seed = 42) {

  bhat <- fit2$coefficients
  shat <- fit2$coefficients / fit2$t

  # Drop features with non-finite or zero SE in any contrast
  finite_rows <- apply(shat, 1, function(x) all(is.finite(x) & x > 0))
  cat(sprintf("[%s] %ss with valid SEs in all contrasts: %d of %d\n",
              level_label, level_label, sum(finite_rows), nrow(shat)))
  bhat <- bhat[finite_rows, , drop = FALSE]
  shat <- shat[finite_rows, , drop = FALSE]
  stopifnot(identical(dim(bhat), dim(shat)))

  # Strong signals (used to seed data-driven covariances)
  mash_data_init <- mash_set_data(Bhat = bhat, Shat = shat)
  m_1by1 <- mash_1by1(mash_data_init)
  strong <- get_significant_results(m_1by1, thresh = 0.05)
  cat(sprintf("[%s] Strong (1-by-1) signals: %d\n", level_label, length(strong)))

  # Estimate null correlation from a random subset
  set.seed(seed)
  n_use <- min(n_random, nrow(bhat))
  random_idx <- sample(seq_len(nrow(bhat)), n_use)
  mash_data_random <- mash_set_data(Bhat = bhat[random_idx, , drop = FALSE],
                                    Shat = shat[random_idx, , drop = FALSE])
  Vhat <- estimate_null_correlation_simple(mash_data_random)
  cat(sprintf("[%s] Null correlation Vhat:\n", level_label))
  print(round(Vhat, 3))

  # Final mash data with Vhat
  mash_data <- mash_set_data(Bhat = bhat, Shat = shat, V = Vhat)

  # Covariance library: data-driven PCA (if enough strong signals) + canonical
  if (length(strong) >= 5) {
    U_pca <- cov_pca(mash_data, npc = min(5, ncol(bhat) - 1), subset = strong)
  } else {
    U_pca <- list()
    cat(sprintf("[%s] Too few strong signals; skipping data-driven covariances\n",
                level_label))
  }
  U_c   <- cov_canonical(mash_data)
  U_all <- c(U_pca, U_c)
  cat(sprintf("[%s] Total covariance matrices: %d\n", level_label, length(U_all)))

  # Fit
  t_start <- Sys.time()
  m <- mash(mash_data, Ulist = U_all, outputlevel = 2)
  cat(sprintf("[%s] mashr fit complete in %.1f min\n",
              level_label,
              as.numeric(difftime(Sys.time(), t_start, units = "mins"))))

  pi_vals <- get_estimated_pi(m)
  cat(sprintf("[%s] Estimated mixture weights (>1%%):\n", level_label))
  print(round(sort(pi_vals[pi_vals > 0.01], decreasing = TRUE), 3))

  pm   <- get_pm(m)
  lfsr <- get_lfsr(m)

  # Per-contrast significance
  cat(sprintf("[%s] Significant features per contrast (lfsr < %.2f):\n",
              level_label, LFSR_CUT))
  print(apply(lfsr, 2, function(x) sum(x < LFSR_CUT)))

  # Sharing
  n_sig_total <- length(get_significant_results(m, thresh = LFSR_CUT))
  cat(sprintf("[%s] Total significant features (any contrast): %d\n",
              level_label, n_sig_total))

  if (n_sig_total >= 2) {
    sharing      <- get_pairwise_sharing(m, factor = 0.5, lfsr_thresh = LFSR_CUT)
    sharing_sign <- get_pairwise_sharing(m, factor = 0,   lfsr_thresh = LFSR_CUT)
  } else {
    sharing <- sharing_sign <- NULL
  }

  list(model = m, pm = pm, lfsr = lfsr,
       sharing = sharing, sharing_sign = sharing_sign,
       Vhat = Vhat, bhat = bhat, shat = shat,
       strong = strong)
}
```

## LR Isoform Mashr (Primary)

```{r mashr-iso}
mash_iso <- run_mashr_on_fit(fit2_iso, level_label = "isoform")
```

## SR Gene Mashr (Secondary)

```{r mashr-gene}
mash_gene <- run_mashr_on_fit(fit2_gene, level_label = "gene")
```

# Mashr-Shrunken CT-Marker Tables

Build long tables using shrunken effects. Marker classification:

- `is_marker_mash`   = `pm > LOGFC_MARKER  & lfsr < LFSR_CUT` (CT-enriched)
- `is_depleted_mash` = `pm < -LOGFC_MARKER & lfsr < LFSR_CUT` (CT-depleted)

```{r mashr-long-tables}
mash_long <- function(mash_res, feature_col = "feature_id") {
  pm   <- mash_res$pm
  lfsr <- mash_res$lfsr
  stopifnot(identical(rownames(pm), rownames(lfsr)))
  stopifnot(identical(colnames(pm), colnames(lfsr)))

  out <- rbindlist(lapply(colnames(pm), function(cn) {
    data.table(
      feature_id  = rownames(pm),
      ct_contrast = cn,
      pm          = pm[, cn],
      lfsr        = lfsr[, cn]
    )
  }))
  setnames(out, "feature_id", feature_col)

  out[, ct := sub("_vs_rest$", "", ct_contrast)]
  out[, is_marker_mash   := pm >  LOGFC_MARKER & lfsr < LFSR_CUT]
  out[, is_depleted_mash := pm < -LOGFC_MARKER & lfsr < LFSR_CUT]
  out[]
}

ovr_iso_mash_dt  <- mash_long(mash_iso)
ovr_gene_mash_dt <- mash_long(mash_gene)

# Attach gene annotation to isoform-level mashr table (matches non-mashr table)
iso2gene <- as.data.table(y_iso_dmso$genes)[
  , .(feature_id = txid,
      gene_ensg  = sub("\\.\\d+$", "", gene_id),
      gene_sym   = hgnc_symbol)
]
ovr_iso_mash_dt <- merge(ovr_iso_mash_dt, iso2gene, by = "feature_id", all.x = TRUE)

cat("\n=== LR Isoform CT-marker counts (mashr-shrunken) ===\n")
print(ovr_iso_mash_dt[, .(
  n_marker_up   = sum(is_marker_mash),
  n_marker_down = sum(is_depleted_mash),
  n_tested      = .N
), by = ct])

cat("\n=== SR Gene CT-marker counts (mashr-shrunken) ===\n")
print(ovr_gene_mash_dt[, .(
  n_marker_up   = sum(is_marker_mash),
  n_marker_down = sum(is_depleted_mash),
  n_tested      = .N
), by = ct])
```

## Unshrunken vs Shrunken Comparison

```{r mashr-vs-limma}
cmp_iso <- merge(
  ovr_iso_dt[,      .(n_limma_marker = sum(is_marker)),      by = ct],
  ovr_iso_mash_dt[, .(n_mash_marker  = sum(is_marker_mash)), by = ct],
  by = "ct"
)
invisible(cmp_iso[, ratio := round(n_mash_marker / pmax(n_limma_marker, 1), 2)])
cat("\n=== LR Isoform: limma vs mashr CT-marker counts ===\n"); print(cmp_iso)

cmp_gene <- merge(
  ovr_gene_dt[,      .(n_limma_marker = sum(is_marker)),      by = ct],
  ovr_gene_mash_dt[, .(n_mash_marker  = sum(is_marker_mash)), by = ct],
  by = "ct"
)
invisible(cmp_gene[, ratio := round(n_mash_marker / pmax(n_limma_marker, 1), 2)])
cat("\n=== SR Gene: limma vs mashr CT-marker counts ===\n"); print(cmp_gene)
```

## Pairwise Sharing Heatmap (LR Isoform)

How often does a CT-marker isoform also reach significance (same sign, within 2-fold magnitude) in another CT's one-vs-rest contrast? Because the contrasts are by construction zero-sum — being up in CT_X means being relatively down in the other three — sharing values here are expected to be **low and slightly negative**, not high. Anything > 0.2 indicates a feature whose magnitude is similar in two CTs (e.g., shared by AT2 and LAE relative to FB+MV).

```{r mashr-sharing-iso, fig.width=8, fig.height=7}
plot_sharing <- function(mash_res, level_label, fig_path) {
  if (is.null(mash_res$sharing)) {
    cat(sprintf("[%s] No sharing matrix (too few significant); skipping plot\n", level_label))
    return(invisible(NULL))
  }
  s <- mash_res$sharing
  rownames(s) <- gsub("_vs_rest$", "", rownames(s))
  colnames(s) <- gsub("_vs_rest$", "", colnames(s))

  ct_order <- intersect(CT_KEEP, rownames(s))

  df <- melt(s)
  colnames(df) <- c("CT1", "CT2", "Sharing")
  df$CT1 <- factor(df$CT1, levels = ct_order)
  df$CT2 <- factor(df$CT2, levels = ct_order)

  p <- ggplot(df, aes(x = CT1, y = CT2, fill = Sharing)) +
    geom_tile(color = "white", linewidth = 0.8) +
    geom_text(aes(label = round(Sharing, 2)),
              size = 4, fontface = "bold",
              color = ifelse(df$Sharing > 0.6, "white", "grey20")) +
    scale_fill_gradient2(low = "#D73027", mid = "white", high = "#185FA5",
                         midpoint = 0, limits = c(-1, 1), name = "Sharing") +
    labs(title    = sprintf("Pairwise Effect Sharing — %s", level_label),
         subtitle = sprintf("Same sign & within 2-fold magnitude (lfsr < %.2f)",
                            LFSR_CUT),
         x = NULL, y = NULL) +
    theme_minimal(base_size = 13) +
    theme(axis.text.x   = element_text(angle = 45, hjust = 1, face = "bold"),
          axis.text.y   = element_text(face = "bold"),
          panel.grid    = element_blank(),
          plot.title    = element_text(face = "bold", hjust = 0.5),
          plot.subtitle = element_text(color = "grey40", hjust = 0.5))

  ggsave(fig_path, p, width = 8, height = 7, dpi = 300, bg = "white")
  print(p)
}

plot_sharing(mash_iso,  "LR Isoform",
             file.path(FIG_DIR, paste0("de_by_ct_dmso_mashr_sharing_iso_",  DATE_STAMP, ".png")))
plot_sharing(mash_gene, "SR Gene",
             file.path(FIG_DIR, paste0("de_by_ct_dmso_mashr_sharing_gene_", DATE_STAMP, ".png")))
```

## logFC Correlation Across CTs (Mashr Posterior Means)

```{r mashr-logfc-cor, fig.width=8, fig.height=7}
plot_pm_cor <- function(mash_res, level_label, fig_path) {
  pm   <- mash_res$pm
  lfsr <- mash_res$lfsr

  any_sig <- rowSums(lfsr < LFSR_CUT) >= 1
  cat(sprintf("[%s] Features significant in any contrast: %d of %d\n",
              level_label, sum(any_sig), nrow(pm)))
  if (sum(any_sig) < 2) return(invisible(NULL))

  pm_sig <- pm[any_sig, , drop = FALSE]
  cor_mat <- cor(pm_sig, method = "pearson")
  rownames(cor_mat) <- gsub("_vs_rest$", "", rownames(cor_mat))
  colnames(cor_mat) <- gsub("_vs_rest$", "", colnames(cor_mat))
  ct_order <- intersect(CT_KEEP, rownames(cor_mat))

  df <- melt(cor_mat)
  colnames(df) <- c("CT1", "CT2", "r")
  df$CT1 <- factor(df$CT1, levels = ct_order)
  df$CT2 <- factor(df$CT2, levels = ct_order)

  p <- ggplot(df, aes(x = CT1, y = CT2, fill = r)) +
    geom_tile(color = "white", linewidth = 0.8) +
    geom_text(aes(label = round(r, 2)),
              size = 4, fontface = "bold",
              color = ifelse(abs(df$r) > 0.6, "white", "grey20")) +
    scale_fill_gradient2(low = "#D73027", mid = "white", high = "#185FA5",
                         midpoint = 0, limits = c(-1, 1), name = "Pearson r") +
    labs(title    = sprintf("logFC Correlation Across CTs — %s", level_label),
         subtitle = "Pearson r of posterior mean logFC (one-vs-rest contrasts)",
         x = NULL, y = NULL) +
    theme_minimal(base_size = 13) +
    theme(axis.text.x   = element_text(angle = 45, hjust = 1, face = "bold"),
          axis.text.y   = element_text(face = "bold"),
          panel.grid    = element_blank(),
          plot.title    = element_text(face = "bold", hjust = 0.5),
          plot.subtitle = element_text(color = "grey40", hjust = 0.5))

  ggsave(fig_path, p, width = 8, height = 7, dpi = 300, bg = "white")
  print(p)
}

plot_pm_cor(mash_iso,  "LR Isoform",
            file.path(FIG_DIR, paste0("de_by_ct_dmso_mashr_pmcor_iso_",  DATE_STAMP, ".png")))
plot_pm_cor(mash_gene, "SR Gene",
            file.path(FIG_DIR, paste0("de_by_ct_dmso_mashr_pmcor_gene_", DATE_STAMP, ".png")))
```

## Save Mashr Outputs

```{r mashr-save}
# Models
saveRDS(mash_iso$model,
        file.path(MASH_DIR, paste0("mashr_model_iso_",  DATE_STAMP, ".rds")))
saveRDS(mash_gene$model,
        file.path(MASH_DIR, paste0("mashr_model_gene_", DATE_STAMP, ".rds")))

# Long tables (the user-facing artifacts)
fwrite(ovr_iso_mash_dt,
       file.path(MASH_DIR, paste0("ovr_iso_mashr_results_",  DATE_STAMP, ".csv")))
fwrite(ovr_gene_mash_dt,
       file.path(MASH_DIR, paste0("ovr_gene_mashr_results_", DATE_STAMP, ".csv")))

# Wide pm / lfsr matrices (mirrors SR DGE pipeline format)
write_pm_lfsr <- function(mash_res, gene_anno, level_tag) {
  pm_df   <- as.data.frame(mash_res$pm);   pm_df$feature_id   <- rownames(pm_df)
  lfsr_df <- as.data.frame(mash_res$lfsr); lfsr_df$feature_id <- rownames(lfsr_df)
  if (!is.null(gene_anno)) {
    pm_df   <- merge(pm_df,   gene_anno, by = "feature_id", all.x = TRUE)
    lfsr_df <- merge(lfsr_df, gene_anno, by = "feature_id", all.x = TRUE)
  }
  fwrite(pm_df,
         file.path(MASH_DIR, sprintf("mashr_posterior_means_%s_%s.csv",
                                     level_tag, DATE_STAMP)))
  fwrite(lfsr_df,
         file.path(MASH_DIR, sprintf("mashr_lfsr_%s_%s.csv",
                                     level_tag, DATE_STAMP)))
}

# Isoform: include gene-symbol annotation
iso_anno <- iso2gene[, .(feature_id, gene_sym, gene_ensg)]
write_pm_lfsr(mash_iso, iso_anno, "iso")

# Gene: include external_gene_name from DGEList
gene_anno_dt <- as.data.table(y_gene_dmso_filt$genes)[
  , .(feature_id = ensembl_gene_id_version,
      external_gene_name)
]
gene_anno_dt <- gene_anno_dt[!duplicated(feature_id)]
write_pm_lfsr(mash_gene, gene_anno_dt, "gene")

# Sharing matrices
if (!is.null(mash_iso$sharing)) {
  fwrite(as.data.frame(mash_iso$sharing),
         file.path(MASH_DIR, paste0("mashr_sharing_2fold_iso_", DATE_STAMP, ".csv")),
         row.names = TRUE)
  fwrite(as.data.frame(mash_iso$sharing_sign),
         file.path(MASH_DIR, paste0("mashr_sharing_sign_iso_", DATE_STAMP, ".csv")),
         row.names = TRUE)
}
if (!is.null(mash_gene$sharing)) {
  fwrite(as.data.frame(mash_gene$sharing),
         file.path(MASH_DIR, paste0("mashr_sharing_2fold_gene_", DATE_STAMP, ".csv")),
         row.names = TRUE)
  fwrite(as.data.frame(mash_gene$sharing_sign),
         file.path(MASH_DIR, paste0("mashr_sharing_sign_gene_", DATE_STAMP, ".csv")),
         row.names = TRUE)
}

# Vhat
fwrite(as.data.frame(mash_iso$Vhat),
       file.path(MASH_DIR, paste0("mashr_Vhat_iso_",  DATE_STAMP, ".csv")),
       row.names = TRUE)
fwrite(as.data.frame(mash_gene$Vhat),
       file.path(MASH_DIR, paste0("mashr_Vhat_gene_", DATE_STAMP, ".csv")),
       row.names = TRUE)

cat("\nmashr outputs saved to:", MASH_DIR, "\n")
print(list.files(MASH_DIR))
```

## Overlap Across Levels (Mashr-Shrunken)

Genes whose mRNA is a CT-marker at both the SR gene level (mashr) and the LR isoform level (mashr, any isoform).

```{r mashr-overlap}
# Harmonize both sides on bare ENSG (strip version suffix) so the intersect works.
ovr_iso_dt <- merge(ovr_iso_dt, iso2gene, by = "feature_id", all.x = TRUE)

iso_marker_genes <- ovr_iso_dt[is_marker == TRUE,
  .(ensg = unique(gene_ensg)), by = ct][!is.na(ensg) & ensg != ""]

gene_marker_genes <- ovr_gene_dt[is_marker == TRUE,
  .(ensg = sub("\\.\\d+$", "", feature_id)), by = ct]

overlap_tab <- merge(
  iso_marker_genes[, .N, by = ct][, .(ct, n_iso_marker_genes = N)],
  gene_marker_genes[, .N, by = ct][, .(ct, n_gene_marker = N)],
  by = "ct"
)
invisible(overlap_tab[, n_both := mapply(function(c) {
  length(intersect(iso_marker_genes[ct == c, ensg],
                   gene_marker_genes[ct == c, ensg]))
}, ct)])
invisible(overlap_tab[, `:=`(
  pct_iso_in_gene = round(100 * n_both / n_iso_marker_genes, 1),
  pct_gene_in_iso = round(100 * n_both / n_gene_marker,      1)
)])

cat("\n=== Iso↔Gene marker overlap (bare ENSG) ===\n")
print(overlap_tab)

fwrite(overlap_tab, file.path(OUT_DIR, "iso_gene_marker_overlap.csv"))
```



## CT-Marker Counts Per CT

```{r counts_barplot, fig.width=10, fig.height=5}
counts_iso  <- ovr_iso_dt[, .(level = "Isoform (LR)", n = sum(is_marker)), by = ct]
counts_gene <- ovr_gene_dt[, .(level = "Gene (SR)",   n = sum(is_marker)), by = ct]
counts_all  <- rbind(counts_iso, counts_gene)
counts_all[, ct := factor(ct, levels = CT_KEEP)]

p_counts <- ggplot(counts_all, aes(x = ct, y = n, fill = ct)) +
  geom_col(width = 0.7) +
  geom_text(aes(label = scales::comma(n)), vjust = -0.4, size = 3.2) +
  facet_wrap(~ level, scales = "free_y") +
  scale_fill_brewer(palette = "Set2", guide = "none") +
  scale_y_continuous(labels = scales::comma, expand = expansion(mult = c(0, 0.12))) +
  labs(
    title    = "Cell-Type Marker Features Under DMSO",
    subtitle = sprintf(
      "One-vs-rest contrast | adj.P < %.2f & logFC > %.0f",
      PADJ_CUT, LOGFC_MARKER
    ),
    x = "Cell Type", y = "Number of Markers"
  ) +
  theme_minimal(base_size = 11) +
  theme(panel.grid.major.x = element_blank())

print(p_counts)
ggsave(file.path(FIG_DIR, "de_by_ct_dmso_counts.png"),
       p_counts, width = 10, height = 5, dpi = 300)
```

### Limma Vs Mashr CT-Marker Counts

```{r counts_barplot_mash, fig.width=10, fig.height=5}
counts_compare <- rbind(
  ovr_iso_dt[,       .(level = "Isoform (LR)", method = "limma", n = sum(is_marker)),       by = ct],
  ovr_iso_mash_dt[,  .(level = "Isoform (LR)", method = "mashr", n = sum(is_marker_mash)),  by = ct],
  ovr_gene_dt[,      .(level = "Gene (SR)",    method = "limma", n = sum(is_marker)),       by = ct],
  ovr_gene_mash_dt[, .(level = "Gene (SR)",    method = "mashr", n = sum(is_marker_mash)),  by = ct]
)
counts_compare[, ct := factor(ct, levels = CT_KEEP)]
counts_compare[, method := factor(method, levels = c("limma", "mashr"))]

p_counts_cmp <- ggplot(counts_compare, aes(x = ct, y = n, fill = method)) +
  geom_col(position = position_dodge(width = 0.8), width = 0.75) +
  geom_text(aes(label = scales::comma(n)),
            position = position_dodge(width = 0.8),
            vjust = -0.4, size = 2.8) +
  facet_wrap(~ level, scales = "free_y") +
  scale_fill_manual(values = c("limma" = "#9ECAE1", "mashr" = "#08519C"),
                    name = NULL) +
  scale_y_continuous(labels = scales::comma, expand = expansion(mult = c(0, 0.15))) +
  labs(
    title    = "CT-Marker Counts: Unshrunken Vs Mashr-Shrunken",
    subtitle = sprintf("limma: adj.P < %.2f & logFC > %.0f | mashr: lfsr < %.2f & pm > %.0f",
                       PADJ_CUT, LOGFC_MARKER, LFSR_CUT, LOGFC_MARKER),
    x = "Cell Type", y = "Number of CT-Enriched Markers"
  ) +
  theme_minimal(base_size = 11) +
  theme(panel.grid.major.x = element_blank(),
        legend.position = "top")

print(p_counts_cmp)
ggsave(file.path(FIG_DIR, paste0("de_by_ct_dmso_counts_limma_vs_mashr_", DATE_STAMP, ".png")),
       p_counts_cmp, width = 10, height = 5, dpi = 300, bg = "white")
```

## Top Markers Per CT

Top 10 isoform markers per CT by `logFC`, with gene symbol annotation.

```{r top_markers}
top_iso_markers <- ovr_iso_dt[is_marker == TRUE,
  .SD[order(-logFC)][1:10], by = ct,
  .SDcols = c("feature_id", "gene_sym", "gene_ensg", "logFC", "AveExpr", "adj.P.Val")
]
setnames(top_iso_markers, "gene_sym", "gene")
print(top_iso_markers)

fwrite(top_iso_markers,
       file.path(OUT_DIR, "top10_iso_markers_per_ct.csv"))

top_gene_markers <- ovr_gene_dt[is_marker == TRUE,
  .SD[order(-logFC)][1:10], by = ct,
  .SDcols = c("feature_id", "logFC", "AveExpr", "adj.P.Val")
]
setnames(top_gene_markers, "feature_id", "gene")
print(top_gene_markers)

fwrite(top_gene_markers,
       file.path(OUT_DIR, "top10_gene_markers_per_ct.csv"))
```

### Top Markers Per CT (Mashr)

```{r top_markers_mash}
top_iso_markers_mash <- ovr_iso_mash_dt[is_marker_mash == TRUE,
  .SD[order(-pm)][1:10], by = ct,
  .SDcols = c("feature_id", "gene_sym", "gene_ensg", "pm", "lfsr")
]
setnames(top_iso_markers_mash, "gene_sym", "gene")
print(top_iso_markers_mash)
fwrite(top_iso_markers_mash,
       file.path(MASH_DIR, paste0("top10_iso_markers_per_ct_mashr_", DATE_STAMP, ".csv")))

top_gene_markers_mash <- ovr_gene_mash_dt[is_marker_mash == TRUE,
  .SD[order(-pm)][1:10], by = ct,
  .SDcols = c("feature_id", "pm", "lfsr")
]
setnames(top_gene_markers_mash, "feature_id", "gene")
print(top_gene_markers_mash)
fwrite(top_gene_markers_mash,
       file.path(MASH_DIR, paste0("top10_gene_markers_per_ct_mashr_", DATE_STAMP, ".csv")))
```

## Heatmap of CT-Marker Expression

Take the union of the top 25 markers per CT (isoform level) and plot mean log-CPM by CT × donor.

```{r marker_heatmap, fig.width=8, fig.height=10}
top25_per_ct <- ovr_iso_dt[is_marker == TRUE,
  .SD[order(-logFC)][1:25], by = ct]
marker_set <- unique(top25_per_ct$feature_id)

# Mean log-CPM per CT × sample
log_cpm <- cpm(y_iso_dmso, log = TRUE, prior.count = 1)
log_cpm_mark <- log_cpm[rownames(log_cpm) %in% marker_set, , drop = FALSE]

# Z-score per row
z <- t(scale(t(log_cpm_mark)))

# Annotation
ann_col <- data.frame(
  CellType = y_iso_dmso$samples$ct,
  Donor    = y_iso_dmso$samples$donor_factor,
  row.names = colnames(z)
)
ann_colors <- list(
  CellType = setNames(brewer.pal(4, "Set2"), CT_KEEP)
)

# Order columns by CT then donor
ord <- order(ann_col$CellType, ann_col$Donor)
z   <- z[, ord]
ann_col <- ann_col[ord, , drop = FALSE]

# Order rows by CT-marker block
row_ann <- top25_per_ct[!duplicated(feature_id),
  .(feature_id, marker_ct = ct)]
row_order <- match(rownames(z), row_ann$feature_id)
z <- z[order(row_ann$marker_ct[row_order]), , drop = FALSE]

pheatmap(
  z,
  cluster_rows = FALSE, cluster_cols = FALSE,
  annotation_col = ann_col, annotation_colors = ann_colors,
  show_rownames = FALSE, show_colnames = FALSE,
  color = colorRampPalette(rev(brewer.pal(11, "RdBu")))(50),
  breaks = seq(-3, 3, length.out = 51),
  main = "Top 25 CT-Marker Isoforms Per CT (Z-scored log-CPM, DMSO)",
  filename = file.path(FIG_DIR, "de_by_ct_dmso_heatmap.png"),
  width = 8, height = 10
)
```

# GSEA — topGO Overrepresentation On SR Gene Markers

We test whether CT-marker genes (one-vs-rest, SR DGE level, adj.P < 0.05 & logFC > 1) are enriched for GO terms relative to the universe of all SR genes tested. Run separately per CT and per ontology (BP, MF, CC). Classic Fisher test; FDR-adjusted across terms within each (CT, ontology). topGO requires Ensembl gene IDs without version suffix or HGNC symbols — we map via the SR DGEList annotation.

```{r topgo_libs}
suppressPackageStartupMessages({
  library(topGO)
  library(org.Hs.eg.db)
})
```

```{r topgo_prep}
# SR gene IDs in this project: rownames(y_gene_dmso) are ensembl_gene_id_version
# topGO works best with bare ENSG IDs or symbols; strip version suffix.
sr_universe_ensg <- sub("\\.\\d+$", "", rownames(y_gene_dmso))
names(sr_universe_ensg) <- rownames(y_gene_dmso)

# Map per-CT marker gene IDs (versioned) → bare ENSG
get_marker_ensg <- function(ct_x) {
  vers <- ovr_gene_dt[ct == ct_x & is_marker == TRUE, feature_id]
  unname(sub("\\.\\d+$", "", vers))
}

# Universe: every gene that survived filtering and was tested
all_ensg_bare <- unique(sr_universe_ensg)
cat("topGO universe size:", length(all_ensg_bare), "\n")
for (ct_x in CT_KEEP) {
  cat(sprintf("  %s markers: %d\n", ct_x, length(get_marker_ensg(ct_x))))
}
```

```{r topgo_run}
run_topgo_one <- function(ct_x, ontology = "BP") {
  marker_ensg <- get_marker_ensg(ct_x)
  if (length(marker_ensg) == 0) return(NULL)

  gene_list <- factor(as.integer(all_ensg_bare %in% marker_ensg))
  names(gene_list) <- all_ensg_bare

  GOdata <- new(
    "topGOdata",
    ontology     = ontology,
    allGenes     = gene_list,
    annot        = annFUN.org,
    mapping      = "org.Hs.eg.db",
    ID           = "ensembl",
    nodeSize     = 10
  )

  result_fisher <- runTest(GOdata, algorithm = "classic", statistic = "fisher")

  # Pull all tested terms so FDR is meaningful and N is honest
  all_terms <- usedGO(GOdata)
  tab <- GenTable(GOdata, classicFisher = result_fisher,
                  orderBy = "classicFisher", topNodes = length(all_terms))

  # GenTable returns Fisher p as character (sometimes "<1e-30") — coerce safely
  tab$pvalue <- as.numeric(gsub("<", "", tab$classicFisher))
  tab$FDR    <- p.adjust(tab$pvalue, method = "BH")
  tab$ct     <- ct_x
  tab$ontology <- ontology

  setDT(tab)
  tab[order(FDR)]
}

topgo_all <- list()
for (ct_x in CT_KEEP) {
  for (ont in c("BP", "MF", "CC")) {
    cat(sprintf("topGO %s | %s ...\n", ct_x, ont))
    topgo_all[[paste(ct_x, ont, sep = "_")]] <-
      run_topgo_one(ct_x, ontology = ont)
  }
}

topgo_dt <- rbindlist(topgo_all, fill = TRUE)

cat("\nAll tested terms per CT × ontology:\n")
print(topgo_dt[, .N, by = .(ct, ontology)])

cat("\nSignificant terms (FDR < 0.05) per CT × ontology:\n")
print(topgo_dt[FDR < 0.05, .N, by = .(ct, ontology)])
```

## Top Terms Per CT (BP)

```{r topgo_top_bp}
top_bp <- topgo_dt[ontology == "BP" & FDR < 0.05][
  order(ct, FDR)][, head(.SD, 25), by = ct]
print(top_bp[, .(ct, GO.ID, Term, Significant, Annotated, pvalue, FDR)])
```

## Save topGO Results

```{r topgo_save}
fwrite(topgo_dt, file.path(OUT_DIR, "topgo_results_all.csv"))
fwrite(topgo_dt[FDR < 0.05],
       file.path(OUT_DIR, "topgo_results_FDR05.csv"))
cat("topGO results saved to:", OUT_DIR, "\n")
```
# Session Info

```{r session_info}
sessionInfo()
```