---
title: "Isoform-Level DIE: Pattern Interpretation"
author: "Leshem, Yul"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: lumen
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
    code_folding: hide
---

Code location: `/udd/reyle/nmd_lungcells_2026/code/final/interpret_isoform_patterns_mashr_2026.3.10.Rmd`

```{r setup, include=FALSE}
options(bitmapType = "cairo")
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, dev = "png")

```

# Analysis Overview

This report interprets isoform-proportion patterns to characterize NMD
inhibition effects on transcript diversity. Four questions are addressed:

- **Q1.** How does the number of detected isoforms per gene change under Smg1i?
- **Q2.** How prevalent are dominant isoforms (>50% of gene CPM) under each condition?
- **Q3.** What are the baseline (DMSO) proportions of NMD susceptible transcripts?
- **Q6.** Which pathways are enriched among genes with high NMD output?

NMD susceptible transcripts are defined as posterior mean logFC > 0 AND lfsr <
0.05. Proportion calculations use raw normalized CPM from the DGEList.

```{r config, include=FALSE}
DATE_STAMP <- "2026.3.10"

INPUT_DGELIST          <- "/udd/reyle/nmd_lungcells_2026/long_read/dge_isoform_longread_2026.3.3.rds"
INPUT_DGELIST_FILTERED <- "/udd/reyle/nmd_lungcells_2026/long_read/dge_isoform_longread_filtered_2026.3.3.rds"

INPUT_DGE_DIR    <- "/udd/reyle/nmd_lungcells_2026/long_read/mashr"
DGE_FILE_PATTERN <- paste0("^nmd_mashr_die_.*_", gsub("\\.", "\\\\.", DATE_STAMP), "\\.csv$")

OUTPUT_DIR  <- file.path("/udd/reyle/nmd_lungcells_2026/tmp/mashr_die_interpret")
OUT_DIR_FIG <- "/udd/reyle/nmd_lungcells_2026/final_figures/lr_iso"
dir.create(OUTPUT_DIR,  recursive = TRUE, showWarnings = FALSE)
dir.create(OUT_DIR_FIG, recursive = TRUE, showWarnings = FALSE)

OUTPUT_Q1      <- file.path(OUTPUT_DIR, paste0("mashr_isoform_q1_isoforms_per_gene_",    DATE_STAMP, ".tsv"))
OUTPUT_Q2      <- file.path(OUTPUT_DIR, paste0("mashr_isoform_q2_dominant_isoforms_",    DATE_STAMP, ".tsv"))
OUTPUT_Q3      <- file.path(OUTPUT_DIR, paste0("mashr_isoform_q3_upregulated_baseline_", DATE_STAMP, ".tsv"))
OUTPUT_Q6_GSEA <- file.path(OUTPUT_DIR, paste0("mashr_isoform_q6_gsea_nmd_output_",      DATE_STAMP, ".tsv"))

CPM_THRESHOLD          <- 0.25
SIGNIFICANCE_THRESHOLD <- 0.05
GSEA_MIN_SIZE          <- 15
GSEA_MAX_SIZE          <- 500
GSEA_SEED              <- 42  # re-seeded before fgsea() for reproducible nulls

# 4 cell types only. Pipeline writes nmd_mashr_die_at2/lae/fb/mv_<date>.csv, so
# the keys here are the lower-case file codes, mapped to display names.
CELLTYPE_MAP     <- c("at2" = "AT2", "lae" = "LAE", "fb" = "FB", "mv" = "MV")
CT_DISPLAY_ORDER <- c("AT2", "LAE", "FB", "MV")

stopifnot("DGEList (unfiltered) not found" = file.exists(INPUT_DGELIST))
stopifnot("DGEList (filtered) not found"   = file.exists(INPUT_DGELIST_FILTERED))
stopifnot("DGE directory not found"        = dir.exists(INPUT_DGE_DIR))
```

<style>
body { font-size: 16px; }
p, li { font-size: 16px; line-height: 1.6; }
</style>

# Libraries

```{r load-libs}
library(tidyverse)
library(data.table)
library(DT)
library(edgeR)

# Bioconductor namespace masking guard
rename <- dplyr::rename
select <- dplyr::select
filter <- dplyr::filter
mutate <- dplyr::mutate
first  <- dplyr::first
count  <- dplyr::count

# Fix RNG once (fgsea draws permutations; matches the short-read pipeline).
set.seed(GSEA_SEED)
```

# Load Data

## DGEList Objects And Normalized CPM

```{r load-dgelist}
dge_isoform  <- readRDS(INPUT_DGELIST)
dge_filtered <- readRDS(INPUT_DGELIST_FILTERED)

cat("Unfiltered:", nrow(dge_isoform),  "transcripts x", ncol(dge_isoform),  "samples\n")
cat("Filtered:  ", nrow(dge_filtered), "transcripts x", ncol(dge_filtered), "samples\n")

# Repair / harmonize cell-type column from sample names. Sample-name suffixes
# still carry the original DD/AT codes (kept so they match the count columns);
# parse them first, then relabel to display names AFTER matching:
#   DD -> LAE, AT -> AT2 (DD_ALI / DO_ALI untouched, dropped downstream).
fix_ct <- function(samples_df) {
  samples_df$ct <- sub("^.*_(DD_ALI|DD|AT|DO|FB|MV)$", "\\1", samples_df$sample_name)
  samples_df$ct <- ifelse(samples_df$ct == "DO", "DO_ALI", samples_df$ct)
  samples_df$ct[samples_df$ct == "DD"] <- "LAE"
  samples_df$ct[samples_df$ct == "AT"] <- "AT2"
  samples_df
}
dge_isoform$samples  <- fix_ct(dge_isoform$samples)
dge_filtered$samples <- fix_ct(dge_filtered$samples)

cat("Cell types in unfiltered:", paste(sort(unique(dge_isoform$samples$ct)),  collapse = ", "), "\n")
cat("Cell types in filtered:  ", paste(sort(unique(dge_filtered$samples$ct)), collapse = ", "), "\n")

cpm_normalized <- cpm(dge_filtered, normalized = TRUE, log = FALSE)
```

## Calculate Isoform Proportions

```{r calculate-proportions}
sample_meta            <- dge_filtered$samples
sample_meta$cpm_col_id <- rownames(sample_meta)

cpm_long <- as.data.frame(cpm_normalized) %>%
  mutate(transcript_id = dge_filtered$genes$txid) %>%
  pivot_longer(cols = -transcript_id, names_to = "cpm_col_id", values_to = "cpm") %>%
  left_join(sample_meta %>% select(cpm_col_id, sample_name, treatment, ct),
            by = "cpm_col_id") %>%
  rename(cell_type = ct)

stopifnot("Sample metadata join failed" = sum(is.na(cpm_long$treatment)) == 0)

cpm_long <- cpm_long %>%
  left_join(dge_filtered$genes %>% select(txid, gene_id),
            by = c("transcript_id" = "txid")) %>%
  rename(gene_group_id = gene_id)

# FILTER: restrict to the 4-cell-type universe (drops ALI) at the long-frame boundary.
n_before <- nrow(cpm_long)
cpm_long <- cpm_long %>% filter(cell_type %in% CT_DISPLAY_ORDER)
cat(sprintf("FILTER: kept rows in {%s}.\n  Rows before: %d / Rows now: %d.\n",
            paste(CT_DISPLAY_ORDER, collapse = ", "), n_before, nrow(cpm_long)))

all_proportions <- cpm_long %>%
  group_by(cell_type, treatment, cpm_col_id, gene_group_id) %>%
  mutate(
    gene_total_cpm     = sum(cpm),
    isoform_proportion = cpm / gene_total_cpm,
    isoform_proportion = ifelse(is.nan(isoform_proportion), 0, isoform_proportion),
    n_isoforms         = n_distinct(transcript_id)
  ) %>%
  ungroup() %>%
  rename(normalized_cpm = cpm)

cat("Proportion data frame:", nrow(all_proportions), "rows\n")
cat("Cell types:", paste(unique(all_proportions$cell_type), collapse = ", "), "\n")
cat("Treatments:", paste(unique(all_proportions$treatment), collapse = ", "), "\n")

```

## Apply Abundance Filter

```{r apply-filter}
filtered_dict <- all_proportions %>%
  group_by(cell_type, transcript_id) %>%
  summarize(max_cpm = max(normalized_cpm), .groups = "drop") %>%
  filter(max_cpm >= CPM_THRESHOLD)

n_iso_before <- n_distinct(all_proportions$transcript_id)
all_proportions <- all_proportions %>%
  inner_join(filtered_dict %>% select(cell_type, transcript_id),
             by = c("cell_type", "transcript_id"))

cat("FILTER: abundance >= ", CPM_THRESHOLD, " CPM in either condition (per cell type).\n", sep = "")
cat("  Unique isoforms before:", n_iso_before,
    "/ after:", n_distinct(all_proportions$transcript_id), "\n")
cat("  Rows:", nrow(all_proportions), "\n")
cat("  Unique genes:   ", n_distinct(all_proportions$gene_group_id), "\n")
```

## Load DIE Results

```{r load-die}
dge_files <- list.files(INPUT_DGE_DIR, pattern = DGE_FILE_PATTERN, full.names = TRUE)
cat("Loading", length(dge_files), "DIE files:\n"); print(basename(dge_files))

# Cell type lives at parts[4]: nmd_mashr_die_<ct>_<date>.csv
all_dge <- map_df(dge_files, ~{
  df    <- read_csv(.x, show_col_types = FALSE)
  parts <- str_split(basename(.x), "_")[[1]]
  df$cell_type_file <- parts[4]
  df
})

all_dge$cell_type <- CELLTYPE_MAP[all_dge$cell_type_file]

# FILTER: drop rows with an unmapped cell-type code (stale at/dd keys would
# map every row to NA and empty the report).
n_before <- nrow(all_dge)
all_dge  <- all_dge %>% filter(!is.na(cell_type))
cat(sprintf("FILTER: dropped %d row(s) with an unmapped cell-type code.\n  Rows before: %d / Rows now: %d.\n",
            n_before - nrow(all_dge), n_before, nrow(all_dge)))
stopifnot(nrow(all_dge) > 0)

cat("Columns present in DIE files:\n  ", paste(colnames(all_dge), collapse = ", "), "\n")

if ("txid" %in% colnames(all_dge)) {
  colnames(all_dge)[colnames(all_dge) == "txid"] <- "transcript_id"
}

# Resolve the gene-id column to a canonical `gene_id` (fail loudly if absent).
gene_id_candidates <- c("gene_id", "gene_group_id", "gene", "ensembl_gene_id",
                        "geneid", "ensembl_gene_id_version")
gid <- intersect(gene_id_candidates, colnames(all_dge))
if (length(gid) == 0) {
  stop("No gene-id column found in DIE files. Columns are: ",
       paste(colnames(all_dge), collapse = ", "))
}
if (!"gene_id" %in% colnames(all_dge)) {
  all_dge$gene_id <- all_dge[[gid[1]]]
  cat("Using '", gid[1], "' as gene_id.\n", sep = "")
}

all_dge$gene_symbol <- if ("hgnc_symbol" %in% colnames(all_dge)) {
  ifelse(!is.na(all_dge$hgnc_symbol) & all_dge$hgnc_symbol != "",
         all_dge$hgnc_symbol, all_dge$gene_id)
} else {
  all_dge$gene_id
}

cat("DIE results:", nrow(all_dge), "rows,",
    length(unique(all_dge$cell_type)), "cell types\n")

```

## Build Gene Symbol Map

```{r gene-symbol-map}
gene_symbol_map <- all_dge %>%
  filter(!is.na(hgnc_symbol), hgnc_symbol != "", !grepl("^ENSG", hgnc_symbol)) %>%
  mutate(gene_id_no_version = str_replace(gene_id, "\\.\\d+$", "")) %>%
  select(gene_id_no_version, hgnc_symbol) %>%
  distinct()

cat("Gene symbol map:", nrow(gene_symbol_map), "genes\n")
```

# Q1: Isoforms Per Gene

```{r q1-isoforms-per-gene, fig.width=8, fig.height=5}
isoforms_per_gene <- all_proportions %>%
  filter(normalized_cpm > 0) %>%
  group_by(cell_type, treatment, gene_group_id) %>%
  summarize(n_isoforms_detected = n_distinct(transcript_id), .groups = "drop")

isoform_comparison <- isoforms_per_gene %>%
  pivot_wider(names_from = treatment, values_from = n_isoforms_detected, values_fill = 0) %>%
  mutate(
    delta_isoforms = Smg1i - DMSO,
    more_tx_smg1i  = delta_isoforms > 0,
    fewer_tx_smg1i = delta_isoforms < 0,
    same_tx_count  = delta_isoforms == 0
  )

q1_summary <- isoform_comparison %>%
  group_by(cell_type) %>%
  summarize(
    n_genes                  = n(),
    mean_dmso                = mean(DMSO),
    mean_smg1i               = mean(Smg1i),
    mean_delta               = mean(delta_isoforms),
    genes_with_more_tx_smg1i = sum(more_tx_smg1i),
    pct_more_tx_smg1i        = 100 * sum(more_tx_smg1i) / n(),
    .groups = "drop"
  )

datatable(q1_summary,
          caption = "Average Isoforms Per Gene: DMSO Vs Smg1i") %>%
  formatRound(columns = c("mean_dmso", "mean_smg1i", "mean_delta",
                          "pct_more_tx_smg1i"), digits = 2)

category_summary <- isoform_comparison %>%
  mutate(category = case_when(
    more_tx_smg1i  ~ "More In Smg1i",
    fewer_tx_smg1i ~ "Fewer In Smg1i",
    same_tx_count  ~ "Unchanged"
  )) %>%
  count(cell_type, category) %>%
  filter(category != "Unchanged")

category_summary$cell_type <- factor(category_summary$cell_type, levels = CT_DISPLAY_ORDER)

p1 <- ggplot(category_summary, aes(x = cell_type, y = n, fill = category)) +
  geom_col(position = "dodge") +
  scale_fill_manual(values = c("More In Smg1i" = "steelblue", "Fewer In Smg1i" = "coral")) +
  labs(title = "Genes With Changed Transcript Counts (Smg1i Vs DMSO)",
       x = "Cell Type", y = "Number Of Genes", fill = NULL) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "top",
        plot.title      = element_text(face = "bold"))

ggsave(file.path(OUT_DIR_FIG, paste0("q1_isoform_count_change_", DATE_STAMP, ".png")),
       p1, width = 8, height = 5, dpi = 300, bg = "white")

print(p1)
write_tsv(isoform_comparison, OUTPUT_Q1)

```

# Q2: Dominant Isoforms

A gene is counted as having a "dominant isoform" only if a single transcript
exceeds 50% of total gene CPM in **every** sample of that cell type +
treatment combination. Genes with fewer than 3 detected isoforms are excluded.

```{r q2-dominant-isoforms, fig.width=8, fig.height=5}
isoform_props <- all_proportions %>%
  filter(normalized_cpm > 0) %>%
  select(cell_type, treatment, cpm_col_id, gene_group_id, transcript_id, isoform_proportion)

genes_with_3plus_isoforms <- isoform_props %>%
  group_by(cell_type, treatment, gene_group_id) %>%
  summarize(n_isoforms = n_distinct(transcript_id), .groups = "drop") %>%
  filter(n_isoforms >= 3)

isoform_props <- isoform_props %>%
  semi_join(genes_with_3plus_isoforms, by = c("cell_type", "treatment", "gene_group_id"))

dominant_genes <- isoform_props %>%
  group_by(cell_type, treatment, gene_group_id, transcript_id) %>%
  summarize(n_samples            = n(),
            n_dominant_samples   = sum(isoform_proportion > 0.5),
            mean_proportion      = mean(isoform_proportion),
            .groups = "drop") %>%
  mutate(is_dominant_all_samples = n_dominant_samples == n_samples & n_samples > 0) %>%
  group_by(cell_type, treatment, gene_group_id) %>%
  summarize(n_isoforms          = n(),
            has_dominant_isoform = any(is_dominant_all_samples),
            max_mean_proportion  = max(mean_proportion),
            .groups = "drop")

q2_summary <- dominant_genes %>%
  group_by(cell_type, treatment) %>%
  summarize(n_genes              = n(),
            n_with_dominant      = sum(has_dominant_isoform),
            pct_with_dominant    = 100 * n_with_dominant / n_genes,
            mean_top_proportion  = mean(max_mean_proportion),
            .groups = "drop")

q2_summary$cell_type <- factor(q2_summary$cell_type, levels = CT_DISPLAY_ORDER)

p2 <- ggplot(q2_summary, aes(x = cell_type, y = pct_with_dominant, fill = treatment)) +
  geom_col(position = "dodge") +
  scale_fill_manual(values = c("DMSO" = "gray70", "Smg1i" = "steelblue")) +
  labs(title    = "Genes With Dominant Isoform (>50% In All Samples)",
       subtitle = "Fraction of >=3-isoform genes with one transcript dominant in all samples",
       x = "Cell Type", y = "% Of Genes", fill = "Treatment") +
  theme_minimal(base_size = 13) +
  theme(legend.position = "top",
        plot.title      = element_text(face = "bold"),
        plot.subtitle   = element_text(color = "grey40"))

ggsave(file.path(OUT_DIR_FIG, paste0("q2_dominant_isoforms_", DATE_STAMP, ".png")),
       p2, width = 8, height = 5, dpi = 300, bg = "white")

print(p2)
write_tsv(q2_summary, OUTPUT_Q2)
```

# Q3: Baseline Proportions Of NMD Susceptible Transcripts

## Identify NMD Susceptible Transcripts

```{r q3-identify}
nmd_responsive <- all_dge %>%
  filter(logFC > 0, adj.P.Val < SIGNIFICANCE_THRESHOLD) %>%
  select(cell_type, transcript_id, logFC, adj.P.Val)

cat("NMD susceptible transcripts (lfsr < 0.05) by cell type:\n")
print(nmd_responsive %>% group_by(cell_type) %>% summarize(n = n(), .groups = "drop"))
cat("\nUnique transcripts:", n_distinct(nmd_responsive$transcript_id), "\n")

```

## Baseline (DMSO) Proportions

```{r q3-baseline}
baseline_proportions <- all_proportions %>%
  filter(treatment == "DMSO") %>%
  inner_join(nmd_responsive, by = c("cell_type", "transcript_id")) %>%
  select(cell_type, cpm_col_id, gene_group_id, transcript_id,
         normalized_cpm, isoform_proportion, n_isoforms, logFC, adj.P.Val)

coverage_summary <- nmd_responsive %>%
  select(cell_type, transcript_id) %>% distinct() %>%
  left_join(
    baseline_proportions %>% select(cell_type, transcript_id) %>% distinct() %>%
      mutate(has_baseline = TRUE),
    by = c("cell_type", "transcript_id")
  ) %>%
  group_by(cell_type) %>%
  summarize(n_nmd_susceptible = n(),
            n_with_baseline   = sum(has_baseline, na.rm = TRUE),
            pct_coverage      = 100 * n_with_baseline / n(),
            .groups = "drop")

cat("Baseline coverage:\n")
print(coverage_summary)

transcript_summary <- baseline_proportions %>%
  group_by(cell_type, transcript_id, gene_group_id, logFC, adj.P.Val) %>%
  summarize(mean_cpm        = mean(normalized_cpm),
            mean_proportion = mean(isoform_proportion),
            n_samples       = n(),
            n_zero_samples  = sum(normalized_cpm == 0),
            .groups = "drop") %>%
  mutate(is_zero_all  = n_zero_samples == n_samples,
         is_low       = mean_cpm > 0 & mean_cpm < 0.1,
         is_expressed = mean_cpm >= 0.1)

q3_summary <- transcript_summary %>%
  group_by(cell_type) %>%
  summarize(
    n_genes              = n_distinct(gene_group_id),
    n_transcripts        = n(),
    n_zero_dmso          = sum(is_zero_all),
    n_low_dmso           = sum(is_low),
    n_expressed_dmso     = sum(is_expressed),
    pct_zero             = 100 * n_zero_dmso / n(),
    pct_low              = 100 * n_low_dmso  / n(),
    mean_baseline_cpm    = mean(mean_cpm[mean_cpm > 0]),
    median_baseline_cpm  = median(mean_cpm[mean_cpm > 0]),
    mean_baseline_prop   = mean(mean_proportion),
    median_baseline_prop = median(mean_proportion),
    pct_minor_isoforms   = 100 * sum(mean_proportion < 0.5) / n(),
    pct_very_minor       = 100 * sum(mean_proportion < 0.1) / n(),
    mean_logFC           = mean(logFC),
    .groups = "drop"
  )

knitr::kable(q3_summary, digits = 2,
             caption = "NMD Susceptible Transcripts: Baseline DMSO Expression")

write_tsv(baseline_proportions, OUTPUT_Q3)
```

## All Isoforms — Proportion Vs Expression In DMSO

```{r q3-all-iso, fig.width=8, fig.height=5}
all_dmso_proportions <- all_proportions %>%
  filter(treatment == "DMSO") %>%
  group_by(transcript_id, gene_group_id) %>%
  summarize(mean_cpm        = mean(normalized_cpm),
            mean_proportion = mean(isoform_proportion),
            .groups = "drop") %>%
  filter(mean_cpm > 0)

p_all <- ggplot(all_dmso_proportions, aes(x = mean_cpm, y = mean_proportion)) +
  geom_point(alpha = 0.3, size = 0.5, color = "gray40") +
  geom_hline(yintercept = 0.5, linetype = "dashed", color = "red") +
  geom_hline(yintercept = 0.1, linetype = "dashed", color = "orange") +
  scale_x_log10() +
  labs(title    = "All Isoforms — Proportion Vs Expression In DMSO",
       subtitle = "Lines: 50% (red), 10% (orange)",
       x = "Mean Normalized CPM In DMSO (log10)",
       y = "Mean Isoform Proportion In DMSO") +
  theme_minimal(base_size = 13) +
  theme(plot.title    = element_text(face = "bold"),
        plot.subtitle = element_text(color = "grey40"))

ggsave(file.path(OUT_DIR_FIG, paste0("q3_all_iso_dmso_", DATE_STAMP, ".png")),
       p_all, width = 8, height = 5, dpi = 300, bg = "white")

print(p_all)

```

## NMD Susceptible — Proportion Vs Expression In DMSO

```{r q3-nmd-dmso, fig.width=8, fig.height=5}
nmd_dmso_proportions <- baseline_proportions %>%
  group_by(transcript_id, gene_group_id) %>%
  summarize(mean_cpm        = mean(normalized_cpm),
            mean_proportion = mean(isoform_proportion),
            .groups = "drop") %>%
  filter(mean_cpm > 0)

p_nmd <- ggplot(nmd_dmso_proportions, aes(x = mean_cpm, y = mean_proportion)) +
  geom_point(alpha = 0.4, size = 0.8, color = "gray70") +
  geom_hline(yintercept = 0.5, linetype = "dashed", color = "red") +
  geom_hline(yintercept = 0.1, linetype = "dashed", color = "orange") +
  scale_x_log10() +
  labs(title    = "NMD Susceptible Isoforms — Proportion Vs Expression In DMSO",
       subtitle = "Lines: 50% (red), 10% (orange)",
       x = "Mean Normalized CPM In DMSO (log10)",
       y = "Mean Isoform Proportion In DMSO") +
  theme_minimal(base_size = 13) +
  theme(plot.title    = element_text(face = "bold"),
        plot.subtitle = element_text(color = "grey40"))

ggsave(file.path(OUT_DIR_FIG, paste0("q3_nmd_dmso_", DATE_STAMP, ".png")),
       p_nmd, width = 8, height = 5, dpi = 300, bg = "white")

print(p_nmd)
```

## NMD Susceptible — DMSO Vs Smg1i

```{r q3-dmso-vs-smg1i, fig.width=8, fig.height=5}
nmd_dmso_data  <- nmd_dmso_proportions %>% mutate(treatment = "DMSO")

nmd_smg1i_data <- all_proportions %>%
  filter(treatment == "Smg1i") %>%
  inner_join(nmd_responsive, by = c("cell_type", "transcript_id")) %>%
  group_by(transcript_id, gene_group_id) %>%
  summarize(mean_cpm        = mean(normalized_cpm),
            mean_proportion = mean(isoform_proportion),
            .groups = "drop") %>%
  filter(mean_cpm > 0) %>%
  mutate(treatment = "Smg1i")

nmd_combined_data <- bind_rows(nmd_dmso_data, nmd_smg1i_data)

p_compare <- ggplot(nmd_combined_data,
                    aes(x = mean_cpm, y = mean_proportion, color = treatment)) +
  geom_point(alpha = 0.4, size = 0.8) +
  geom_hline(yintercept = 0.5, linetype = "dashed", color = "black") +
  geom_hline(yintercept = 0.1, linetype = "dashed", color = "gray50") +
  scale_x_log10() +
  scale_color_manual(values = c("DMSO" = "gray70", "Smg1i" = "steelblue")) +
  labs(title    = "NMD Susceptible Isoforms — DMSO Vs Smg1i",
       subtitle = "Lines: 50% (black), 10% (gray)",
       x = "Mean Normalized CPM (log10)",
       y = "Isoform Proportion",
       color = "Treatment") +
  theme_minimal(base_size = 13) +
  theme(legend.position = "top",
        plot.title      = element_text(face = "bold"),
        plot.subtitle   = element_text(color = "grey40"))

ggsave(file.path(OUT_DIR_FIG, paste0("q3_nmd_dmso_vs_smg1i_", DATE_STAMP, ".png")),
       p_compare, width = 8, height = 5, dpi = 300, bg = "white")

print(p_compare)

```

# Q6: Pathway Enrichment Of NMD Output

Genes are ranked by `pct_to_nmd` — the fraction of each gene's transcriptional
output (mean Smg1i CPM summed across its isoforms) that goes to NMD susceptible
isoforms. Genes with no NMD susceptible isoforms get a score of 0. fgsea then
asks which pathways concentrate at the top of this ranking.

```{r q6-build-pct-to-nmd}
nmd_responsive_set <- nmd_responsive %>%
  select(cell_type, transcript_id) %>%
  distinct() %>%
  mutate(is_nmd_responsive = TRUE)

# Per-isoform mean Smg1i CPM, tagged with NMD susceptible status
smg1i_expression <- all_proportions %>%
  filter(treatment == "Smg1i") %>%
  group_by(cell_type, gene_group_id, transcript_id) %>%
  summarize(mean_cpm = mean(normalized_cpm), .groups = "drop") %>%
  left_join(nmd_responsive_set, by = c("cell_type", "transcript_id")) %>%
  mutate(is_nmd_responsive = replace_na(is_nmd_responsive, FALSE))

# Per-gene NMD output share
gene_nmd_output <- smg1i_expression %>%
  group_by(cell_type, gene_group_id) %>%
  summarize(total_output    = sum(mean_cpm),
            nmd_output      = sum(mean_cpm[is_nmd_responsive]),
            has_nmd_isoform = any(is_nmd_responsive),
            .groups = "drop") %>%
  filter(has_nmd_isoform, total_output > 0) %>%
  mutate(pct_to_nmd          = 100 * nmd_output / total_output,
         gene_id_no_version  = str_replace(gene_group_id, "\\.\\d+$", "")) %>%
  left_join(gene_symbol_map, by = "gene_id_no_version") %>%
  select(-gene_id_no_version)
```

```{r q6-setup}
if (!require("fgsea",   quietly = TRUE)) BiocManager::install("fgsea")
if (!require("msigdbr", quietly = TRUE)) install.packages("msigdbr")
library(fgsea); library(msigdbr); library(htmltools)

```

```{r q6-load-genesets}
msigdb_h        <- msigdbr(species = "Homo sapiens", collection = "H")
msigdb_bp       <- msigdbr(species = "Homo sapiens", collection = "C5", subcollection = "GO:BP")
msigdb_kegg     <- msigdbr(species = "Homo sapiens", collection = "C2", subcollection = "CP:KEGG_LEGACY")
msigdb_reactome <- msigdbr(species = "Homo sapiens", collection = "C2", subcollection = "CP:REACTOME")

all_genesets <- bind_rows(
  msigdb_h        %>% mutate(collection = "Hallmark"),
  msigdb_bp       %>% mutate(collection = "GO_BP"),
  msigdb_kegg     %>% mutate(collection = "KEGG"),
  msigdb_reactome %>% mutate(collection = "Reactome")
)

pathways <- all_genesets %>% split(x = .$gene_symbol, f = .$gs_name)

# FILTER: keep only gene sets within the GSEA size window.
n_pathways_before <- length(pathways)
pathways_filtered <- pathways[
  sapply(pathways, length) >= GSEA_MIN_SIZE &
  sapply(pathways, length) <= GSEA_MAX_SIZE
]
cat(sprintf("FILTER: gene sets restricted to [%d, %d] genes.\n  Sets before: %d / Sets now: %d.\n",
            GSEA_MIN_SIZE, GSEA_MAX_SIZE, n_pathways_before, length(pathways_filtered)))
```

```{r q6-prepare-rankings}
MIN_NMD_GENES             <- 100
ranked_lists_per_celltype <- list()

for (ct in unique(gene_nmd_output$cell_type)) {
  all_genes_ct <- smg1i_expression %>%
    filter(cell_type == ct) %>%
    group_by(gene_group_id) %>%
    summarize(total_output = sum(mean_cpm), .groups = "drop") %>%
    mutate(gene_id_no_version = str_replace(gene_group_id, "\\.\\d+$", "")) %>%
    left_join(gene_symbol_map, by = "gene_id_no_version")

  genes_with_nmd <- gene_nmd_output %>%
    filter(cell_type == ct) %>%
    select(hgnc_symbol, pct_to_nmd)

  ranked_data <- all_genes_ct %>%
    left_join(genes_with_nmd, by = "hgnc_symbol") %>%
    mutate(pct_to_nmd = replace_na(pct_to_nmd, 0)) %>%
    filter(!is.na(hgnc_symbol), hgnc_symbol != "", !str_detect(hgnc_symbol, "^ENSG")) %>%
    group_by(hgnc_symbol) %>%
    slice_max(order_by = pct_to_nmd, n = 1, with_ties = FALSE) %>%
    ungroup() %>%
    arrange(desc(pct_to_nmd))

  ranked_lists_per_celltype[[as.character(ct)]] <- setNames(
    ranked_data$pct_to_nmd, ranked_data$hgnc_symbol
  )
  cat(as.character(ct),
      "- Genes:", length(ranked_lists_per_celltype[[as.character(ct)]]),
      "| With NMD signal:", sum(ranked_data$pct_to_nmd > 0), "\n")
}

```

```{r q6-run-gsea, results='asis'}
gsea_results_all <- list()
table_num        <- 1

for (ct in names(ranked_lists_per_celltype)) {
  cat("###", ct, "\n\n")
  ranked_list         <- ranked_lists_per_celltype[[ct]]
  n_genes_with_signal <- sum(ranked_list > 0)

  if (n_genes_with_signal < MIN_NMD_GENES) {
    cat("**Skipping GSEA** — insufficient NMD signal (", n_genes_with_signal, "genes)\n\n")
    next
  }

  set.seed(GSEA_SEED)  # reproducible fgsea permutation null
  gsea_res <- fgsea(pathways = pathways_filtered, stats = ranked_list,
                    minSize  = GSEA_MIN_SIZE, maxSize = GSEA_MAX_SIZE)

  gsea_res$cell_type     <- ct
  gsea_results_all[[ct]] <- gsea_res
  cat("Pathways with FDR < 0.05:", sum(gsea_res$padj < 0.05, na.rm = TRUE), "\n\n")

  top_100 <- gsea_res %>%
    arrange(pval) %>%
    head(100) %>%
    mutate(pathway_short = str_trunc(pathway, 60),
           leadingEdge   = sapply(leadingEdge, function(x) paste(head(x, 5), collapse = ", "))) %>%
    select(pathway_short, pval, padj, ES, NES, size, leadingEdge)

  dt <- datatable(
    top_100,
    options = list(pageLength = 10, scrollX = TRUE),
    caption = paste0("Table ", table_num, ". GSEA Results — ", ct)
  ) %>%
    formatRound(columns = c("pval", "padj", "ES", "NES"), digits = 4)

  print(htmltools::tagList(dt))
  cat("\n\n")
  table_num <- table_num + 1
}

if (length(gsea_results_all) > 0) {
  gsea_combined <- bind_rows(gsea_results_all)
  gsea_combined$leadingEdge <- sapply(gsea_combined$leadingEdge, paste, collapse = ", ")
  write_tsv(gsea_combined, OUTPUT_Q6_GSEA)
} else {
  gsea_combined <- tibble()
}
```

# Key Findings

```{r findings}
cat("**Q1: Isoform Diversity**\n")
cat("- Mean change in detected isoforms per gene:",
    round(mean(q1_summary$mean_delta), 2), "\n")
cat("- Mean fraction of genes with more isoforms in Smg1i:",
    round(mean(q1_summary$pct_more_tx_smg1i), 1), "%\n\n")

cat("**Q2: Dominant Isoforms**\n")
cat("- Mean fraction of genes with a dominant isoform:",
    round(mean(q2_summary$pct_with_dominant), 1), "%\n\n")

cat("**Q3: NMD Target Characteristics**\n")
cat("- Median baseline proportion across cell types:",
    round(mean(q3_summary$median_baseline_prop), 3), "\n")
cat("- Mean fraction of NMD targets that are very minor isoforms (<10%):",
    round(mean(q3_summary$pct_very_minor), 1), "%\n\n")

if (exists("gsea_combined") && nrow(gsea_combined) > 0) {
  cat("**Q6: GSEA**\n")
  cat("- Total significant pathways across cell types (FDR < 0.05):",
      sum(gsea_combined$padj < 0.05, na.rm = TRUE), "\n")
}

```

# Session Info

```{r session-info}
sessionInfo()
```
