---
title: "SQANTI3 Isoform Characterization"
subtitle: "Long-read isoform landscape across cell types and treatments"
author: "Leshem, Yul"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: lumen
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 12, fig.height = 8, dev = "png")
options(bitmapType = "cairo")

```

# Overview

Code location: `/udd/reyle/nmd_lungcells_2026/code/final/Isoform_Landscape.Rmd`

This report characterizes the long-read isoform landscape across the four
primary lung cell types (AT2, LAE, FB, MV): structural category (SQANTI3),
cell-type expression sharing, length distributions, and pairwise expression
similarity within and across treatment conditions.

Note: the ALI cell types (DD_ALI, DO_ALI) are excluded; this is a descriptive
isoform landscape with no donor-paired modeling. The DGEList stores the
original ct codes `AT` (= AT2) and `DD` (= LAE); we relabel to the display
names right after loading and use AT2 / LAE throughout.

```{r libraries}
suppressPackageStartupMessages({
  library(data.table)
  library(dplyr)
  library(ggplot2)
  library(tidyr)
  library(reshape2)
  library(patchwork)
  library(scales)
  library(edgeR)
})

# Bioconductor namespace masking guard
select <- dplyr::select
filter <- dplyr::filter
rename <- dplyr::rename
count  <- dplyr::count
slice  <- dplyr::slice
first  <- dplyr::first
```

# Configuration

```{r config}
BASE          <- "/udd/reyle/nmd_lungcells_2026"
DGE_FILT_PATH <- file.path(BASE, "long_read/dge_isoform_longread_filtered_2026.3.3.rds")

# SQANTI3 classification (for structural_category)
SQ_CLASS_PATH <- "/proj/regeps/regep00/studies/ExternalCellLines/analyses/repjc/Randell_Lung_Cells_2025/code/sqanti3_by_chromosome/projects/nmd_lungcells/results/nmd_lungcells_classification.txt"

OUTPUT_DIR <- file.path(BASE, "long_read/isoform_characterization")
dir.create(OUTPUT_DIR, recursive = TRUE, showWarnings = FALSE)

# The four primary lung cell types (ALI types excluded), display naming.
# AT2 = old AT, LAE = old DD.
CT_KEEP   <- c("LAE", "AT2", "FB", "MV")
CAT_ORDER <- c("FSM", "ISM", "NIC", "NNC", "Other")

# Relabel helper: rename the DGEList's internal ct codes to display names.
# Safe on the loaded object — only a metadata column is renamed.
relabel_ct <- function(ct_vec) {
  ct_vec <- as.character(ct_vec)
  ct_vec[ct_vec == "AT"] <- "AT2"
  ct_vec[ct_vec == "DD"] <- "LAE"
  ct_vec
}

```

# Load Data

```{r load-dgelist}
dge_filt <- readRDS(DGE_FILT_PATH)
cat("Filtered DGEList:", nrow(dge_filt), "isoforms x", ncol(dge_filt), "samples\n")

# Relabel ct to display names AFTER loading (AT -> AT2, DD -> LAE)
dge_filt$samples$ct <- relabel_ct(dge_filt$samples$ct)

# Keep the 4 primary cell types; restrict to DMSO only for the baseline isoform landscape.
# (ALI types are dropped here via CT_KEEP, so no explicit 029T exclusion is needed.)
ct_char   <- as.character(dge_filt$samples$ct)
trt_char  <- as.character(dge_filt$samples$treatment)
keep_samp <- ct_char %in% CT_KEEP & trt_char == "DMSO"
cat(sprintf("FILTER: 4 primary CTs {%s} + DMSO only. Samples before: %d / after: %d.\n",
            paste(CT_KEEP, collapse = ", "), ncol(dge_filt), sum(keep_samp)))
dge_filt  <- dge_filt[, keep_samp, keep.lib.sizes = FALSE]

cat("After restricting to the 4 primary cell types (DMSO only):\n")
cat("  ", nrow(dge_filt), "isoforms x", ncol(dge_filt), "samples\n")
cat("  Cell types:", paste(sort(unique(as.character(dge_filt$samples$ct))), collapse = ", "), "\n")
cat("  Treatments:", paste(sort(unique(as.character(dge_filt$samples$treatment))), collapse = ", "), "\n")

# Hoist sample-metadata reference and cell-type vector
samp_filt <- dge_filt$samples
cts       <- sort(unique(as.character(samp_filt$ct)))
```

```{r load-sqanti}
sq <- fread(SQ_CLASS_PATH)
cat("SQANTI3 classification:", nrow(sq), "rows\n")

sq_filt <- sq %>%
  filter(isoform %in% rownames(dge_filt$counts)) %>%
  mutate(
    novelty = case_when(
      structural_category == "full-splice_match"       ~ "Known (FSM)",
      structural_category == "incomplete-splice_match" ~ "Known Variant (ISM)",
      structural_category == "novel_in_catalog"        ~ "Novel (NIC)",
      structural_category == "novel_not_in_catalog"    ~ "Novel (NNC)",
      TRUE                                             ~ structural_category
    ),
    known_novel = case_when(
      structural_category %in% c("full-splice_match", "incomplete-splice_match") ~ "Known",
      structural_category %in% c("novel_in_catalog",  "novel_not_in_catalog")    ~ "Novel",
      TRUE                                                                       ~ "Other"
    ),
    category_short = case_when(
      structural_category == "full-splice_match"       ~ "FSM",
      structural_category == "incomplete-splice_match" ~ "ISM",
      structural_category == "novel_in_catalog"        ~ "NIC",
      structural_category == "novel_not_in_catalog"    ~ "NNC",
      TRUE                                             ~ "Other"
    )
  )

cat("Matched to filtered DGEList:", nrow(sq_filt), "isoforms\n")

```

```{r helper-get-expressed}
# Returns isoform IDs with at least one count across the given sample indices
get_expressed <- function(dge, sample_idx) {
  counts_sub <- dge$counts[, sample_idx, drop = FALSE]
  rownames(counts_sub)[rowSums(counts_sub) > 0]
}
```

# Total Isoforms — Known Vs Novel

```{r q1-total}
cat_table <- sq_filt %>%
  count(structural_category, novelty, known_novel) %>%
  arrange(desc(n)) %>%
  mutate(pct = round(100 * n / sum(n), 2))

cat("Structural category breakdown:\n")
print(cat_table)

kn_table <- sq_filt %>%
  count(known_novel) %>%
  mutate(pct = round(100 * n / sum(n), 2))

cat("\nKnown vs Novel:\n")
print(kn_table)

cat("\nTotal isoforms:", nrow(sq_filt), "\n")
cat("Unique genes:",    length(unique(sq_filt$associated_gene)), "\n")

```

```{r q1-barplot, fig.width=10, fig.height=6}
p_cat <- ggplot(cat_table, aes(x = reorder(novelty, n), y = n, fill = known_novel)) +
  geom_col(alpha = 0.8) +
  geom_text(aes(label = paste0(comma(n), " (", pct, "%)")), hjust = -0.05, size = 3.5) +
  coord_flip() +
  scale_fill_manual(values = c("Known" = "steelblue",
                               "Novel" = "coral",
                               "Other" = "gray60")) +
  scale_y_continuous(labels = comma) +
  labs(title    = "SQANTI3 Structural Categories",
       subtitle = paste0("Total: ", comma(nrow(sq_filt)),
                         " isoforms from ",
                         comma(length(unique(sq_filt$associated_gene))),
                         " genes"),
       x = NULL, y = "Number Of Isoforms", fill = NULL) +
  theme_minimal(base_size = 13) +
  theme(plot.title      = element_text(face = "bold"),
        plot.subtitle   = element_text(color = "grey40"),
        legend.position = "top") +
  expand_limits(y = max(cat_table$n) * 1.25)

print(p_cat)

ggsave(file.path(OUTPUT_DIR, "structural_categories.png"),
       p_cat, width = 10, height = 6, dpi = 300, bg = "white")
```

# Structural Classes By Cell Type

```{r q4-by-ct}
cat_by_ct <- list()
for (ct_i in cts) {
  idx       <- which(as.character(samp_filt$ct) == ct_i)
  expressed <- get_expressed(dge_filt, idx)

  ct_sq <- sq_filt %>% filter(isoform %in% expressed)
  cat_by_ct[[ct_i]] <- ct_sq %>%
    count(category_short) %>%
    mutate(ct  = ct_i,
           pct = round(100 * n / sum(n), 2),
           category_short = factor(category_short, levels = CAT_ORDER))
}

cat_by_ct_df <- bind_rows(cat_by_ct)
print(cat_by_ct_df %>% arrange(ct, category_short))

```

```{r q4-stacked, fig.width=12, fig.height=7}
p_cat_ct <- ggplot(cat_by_ct_df, aes(x = ct, y = n, fill = category_short)) +
  geom_col(alpha = 0.8) +
  geom_text(aes(label = comma(n)),
            position = position_stack(vjust = 0.5), size = 3.5) +
  scale_fill_manual(values = c("FSM"   = "#2166AC",
                               "ISM"   = "#67A9CF",
                               "NIC"   = "#EF8A62",
                               "NNC"   = "#B2182B",
                               "Other" = "gray70")) +
  scale_y_continuous(labels = comma) +
  labs(title    = "Structural Categories By Cell Type",
       subtitle = "Number of expressed isoforms in analysis set",
       x = "Cell Type", y = "Number Of Isoforms", fill = "Category") +
  theme_minimal(base_size = 13) +
  theme(plot.title      = element_text(face = "bold"),
        plot.subtitle   = element_text(color = "grey40"),
        legend.position = "top")

print(p_cat_ct)

ggsave(file.path(OUTPUT_DIR, "structural_categories_by_ct.png"),
       p_cat_ct, width = 12, height = 7, dpi = 300, bg = "white")
```

## Isoforms Per Gene

```{r iso-per-gene}
# Verify the gene column exists. SQANTI3 typically uses `associated_gene`
# but some pipelines rename it. Print available cols on failure.
gene_col <- intersect(c("associated_gene", "associated_genes", "gene", "gene_id"),
                      colnames(sq_filt))[1]
if (is.na(gene_col)) {
  stop("No gene column found in sq_filt. Available columns:\n  ",
       paste(colnames(sq_filt), collapse = ", "))
}
cat("Using gene column:", gene_col, "\n\n")

# Per-gene isoform count, base R to dodge namespace-masking traps
gene_vec <- sq_filt[[gene_col]]
gene_vec <- gene_vec[!is.na(gene_vec) & gene_vec != ""]

iso_per_gene <- as.data.frame(
  table(associated_gene = gene_vec),
  stringsAsFactors = FALSE
)
colnames(iso_per_gene) <- c("associated_gene", "n_isoforms")
iso_per_gene <- iso_per_gene[order(-iso_per_gene$n_isoforms), ]
rownames(iso_per_gene) <- NULL

cat("Isoforms per gene (filtered set):\n")
cat("  Genes:                       ", nrow(iso_per_gene), "\n")
cat("  Total isoforms accounted for:", sum(iso_per_gene$n_isoforms), "\n")
cat("  Median:                      ", median(iso_per_gene$n_isoforms), "\n")
cat("  Mean:                        ", round(mean(iso_per_gene$n_isoforms), 2), "\n")
cat("  IQR (Q1, Q3):                ",
    quantile(iso_per_gene$n_isoforms, 0.25), ", ",
    quantile(iso_per_gene$n_isoforms, 0.75), "\n")
cat("  Max:                         ", max(iso_per_gene$n_isoforms), "\n\n")

multi_iso  <- sum(iso_per_gene$n_isoforms >= 2)
single_iso <- sum(iso_per_gene$n_isoforms == 1)
ge10_iso   <- sum(iso_per_gene$n_isoforms >= 10)

cat("Genes with multiple isoforms (>=2):",
    multi_iso, " (",
    round(100 * multi_iso / nrow(iso_per_gene), 1), "%)\n", sep = "")
cat("Genes with a single isoform:       ",
    single_iso, " (",
    round(100 * single_iso / nrow(iso_per_gene), 1), "%)\n", sep = "")
cat("Genes with >=10 isoforms:          ",
    ge10_iso, " (",
    round(100 * ge10_iso / nrow(iso_per_gene), 1), "%)\n", sep = "")

# Distribution table (capped at 20+) — base R aggregate
bin_vec  <- ifelse(iso_per_gene$n_isoforms >= 20, "20+",
                   as.character(iso_per_gene$n_isoforms))
dist_tbl <- as.data.frame(table(bin = bin_vec), stringsAsFactors = FALSE)
colnames(dist_tbl) <- c("bin", "n_genes")
dist_tbl$bin_num <- ifelse(dist_tbl$bin == "20+", 20L, as.integer(dist_tbl$bin))
dist_tbl <- dist_tbl[order(dist_tbl$bin_num), ]
dist_tbl$pct <- round(100 * dist_tbl$n_genes / sum(dist_tbl$n_genes), 2)
rownames(dist_tbl) <- NULL

cat("\nDistribution of isoforms per gene (capped at 20+):\n")
print(dist_tbl[, c("bin", "n_genes", "pct")], row.names = FALSE)

# Bind into the names the histogram chunk expects
dist_table <- dist_tbl

```

```{r iso-per-gene-hist, fig.width=10, fig.height=6}
p_ipg <- ggplot(iso_per_gene %>% dplyr::mutate(capped = pmin(n_isoforms, 20)),
                aes(x = capped)) +
  geom_histogram(binwidth = 1, fill = "steelblue", color = "white", alpha = 0.85) +
  scale_x_continuous(breaks = c(1, 5, 10, 15, 20),
                     labels = c("1", "5", "10", "15", "20+")) +
  scale_y_continuous(labels = comma) +
  labs(title    = "Isoforms Per Gene",
       subtitle = paste0(comma(nrow(iso_per_gene)), " genes | ",
                         comma(sum(iso_per_gene$n_isoforms)), " isoforms | ",
                         "median ", median(iso_per_gene$n_isoforms),
                         ", mean ", round(mean(iso_per_gene$n_isoforms), 1),
                         ", max ", max(iso_per_gene$n_isoforms)),
       x = "Number Of Isoforms (capped at 20+)",
       y = "Number Of Genes") +
  theme_minimal(base_size = 13) +
  theme(plot.title    = element_text(face = "bold"),
        plot.subtitle = element_text(color = "grey40"))

print(p_ipg)

ggsave(file.path(OUTPUT_DIR, "isoforms_per_gene.png"),
       p_ipg, width = 10, height = 6, dpi = 300, bg = "white")
```

```{r iso-per-gene-save}
fwrite(iso_per_gene, file.path(OUTPUT_DIR, "isoforms_per_gene.csv"))
fwrite(dist_table %>% dplyr::select(bin, n_genes, pct),
       file.path(OUTPUT_DIR, "isoforms_per_gene_distribution.csv"))

```
# Expressed Isoforms By Cell Type

## Per Cell Type, Per Treatment

```{r q2-unique}
ct_isoforms <- list()
for (ct_i in cts) {
  idx <- which(as.character(samp_filt$ct) == ct_i)
  ct_isoforms[[ct_i]] <- get_expressed(dge_filt, idx)
}

ct_summary <- data.frame(
  ct          = names(ct_isoforms),
  n_expressed = sapply(ct_isoforms, length),
  stringsAsFactors = FALSE
) %>% arrange(desc(n_expressed))

cat("Expressed isoforms per cell type:\n"); print(ct_summary)

ct_treat_list <- list()
for (ct_i in cts) {
  for (trt in c("DMSO", "Smg1i")) {
    idx <- which(as.character(samp_filt$ct) == ct_i &
                 as.character(samp_filt$treatment) == trt)
    if (length(idx) == 0) next
    n_expr <- length(get_expressed(dge_filt, idx))
    ct_treat_list[[paste(ct_i, trt)]] <- data.frame(
      ct = ct_i, treatment = trt, n_expressed = n_expr, stringsAsFactors = FALSE
    )
  }
}
ct_treat_df <- bind_rows(ct_treat_list)
cat("\nBy cell type and treatment:\n")
print(ct_treat_df %>% arrange(ct, treatment))
```

## Cell-Type-Specific Vs Shared in Both Conditions

```{r q2-sharing-counts}
all_iso <- unique(unlist(ct_isoforms))

# Long table: one row per (isoform, ct) where the isoform was expressed in ct
iso_long <- rbindlist(lapply(names(ct_isoforms), function(ct_i) {
  data.table(isoform = ct_isoforms[[ct_i]], ct = ct_i)
}))

iso_ct_count <- iso_long[, .(n_cts = uniqueN(ct)), by = isoform]

cat("Isoforms specific to a single cell type:\n")
for (ct_i in names(ct_isoforms)) {
  specific <- iso_ct_count[n_cts == 1 & isoform %in% ct_isoforms[[ct_i]], .N]
  cat("  ", ct_i, ":", specific, "\n")
}

cat("\nExpressed across any cell type:", length(all_iso), "\n")
cat("Expressed in all", length(ct_isoforms), "cell types:",
    iso_ct_count[n_cts == length(ct_isoforms), .N], "\n")

```

```{r seem ct specific}
# Does NMD make isoforms look cell-type-restricted? Compare DMSO-only vs SMG1i-only
# expression breadth (equal sample numbers per CT per treatment).
MINCOUNT <- 1   # reads required...
MINSAMP  <- 1   # ...in at least this many samples of a CT×treatment to call "expressed"

# Reload the FULL DGEList (both treatments), drop DO_ALI 029T, keep the 4 CTs
dge_all <- readRDS(DGE_FILT_PATH)
# Relabel ct to display names AFTER loading (AT -> AT2, DD -> LAE)
dge_all$samples$ct <- relabel_ct(dge_all$samples$ct)

sc <- as.character(dge_all$samples$ct); ic <- as.character(dge_all$samples$id)
n_before <- ncol(dge_all)
dge_all <- dge_all[, sc %in% CT_KEEP & !(sc == "DO_ALI" & ic == "029T"), keep.lib.sizes = FALSE]
cat(sprintf("FILTER: 4 CTs (drop DO_ALI 029T). Samples before: %d / after: %d.\n",
            n_before, ncol(dge_all)))
sampA   <- dge_all$samples
ctsA    <- sort(unique(as.character(sampA$ct)))
all_iso <- rownames(dge_all$counts)

# expressed-isoform logical vector for one CT × treatment
expr_vec <- function(ct_i, trt){
  idx <- which(as.character(sampA$ct) == ct_i & as.character(sampA$treatment) == trt)
  if (length(idx) == 0) return(setNames(rep(FALSE, length(all_iso)), all_iso))
  rowSums(dge_all$counts[, idx, drop = FALSE] >= MINCOUNT) >= MINSAMP
}
mat_dmso <- sapply(ctsA, expr_vec, trt = "DMSO")   # isoform × CT logical
mat_smg  <- sapply(ctsA, expr_vec, trt = "Smg1i")
rownames(mat_dmso) <- all_iso; rownames(mat_smg) <- all_iso

n_dmso <- rowSums(mat_dmso); n_smg <- rowSums(mat_smg)
cat(sprintf("Samples per CT×treatment:\n")); print(table(sampA$ct, sampA$treatment))

# ---- isoforms that look CT-restricted under DMSO (expressed in exactly 1 CT) ----
restr_dmso <- all_iso[n_dmso == 1]
# for each, the CTs it gains under SMG1i that it did NOT have in DMSO
gained_other <- rowSums(mat_smg[restr_dmso, , drop = FALSE] & !mat_dmso[restr_dmso, , drop = FALSE])
res <- data.frame(
  isoform   = restr_dmso,
  dmso_ct   = ctsA[apply(mat_dmso[restr_dmso, , drop = FALSE], 1, which.max)],
  n_ct_smg  = n_smg[restr_dmso],
  gained_cts_under_smg = gained_other,
  unmasked  = gained_other >= 1,         # appears in >=1 OTHER CT only under SMG1i
  stringsAsFactors = FALSE
)
cat(sprintf("\nDMSO-restricted isoforms (1 CT): %d\n", nrow(res)))
cat(sprintf("  ...appearing in >=1 OTHER CT under SMG1i (NMD-masked elsewhere): %d (%.1f%%)\n",
            sum(res$unmasked), 100*mean(res$unmasked)))
cat("  Breakdown by the cell type they were 'restricted' to (DMSO):\n")
print(res %>% group_by(dmso_ct) %>%
        summarise(n_restricted = n(), n_unmasked = sum(unmasked),
                  pct_unmasked = round(100*mean(unmasked),1), .groups="drop"))

# ---- symmetric control: restricted under SMG1i, do they expand under DMSO? ----
restr_smg <- all_iso[n_smg == 1]
gained_d  <- rowSums(mat_dmso[restr_smg, , drop = FALSE] & !mat_smg[restr_smg, , drop = FALSE])
cat(sprintf("\nCONTROL — SMG1i-restricted isoforms (1 CT): %d; appearing in >=1 OTHER CT under DMSO: %d (%.1f%%)\n",
            length(restr_smg), sum(gained_d >= 1), 100*mean(gained_d >= 1)))
cat("If the DMSO->SMG1i unmasking rate is much higher than this control, the apparent CT-restriction is NMD-driven, not detection noise.\n")

fwrite(res, file.path(OUTPUT_DIR, "ct_restriction_nmd_unmasking.csv"))
```

```{r q2-unique-barplot-split, fig.width=10, fig.height=8}
unique_sharing <- data.frame(
  ct      = names(ct_isoforms),
  n_total = sapply(ct_isoforms, length),
  n_specific = sapply(names(ct_isoforms), function(ct_i) {
    specific_isos <- iso_ct_count[n_cts == 1, isoform]
    sum(ct_isoforms[[ct_i]] %in% specific_isos)
  }),
  stringsAsFactors = FALSE
) %>%
  mutate(n_shared = n_total - n_specific) %>%
  pivot_longer(cols = c(n_shared, n_specific),
               names_to = "category", values_to = "n") %>%
  mutate(category = ifelse(category == "n_specific",
                           "Cell-Type-Specific",
                           "Shared (2+ Cell Types)"))

# Order x-axis by total count
ct_order <- unique_sharing %>%
  group_by(ct) %>%
  summarise(total = sum(n), .groups = "drop") %>%
  arrange(desc(total)) %>%
  pull(ct)
unique_sharing$ct <- factor(unique_sharing$ct, levels = ct_order)

fill_vals <- c("Cell-Type-Specific"     = "#F4A582",
               "Shared (2+ Cell Types)" = "#2166AC")

# Split-axis layout: top panel zooms into the shared bulk, bottom panel
# resolves the much smaller cell-type-specific counts.
p_top <- ggplot(unique_sharing, aes(x = ct, y = n, fill = category)) +
  geom_col(alpha = 0.8) +
  geom_text(aes(label = comma(n)), position = position_stack(vjust = 0.5), size = 3.5) +
  scale_fill_manual(values = fill_vals) +
  scale_y_continuous(labels = comma) +
  coord_cartesian(ylim = c(125000, NA)) +
  labs(title    = "Expressed Isoforms Per Cell Type: Shared Vs Cell-Type-Specific",
       subtitle = "Cell-type-specific = expressed in only one cell type",
       y = NULL, x = NULL, fill = NULL) +
  theme_minimal(base_size = 14) +
  theme(plot.title      = element_text(face = "bold"),
        plot.subtitle   = element_text(color = "grey40"),
        legend.position = "top",
        axis.text.x     = element_blank(),
        axis.ticks.x    = element_blank(),
        plot.margin     = margin(5, 5, 2, 5))

p_bottom <- ggplot(unique_sharing, aes(x = ct, y = n, fill = category)) +
  geom_col(alpha = 0.8) +
  scale_fill_manual(values = fill_vals) +
  scale_y_continuous(labels = comma) +
  coord_cartesian(ylim = c(0, 50000)) +
  labs(x = "Cell Type", y = NULL, fill = NULL) +
  theme_minimal(base_size = 14) +
  theme(legend.position = "none",
        plot.margin     = margin(2, 5, 5, 5))

p_combined <- p_top / plot_spacer() / p_bottom +
  plot_layout(heights = c(3, 0.15, 1))

p_final <- wrap_elements(p_combined) +
  labs(tag = "Number Of Expressed Isoforms") +
  theme(plot.tag          = element_text(size = 13, angle = 90),
        plot.tag.position = "left")

print(p_final)

ggsave(file.path(OUTPUT_DIR, "expressed_isoforms_shared_vs_specific.png"),
       p_final, width = 10, height = 8, dpi = 300, bg = "white")

```

# Isoform Length Distribution

## Overall — Known Vs Novel

```{r q3-overall-density, fig.width=12, fig.height=6}
sq_filt_len <- sq_filt %>% filter(!is.na(length))

p_len_overall <- ggplot(sq_filt_len, aes(x = log10(length), color = known_novel)) +
  geom_freqpoly(bins = 80, linewidth = 1) +
  scale_color_manual(values = c("Known" = "steelblue",
                                "Novel" = "coral",
                                "Other" = "gray60")) +
  labs(title    = "Isoform Length Distribution",
       subtitle = paste0("All ", comma(nrow(sq_filt_len)),
                         " isoforms — SQANTI3 transcript length"),
       x = "log10(Isoform Length, bp)", y = "Number Of Isoforms", color = NULL) +
  theme_minimal(base_size = 13) +
  theme(plot.title      = element_text(face = "bold"),
        plot.subtitle   = element_text(color = "grey40"),
        legend.position = "top")

print(p_len_overall)

ggsave(file.path(OUTPUT_DIR, "length_distribution.png"),
       p_len_overall, width = 12, height = 6, dpi = 300, bg = "white")

cat("Length by known/novel:\n")
sq_filt %>%
  filter(!is.na(length)) %>%
  group_by(known_novel) %>%
  summarise(n      = n(),
            median = median(length),
            mean   = round(mean(length), 1),
            Q1     = quantile(length, 0.25),
            Q3     = quantile(length, 0.75),
            .groups = "drop") %>%
  print()
```

## Build Per-Treatment Length Table

```{r build-length-data}
length_data <- list()
for (ct_i in cts) {
  for (trt in c("DMSO", "Smg1i")) {
    idx <- which(as.character(samp_filt$ct) == ct_i &
                 as.character(samp_filt$treatment) == trt)
    if (length(idx) == 0) next
    expressed <- get_expressed(dge_filt, idx)
    length_data[[paste(ct_i, trt)]] <- sq_filt %>%
      filter(isoform %in% expressed, !is.na(length)) %>%
      select(isoform, length, known_novel, category_short) %>%
      mutate(ct = ct_i, treatment = trt)
  }
}
length_data <- bind_rows(length_data)

cat("length_data:", nrow(length_data), "rows\n")
cat("Cell types:", paste(sort(unique(length_data$ct)), collapse = ", "), "\n")
cat("Treatments:", paste(unique(length_data$treatment), collapse = ", "), "\n")

```


## By Cell Type And Treatment

```{r q3-by-ct-density, fig.width=14, fig.height=10}
p_len_ct <- ggplot(length_data, aes(x = log10(length), color = treatment)) +
  geom_freqpoly(bins = 80, linewidth = 0.8) +
  facet_wrap(~ ct, ncol = 2, scales = "free_y") +
  scale_color_manual(values = c("DMSO" = "orange", "Smg1i" = "purple")) +
  labs(title = "Isoform Length By Cell Type And Treatment",
       x = "log10(Isoform Length, bp)", y = "Number Of Isoforms", color = "Treatment") +
  theme_minimal(base_size = 12) +
  theme(plot.title      = element_text(face = "bold"),
        strip.text      = element_text(face = "bold", size = 13),
        legend.position = "top")

print(p_len_ct)

ggsave(file.path(OUTPUT_DIR, "length_distribution_by_ct_treatment.png"),
       p_len_ct, width = 14, height = 10, dpi = 300, bg = "white")
```

```{r q3-stats}
cat("--- Length Stats By Treatment ---\n")
length_data %>%
  group_by(treatment) %>%
  summarise(n = n(), min = min(length), Q1 = quantile(length, 0.25),
            median = median(length), mean = round(mean(length), 1),
            Q3 = quantile(length, 0.75), max = max(length),
            .groups = "drop") %>%
  print()

cat("\n--- Length Stats By Cell Type And Treatment ---\n")
length_data %>%
  group_by(ct, treatment) %>%
  summarise(n = n(), min = min(length), Q1 = quantile(length, 0.25),
            median = median(length), mean = round(mean(length), 1),
            Q3 = quantile(length, 0.75), max = max(length),
            .groups = "drop") %>%
  print()

```

# Pairwise Expression Similarity Across Cell Types

## Build Mean Log-CPM Matrices

```{r build-cpm-matrices}
# Helper: mean log-CPM per cell type, optionally restricted to one treatment
build_cpm_matrix <- function(dge, samp, cts, treatment = NULL) {
  out <- list()
  for (ct_i in cts) {
    idx <- if (is.null(treatment)) {
      which(as.character(samp$ct) == ct_i)
    } else {
      which(as.character(samp$ct) == ct_i &
            as.character(samp$treatment) == treatment)
    }
    if (length(idx) == 0) next
    ct_cpm    <- cpm(dge[, idx], log = TRUE, prior.count = 1)
    out[[ct_i]] <- rowMeans(ct_cpm)
  }
  do.call(cbind, out)
}

cpm_ct_mat    <- build_cpm_matrix(dge_filt, samp_filt, cts)
cpm_dmso_mat  <- build_cpm_matrix(dge_filt, samp_filt, cts, treatment = "DMSO")
cpm_smg1i_mat <- build_cpm_matrix(dge_filt, samp_filt, cts, treatment = "Smg1i")

# Combined: each isoform contributes one point per treatment, stacked vertically
cts_shared <- intersect(colnames(cpm_dmso_mat), colnames(cpm_smg1i_mat))
cpm_combined_mat <- rbind(cpm_dmso_mat[, cts_shared, drop = FALSE],
                          cpm_smg1i_mat[, cts_shared, drop = FALSE])

cat("All-samples matrix:", dim(cpm_ct_mat),    "\n")
cat("DMSO matrix:       ", dim(cpm_dmso_mat),  "\n")
cat("Smg1i matrix:      ", dim(cpm_smg1i_mat), "\n")
cat("Combined matrix:   ", dim(cpm_combined_mat), "\n")
```

## Spearman And Pearson Across All Samples

```{r corr-all}
cor_pearson  <- cor(cpm_ct_mat, method = "pearson")
cor_spearman <- cor(cpm_ct_mat, method = "spearman")

cat("Pearson correlation:\n");  print(round(cor_pearson,  3))
cat("\nSpearman correlation:\n"); print(round(cor_spearman, 3))

```

## Jaccard Similarity Of Expressed Isoform Sets

```{r jaccard}
jaccard_mat <- matrix(NA, length(cts), length(cts), dimnames = list(cts, cts))
for (i in seq_along(cts)) {
  for (j in seq_along(cts)) {
    a <- ct_isoforms[[cts[i]]]
    b <- ct_isoforms[[cts[j]]]
    jaccard_mat[i, j] <- length(intersect(a, b)) / length(union(a, b))
  }
}
print(round(jaccard_mat, 3))
```

```{r corr-jaccard-heatmaps, fig.width=10, fig.height=8}
# --- Spearman heatmap (all samples) ---
cor_df <- melt(cor_spearman)
colnames(cor_df) <- c("CT1", "CT2", "Correlation")

p_cor <- ggplot(cor_df, aes(x = CT1, y = CT2, fill = Correlation)) +
  geom_tile(color = "white", linewidth = 0.8) +
  geom_text(aes(label = round(Correlation, 2)),
            size = 4, fontface = "bold",
            color = ifelse(cor_df$Correlation > 0.85, "white", "grey20")) +
  scale_fill_gradient2(low = "#D73027", mid = "white", high = "#2166AC",
                       midpoint = min(cor_spearman),
                       limits   = c(min(cor_spearman), 1),
                       name = "Spearman r") +
  labs(title    = "Pairwise Expression Similarity Across Cell Types",
       subtitle = "Spearman correlation of mean log-CPM per isoform",
       x = NULL, y = NULL) +
  theme_minimal(base_size = 13) +
  theme(axis.text.x   = element_text(angle = 45, hjust = 1, face = "bold", size = 14),
        axis.text.y   = element_text(face = "bold", size = 14),
        panel.grid    = element_blank(),
        plot.title    = element_text(face = "bold", size = 16),
        plot.subtitle = element_text(color = "grey40", size = 13))

print(p_cor)

ggsave(file.path(OUTPUT_DIR, "pairwise_expression_similarity_spearman.png"),
       p_cor, width = 10, height = 8, dpi = 300, bg = "white")

# --- Jaccard heatmap ---
jac_df <- melt(jaccard_mat)
colnames(jac_df) <- c("CT1", "CT2", "Jaccard")

p_jac <- ggplot(jac_df, aes(x = CT1, y = CT2, fill = Jaccard)) +
  geom_tile(color = "white", linewidth = 0.8) +
  geom_text(aes(label = round(Jaccard, 2)),
            size = 6, fontface = "bold", color = "black") +
  scale_fill_gradient(low = "white", high = "#185FA5",
                      limits = c(min(jaccard_mat), 1), name = "Jaccard") +
  labs(title    = "Pairwise Isoform Overlap Across Cell Types",
       subtitle = "Jaccard index of expressed isoform sets",
       x = NULL, y = NULL) +
  theme_minimal(base_size = 13) +
  theme(axis.text.x   = element_text(angle = 45, hjust = 1, face = "bold", size = 14),
        axis.text.y   = element_text(face = "bold", size = 14),
        panel.grid    = element_blank(),
        plot.title    = element_text(face = "bold", size = 16),
        plot.subtitle = element_text(color = "grey40", size = 13))

print(p_jac)

ggsave(file.path(OUTPUT_DIR, "pairwise_isoform_overlap_jaccard.png"),
       p_jac, width = 10, height = 8, dpi = 300, bg = "white")

```

## Pairwise Scatterplots

Helper makes a single 3-column patchwork of scatterplots for any CPM matrix.
Called three times: all samples, DMSO only, Smg1i only, and combined-stacked.

```{r scatter-helper}
make_pairwise_scatter <- function(cpm_mat, title, subtitle) {
  ct_pairs <- combn(colnames(cpm_mat), 2, simplify = FALSE)

  scatter_list <- lapply(ct_pairs, function(pair) {
    ct1 <- pair[1]; ct2 <- pair[2]

    df_pair <- data.frame(x = cpm_mat[, ct1], y = cpm_mat[, ct2])

    r_s <- round(cor(df_pair$x, df_pair$y, method = "spearman"), 3)
    r_p <- round(cor(df_pair$x, df_pair$y, method = "pearson"),  3)

    ggplot(df_pair, aes(x = x, y = y)) +
      geom_point(alpha = 0.05, size = 0.3, color = "grey30") +
      geom_abline(slope = 1, intercept = 0,
                  linetype = "dashed", color = "red", linewidth = 0.5) +
      annotate("text",
               x = min(df_pair$x) + 0.5, y = max(df_pair$y) - 0.5,
               label = paste0("r[s] == ", r_s),
               parse = TRUE, size = 4.5, hjust = 0, color = "#2166AC") +
      annotate("text",
               x = min(df_pair$x) + 0.5, y = max(df_pair$y) - 1.5,
               label = paste0("r[p] == ", r_p),
               parse = TRUE, size = 4.5, hjust = 0, color = "#D73027") +
      labs(x = paste0(ct1, " (Mean log-CPM)"),
           y = paste0(ct2, " (Mean log-CPM)")) +
      theme_minimal(base_size = 11) +
      theme(plot.margin = margin(5, 5, 5, 5))
  })

  wrap_plots(scatter_list, ncol = 3) +
    plot_annotation(
      title    = title,
      subtitle = subtitle,
      theme    = theme(plot.title    = element_text(face = "bold", size = 16),
                       plot.subtitle = element_text(color = "grey40", size = 12))
    )
}
```

```{r pairwise-scatter-dmso, fig.width=14, fig.height=14}
p_scatter_dmso <- make_pairwise_scatter(
  cpm_dmso_mat,
  title    = "Pairwise Isoform Expression: DMSO Only",
  subtitle = "Mean log-CPM per isoform across DMSO samples"
)
print(p_scatter_dmso)

ggsave(file.path(OUTPUT_DIR, "pairwise_expression_scatterplots_dmso.png"),
       p_scatter_dmso, width = 14, height = 14, dpi = 300, bg = "white")

cat("Spearman (DMSO):\n");  print(round(cor(cpm_dmso_mat,  method = "spearman"), 3))
cat("\nPearson (DMSO):\n"); print(round(cor(cpm_dmso_mat,  method = "pearson"),  3))

```
```{r long_read_summary}
library(dplyr)
library(readr)
library(tibble)

counts_path <- "/proj/regeps/regep00/studies/ExternalCellLines/analyses/repjc/Randell_Lung_Cells_2025/code/isocall_v1/projects/nmd_lungcells/results/call/nmd_isocall.count_matrix.txt"

counts <- read_csv(counts_path, show_col_types = FALSE)

id_col <- names(counts)[1]
sample_cols <- setdiff(names(counts), id_col)

per_sample <- tibble(
  sample = sample_cols,
  assigned_reads = colSums(counts[sample_cols])
) %>%
  mutate(
    # Sample names still carry the original _AT_/_DD_ tokens; map to display names.
    cell_type = case_when(
      grepl("_DD_ALI_", sample) ~ "DD_ALI",
      grepl("_DO_",     sample) ~ "DO_ALI",
      grepl("_AT_",     sample) ~ "AT2",
      grepl("_DD_",     sample) ~ "LAE",
      grepl("_FB_",     sample) ~ "FB",
      grepl("_MV_",     sample) ~ "MV"
    ),
    treatment = if_else(grepl("Smg1i$", sample), "Smg1i", "DMSO"),
    donor     = sub(".*_([0-9]{3}[A-Z])_.*", "\\1", sample)
  )

per_sample_kept <- per_sample %>%
  filter(cell_type %in% c("AT2", "LAE", "FB", "MV"))

# Methods numbers
n_samples    <- nrow(per_sample_kept)
mean_reads   <- mean(per_sample_kept$assigned_reads)
median_reads <- median(per_sample_kept$assigned_reads)
total_reads  <- sum(per_sample_kept$assigned_reads)
range_reads  <- range(per_sample_kept$assigned_reads)

cat(sprintf(
  "Across %d samples (4 cell types × Smg1i/DMSO, matched donors), an average of %s long reads per sample (median %s, range %s–%s) were assigned to annotated isoforms by IsoQuant, totaling %s reads.\n",
  n_samples,
  format(round(mean_reads), big.mark = ","),
  format(round(median_reads), big.mark = ","),
  format(range_reads[1], big.mark = ","),
  format(range_reads[2], big.mark = ","),
  format(total_reads, big.mark = ",")
))

per_sample_kept %>% arrange(cell_type, treatment, donor)
```

# Save Results

```{r save-results}
# Tables — structural and expression summaries
fwrite(cat_table,      file.path(OUTPUT_DIR, "structural_categories.csv"))
fwrite(kn_table,       file.path(OUTPUT_DIR, "known_vs_novel.csv"))
fwrite(ct_summary,     file.path(OUTPUT_DIR, "expressed_isoforms_per_ct.csv"))
fwrite(ct_treat_df,    file.path(OUTPUT_DIR, "expressed_isoforms_per_ct_treatment.csv"))
fwrite(cat_by_ct_df,   file.path(OUTPUT_DIR, "structural_categories_by_ct.csv"))

# Sharing / overlap
fwrite(unique_sharing %>% pivot_wider(id_cols = c(ct, n_total),
                                       names_from = category, values_from = n),
       file.path(OUTPUT_DIR, "expressed_isoforms_sharing.csv"))

# Length stats
fwrite(length_data,
       file.path(OUTPUT_DIR, "length_data_long.csv"))

# Pairwise similarity matrices
fwrite(as.data.frame(cor_pearson),  file.path(OUTPUT_DIR, "pairwise_pearson.csv"),  row.names = TRUE)
fwrite(as.data.frame(cor_spearman), file.path(OUTPUT_DIR, "pairwise_spearman.csv"), row.names = TRUE)
fwrite(as.data.frame(jaccard_mat),  file.path(OUTPUT_DIR, "pairwise_jaccard.csv"),  row.names = TRUE)

cat("All saved to:", OUTPUT_DIR, "\n")
print(list.files(OUTPUT_DIR, pattern = "\\.csv$"))

```

# Session Info

```{r session-info}
sessionInfo()
```

