---
title: "Isoform-Level DIE Summary: NMD Inhibition Effects (Part 1)"
author: "Leshem, Yul"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: lumen
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
    code_folding: hide
---

Code location: `/udd/reyle/nmd_lungcells_2026/code/final/die_mashr_summary_part1_2026.3.10.Rmd`



```{r setup, include=FALSE}
options(bitmapType = "cairo")
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      cache = FALSE, dev = "png")

```

# Analysis Overview

This report summarizes Differential Isoform Expression (DIE) results from the
NMD inhibition study, using mashr posterior estimates.

Effect sizes (`logFC`) are posterior means from multivariate adaptive shrinkage,
which borrows strength across cell types. Significance is the local false sign
rate (lfsr); lfsr < 0.05 means there is < 5% probability that the sign of the
effect estimate is wrong. lfsr is carried in the `adj.P.Val` column for
column-format compatibility with limma topTables.

**Experimental Design:**

- **Treatment:** Smg1i (NMD inhibitor) vs DMSO (control)
- **Cell Types:** AT2, LAE, FB, MV (DD_ALI / DO_ALI excluded — donor confounding / small n)
- **Data Type:** Long-read PacBio Kinnex/MAS-Seq, SQANTI3-filtered isoforms
- **Analysis:** limma-voom + mashr at transcript level

**Definitions:**

- **NMD susceptible transcript:** posterior mean logFC > 0 AND lfsr < 0.05
  (rises when NMD is inhibited).
- **NMD suppressed transcript:** posterior mean logFC < 0 AND lfsr < 0.05.

# Libraries

```{r load-libs}
library(tidyverse)
library(data.table)
library(DT)
library(edgeR)
library(pheatmap)

# Bioconductor namespace masking guard
rename <- dplyr::rename
select <- dplyr::select
filter <- dplyr::filter
mutate <- dplyr::mutate
count  <- dplyr::count
```

```{r configuration}
DATE_STAMP <- "2026.3.10"

INPUT_DIE_DIR <- "/udd/reyle/nmd_lungcells_2026/long_read/mashr"
OUTPUT_DIR    <- file.path("/udd/reyle/nmd_lungcells_2026/tmp/mashr_die")
OUT_DIR_FIG   <- "/udd/reyle/nmd_lungcells_2026/final_figures/lr_iso"

dir.create(OUTPUT_DIR,  recursive = TRUE, showWarnings = FALSE)
dir.create(OUT_DIR_FIG, recursive = TRUE, showWarnings = FALSE)

# 4 cell types only — DD_ALI / DO_ALI dropped upstream.
# Pipeline now writes nmd_mashr_die_at2/lae/fb/mv_<date>.csv, so the keys here
# are the lower-case file codes (at2/lae/fb/mv), mapped to display names.
CELLTYPE_MAP     <- c("at2" = "AT2", "lae" = "LAE", "fb" = "FB", "mv" = "MV")
CT_DISPLAY_ORDER <- c("AT2", "LAE", "FB", "MV")

SIGNIFICANCE_THRESHOLD <- 0.05
LOGFC_THRESHOLD        <- 0
TOP_HITS_N             <- 50

cat("Input DIE directory:", INPUT_DIE_DIR, "\n")
cat("Output directory:",     OUTPUT_DIR,    "\n\n")

```

# Load DIE Results

```{r load-die}
# File pattern: nmd_mashr_die_<ct>_<date>.csv
# Splitting by underscore: parts[1]="nmd", [2]="mashr", [3]="die",
# [4]=<ct>, [5]="<date>.csv"  -> cell type lives at parts[4]
die_files <- list.files(
  INPUT_DIE_DIR,
  pattern = paste0("^nmd_mashr_die_.*_", gsub("\\.", "\\\\.", DATE_STAMP), "\\.csv$"),
  full.names = TRUE
)

cat("Found", length(die_files), "DIE files:\n")
print(basename(die_files))

# Original transcript counts per file (before any filtering in this report)
original_counts <- map_df(die_files, ~{
  parts        <- str_split(basename(.x), "_")[[1]]
  celltype     <- parts[4]
  n_transcripts <- nrow(fread(.x, select = 1L, header = TRUE))
  tibble(cell_type_file = celltype,
         cell_type      = CELLTYPE_MAP[celltype],
         original_count = n_transcripts)
})

cat("Transcripts loaded per cell type (before report-level filtering):\n")
print(original_counts %>% select(cell_type_file, cell_type, original_count))

all_dge <- map_df(die_files, ~{
  df    <- read_csv(.x, show_col_types = FALSE)
  parts <- str_split(basename(.x), "_")[[1]]
  df$cell_type_file <- parts[4]
  df
})

all_dge$cell_type <- CELLTYPE_MAP[all_dge$cell_type_file]

# FILTER: drop rows whose file code is not in CELLTYPE_MAP. This is the
# load-bearing sanity check — if CELLTYPE_MAP does not match the filenames
# (e.g. stale at/dd keys), every row maps to NA and the report empties out.
n_before <- nrow(all_dge)
all_dge  <- all_dge %>% filter(!is.na(cell_type))
cat(sprintf(
  "FILTER: dropped %d row(s) with an unmapped cell-type code (check CELLTYPE_MAP vs filenames).\n  Rows before: %d / Rows now: %d.\n",
  n_before - nrow(all_dge), n_before, nrow(all_dge)))
stopifnot(nrow(all_dge) > 0)

# Standardize columns
cat("Columns present in DIE files:\n  ", paste(colnames(all_dge), collapse = ", "), "\n")

if ("txid" %in% colnames(all_dge) && !"transcript_id" %in% colnames(all_dge)) {
  colnames(all_dge)[colnames(all_dge) == "txid"] <- "transcript_id"
}

# Gene-id column may be written under several names by the pipeline. Resolve it
# to a canonical `gene_id`, failing loudly if none is found (a silent NULL
# assignment would leave gene_group_id missing and break the overlap chunks).
gene_id_candidates <- c("gene_id", "gene_group_id", "gene", "ensembl_gene_id",
                        "geneid", "ensembl_gene_id_version")
gid <- intersect(gene_id_candidates, colnames(all_dge))
if (length(gid) == 0) {
  stop("No gene-id column found in DIE files. Columns are: ",
       paste(colnames(all_dge), collapse = ", "))
}
if (!"gene_id" %in% colnames(all_dge)) {
  all_dge$gene_id <- all_dge[[gid[1]]]
  cat("Using '", gid[1], "' as gene_id.\n", sep = "")
}

if ("hgnc_symbol" %in% colnames(all_dge)) {
  all_dge <- all_dge %>%
    mutate(gene_symbol = ifelse(!is.na(hgnc_symbol) & hgnc_symbol != "",
                                hgnc_symbol, gene_id))
} else if (!"gene_symbol" %in% colnames(all_dge)) {
  all_dge$gene_symbol <- all_dge$gene_id
}
if (!"gene_group_id" %in% colnames(all_dge)) {
  all_dge$gene_group_id <- all_dge$gene_id
}

cat("Combined DIE data:", nrow(all_dge), "rows,",
    length(unique(all_dge$cell_type)), "cell types\n")
cat("Cell types present:", paste(sort(unique(all_dge$cell_type)), collapse = ", "), "\n")
cat("Example transcript IDs:\n  ",
    paste(head(unique(all_dge$transcript_id), 3), collapse = ", "), "\n")

n_with_symbol <- sum(!is.na(all_dge$gene_symbol) &
                     all_dge$gene_symbol != "" &
                     !grepl("^ENSG", all_dge$gene_symbol))
cat("Gene symbol coverage:", n_with_symbol, "/", nrow(all_dge),
    "(", round(100 * n_with_symbol / nrow(all_dge), 1), "%)\n")
```

# Summary Statistics

```{r summary-stats}
# Column definitions (first use):
cat("Table 1 column definitions:\n")
cat("  total_transcripts   = isoforms tested in this cell type (mashr output).\n")
cat("  sig_transcripts     = isoforms with lfsr < 0.05.\n")
cat("  nmd_susceptible     = significant isoforms with logFC > 0 (Smg1i-up; NMD susceptible).\n")
cat("  nmd_suppressed      = significant isoforms with logFC < 0.\n")
cat("  pct_sig             = sig_transcripts as % of total tested.\n")
cat("  pct_nmd_susceptible = NMD susceptible as % of significant.\n\n")

die_summary <- all_dge %>%
  group_by(cell_type) %>%
  summarize(
    sig_transcripts  = sum(adj.P.Val < SIGNIFICANCE_THRESHOLD, na.rm = TRUE),
    nmd_susceptible  = sum(adj.P.Val < SIGNIFICANCE_THRESHOLD &
                           logFC     > LOGFC_THRESHOLD, na.rm = TRUE),
    nmd_suppressed   = sum(adj.P.Val < SIGNIFICANCE_THRESHOLD &
                           logFC    < -LOGFC_THRESHOLD, na.rm = TRUE),
    mean_logFC_nmd   = mean(logFC[adj.P.Val < SIGNIFICANCE_THRESHOLD &
                                  logFC > LOGFC_THRESHOLD], na.rm = TRUE),
    median_logFC_nmd = median(logFC[adj.P.Val < SIGNIFICANCE_THRESHOLD &
                                    logFC > LOGFC_THRESHOLD], na.rm = TRUE),
    .groups = "drop"
  ) %>%
  left_join(original_counts %>% select(cell_type, original_count), by = "cell_type") %>%
  rename(total_transcripts = original_count) %>%
  mutate(
    pct_sig             = round(100 * sig_transcripts / total_transcripts, 2),
    pct_nmd_susceptible = round(100 * nmd_susceptible  / sig_transcripts,   2)
  ) %>%
  select(cell_type, total_transcripts, sig_transcripts, nmd_susceptible, nmd_suppressed,
         pct_sig, pct_nmd_susceptible, mean_logFC_nmd, median_logFC_nmd)

knitr::kable(die_summary, digits = 2,
             caption = "Table 1. DIE Summary: Significant Transcripts And NMD Susceptible Counts Per Cell Type")

write_tsv(die_summary, file.path(OUTPUT_DIR, paste0("mashr_die_summary_stats_", DATE_STAMP, ".tsv")))

```

## NMD Susceptible Transcripts

```{r nmd-responsive}
nmd_responsive <- all_dge %>%
  filter(logFC > LOGFC_THRESHOLD, adj.P.Val < SIGNIFICANCE_THRESHOLD)

cat(sprintf("Of %d tested isoform x cell-type rows, %d are NMD susceptible (logFC > 0 & lfsr < 0.05).\n",
            nrow(all_dge), nrow(nmd_responsive)))
cat(sprintf("Unique NMD susceptible transcripts (any cell type): %d.\n",
            n_distinct(nmd_responsive$transcript_id)))

nmd_summary <- nmd_responsive %>%
  group_by(cell_type) %>%
  summarize(
    n_transcripts = n(),
    n_genes       = n_distinct(gene_group_id, na.rm = TRUE),
    mean_logFC    = mean(logFC),
    median_logFC  = median(logFC),
    max_logFC     = max(logFC),
    .groups = "drop"
  )

knitr::kable(nmd_summary, digits = 2,
             caption = "Table 2. NMD Susceptible Transcripts By Cell Type")

write_tsv(nmd_responsive, file.path(OUTPUT_DIR, paste0("mashr_die_nmd_responsive_all_", DATE_STAMP, ".tsv")))
```

## Top Hits Per Cell Type

```{r top-hits, results='asis'}
table_num <- 3

for (ct in CT_DISPLAY_ORDER) {
  if (!ct %in% nmd_responsive$cell_type) next
  cat("\n###", ct, "\n\n")

  available_cols <- intersect(
    c("transcript_id", "gene_symbol", "gene_id", "logFC", "adj.P.Val", "nmd_responsive"),
    colnames(nmd_responsive)
  )

  top_hits <- nmd_responsive %>%
    filter(cell_type == ct) %>%
    arrange(adj.P.Val) %>%
    head(TOP_HITS_N) %>%
    select(all_of(available_cols))

  print(knitr::kable(
    top_hits, digits = 4, row.names = FALSE,
    caption = paste0("Table ", table_num, ". Top ", TOP_HITS_N,
                     " NMD Susceptible Transcripts In ", ct)
  ))
  cat("\n\n")
  table_num <- table_num + 1

  output_file <- file.path(OUTPUT_DIR,
                           paste0("mashr_die_top_hits_", tolower(ct), "_", DATE_STAMP, ".tsv"))
  write_tsv(top_hits, output_file)
}

```

```{r top10-parallel-with-transcript-name-standalone-fixed}
# ============================================================
# STANDALONE: Top 10 NMD susceptible isoforms per cell type
# Rebuilds nmd_responsive from DIE CSVs, then renders Option-4 table.
# FIXED: deduplicate gmap lookup; vectorized case_when on scalar inputs.
# ============================================================

library(tidyverse)
library(data.table)

# --- Config (match your main report) ---
DATE_STAMP    <- "2026.3.10"
INPUT_DIE_DIR <- "/udd/reyle/nmd_lungcells_2026/long_read/mashr"
OUTPUT_DIR    <- "/udd/reyle/nmd_lungcells_2026/tmp/mashr_die"
dir.create(OUTPUT_DIR, recursive = TRUE, showWarnings = FALSE)

CELLTYPE_MAP     <- c("at2" = "AT2", "lae" = "LAE", "fb" = "FB", "mv" = "MV")
CT_DISPLAY_ORDER <- c("AT2", "LAE", "FB", "MV")

SIGNIFICANCE_THRESHOLD <- 0.05
LOGFC_THRESHOLD        <- 0
TOP_N                  <- 10

# --- 1. Load DIE results ---
die_files <- list.files(
  INPUT_DIE_DIR,
  pattern = paste0("^nmd_mashr_die_.*_", gsub("\\.", "\\\\.", DATE_STAMP), "\\.csv$"),
  full.names = TRUE
)

all_dge <- map_df(die_files, ~{
  df    <- read_csv(.x, show_col_types = FALSE)
  parts <- str_split(basename(.x), "_")[[1]]
  df$cell_type_file <- parts[4]
  df
})

all_dge$cell_type <- CELLTYPE_MAP[all_dge$cell_type_file]
n_before <- nrow(all_dge)
all_dge  <- all_dge %>% filter(!is.na(cell_type))
cat(sprintf("FILTER: dropped %d row(s) with an unmapped cell-type code; %d rows retained.\n",
            n_before - nrow(all_dge), nrow(all_dge)))

if ("txid" %in% colnames(all_dge) && !"transcript_id" %in% colnames(all_dge)) {
  colnames(all_dge)[colnames(all_dge) == "txid"] <- "transcript_id"
}
gene_id_candidates <- c("gene_id", "gene_group_id", "gene", "ensembl_gene_id",
                        "geneid", "ensembl_gene_id_version")
gid <- intersect(gene_id_candidates, colnames(all_dge))
if (length(gid) == 0) {
  stop("No gene-id column found in DIE files. Columns are: ",
       paste(colnames(all_dge), collapse = ", "))
}
if (!"gene_id" %in% colnames(all_dge)) all_dge$gene_id <- all_dge[[gid[1]]]

if ("hgnc_symbol" %in% colnames(all_dge)) {
  all_dge <- all_dge %>%
    mutate(gene_symbol = ifelse(!is.na(hgnc_symbol) & hgnc_symbol != "",
                                hgnc_symbol, gene_id))
} else if (!"gene_symbol" %in% colnames(all_dge)) {
  all_dge$gene_symbol <- all_dge$gene_id
}
# Downstream overlap chunks group on gene_group_id, and this chunk overwrites
# the global nmd_responsive, so create it here too.
if (!"gene_group_id" %in% colnames(all_dge)) all_dge$gene_group_id <- all_dge$gene_id

nmd_responsive <- all_dge %>%
  filter(logFC > LOGFC_THRESHOLD, adj.P.Val < SIGNIFICANCE_THRESHOLD)

cat("Loaded", nrow(nmd_responsive), "NMD susceptible isoform-celltype rows\n")

# --- 2. Load gmap and build *deduplicated* lookup ---
GMAP_PATH <- "/udd/repjc/RESEARCH/REUSABLE_CODE/rds/gmap_txlevel_ENSGv115_2025.08.12.rds"
gmap <- readRDS(GMAP_PATH)

# Dedupe: one row per key. If multiple transcript_names exist for the same ID,
# keep the first non-empty one.
tx_lookup_ver <- gmap %>%
  as.data.frame() %>%
  transmute(
    tx_key          = ensembl_transcript_id_version,
    transcript_name = external_transcript_name
  ) %>%
  filter(!is.na(transcript_name), transcript_name != "",
         !is.na(tx_key), tx_key != "") %>%
  distinct(tx_key, .keep_all = TRUE)

tx_lookup_nover <- gmap %>%
  as.data.frame() %>%
  transmute(
    tx_key          = sub("\\.\\d+$", "", ensembl_transcript_id_version),
    transcript_name = external_transcript_name
  ) %>%
  filter(!is.na(transcript_name), transcript_name != "",
         !is.na(tx_key), tx_key != "") %>%
  distinct(tx_key, .keep_all = TRUE)

cat("Deduped gmap lookups: ", nrow(tx_lookup_ver), " versioned, ",
    nrow(tx_lookup_nover), " unversioned\n", sep = "")

# --- 3. Vectorized label builder (rowwise-safe) ---
make_clean_label <- function(tx_id, gene_symbol, transcript_name) {
  # Coerce all to character of equal length up front
  tx_id           <- as.character(tx_id)
  gene_symbol     <- as.character(gene_symbol)
  transcript_name <- as.character(transcript_name)

  # Element-wise novel detection
  novel_match <- regmatches(tx_id, regexpr("\\.novel(\\d+)$", tx_id))
  is_novel    <- grepl("\\.novel\\d+$", tx_id)
  novel_num   <- ifelse(is_novel, sub(".*\\.novel(\\d+)$", "\\1", tx_id), NA_character_)

  has_symbol  <- !is.na(gene_symbol) & gene_symbol != "" & !grepl("^ENSG", gene_symbol)
  has_txname  <- !is.na(transcript_name) & transcript_name != ""
  bare_ensg   <- sub("\\.\\d+$", "", tx_id)

  dplyr::case_when(
    is_novel & has_symbol   ~ paste0(gene_symbol, " (novel ", novel_num, ")"),
    is_novel                ~ paste0(bare_ensg, " (novel ", novel_num, ")"),
    has_txname              ~ transcript_name,
    grepl("^ENST", tx_id)   ~ tx_id,
    has_symbol              ~ gene_symbol,
    TRUE                    ~ bare_ensg
  )
}

# --- 4. Rank, join transcript_name, build display ---
top10_long <- nmd_responsive %>%
  group_by(cell_type) %>%
  arrange(adj.P.Val, desc(logFC), .by_group = TRUE) %>%
  slice_head(n = TOP_N) %>%
  ungroup() %>%
  # First try versioned key
  left_join(tx_lookup_ver, by = c("transcript_id" = "tx_key")) %>%
  # Then unversioned (under a different column name so we can coalesce)
  left_join(tx_lookup_nover %>% rename(transcript_name_nover = transcript_name),
            by = c("transcript_id" = "tx_key")) %>%
  mutate(
    transcript_name = coalesce(transcript_name, transcript_name_nover)
  ) %>%
  select(-transcript_name_nover) %>%
  mutate(
    label      = make_clean_label(transcript_id, gene_symbol, transcript_name),
    cell_value = sprintf("%s — %.2f", label, logFC)
  ) %>%
  group_by(cell_type) %>%
  mutate(rank = row_number()) %>%
  ungroup() %>%
  select(rank, cell_type, cell_value)

# Sanity: confirm exactly TOP_N rows per cell type
top10_long %>% count(cell_type) %>% print()

# --- 5. Pivot wide and render ---
top10_wide <- top10_long %>%
  pivot_wider(names_from = cell_type, values_from = cell_value) %>%
  select(rank, all_of(CT_DISPLAY_ORDER))

knitr::kable(
  top10_wide,
  col.names = c("Rank", CT_DISPLAY_ORDER),
  align     = c("c", rep("l", length(CT_DISPLAY_ORDER))),
  caption   = paste0(
    "Table 2. Top ", TOP_N, " NMD susceptible isoforms per cell type. ",
    "Cells show isoform identifier and Smg1i vs DMSO log2 fold change, ",
    "ranked by lfsr ascending. Novel isoforms denoted \"GENE (novel N)\"; ",
    "known isoforms shown by Ensembl transcript name (e.g., SRSF2-208)."
  )
)

# --- 6. Sanity check: which known ENST IDs failed lookup? ---
missing_txname <- nmd_responsive %>%
  group_by(cell_type) %>%
  arrange(adj.P.Val, desc(logFC), .by_group = TRUE) %>%
  slice_head(n = TOP_N) %>%
  ungroup() %>%
  filter(grepl("^ENST", transcript_id)) %>%
  left_join(tx_lookup_ver, by = c("transcript_id" = "tx_key")) %>%
  left_join(tx_lookup_nover %>% rename(transcript_name_nover = transcript_name),
            by = c("transcript_id" = "tx_key")) %>%
  mutate(transcript_name = coalesce(transcript_name, transcript_name_nover)) %>%
  filter(is.na(transcript_name)) %>%
  select(cell_type, transcript_id, gene_symbol)

if (nrow(missing_txname) > 0) {
  cat("\nKnown ENST IDs in top 10 with no transcript_name in gmap:\n")
  print(missing_txname)
} else {
  cat("\nAll known ENST IDs in top 10 resolved to transcript_name.\n")
}

# --- 7. Write TSV ---
write_tsv(
  top10_wide,
  file.path(OUTPUT_DIR, paste0("mashr_die_top10_parallel_option4_", DATE_STAMP, ".tsv"))
)
```

# Visualizations

## Volcano Plots

```{r volcano-plots, fig.width=10, fig.height=10}
volcano_data <- all_dge %>%
  mutate(
    neg_log10_lfsr = -log10(adj.P.Val + 1e-300),
    significance = case_when(
      logFC > 0 & adj.P.Val < SIGNIFICANCE_THRESHOLD ~ "NMD susceptible",
      logFC < 0 & adj.P.Val < SIGNIFICANCE_THRESHOLD ~ "NMD suppressed",
      TRUE                                            ~ "Not significant"
    )
  )
volcano_data$cell_type <- factor(volcano_data$cell_type, levels = CT_DISPLAY_ORDER)

p_volcano <- ggplot(volcano_data,
                    aes(x = logFC, y = neg_log10_lfsr, color = significance)) +
  geom_point(alpha = 0.5, size = 0.5) +
  facet_wrap(~ cell_type, ncol = 2) +
  scale_color_manual(values = c(
    "NMD susceptible" = "steelblue",
    "NMD suppressed"  = "coral",
    "Not significant" = "gray70"
  )) +
  coord_cartesian(ylim = c(0, 50)) +
  geom_hline(yintercept = -log10(SIGNIFICANCE_THRESHOLD), linetype = "dashed", color = "black") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "black") +
  labs(
    title    = "Figure 1: Volcano Plots",
    subtitle = "Posterior mean logFC vs −log10(lfsr); y-axis capped at 50",
    x        = "Posterior Mean log2 Fold Change (Smg1i Vs DMSO)",
    y        = expression(-log[10]("lfsr")),
    color    = "Status"
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "top",
        strip.text      = element_text(face = "bold", size = 13),
        plot.title      = element_text(size = 14, face = "bold", hjust = 0.5),
        plot.subtitle   = element_text(color = "grey40", hjust = 0.5))

ggsave(file.path(OUT_DIR_FIG, paste0("volcano_2x2_lr_iso_", DATE_STAMP, ".png")),
       p_volcano, width = 10, height = 10, dpi = 300, bg = "white")

print(p_volcano)

```

## logFC Distribution

```{r logfc-distribution, fig.width=10, fig.height=6}
median_logfc <- nmd_responsive %>%
  group_by(cell_type) %>%
  summarize(median_logFC = median(logFC), .groups = "drop")

nmd_responsive$cell_type <- factor(nmd_responsive$cell_type, levels = CT_DISPLAY_ORDER)
median_logfc$cell_type   <- factor(median_logfc$cell_type,   levels = CT_DISPLAY_ORDER)

p_dens <- ggplot(nmd_responsive, aes(x = logFC, fill = cell_type, color = cell_type)) +
  geom_density(alpha = 0.3, linewidth = 1) +
  geom_vline(data = median_logfc,
             aes(xintercept = median_logFC, color = cell_type),
             linetype = "dashed", linewidth = 0.8) +
  labs(title    = "Figure 3: logFC Distribution For NMD Susceptible Transcripts",
       subtitle = "Posterior mean logFC; dashed lines = per-cell-type median",
       x        = "Posterior Mean log2 Fold Change (Smg1i Vs DMSO)",
       y        = "Density",
       fill     = "Cell Type", color = "Cell Type") +
  theme_minimal(base_size = 13) +
  theme(legend.position = "top",
        plot.title      = element_text(face = "bold"),
        plot.subtitle   = element_text(color = "grey40"))

ggsave(file.path(OUT_DIR_FIG, paste0("logfc_density_lr_iso_", DATE_STAMP, ".png")),
       p_dens, width = 10, height = 6, dpi = 300, bg = "white")

print(p_dens)
```

# Isoform-Level Cell-Type Overlap

## Overlap Counts

```{r isoform-overlap}
cat("Cell-type sharing of NMD susceptible isoforms.\n")
cat("  n_celltypes = number of cell types in which a transcript is NMD susceptible.\n")
cat("  core        = NMD susceptible in all 4 cell types; cell-type-specific = exactly 1.\n\n")

transcript_celltype_counts <- nmd_responsive %>%
  select(cell_type, transcript_id, gene_group_id, gene_symbol) %>%
  group_by(transcript_id, gene_group_id, gene_symbol) %>%
  summarize(
    n_celltypes = n_distinct(cell_type),
    celltypes   = paste(sort(unique(as.character(cell_type))), collapse = ", "),
    .groups = "drop"
  ) %>%
  arrange(desc(n_celltypes), transcript_id)

isoform_overlap_summary <- transcript_celltype_counts %>%
  count(n_celltypes, name = "n_transcripts") %>%
  mutate(pct_transcripts = 100 * n_transcripts / sum(n_transcripts))

knitr::kable(isoform_overlap_summary, digits = 1,
             caption = "Isoform Overlap Summary By Number Of Cell Types")

p_overlap <- ggplot(isoform_overlap_summary,
                    aes(x = factor(n_celltypes), y = n_transcripts, fill = factor(n_celltypes))) +
  geom_col() +
  geom_text(aes(label = n_transcripts), vjust = -0.5, size = 4) +
  scale_fill_brewer(palette = "Set3") +
  labs(title = "Figure 4A: NMD Susceptible Transcripts By Cell-Type Consistency",
       x = "Number Of Cell Types With NMD Response",
       y = "Number Of Transcripts") +
  theme_minimal(base_size = 13) +
  theme(legend.position = "none",
        plot.title      = element_text(face = "bold"))

ggsave(file.path(OUT_DIR_FIG, paste0("isoform_overlap_lr_iso_", DATE_STAMP, ".png")),
       p_overlap, width = 8, height = 5, dpi = 300, bg = "white")

print(p_overlap)

write_tsv(transcript_celltype_counts,
          file.path(OUTPUT_DIR, paste0("mashr_die_transcript_celltype_overlap_", DATE_STAMP, ".tsv")))

```

## Multi-Cell-Type And Core Transcripts

```{r multi-and-core-transcripts}
multi_celltype_transcripts <- transcript_celltype_counts %>%
  filter(n_celltypes >= 2) %>%
  arrange(desc(n_celltypes))

cat("Transcripts present in 2+ cell types:", nrow(multi_celltype_transcripts), "\n")

# "Core" = present in all 4 cell types (was 4 before; matches new design)
core_transcripts <- transcript_celltype_counts %>%
  filter(n_celltypes >= 4) %>%
  arrange(desc(n_celltypes), transcript_id)

cat("Core NMD target transcripts (all 4 cell types):", nrow(core_transcripts), "\n")

if (nrow(core_transcripts) > 0) {
  knitr::kable(head(core_transcripts, 50),
               caption = "Core NMD Target Transcripts (Present In All 4 Cell Types)")

  write_tsv(core_transcripts,
            file.path(OUTPUT_DIR, paste0("mashr_die_core_nmd_transcripts_", DATE_STAMP, ".tsv")))

  detail_cols <- intersect(
    c("cell_type", "transcript_id", "gene_group_id", "gene_symbol", "logFC", "adj.P.Val"),
    colnames(nmd_responsive)
  )
  core_transcript_details <- nmd_responsive %>%
    filter(transcript_id %in% core_transcripts$transcript_id) %>%
    select(all_of(detail_cols)) %>%
    arrange(transcript_id, cell_type)
  write_tsv(core_transcript_details,
            file.path(OUTPUT_DIR, paste0("mashr_die_core_transcript_details_", DATE_STAMP, ".tsv")))
}
```

## Cell-Type-Specific Transcripts

```{r celltype-specific-transcripts, results='asis'}
celltype_specific_transcripts <- transcript_celltype_counts %>%
  filter(n_celltypes == 1) %>%
  left_join(nmd_responsive %>% select(transcript_id, cell_type, logFC, adj.P.Val),
            by = "transcript_id") %>%
  arrange(cell_type, desc(logFC))

cat("Cell-type-specific transcripts:", nrow(celltype_specific_transcripts), "\n\n")

specific_transcript_summary <- celltype_specific_transcripts %>%
  group_by(cell_type) %>%
  summarize(n_transcripts = n(),
            mean_logFC    = mean(logFC),
            .groups = "drop")

knitr::kable(specific_transcript_summary, digits = 2,
             caption = "Cell-Type-Specific NMD Susceptible Transcripts By Cell Type")

for (ct in CT_DISPLAY_ORDER) {
  ct_specific <- celltype_specific_transcripts %>%
    filter(cell_type == ct) %>%
    select(transcript_id, gene_symbol, logFC, adj.P.Val) %>%
    head(10)
  if (nrow(ct_specific) > 0) {
    cat("\n**", ct, "** (",
        sum(celltype_specific_transcripts$cell_type == ct), " total)\n\n")
    print(knitr::kable(ct_specific, digits = 4, row.names = FALSE))
    cat("\n")
  }
}

write_tsv(celltype_specific_transcripts,
          file.path(OUTPUT_DIR, paste0("mashr_die_celltype_specific_transcripts_", DATE_STAMP, ".tsv")))

```

```{r isoform-overlap-interpretation}
total_transcripts  <- n_distinct(nmd_responsive$transcript_id)
transcripts_in_all <- sum(transcript_celltype_counts$n_celltypes == length(CT_DISPLAY_ORDER))
transcripts_in_1   <- sum(transcript_celltype_counts$n_celltypes == 1)

cat("**Summary:**\n")
cat("- Total unique NMD susceptible transcripts:", total_transcripts, "\n")
cat("- Transcripts in all 4 cell types:", transcripts_in_all,
    "(", round(100 * transcripts_in_all / total_transcripts, 1), "%)\n")
cat("- Transcripts in only 1 cell type:", transcripts_in_1,
    "(", round(100 * transcripts_in_1 / total_transcripts, 1), "%)\n")
```

# Gene-Level Cell-Type Overlap

## Overlap Counts

```{r gene-overlap}
cat("Cell-type sharing of genes with NMD susceptible isoforms.\n")
cat("  n_celltypes = number of cell types in which a gene has an NMD susceptible isoform.\n")
cat("  core        = susceptible in all 4 cell types; cell-type-specific = exactly 1.\n\n")

nmd_responsive_genes <- nmd_responsive %>%
  select(cell_type, gene_group_id, gene_symbol) %>%
  distinct() %>%
  filter(!is.na(gene_group_id), gene_group_id != "")

gene_celltype_counts <- nmd_responsive_genes %>%
  group_by(gene_group_id, gene_symbol) %>%
  summarize(
    n_celltypes = n_distinct(cell_type),
    celltypes   = paste(sort(unique(as.character(cell_type))), collapse = ", "),
    .groups = "drop"
  ) %>%
  arrange(desc(n_celltypes), gene_symbol)

overlap_summary <- gene_celltype_counts %>%
  count(n_celltypes, name = "n_genes") %>%
  mutate(pct_genes = 100 * n_genes / sum(n_genes))

knitr::kable(overlap_summary, digits = 1, caption = "Gene Overlap Summary By Number Of Cell Types")

p_gene_overlap <- ggplot(overlap_summary,
                         aes(x = factor(n_celltypes), y = n_genes, fill = factor(n_celltypes))) +
  geom_col() +
  geom_text(aes(label = n_genes), vjust = -0.5, size = 4) +
  scale_fill_brewer(palette = "Set3") +
  labs(title = "NMD Susceptible Genes By Cell-Type Consistency",
       x = "Number Of Cell Types With NMD Response",
       y = "Number Of Genes") +
  theme_minimal(base_size = 13) +
  theme(legend.position = "none",
        plot.title      = element_text(face = "bold"))

ggsave(file.path(OUT_DIR_FIG, paste0("gene_overlap_lr_iso_", DATE_STAMP, ".png")),
       p_gene_overlap, width = 8, height = 5, dpi = 300, bg = "white")

print(p_gene_overlap)

write_tsv(gene_celltype_counts,
          file.path(OUTPUT_DIR, paste0("mashr_die_gene_celltype_overlap_", DATE_STAMP, ".tsv")))

```

## Core NMD Target Genes

```{r core-nmd-genes}
core_genes <- gene_celltype_counts %>%
  filter(n_celltypes >= length(CT_DISPLAY_ORDER)) %>%
  arrange(desc(n_celltypes), gene_symbol)

cat("Core NMD target genes (all", length(CT_DISPLAY_ORDER), "cell types):", nrow(core_genes), "\n")

if (nrow(core_genes) > 0) {
  knitr::kable(head(core_genes, 50), caption = "Core NMD Target Genes (All Cell Types)")

  write_tsv(core_genes, file.path(OUTPUT_DIR, paste0("mashr_die_core_nmd_genes_", DATE_STAMP, ".tsv")))

  core_gene_transcripts <- nmd_responsive %>%
    filter(gene_group_id %in% core_genes$gene_group_id) %>%
    select(cell_type, gene_group_id, gene_symbol, transcript_id, logFC, adj.P.Val) %>%
    arrange(gene_group_id, cell_type, desc(logFC))
  write_tsv(core_gene_transcripts,
            file.path(OUTPUT_DIR, paste0("mashr_die_core_gene_transcripts_", DATE_STAMP, ".tsv")))
}
```

## Cell-Type-Specific Genes

```{r celltype-specific-genes, results='asis'}
celltype_specific_genes <- gene_celltype_counts %>%
  filter(n_celltypes == 1) %>%
  left_join(nmd_responsive_genes %>% select(gene_group_id, cell_type),
            by = "gene_group_id") %>%
  arrange(cell_type, gene_symbol)

specific_summary <- celltype_specific_genes %>%
  group_by(cell_type) %>%
  summarize(n_genes = n(), .groups = "drop")

knitr::kable(specific_summary, caption = "Cell-Type-Specific NMD Susceptible Genes")

for (ct in CT_DISPLAY_ORDER) {
  ct_specific <- celltype_specific_genes %>%
    filter(cell_type == ct) %>%
    select(gene_symbol, gene_group_id) %>%
    head(20)
  if (nrow(ct_specific) > 0) {
    cat("\n**", ct, "** (",
        sum(celltype_specific_genes$cell_type == ct), " total)\n\n")
    print(knitr::kable(ct_specific, row.names = FALSE))
    cat("\n")
  }
}

write_tsv(celltype_specific_genes,
          file.path(OUTPUT_DIR, paste0("mashr_die_celltype_specific_genes_", DATE_STAMP, ".tsv")))

```

```{r overlap-interpretation}
total_genes  <- n_distinct(nmd_responsive_genes$gene_group_id)
genes_in_all <- sum(gene_celltype_counts$n_celltypes == length(CT_DISPLAY_ORDER))
genes_in_1   <- sum(gene_celltype_counts$n_celltypes == 1)

cat("**Summary:**\n")
cat("- Total unique genes with NMD susceptible transcripts:", total_genes, "\n")
cat("- Genes in all 4 cell types:", genes_in_all,
    "(", round(100 * genes_in_all / total_genes, 1), "%)\n")
cat("- Genes in only 1 cell type:", genes_in_1,
    "(", round(100 * genes_in_1 / total_genes, 1), "%)\n")
```

# Session Info

```{r session-info}
sessionInfo()
```
