---
title: "Isoform-Level DIE: limma-voom + mashr"
author: "Leshem, Yul"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: lumen
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
    code_folding: hide
---

Code location: `/udd/reyle/nmd_lungcells_2026/code/final/Isoform_Level_Quantification.Rmd`

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, dev = "pdf")
options(bitmapType = "cairo")
```

# Overview

This report produces isoform-level differential expression results for the
NMD-inhibitor (Smg1i) vs DMSO contrast across four primary human lung cell
types (AT2, LAE, FB, MV), using long-read PacBio Kinnex/MAS-Seq counts
quantified by Isocall and filtered through SQANTI3.

Cell-type codes in the raw data use the internal labels DD (= LAE) and AT (= AT2);
sample names keep these prefixes to match the count-matrix columns, and the cell-type
factor is relabeled to LAE/AT2 for analysis and all saved outputs (matching the
short-read pipeline).

```{r libraries}
suppressPackageStartupMessages({
  library(data.table)
  library(ggplot2)
  library(edgeR)
  library(limma)
  library(DT)
  library(BiocParallel)
  library(htmltools)
  library(dplyr)
  library(ggrepel)
  library(gridExtra)
  library(tidyr)
})

# Bioconductor namespace masking guard
select <- dplyr::select
filter <- dplyr::filter
rename <- dplyr::rename
count  <- dplyr::count
slice  <- dplyr::slice

# Output paths
OUT_DIR_FIGURES <- "/udd/reyle/nmd_lungcells_2026/final_figures/lr_iso"
DATE_STAMP      <- "4.30.26"
dir.create(OUT_DIR_FIGURES, recursive = TRUE, showWarnings = FALSE)

# Note: per-cell-type volcano plots, top-50 tables, and the logFC-correlation
# heatmap are produced in the downstream LR summary RMDs, not here.
```

# Step 1 — Load Counts and GTF

```{r load-data}
counts_iso <- fread(
  "/proj/regeps/regep00/studies/ExternalCellLines/analyses/repjc/Randell_Lung_Cells_2025/code/sqanti3_by_chromosome/projects/nmd_lungcells/results/nmd_lungcells_filtered.count_matrix.txt",
  data.table = FALSE
)
cat(sprintf("Loading isoform count matrix. Dimensions: %d isoforms x %d sample columns.\n",
            nrow(counts_iso), ncol(counts_iso) - 1L))

stopifnot("Duplicate transcript IDs in count matrix" = sum(duplicated(counts_iso$id)) == 0)

n_na <- sum(is.na(counts_iso[, -1]))
if (n_na > 0) cat(sprintf("WARNING: count matrix contains %d NA values (treated as 0).\n", n_na))

gtf_raw <- fread(
  "/proj/regeps/regep00/studies/ExternalCellLines/analyses/repjc/Randell_Lung_Cells_2025/code/sqanti3_by_chromosome/projects/nmd_lungcells/results/nmd_lungcells_filtered.gtf",
  header = FALSE, sep = "\t", quote = "",
  col.names = c("seqname", "source", "type", "start", "end",
                "score", "strand", "frame", "attributes")
)
cat("Loaded GTF. Features by type:\n")
print(table(gtf_raw$type))
```

# Step 2 — Build Transcript / Gene Annotation

Parse the GTF to get transcript-to-gene mapping, exonic length per transcript,
and (where available) transcript biotype. Symbols come from the project's
gmap RDS; biotype is preferred from the GTF and falls back to biomaRt only if
the GTF lacks that attribute.

```{r annotation}
parse_attr <- function(attrs, key) {
  pattern <- paste0(key, ' "([^"]+)"')
  m       <- regmatches(attrs, regexpr(pattern, attrs, perl = TRUE))
  out     <- rep(NA_character_, length(attrs))
  has     <- nchar(m) > 0
  out[has] <- gsub(paste0(key, ' "|"'), "", m[has])
  out
}

# -------- Transcript-to-gene mapping --------
gtf_tx <- gtf_raw[gtf_raw$type == "transcript", ]
gtf_tx$transcript_id <- parse_attr(gtf_tx$attributes, "transcript_id")
gtf_tx$gene_id       <- parse_attr(gtf_tx$attributes, "gene_id")

tx2gene <- unique(gtf_tx[!is.na(gtf_tx$transcript_id), c("transcript_id", "gene_id")])

# Defensive: collapse transcripts mapping to multiple genes (rare)
tx_multi_gene <- tx2gene %>% group_by(transcript_id) %>% filter(n() > 1)
if (nrow(tx_multi_gene) > 0) {
  cat(sprintf("FILTER: %d transcript(s) map to multiple genes; keeping the FIRST gene for each.\n",
              n_distinct(tx_multi_gene$transcript_id)))
  tx2gene <- tx2gene[!duplicated(tx2gene$transcript_id), ]
}

# -------- Exonic length per transcript --------
gtf_exon <- gtf_raw[gtf_raw$type == "exon", ]
gtf_exon$transcript_id <- parse_attr(gtf_exon$attributes, "transcript_id")
gtf_exon$width         <- gtf_exon$end - gtf_exon$start + 1

exon_lengths <- gtf_exon %>%
  group_by(transcript_id) %>%
  summarise(exonic_length = sum(width), .groups = "drop")

gdf <- merge(tx2gene, exon_lengths, by = "transcript_id", all.x = TRUE)
names(gdf)[1:2] <- c("txid", "gene_id")

# -------- Add gene symbol via project gmap --------
gmap            <- readRDS("/udd/repjc/RESEARCH/REUSABLE_CODE/rds/gmap_txlevel_ENSGv115_2025.08.12.rds")
gene_symbol_map <- unique(gmap[, c("ensembl_gene_id_version", "hgnc_symbol")])
gdf <- merge(gdf, gene_symbol_map,
             by.x = "gene_id", by.y = "ensembl_gene_id_version",
             all.x = TRUE)

# -------- Add transcript biotype (prefer GTF, fall back to biomaRt) --------
sample_attrs <- gtf_tx$attributes[1:5]
has_transcript_type <- any(grepl('transcript_type', sample_attrs))
has_gene_type       <- any(grepl('gene_type',       sample_attrs))

if (has_transcript_type) {
  biotype_df <- unique(data.frame(
    txid = gtf_tx$transcript_id,
    transcript_biotype = parse_attr(gtf_tx$attributes, "transcript_type"),
    stringsAsFactors = FALSE
  ))
  gdf <- merge(gdf, biotype_df, by = "txid", all.x = TRUE)
} else if (has_gene_type) {
  biotype_df <- unique(data.frame(
    txid = gtf_tx$transcript_id,
    transcript_biotype = parse_attr(gtf_tx$attributes, "gene_type"),
    stringsAsFactors = FALSE
  ))
  gdf <- merge(gdf, biotype_df, by = "txid", all.x = TRUE)
} else {
  if (!require("biomaRt", quietly = TRUE)) BiocManager::install("biomaRt")
  library(biomaRt)
  ensembl <- useEnsembl(biomart = "genes", dataset = "hsapiens_gene_ensembl")

  tx_ids_noversion <- unique(sub("\\..*", "", gdf$txid))
  batch_size       <- 5000
  biotype_list     <- list()
  for (b in seq(1, length(tx_ids_noversion), by = batch_size)) {
    batch_end <- min(b + batch_size - 1, length(tx_ids_noversion))
    biotype_list[[ceiling(b / batch_size)]] <- getBM(
      attributes = c("ensembl_transcript_id", "transcript_biotype"),
      filters    = "ensembl_transcript_id",
      values     = tx_ids_noversion[b:batch_end],
      mart       = ensembl
    )
  }
  biotype_df <- unique(do.call(rbind, biotype_list))

  gdf$txid_noversion <- sub("\\..*", "", gdf$txid)
  gdf <- merge(gdf, biotype_df,
               by.x = "txid_noversion", by.y = "ensembl_transcript_id",
               all.x = TRUE)
  gdf$txid_noversion <- NULL
}

if (!"transcript_biotype" %in% colnames(gdf)) gdf$transcript_biotype <- NA

cat(sprintf("Annotation built: %d transcripts. Symbol coverage %d/%d; biotype coverage %d/%d.\n",
            nrow(gdf), sum(!is.na(gdf$hgnc_symbol)), nrow(gdf),
            sum(!is.na(gdf$transcript_biotype)), nrow(gdf)))
cat(sprintf("Count-matrix transcripts found in annotation: %d of %d.\n",
            sum(counts_iso$id %in% gdf$txid), nrow(counts_iso)))

rm(gtf_raw, gtf_tx, gtf_exon)
```

# Step 3 — Reformat Sample Names and Build DGEList

Original sample names from the count matrix look like `Sample##_CT_ID_Treatment`.
We strip the `Sample##` prefix and rebuild as `ID_Treatment_CT`, keeping the
internal CT code (DD/AT) in the name so downstream regex parsing can pull the
cell type off the end. ALI/submerged cultures (DO and the three DD ALI donors)
are dropped, and the cell-type factor is then relabeled DD→LAE and AT→AT2 for
analysis (sample names keep their DD/AT prefixes).

```{r prepare-counts}
rownames(counts_iso) <- counts_iso$id
counts_iso$id        <- NULL
count_mat            <- round(as.matrix(counts_iso))

cat(sprintf("Formatted isoform count matrix: %d isoforms x %d samples.\n",
            nrow(count_mat), ncol(count_mat)))
cat(sprintf("  Total counts: %s | median library size: %s | all-zero isoforms: %d.\n",
            format(sum(count_mat), big.mark = ","),
            format(median(colSums(count_mat)), big.mark = ","),
            sum(rowSums(count_mat) == 0)))

# Reformat sample names: Sample##_CT_ID_Treatment -> ID_Treatment_CT
nms <- gsub("^Sample\\d+_", "", colnames(count_mat))
new_nms <- sapply(nms, function(x) {
  parts     <- strsplit(x, "_")[[1]]
  treatment <- parts[length(parts)]
  sample_id <- parts[length(parts) - 1]
  ct        <- paste(parts[1:(length(parts) - 2)], collapse = "_")
  paste(sample_id, treatment, ct, sep = "_")
})
colnames(count_mat) <- new_nms
stopifnot("Duplicate sample names after reformatting" = sum(duplicated(colnames(count_mat))) == 0)
cat("Reformatted sample names (ID_Treatment_CT). Examples:\n  ",
    paste(head(colnames(count_mat), 5), collapse = ", "), "\n")
```

```{r build-pheno-and-dge}
pheno <- data.frame(sample_name = colnames(count_mat), stringsAsFactors = FALSE)
pheno$id        <- sapply(strsplit(pheno$sample_name, "_"), function(x) x[1])
pheno$treatment <- sapply(strsplit(pheno$sample_name, "_"), function(x) x[2])
pheno$ct        <- sapply(strsplit(pheno$sample_name, "_"), function(x) paste(x[3:length(x)], collapse = "_"))
pheno$ct[pheno$ct == "DO"] <- "DO_ALI"   # DD_ALI is already labeled as such in the source sample names
cat("Cell-type counts before ALI handling:\n"); print(table(pheno$ct))

# ---- drop ALI / submerged cultures (DO_ALI and DD_ALI are already labeled in the data) ----
n_before  <- nrow(pheno)
drop_ali  <- pheno$ct %in% c("DO_ALI", "DD_ALI")
pheno     <- pheno[!drop_ali, ]
count_mat <- count_mat[, colnames(count_mat) %in% pheno$sample_name]
cat(sprintf("FILTER: dropped DO_ALI and DD_ALI samples (ALI/submerged cultures, not part of the 4-cell-type analysis).\n  Samples before: %d / Samples now: %d.\n",
            n_before, nrow(pheno)))

# ---- analysis labels: DD -> LAE, AT -> AT2 (sample NAMES keep DD/AT prefix to match counts) ----
n_lae <- sum(pheno$ct == "DD"); n_at2 <- sum(pheno$ct == "AT")
pheno$ct[pheno$ct == "DD"] <- "LAE"
pheno$ct[pheno$ct == "AT"] <- "AT2"
cat(sprintf("Relabeled cell types for analysis: DD -> LAE (%d samples), AT -> AT2 (%d samples).\n  Sample names keep their DD/AT prefixes to match count columns.\n",
            n_lae, n_at2))

pheno$ct        <- relevel(factor(pheno$ct),        ref = "LAE")
pheno$treatment <- relevel(factor(pheno$treatment), ref = "DMSO")
cat(sprintf("Final design: %d samples; cell-type reference = %s, treatment reference = %s.\n",
            nrow(pheno), levels(pheno$ct)[1], levels(pheno$treatment)[1]))
cat("Samples per cell type x treatment:\n"); print(table(pheno$ct, pheno$treatment))

# Match annotation to count matrix
gdf_matched <- gdf[match(rownames(count_mat), gdf$txid), ]
n_unmatched <- sum(is.na(gdf_matched$txid))
if (n_unmatched > 0) {
  cat(sprintf("Unmatched transcripts (novel isoforms, no annotation): %d.\n", n_unmatched))
  gdf_matched$txid[is.na(gdf_matched$txid)] <- rownames(count_mat)[is.na(gdf_matched$txid)]
}
rownames(gdf_matched) <- rownames(count_mat)

dge <- DGEList(count_mat, samples = pheno, genes = gdf_matched)
dge_nofilt <- dge

dir.create("/udd/reyle/nmd_lungcells_2026/long_read", recursive = TRUE, showWarnings = FALSE)
saveRDS(dge_nofilt, file = "/udd/reyle/nmd_lungcells_2026/long_read/dge_isoform_longread_2026.3.3.rds")
cat(sprintf("Built DGEList: %d isoforms x %d samples (before expression filter).\n",
            nrow(dge$counts), ncol(dge$counts)))
```

# Step 4 — Filter and Normalize

```{r filter-normalize}
design_all <- model.matrix(~ ct + treatment + ct:treatment, data = dge$samples)
cat(sprintf("Design matrix rank: %d of %d columns.\n", qr(design_all)$rank, ncol(design_all)))

n_before <- nrow(dge)
keep <- filterByExpr(dge, design = design_all, min.count = 5, min.total.count = 10, min.prop = 0)
dge  <- dge[keep, , keep.lib.sizes = FALSE]
cat(sprintf("FILTER: removed low-expression isoforms (filterByExpr; min.count=5, min.total.count=10).\n  Isoforms before: %d / Isoforms now: %d (%.1f%% kept).\n",
            n_before, nrow(dge), 100 * nrow(dge) / n_before))

dge <- calcNormFactors(dge, method = "TMM")
cat(sprintf("Applied TMM normalization. Norm factors: range %.3f-%.3f, median %.3f.\n",
            min(dge$samples$norm.factors), max(dge$samples$norm.factors),
            median(dge$samples$norm.factors)))
extreme_nf <- dge$samples$norm.factors < 0.7 | dge$samples$norm.factors > 1.3
if (any(extreme_nf)) {
  cat(sprintf("  WARNING: extreme norm factors for: %s.\n",
              paste(dge$samples$sample_name[extreme_nf], collapse = ", ")))
}

saveRDS(dge,
        file = "/udd/reyle/nmd_lungcells_2026/long_read/dge_isoform_longread_filtered_2026.3.3.rds")
cat(sprintf("Final filtered + normalized DGEList: %d samples, %d isoforms.\n",
            ncol(dge), nrow(dge)))
```

# Step 5 — Diagnostic PCA

```{r pca}
v <- voom(dge, design = design_all, plot = TRUE)
cat(sprintf("Applied voom (log2-CPM, %d isoforms x %d samples). Running PCA for QC.\n",
            nrow(v$E), ncol(v$E)))

pca     <- prcomp(t(v$E), scale. = TRUE)
pca_df  <- as.data.frame(pca$x[, 1:2])
var_exp <- round(100 * (pca$sdev^2 / sum(pca$sdev^2))[1:2], 1)

pca_df$treatment <- dge$samples$treatment
pca_df$ct        <- dge$samples$ct
pca_df$id        <- dge$samples$id

p_pca <- ggplot(pca_df, aes(x = PC1, y = PC2, color = treatment, shape = ct)) +
  geom_point(size = 3, alpha = 0.8) +
  scale_color_manual(values = c("DMSO" = "orange", "Smg1i" = "purple")) +
  labs(title = "Pre-Adjustment PCA",
       subtitle = "Treatment (color), cell type (shape)",
       x = paste0("PC1 (", var_exp[1], "%)"),
       y = paste0("PC2 (", var_exp[2], "%)")) +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"),
        plot.subtitle = element_text(hjust = 0.5, color = "grey40"))

ggsave(file.path(OUT_DIR_FIGURES, paste0("pca_lr_iso_", DATE_STAMP, ".png")),
       p_pca, width = 10, height = 6, dpi = 300, bg = "white")
print(p_pca)

# Sample-distance outlier check
dist_mat  <- as.matrix(dist(t(v$E)))
mean_dist <- rowMeans(dist_mat)
outliers  <- names(mean_dist[mean_dist > mean(mean_dist) + 2 * sd(mean_dist)])
if (length(outliers) > 0) cat("Distance outliers (> 2 SD):\n  ", paste(outliers, collapse = ", "), "\n")
```

# Step 6 — Fit Model

Model: `~ ct + treatment + ct:treatment` with within-donor blocking via
`duplicateCorrelation`. LAE is the cell-type reference, DMSO is the treatment reference.

```{r fit-model}
corfit <- duplicateCorrelation(v, design_all, block = dge$samples$id)
cat(sprintf("Estimated within-donor correlation using duplicateCorrelation (blocked on donor): consensus = %.4f.\n",
            corfit$consensus.correlation))

fit <- lmFit(v, design_all,
             block       = dge$samples$id,
             correlation = corfit$consensus.correlation)
fit <- eBayes(fit)

plotSA(fit, main = "Residual SD vs Average Expression")
```

# Step 7 — Per-Cell-Type Smg1i Contrasts

```{r contrasts}
colnames(fit$design)       <- make.names(colnames(fit$design))
colnames(fit$coefficients) <- colnames(fit$design)

contrast_matrix <- makeContrasts(
  Smg1i_in_LAE = treatmentSmg1i,
  Smg1i_in_AT2 = treatmentSmg1i + ctAT2.treatmentSmg1i,
  Smg1i_in_FB  = treatmentSmg1i + ctFB.treatmentSmg1i,
  Smg1i_in_MV  = treatmentSmg1i + ctMV.treatmentSmg1i,
  levels = fit$design
)

fit2 <- contrasts.fit(fit, contrast_matrix)
fit2 <- eBayes(fit2)
cat("Fitted limma-voom model and defined per-cell-type contrasts: ",
    paste(colnames(contrast_matrix), collapse = ", "), ".\n", sep = "")

cat("\nSignificance overview (FDR < 0.05):\n")
for (j in seq_len(ncol(fit2$coefficients))) {
  tt     <- topTable(fit2, coef = j, number = Inf, sort.by = "none")
  n_sig  <- sum(tt$adj.P.Val <= 0.05)
  n_up   <- sum(tt$adj.P.Val <= 0.05 & tt$logFC > 0)
  n_down <- sum(tt$adj.P.Val <= 0.05 & tt$logFC < 0)
  cat(sprintf("  %-14s: %5d sig (%4d NMD susceptible / up, %4d down)\n",
              colnames(fit2$coefficients)[j], n_sig, n_up, n_down))
}
```

# Step 8 — Adaptive Shrinkage Across Cell Types

## 8.1 Setup

```{r mashr-setup}
if (!require("ashr",     quietly = TRUE)) install.packages("ashr")
if (!require("mashr",    quietly = TRUE)) BiocManager::install("mashr")
if (!require("reshape2", quietly = TRUE)) install.packages("reshape2")

library(mashr); library(ashr); library(reshape2)
```

## 8.2 Extract Effect Sizes and Standard Errors

```{r mashr-extract}
bhat <- fit2$coefficients
shat <- fit2$coefficients / fit2$t
n_before <- nrow(shat)

finite_rows <- apply(shat, 1, function(x) all(is.finite(x) & x > 0))
bhat <- bhat[finite_rows, ]
shat <- shat[finite_rows, ]
cat(sprintf("FILTER: kept isoforms with a valid (finite, positive) SE in all %d cell types.\n  Isoforms before: %d / Isoforms now: %d (dropped %d).\n",
            ncol(shat), n_before, nrow(bhat), n_before - sum(finite_rows)))

stopifnot("bhat / shat dim mismatch" = identical(dim(bhat), dim(shat)))
cat(sprintf("Effect-size and SE matrices ready for mashr: %d isoforms x %d cell types.\n",
            nrow(bhat), ncol(bhat)))
```

## 8.3 Estimate Null Correlation

```{r mashr-correlation}
mash_data_init <- mash_set_data(Bhat = bhat, Shat = shat)
m_1by1         <- mash_1by1(mash_data_init)
strong         <- get_significant_results(m_1by1, thresh = 0.05)
cat(sprintf("Found %d strong signals (gene-by-gene ash, lfsr < 0.05) to learn data-driven covariance.\n",
            length(strong)))

# Smaller random subset than the gene-level pipeline (10k) — the isoform matrix is much larger.
set.seed(42)
n_random   <- min(10000, nrow(bhat))
random_idx <- sample(seq_len(nrow(bhat)), n_random)
cat(sprintf("Sampling %d random isoforms (seed 42) to estimate the cross-cell-type null correlation.\n", n_random))

mash_data_random <- mash_set_data(Bhat = bhat[random_idx, ], Shat = shat[random_idx, ])
Vhat <- estimate_null_correlation_simple(mash_data_random)
cat("Estimated null correlation (Vhat):\n"); print(round(Vhat, 3))

mash_data <- mash_set_data(Bhat = bhat, Shat = shat, V = Vhat)
```

## 8.4 Build Covariance Library

```{r mashr-covariances}
if (length(strong) >= 5) {
  U_pca <- cov_pca(mash_data, npc = min(5, ncol(bhat) - 1), subset = strong)
} else {
  U_pca <- list()
}
U_c   <- cov_canonical(mash_data)
U_all <- c(U_pca, U_c)
cat(sprintf("Built covariance library: %d data-driven (PCA) + %d canonical = %d total.\n",
            length(U_pca), length(U_c), length(U_all)))
```

## 8.5 Fit Model

```{r mashr-fit}
t_start <- Sys.time()
m       <- mash(mash_data, Ulist = U_all, outputlevel = 2)
cat(sprintf("Fitted mashr model in %.1f min.\n",
            as.numeric(difftime(Sys.time(), t_start, units = "mins"))))

pi_vals <- get_estimated_pi(m)
cat("Estimated mixture weights (components > 1%):\n")
print(round(sort(pi_vals[pi_vals > 0.01], decreasing = TRUE), 3))
```

## 8.6 Posterior Means and lfsr

```{r mashr-results}
pm   <- get_pm(m)
lfsr <- get_lfsr(m)
cat(sprintf("Posterior means and lfsr computed for %d isoforms x %d cell types.\n",
            nrow(pm), ncol(pm)))

n_sig_mash <- apply(lfsr, 2, function(x) sum(x < 0.05))
cat("\nSignificant isoforms (lfsr < 0.05):\n"); print(n_sig_mash)

cat("\nNMD susceptible isoforms (posterior mean logFC > 0 & lfsr < 0.05):\n")
for (j in seq_len(ncol(pm))) {
  ct_name <- colnames(pm)[j]
  n_nmd   <- sum(pm[, j] > 0 & lfsr[, j] < 0.05)
  cat(sprintf("  %-14s: %5d of %5d significant (%.1f%%)\n",
              ct_name, n_nmd, n_sig_mash[ct_name],
              100 * n_nmd / max(n_sig_mash[ct_name], 1)))
}

cat("\nShrunken (mashr) vs unshrunken (limma) significant counts:\n")
for (j in seq_len(ncol(bhat))) {
  ct_name   <- colnames(bhat)[j]
  limma_res <- topTable(fit2, coef = ct_name, number = Inf, sort.by = "none")
  n_limma   <- sum(limma_res$adj.P.Val < 0.05)
  cat(sprintf("  %-14s: unshrunken = %5d, shrunken = %5d (ratio: %.2f)\n",
              ct_name, n_limma, n_sig_mash[ct_name], n_sig_mash[ct_name] / max(n_limma, 1)))
}
```

## 8.7 Effect Sharing

```{r mashr-sharing}
n_sig_total <- length(get_significant_results(m, thresh = 0.05))
cat(sprintf("Total isoforms significant in >=1 cell type (lfsr < 0.05): %d.\n", n_sig_total))

if (n_sig_total >= 2) {
  sharing      <- get_pairwise_sharing(m, factor = 0.5, lfsr_thresh = 0.05)
  sharing_sign <- get_pairwise_sharing(m, factor = 0,   lfsr_thresh = 0.05)
  cat("Computed pairwise effect sharing (same sign & within 2-fold magnitude):\n")
  print(round(sharing, 2))
} else {
  sharing      <- NULL
  sharing_sign <- NULL
  cat("Too few significant isoforms for sharing analysis.\n")
}
```

## 8.8 Visualizations

```{r mashr-sharing-plot, fig.width=8, fig.height=7}
if (!is.null(sharing)) {
  CT_ORDER <- c("LAE", "AT2", "FB", "MV")

  sharing_clean <- sharing
  rownames(sharing_clean) <- gsub("Smg1i_in_", "", rownames(sharing_clean))
  colnames(sharing_clean) <- gsub("Smg1i_in_", "", colnames(sharing_clean))

  sharing_df <- melt(sharing_clean)
  colnames(sharing_df) <- c("CellType1", "CellType2", "Sharing")
  sharing_df$CellType1 <- factor(sharing_df$CellType1, levels = CT_ORDER)
  sharing_df$CellType2 <- factor(sharing_df$CellType2, levels = CT_ORDER)

  p_share <- ggplot(sharing_df, aes(x = CellType1, y = CellType2, fill = Sharing)) +
    geom_tile(color = "white", linewidth = 0.8) +
    geom_text(aes(label = round(Sharing, 2)),
              size = 4, fontface = "bold",
              color = ifelse(sharing_df$Sharing > 0.6, "white", "grey20")) +
    scale_fill_gradient(low = "white", high = "#185FA5",
                        limits = c(0, 1), name = "Sharing") +
    labs(title    = "Pairwise Effect Sharing",
         subtitle = "Same sign & within 2-fold magnitude (lfsr < 0.05)",
         x = NULL, y = NULL) +
    theme_minimal(base_size = 13) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1, face = "bold"),
          axis.text.y = element_text(face = "bold"),
          panel.grid  = element_blank(),
          plot.title  = element_text(face = "bold", hjust = 0.5),
          plot.subtitle = element_text(color = "grey40", hjust = 0.5))

  ggsave(file.path(OUT_DIR_FIGURES, paste0("sharing_heatmap_lr_iso_", DATE_STAMP, ".png")),
         p_share, width = 8, height = 7, dpi = 300, bg = "white")
  print(p_share)
}
```

## 8.9 Save Results

```{r mashr-save}
OUT_DIR_MASH    <- "/udd/reyle/nmd_lungcells_2026/long_read/mashr"
DATE_STAMP_MASH <- "2026.3.10"
dir.create(OUT_DIR_MASH, recursive = TRUE, showWarnings = FALSE)

saveRDS(m, file = file.path(OUT_DIR_MASH, paste0("mashr_isoform_model_", DATE_STAMP_MASH, ".rds")))

gene_anno <- dge$genes[, c("txid", "gene_id", "hgnc_symbol")]
gene_anno <- gene_anno[!duplicated(gene_anno$txid), ]

pm_df <- as.data.frame(pm)
pm_df$txid <- rownames(pm)
pm_df <- merge(pm_df, gene_anno, by = "txid", all.x = TRUE)
pm_df <- pm_df[, c("txid", "gene_id", "hgnc_symbol", colnames(pm))]
pm_df <- pm_df[order(pm_df$txid), ]
fwrite(pm_df, file = file.path(OUT_DIR_MASH,
                               paste0("mashr_isoform_posterior_means_", DATE_STAMP_MASH, ".csv")))

lfsr_df <- as.data.frame(lfsr)
lfsr_df$txid <- rownames(lfsr)
lfsr_df <- merge(lfsr_df, gene_anno, by = "txid", all.x = TRUE)
lfsr_df <- lfsr_df[, c("txid", "gene_id", "hgnc_symbol", colnames(lfsr))]
lfsr_df <- lfsr_df[order(lfsr_df$txid), ]
fwrite(lfsr_df, file = file.path(OUT_DIR_MASH,
                                 paste0("mashr_isoform_lfsr_", DATE_STAMP_MASH, ".csv")))

if (!is.null(sharing)) {
  fwrite(as.data.frame(sharing),
         file = file.path(OUT_DIR_MASH, paste0("mashr_isoform_sharing_2fold_", DATE_STAMP_MASH, ".csv")),
         row.names = TRUE)
  fwrite(as.data.frame(sharing_sign),
         file = file.path(OUT_DIR_MASH, paste0("mashr_isoform_sharing_sign_", DATE_STAMP_MASH, ".csv")),
         row.names = TRUE)
}
fwrite(as.data.frame(Vhat),
       file = file.path(OUT_DIR_MASH, paste0("mashr_isoform_null_correlation_", DATE_STAMP_MASH, ".csv")),
       row.names = TRUE)

for (j in seq_len(ncol(pm))) {
  ct_name <- colnames(pm)[j]
  res <- data.frame(
    txid           = rownames(pm),
    logFC          = pm[, j],
    adj.P.Val      = lfsr[, j],
    nmd_responsive = pm[, j] > 0 & lfsr[, j] < 0.05,   # column name kept for downstream compatibility
    stringsAsFactors = FALSE
  )
  res <- merge(res, gene_anno, by = "txid", all.x = TRUE)
  res <- res[order(res$adj.P.Val), ]

  ct_code <- tolower(gsub("Smg1i_in_", "", ct_name))   # at2, lae, fb, mv
  ct_code <- gsub("_", "", ct_code)

  fwrite(res, file = file.path(OUT_DIR_MASH,
                               paste0("nmd_mashr_die_", ct_code, "_", DATE_STAMP_MASH, ".csv")))

  n_nmd <- sum(res$nmd_responsive)
  n_sig <- sum(res$adj.P.Val < 0.05)
  cat(sprintf("  %-14s: %5d significant (lfsr < 0.05), %5d NMD susceptible\n",
              ct_name, n_sig, n_nmd))
}
cat(sprintf("Saved mashr outputs to %s: model RDS, posterior means, lfsr, sharing, null correlation,\n  and %d per-cell-type CSVs.\n",
            OUT_DIR_MASH, ncol(pm)))
```

# Session Info

```{r session-info}
sessionInfo()
```