---
title: "Long-Read Gene-Level DGE: limma-voom + mashr"
author: "Leshem, Yul"
date: "2026-03-10"
output:
  html_document:
    theme: lumen
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
    code_folding: hide
---

Code location: `/udd/reyle/nmd_lungcells_2026/code/final/long_read_dge_mashR.Rmd`

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(bitmapType = "cairo")
```

# Libraries

```{r libraries}
suppressPackageStartupMessages({
  library(data.table)
  library(ggplot2)
  library(edgeR)
  library(limma)
  library(DT)
  library(EnhancedVolcano)
  library(BiocParallel)
  library(htmltools)
  library(biomaRt)
  library(dplyr)
  library(ggrepel)
  library(gridExtra)
  library(parallel)
  library(topGO)
  library(org.Hs.eg.db)
})

# Bioconductor namespace masking guard
select <- dplyr::select
filter <- dplyr::filter
rename <- dplyr::rename
count  <- dplyr::count
slice  <- dplyr::slice

plot_volcano <- function(df, design_name, coef_name) {
  df <- df %>%
    mutate(
      negLogP = -log10(P.Value),
      sig     = ifelse(adj.P.Val < 0.05, "Significant", "Not Sig"),
      label   = ifelse(is.na(external_gene_name) | external_gene_name == "",
                       ensembl_gene_id_version, external_gene_name)
    )
  top_hits <- df %>%
    filter(sig == "Significant") %>%
    arrange(P.Value) %>%
    slice_head(n = 10)

  ggplot(df, aes(x = logFC, y = negLogP)) +
    geom_point(aes(color = sig), alpha = 0.6, size = 1.5) +
    scale_color_manual(values = c("Significant" = "red", "Not Sig" = "gray")) +
    geom_text_repel(data = top_hits, aes(label = label), size = 3,
                    max.overlaps = Inf, box.padding = 0.3,
                    point.padding = 0.3, segment.color = "grey50") +
    labs(title = paste0("Volcano Plot — ", design_name, "\nCoefficient: ", coef_name),
         x = "log2 Fold Change",
         y = expression(-log[10]("P-value"))) +
    theme_minimal(base_size = 14) +
    theme(legend.position = "bottom",
          plot.title = element_text(hjust = 0.5))
}

OUT_DIR_FIGURES <- "/udd/reyle/nmd_lungcells_2026/final_figures/lr_gene"
DATE_STAMP      <- "4.30.26"
dir.create(OUT_DIR_FIGURES, recursive = TRUE, showWarnings = FALSE)
```

# Step 1 — Load Counts and GTF

```{r load-counts-gtf}
counts_iso <- fread(
  "/proj/regeps/regep00/studies/ExternalCellLines/analyses/repjc/Randell_Lung_Cells_2025/code/sqanti3_by_chromosome/projects/nmd_lungcells/results/nmd_lungcells_filtered.count_matrix.txt",
  data.table = FALSE
)
cat("Count matrix:", nrow(counts_iso), "rows x", ncol(counts_iso), "columns\n")
n_dup <- sum(duplicated(counts_iso$id))
stopifnot("Duplicate transcript IDs found" = n_dup == 0)

gtf_raw <- fread(
  "/proj/regeps/regep00/studies/ExternalCellLines/analyses/repjc/Randell_Lung_Cells_2025/code/sqanti3_by_chromosome/projects/nmd_lungcells/results/nmd_lungcells_filtered.gtf",
  header = FALSE, sep = "\t", quote = "",
  col.names = c("seqname", "source", "type", "start", "end",
                "score", "strand", "frame", "attributes")
)
print(table(gtf_raw$type))
```

# Step 2 — Build Transcript / Gene Annotation

```{r annotation-transcript}
parse_attr <- function(attrs, key) {
  pattern <- paste0(key, ' "([^"]+)"')
  m       <- regmatches(attrs, regexpr(pattern, attrs, perl = TRUE))
  out     <- rep(NA_character_, length(attrs))
  has     <- nchar(m) > 0
  out[has] <- gsub(paste0(key, ' "|"'), "", m[has])
  out
}

gtf_tx <- gtf_raw[gtf_raw$type == "transcript", ]
gtf_tx$transcript_id <- parse_attr(gtf_tx$attributes, "transcript_id")
gtf_tx$gene_id       <- parse_attr(gtf_tx$attributes, "gene_id")

tx2gene <- unique(gtf_tx[!is.na(gtf_tx$transcript_id), c("transcript_id", "gene_id")])
tx_multi_gene <- tx2gene %>% group_by(transcript_id) %>% filter(n() > 1)
if (nrow(tx_multi_gene) > 0) tx2gene <- tx2gene[!duplicated(tx2gene$transcript_id), ]

gtf_exon <- gtf_raw[gtf_raw$type == "exon", ]
gtf_exon$transcript_id <- parse_attr(gtf_exon$attributes, "transcript_id")
gtf_exon$width         <- gtf_exon$end - gtf_exon$start + 1

exon_lengths <- gtf_exon %>%
  group_by(transcript_id) %>%
  summarise(exonic_length = sum(width), .groups = "drop")

gdf <- merge(tx2gene, exon_lengths, by = "transcript_id", all.x = TRUE)
names(gdf)[1:2] <- c("txid", "gene_id")

gmap            <- readRDS("/udd/repjc/RESEARCH/REUSABLE_CODE/rds/gmap_txlevel_ENSGv115_2025.08.12.rds")
gene_symbol_map <- unique(gmap[, c("ensembl_gene_id_version", "hgnc_symbol")])
gdf <- merge(gdf, gene_symbol_map,
             by.x = "gene_id", by.y = "ensembl_gene_id_version",
             all.x = TRUE)

cat("Total transcripts:", nrow(gdf), "\n")
cat("Symbol coverage:",   sum(!is.na(gdf$hgnc_symbol)), "/", nrow(gdf), "\n")

rm(gtf_raw, gtf_tx, gtf_exon)
```

# Step 3 — Aggregate Isoforms To Genes

```{r prepare-counts-gene}
rownames(counts_iso) <- counts_iso$id
counts_iso$id        <- NULL
count_mat <- round(as.matrix(counts_iso))

idx       <- match(rownames(count_mat), gdf$txid)
gene_ids  <- gdf$gene_id[idx]
has_gene  <- !is.na(gene_ids)
count_mat <- count_mat[has_gene, ]
rownames(count_mat) <- gene_ids[has_gene]

count_mat_gene <- rowsum(count_mat, group = rownames(count_mat))
cat("Aggregation:", sum(has_gene), "isoforms ->", nrow(count_mat_gene), "genes\n")

nms <- colnames(count_mat_gene)
nms <- gsub("^Sample\\d+_", "", nms)
new_nms <- sapply(nms, function(x) {
  parts     <- strsplit(x, "_")[[1]]
  treatment <- parts[length(parts)]
  sample_id <- parts[length(parts) - 1]
  ct        <- paste(parts[1:(length(parts) - 2)], collapse = "_")
  paste0(ct, sample_id, "_", treatment)
})
colnames(count_mat_gene) <- new_nms
```

# Step 4 — Phenotype Table and Cell-Type Filtering

```{r gene-annotation-pheno}
pheno              <- data.frame(name = colnames(count_mat_gene), stringsAsFactors = FALSE)
pheno$treatment    <- sapply(strsplit(pheno$name, "_"), function(x) x[length(x)])
ct_id              <- sapply(strsplit(pheno$name, "_"),
                             function(x) paste(x[1:(length(x) - 1)], collapse = "_"))
pheno$sample_id    <- substr(ct_id, nchar(ct_id) - 3, nchar(ct_id))
pheno$ct           <- substr(ct_id, 1, nchar(ct_id) - 4)

pheno$ct[pheno$ct == "DO"] <- "DO_ALI"

drop_ali <- pheno$ct %in% c("DO_ALI", "DD_ALI")
pheno          <- pheno[!drop_ali, ]
count_mat_gene <- count_mat_gene[, colnames(count_mat_gene) %in% pheno$name]

pheno$ct        <- relevel(factor(pheno$ct),        ref = "DD")
pheno$treatment <- relevel(factor(pheno$treatment), ref = "DMSO")

print(table(pheno$ct, pheno$treatment))

gene_anno <- gdf %>%
  group_by(gene_id) %>%
  summarize(
    n_isoforms  = n(),
    hgnc_symbol = hgnc_symbol[!is.na(hgnc_symbol)][1],
    .groups = "drop"
  )

gene_anno_matched <- gene_anno[match(rownames(count_mat_gene), gene_anno$gene_id), ]
unmatched_genes <- is.na(gene_anno_matched$gene_id)
if (any(unmatched_genes)) {
  gene_anno_matched$gene_id[unmatched_genes] <- rownames(count_mat_gene)[unmatched_genes]
}
rownames(gene_anno_matched) <- rownames(count_mat_gene)
gene_anno_matched$ensembl_gene_id_version <- gene_anno_matched$gene_id
gene_anno_matched$external_gene_name      <- gene_anno_matched$hgnc_symbol
```

# Step 5 — DGEList, Filter, TMM-Normalize

```{r dgelist-filter-normalize}
dge <- DGEList(count_mat_gene, samples = pheno, genes = gene_anno_matched)

dir.create("/udd/reyle/nmd_lungcells_2026/long_read", recursive = TRUE, showWarnings = FALSE)
saveRDS(dge, file = "/udd/reyle/nmd_lungcells_2026/long_read/dge_gene_longread_2026.3.10.rds")

design_all <- model.matrix(~ ct + treatment + ct:treatment, data = dge$samples)
keep <- filterByExpr(dge, design = design_all,
                     min.count = 5, min.total.count = 10, min.prop = 0)
cat("Genes retained:", sum(keep), "of", nrow(dge), "\n")

dge <- dge[keep, , keep.lib.sizes = FALSE]
dge <- calcNormFactors(dge, method = "TMM")

saveRDS(dge, file = "/udd/reyle/nmd_lungcells_2026/long_read/dge_gene_longread_filtered_2026.3.10.rds")
```

# Step 6 — Diagnostic PCA

```{r pca}
v <- voom(dge, design = design_all, plot = TRUE)

pca     <- prcomp(t(v$E), scale. = TRUE)
pca_df  <- as.data.frame(pca$x[, 1:2])
var_exp <- round(100 * (pca$sdev^2 / sum(pca$sdev^2))[1:2], 1)

pca_df$treatment <- dge$samples$treatment
pca_df$ct        <- dge$samples$ct
pca_df$sample_id <- dge$samples$sample_id

p_pca <- ggplot(pca_df, aes(x = PC1, y = PC2, color = treatment, shape = ct)) +
  geom_point(size = 3, alpha = 0.8) +
  scale_color_manual(values = c("DMSO" = "orange", "Smg1i" = "purple")) +
  labs(title = "Pre-Adjustment PCA",
       x = paste0("PC1 (", var_exp[1], "%)"),
       y = paste0("PC2 (", var_exp[2], "%)")) +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))

ggsave(file.path(OUT_DIR_FIGURES, paste0("pca_lr_gene_", DATE_STAMP, ".png")),
       p_pca, width = 10, height = 6, dpi = 300, bg = "white")

print(p_pca)
```

# Step 7 — Fit Model

```{r fit-model}
v      <- voom(dge, design = design_all, plot = TRUE)
corfit <- duplicateCorrelation(v, design_all, block = dge$samples$sample_id)
cat("Consensus within-donor correlation:",
    round(corfit$consensus.correlation, 4), "\n")

fit <- lmFit(v, design_all,
             block       = dge$samples$sample_id,
             correlation = corfit$consensus.correlation)
fit <- eBayes(fit)

plotSA(fit, main = "Residual SD vs Average Expression")
```

# Step 8 — Per-Cell-Type Smg1i Contrasts

```{r contrasts}
colnames(fit$design)       <- make.names(colnames(fit$design))
colnames(fit$coefficients) <- colnames(fit$design)

contrast_matrix <- makeContrasts(
  Smg1i_in_DD = treatmentSmg1i,
  Smg1i_in_AT = treatmentSmg1i + ctAT.treatmentSmg1i,
  Smg1i_in_FB = treatmentSmg1i + ctFB.treatmentSmg1i,
  Smg1i_in_MV = treatmentSmg1i + ctMV.treatmentSmg1i,
  levels = fit$design
)

fit2 <- contrasts.fit(fit, contrast_matrix)
fit2 <- eBayes(fit2)

cat("\nSignificance Overview (FDR < 0.05):\n")
for (j in seq_len(ncol(fit2$coefficients))) {
  tt     <- topTable(fit2, coef = j, number = Inf, sort.by = "none")
  n_sig  <- sum(tt$adj.P.Val <= 0.05)
  n_up   <- sum(tt$adj.P.Val <= 0.05 & tt$logFC > 0)
  n_down <- sum(tt$adj.P.Val <= 0.05 & tt$logFC < 0)
  cat(sprintf("  %-20s: %5d sig (%4d up, %4d down)\n",
              colnames(fit2$coefficients)[j], n_sig, n_up, n_down))
}
```

# Step 9 — Per-Cell-Type Results: Tables and Volcano Plots

```{r limma-results-extract}
contrast_names <- c("Smg1i_in_AT", "Smg1i_in_DD", "Smg1i_in_FB", "Smg1i_in_MV")
ct_labels      <- c("AT", "DD", "FB", "MV")

limma_tt <- lapply(contrast_names, function(cn) {
  topTable(fit2, coef = cn, number = Inf, sort.by = "none")
})
names(limma_tt) <- ct_labels

LIMMA_OUT_DIR <- "/udd/reyle/nmd_lungcells_2026/long_read/limma_gene"
dir.create(LIMMA_OUT_DIR, recursive = TRUE, showWarnings = FALSE)

for (ct in ct_labels) {
  tt_save <- limma_tt[[ct]][order(limma_tt[[ct]]$adj.P.Val), ]
  fwrite(tt_save,
         file = file.path(LIMMA_OUT_DIR,
                          paste0("nmd_limma_lr_dge_", tolower(ct), "_", DATE_STAMP, ".csv")))
}
```

```{r limma-volcano-2x2, fig.width=10, fig.height=10}
volcano_df <- bind_rows(lapply(ct_labels, function(ct) {
  tt <- limma_tt[[ct]]
  if (!"ensembl_gene_id_version" %in% colnames(tt)) tt$ensembl_gene_id_version <- rownames(tt)
  if (!"external_gene_name" %in% colnames(tt)) {
    tt$external_gene_name <- if ("hgnc_symbol" %in% colnames(tt)) tt$hgnc_symbol else NA_character_
  }
  tt$cell_type <- ct
  tt[, c("ensembl_gene_id_version", "external_gene_name",
         "logFC", "AveExpr", "t", "P.Value", "adj.P.Val", "B", "cell_type")]
}))

volcano_df <- volcano_df %>%
  mutate(
    negLogP = -log10(P.Value),
    sig     = ifelse(adj.P.Val < 0.05, "Significant", "Not Sig"),
    label   = ifelse(is.na(external_gene_name) | external_gene_name == "",
                     ensembl_gene_id_version, external_gene_name)
  )

top_labels <- volcano_df %>%
  filter(sig == "Significant") %>%
  group_by(cell_type) %>%
  arrange(P.Value, .by_group = TRUE) %>%
  slice_head(n = 10) %>%
  ungroup()

volcano_df$cell_type <- factor(volcano_df$cell_type, levels = c("AT", "DD", "FB", "MV"))
top_labels$cell_type <- factor(top_labels$cell_type, levels = c("AT", "DD", "FB", "MV"))

p_limma_volcano <- ggplot(volcano_df, aes(x = logFC, y = negLogP)) +
  geom_point(aes(color = sig), alpha = 0.5, size = 1.2) +
  scale_color_manual(values = c("Significant" = "red", "Not Sig" = "gray70")) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "black", linewidth = 0.3) +
  geom_text_repel(data = top_labels, aes(label = label),
                  size = 2.8, max.overlaps = Inf, box.padding = 0.3,
                  segment.color = "grey50") +
  facet_wrap(~ cell_type, ncol = 2) +
  labs(
    title    = "Volcano Plots — Smg1i Vs DMSO",
    subtitle = "Per-cell-type contrasts; red = adj.P.Val < 0.05, top 10 labeled per panel",
    x = "log2 Fold Change (Smg1i / DMSO)",
    y = expression(-log[10]("P-value"))
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "bottom",
        strip.text      = element_text(face = "bold", size = 13),
        plot.title      = element_text(face = "bold", hjust = 0.5),
        plot.subtitle   = element_text(color = "grey40", hjust = 0.5))

ggsave(file.path(OUT_DIR_FIGURES,
                 paste0("volcano_2x2_lr_gene_", DATE_STAMP, ".png")),
       p_limma_volcano, width = 10, height = 10, dpi = 300, bg = "white")

print(p_limma_volcano)
```

```{r limma-top-tables}
limma_dt_list <- list()

for (ct in ct_labels) {
  tt <- limma_tt[[ct]]
  if (!"ensembl_gene_id_version" %in% colnames(tt)) tt$ensembl_gene_id_version <- rownames(tt)
  if (!"external_gene_name" %in% colnames(tt)) {
    tt$external_gene_name <- if ("hgnc_symbol" %in% colnames(tt)) tt$hgnc_symbol else NA_character_
  }

  tt_top <- tt %>%
    dplyr::arrange(adj.P.Val) %>%
    dplyr::select(ensembl_gene_id_version, external_gene_name,
                  logFC, AveExpr, t, P.Value, adj.P.Val, B) %>%
    head(50)

  limma_dt_list[[ct]] <- htmltools::tagList(
    htmltools::h3(paste0(ct, " — Top 50 By FDR")),
    datatable(
      tt_top,
      options   = list(pageLength = 10, scrollX = TRUE),
      caption   = paste0("Top 50 — ", ct),
      elementId = paste0("limma_lr_dt_", ct)
    ) %>%
      formatRound(columns = c("logFC", "AveExpr", "t", "B"), digits = 3) %>%
      formatSignif(columns = c("P.Value", "adj.P.Val"),       digits = 3)
  )
}

htmltools::tagList(limma_dt_list)
```

# Step 10 — Adaptive Shrinkage Across Cell Types

## 10.1 Setup

```{r mashr-setup}
if (!require("ashr",     quietly = TRUE)) install.packages("ashr")
if (!require("mashr",    quietly = TRUE)) BiocManager::install("mashr")
if (!require("reshape2", quietly = TRUE)) install.packages("reshape2")

library(mashr); library(ashr); library(reshape2)
```

## 10.2 Extract Effect Sizes and Standard Errors

```{r mashr-extract}
bhat <- fit2$coefficients
shat <- fit2$coefficients / fit2$t

finite_rows <- apply(shat, 1, function(x) all(is.finite(x) & x > 0))
cat("Genes with valid SEs in all cell types:", sum(finite_rows), "of", nrow(shat), "\n")

bhat <- bhat[finite_rows, ]
shat <- shat[finite_rows, ]

stopifnot("bhat / shat dim mismatch" = identical(dim(bhat), dim(shat)))
cat("Final inputs:", nrow(bhat), "genes x", ncol(bhat), "cell types\n")
```

## 10.3 Estimate Null Correlation

```{r mashr-correlation}
mash_data_init <- mash_set_data(Bhat = bhat, Shat = shat)
m_1by1         <- mash_1by1(mash_data_init)
strong         <- get_significant_results(m_1by1, thresh = 0.05)
cat("Strong signals:", length(strong), "\n")

set.seed(42)
n_random   <- min(20000, nrow(bhat))
random_idx <- sample(seq_len(nrow(bhat)), n_random)

mash_data_random <- mash_set_data(Bhat = bhat[random_idx, ],
                                   Shat = shat[random_idx, ])
Vhat <- estimate_null_correlation_simple(mash_data_random)
print(round(Vhat, 3))

mash_data <- mash_set_data(Bhat = bhat, Shat = shat, V = Vhat)
```

## 10.4 Build Covariance Library

```{r mashr-covariances}
if (length(strong) >= 5) {
  U_pca <- cov_pca(mash_data, npc = min(5, ncol(bhat) - 1), subset = strong)
} else {
  U_pca <- list()
}
U_c   <- cov_canonical(mash_data)
U_all <- c(U_pca, U_c)
cat("Total covariance matrices:", length(U_all), "\n")
```

## 10.5 Fit Model

```{r mashr-fit}
t_start <- Sys.time()
m       <- mash(mash_data, Ulist = U_all, outputlevel = 2)
cat("Fit complete in",
    round(difftime(Sys.time(), t_start, units = "mins"), 1), "min\n")

pi_vals <- get_estimated_pi(m)
print(round(sort(pi_vals[pi_vals > 0.01], decreasing = TRUE), 3))
```

## 10.6 Posterior Means and lfsr

```{r mashr-results}
pm   <- get_pm(m)
lfsr <- get_lfsr(m)

n_sig_mash <- apply(lfsr, 2, function(x) sum(x < 0.05))
cat("\nSignificant Genes (lfsr < 0.05):\n"); print(n_sig_mash)

cat("\nNMD-Responsive (logFC > 0 & lfsr < 0.05):\n")
for (j in seq_len(ncol(pm))) {
  ct_name <- colnames(pm)[j]
  n_nmd   <- sum(pm[, j] > 0 & lfsr[, j] < 0.05)
  cat(sprintf("  %-20s: %5d\n", ct_name, n_nmd))
}

cat("\nComparison Vs Unshrunken:\n")
for (j in seq_len(ncol(bhat))) {
  ct_name   <- colnames(bhat)[j]
  limma_res <- topTable(fit2, coef = ct_name, number = Inf, sort.by = "none")
  n_limma   <- sum(limma_res$adj.P.Val < 0.05)
  cat(sprintf("  %-20s: unshrunken=%5d, shrunken=%5d\n",
              ct_name, n_limma, n_sig_mash[ct_name]))
}
```

## 10.7 Effect Sharing

```{r mashr-sharing}
n_sig_total <- length(get_significant_results(m, thresh = 0.05))

if (n_sig_total >= 2) {
  sharing      <- get_pairwise_sharing(m, factor = 0.5, lfsr_thresh = 0.05)
  sharing_sign <- get_pairwise_sharing(m, factor = 0,   lfsr_thresh = 0.05)
  cat("Pairwise sharing (same sign + within 2-fold):\n"); print(round(sharing, 2))
} else {
  sharing      <- NULL
  sharing_sign <- NULL
}
```

## 10.8 Visualizations

```{r mashr-sharing-plot, fig.width=8, fig.height=7}
if (!is.null(sharing)) {
  CT_ORDER <- c("DD", "AT", "FB", "MV")

  sharing_clean <- sharing
  rownames(sharing_clean) <- gsub("Smg1i_in_", "", rownames(sharing_clean))
  colnames(sharing_clean) <- gsub("Smg1i_in_", "", colnames(sharing_clean))

  sharing_df <- melt(sharing_clean)
  colnames(sharing_df) <- c("CellType1", "CellType2", "Sharing")
  sharing_df$CellType1 <- factor(sharing_df$CellType1, levels = CT_ORDER)
  sharing_df$CellType2 <- factor(sharing_df$CellType2, levels = CT_ORDER)

  p_share <- ggplot(sharing_df, aes(x = CellType1, y = CellType2, fill = Sharing)) +
    geom_tile(color = "white", linewidth = 0.8) +
    geom_text(aes(label = round(Sharing, 2)),
              size = 4, fontface = "bold",
              color = ifelse(sharing_df$Sharing > 0.6, "white", "grey20")) +
    scale_fill_gradient(low = "white", high = "#185FA5",
                        limits = c(0, 1), name = "Sharing") +
    labs(title    = "Pairwise Effect Sharing",
         subtitle = "Same sign & within 2-fold magnitude (lfsr < 0.05)",
         x = NULL, y = NULL) +
    theme_minimal(base_size = 13) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1, face = "bold"),
          axis.text.y = element_text(face = "bold"),
          panel.grid  = element_blank(),
          plot.title  = element_text(face = "bold", hjust = 0.5),
          plot.subtitle = element_text(color = "grey40", hjust = 0.5))

  ggsave(file.path(OUT_DIR_FIGURES,
                   paste0("sharing_heatmap_lr_gene_", DATE_STAMP, ".png")),
         p_share, width = 8, height = 7, dpi = 300, bg = "white")
  print(p_share)
}
```


```{r logfc-cor-heatmap, fig.width=8, fig.height=7}
CT_ORDER <- c("DD", "AT", "FB", "MV")

nmd_mask <- (pm > 0) & (lfsr < 0.05)
nmd_any  <- rowSums(nmd_mask) >= 1
cat("NMD-Responsive Genes:", sum(nmd_any), "of", nrow(pm), "\n")
stopifnot("Need >= 2 NMD-responsive genes" = sum(nmd_any) >= 2)

drop_col <- grep("DD_ALI|DO_ALI", colnames(pm), value = TRUE)
pm_nmd   <- pm[nmd_any, !colnames(pm) %in% drop_col]

cor_pearson <- cor(pm_nmd, method = "pearson")
rownames(cor_pearson) <- gsub("Smg1i_in_", "", rownames(cor_pearson))
colnames(cor_pearson) <- gsub("Smg1i_in_", "", colnames(cor_pearson))

cor_df <- melt(cor_pearson)
colnames(cor_df) <- c("CellType1", "CellType2", "Correlation")
cor_df$CellType1 <- factor(cor_df$CellType1, levels = CT_ORDER)
cor_df$CellType2 <- factor(cor_df$CellType2, levels = CT_ORDER)

p_cor <- ggplot(cor_df, aes(x = CellType1, y = CellType2, fill = Correlation)) +
  geom_tile(color = "white", linewidth = 0.8) +
  geom_text(aes(label = round(Correlation, 2)),
            size = 4, fontface = "bold",
            color = ifelse(cor_df$Correlation > 0.6, "white", "grey20")) +
  scale_fill_gradient2(low = "#D73027", mid = "white", high = "#185FA5",
                       midpoint = 0, limits = c(-1, 1), name = "Pearson r") +
  labs(title    = "logFC Correlation: NMD-Responsive Genes",
       subtitle = "Pearson r of posterior mean logFC across cell type pairs",
       x = NULL, y = NULL) +
  theme_minimal(base_size = 13) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, face = "bold"),
        axis.text.y = element_text(face = "bold"),
        panel.grid  = element_blank(),
        plot.title  = element_text(face = "bold", hjust = 0.5),
        plot.subtitle = element_text(color = "grey40", hjust = 0.5))

ggsave(file.path(OUT_DIR_FIGURES,
                 paste0("logfc_cor_heatmap_lr_gene_", DATE_STAMP, ".png")),
       p_cor, width = 8, height = 7, dpi = 300, bg = "white")

print(p_cor)
```

## 10.9 Save Results

```{r mashr-save}
OUT_DIR_MASH    <- "/udd/reyle/nmd_lungcells_2026/long_read/mashr_gene"
DATE_STAMP_MASH <- "2026.3.10"
dir.create(OUT_DIR_MASH, recursive = TRUE, showWarnings = FALSE)

saveRDS(m, file = file.path(OUT_DIR_MASH,
                            paste0("mashr_lr_gene_model_", DATE_STAMP_MASH, ".rds")))

gene_anno_mash <- dge$genes[, c("ensembl_gene_id_version", "external_gene_name")]
gene_anno_mash <- gene_anno_mash[!duplicated(gene_anno_mash$ensembl_gene_id_version), ]

pm_df <- as.data.frame(pm)
pm_df$gene_id <- rownames(pm)
pm_df <- merge(pm_df, gene_anno_mash,
               by.x = "gene_id", by.y = "ensembl_gene_id_version", all.x = TRUE)
pm_df <- pm_df[, c("gene_id", "external_gene_name", colnames(pm))]
fwrite(pm_df, file.path(OUT_DIR_MASH,
                        paste0("mashr_lr_gene_posterior_means_", DATE_STAMP_MASH, ".csv")))

lfsr_df <- as.data.frame(lfsr)
lfsr_df$gene_id <- rownames(lfsr)
lfsr_df <- merge(lfsr_df, gene_anno_mash,
                 by.x = "gene_id", by.y = "ensembl_gene_id_version", all.x = TRUE)
lfsr_df <- lfsr_df[, c("gene_id", "external_gene_name", colnames(lfsr))]
fwrite(lfsr_df, file.path(OUT_DIR_MASH,
                          paste0("mashr_lr_gene_lfsr_", DATE_STAMP_MASH, ".csv")))

if (!is.null(sharing)) {
  fwrite(as.data.frame(sharing),
         file.path(OUT_DIR_MASH,
                   paste0("mashr_lr_gene_sharing_2fold_", DATE_STAMP_MASH, ".csv")),
         row.names = TRUE)
  fwrite(as.data.frame(sharing_sign),
         file.path(OUT_DIR_MASH,
                   paste0("mashr_lr_gene_sharing_sign_", DATE_STAMP_MASH, ".csv")),
         row.names = TRUE)
}
fwrite(as.data.frame(Vhat),
       file.path(OUT_DIR_MASH,
                 paste0("mashr_lr_gene_null_corr_", DATE_STAMP_MASH, ".csv")),
       row.names = TRUE)

for (j in seq_len(ncol(pm))) {
  ct_name <- colnames(pm)[j]

  res <- data.frame(
    ensembl_gene_id_version = rownames(pm),
    logFC                   = pm[, j],
    adj.P.Val               = lfsr[, j],
    nmd_responsive          = pm[, j] > 0 & lfsr[, j] < 0.05,
    stringsAsFactors        = FALSE
  )
  res <- merge(res, gene_anno_mash,
               by.x = "ensembl_gene_id_version",
               by.y = "ensembl_gene_id_version", all.x = TRUE)
  colnames(res)[colnames(res) == "external_gene_name"] <- "hgnc_symbol"

  res <- res[order(res$adj.P.Val), ]

  ct_code <- tolower(gsub("Smg1i_in_", "", ct_name))
  ct_code <- gsub("_", "", ct_code)

  fwrite(res, file.path(OUT_DIR_MASH,
                        paste0("nmd_mashr_lr_dge_", ct_code, "_", DATE_STAMP_MASH, ".csv")))
}
```

# Step 11 — Comparison With Short-Read DGE
 
```{r compare-sr-lr, fig.width=10, fig.height=10}
sr_files <- list.files("/udd/reyle/nmd_lungcells_2026/shortread_dge/mashr",
                       pattern = "^nmd_mashr_dge_.*_2026\\.3\\.10\\.csv$",
                       full.names = TRUE)
 
LR_TO_SR <- c("DD" = "DD", "AT" = "AT", "FB" = "FB", "MV" = "MV")
 
comparison_list <- list()
hex_plots       <- list()
 
for (lr_ct in names(LR_TO_SR)) {
  sr_ct   <- LR_TO_SR[lr_ct]
  lr_coef <- paste0("Smg1i_in_", lr_ct)
  j       <- match(lr_coef, colnames(pm))
  if (is.na(j)) next
 
  lr_res <- data.frame(
    gene_id   = rownames(pm),
    logFC     = pm[, j],
    adj.P.Val = lfsr[, j],
    stringsAsFactors = FALSE
  )
  lr_res$gene_id_clean <- sub("\\..*", "", lr_res$gene_id)
 
  sr_file_code <- tolower(gsub("_", "", sr_ct))
  sr_file      <- sr_files[grepl(paste0("_", sr_file_code, "_"), sr_files)]
  if (length(sr_file) == 0) next
 
  sr_res <- fread(sr_file[1], data.table = FALSE)
  sr_res$gene_id_clean <- sub("\\..*", "", sr_res$ensembl_gene_id_version)
 
  common <- intersect(lr_res$gene_id_clean, sr_res$gene_id_clean)
 
  merged <- merge(
    lr_res[, c("gene_id_clean", "logFC", "adj.P.Val")],
    sr_res[, c("gene_id_clean", "logFC", "adj.P.Val")],
    by = "gene_id_clean", suffixes = c("_LR", "_SR")
  )
 
  cor_p <- round(cor(merged$logFC_LR, merged$logFC_SR, method = "pearson"),  3)
  cor_s <- round(cor(merged$logFC_LR, merged$logFC_SR, method = "spearman"), 3)
 
  lr_nmd <- merged$adj.P.Val_LR < 0.05 & merged$logFC_LR > 0
  sr_nmd <- merged$adj.P.Val_SR < 0.05 & merged$logFC_SR > 0
 
  comparison_list[[lr_ct]] <- data.frame(
    cell_type = lr_ct,
    n_common  = length(common),
    pearson   = cor_p,
    spearman  = cor_s,
    both_nmd  = sum(lr_nmd & sr_nmd),
    LR_only   = sum(lr_nmd & !sr_nmd),
    SR_only   = sum(!lr_nmd & sr_nmd)
  )
 
  merged$cell_type <- lr_ct
  merged$cor_label <- paste0("r = ", cor_p, "\nrho = ", cor_s)
  hex_plots[[lr_ct]] <- merged
}
 
if (length(hex_plots) > 0) {
  hex_df <- bind_rows(hex_plots)
  hex_df$cell_type <- factor(hex_df$cell_type, levels = c("AT", "DD", "FB", "MV"))
 
  label_df <- hex_df %>%
    group_by(cell_type) %>%
    summarise(x = min(logFC_SR) + 0.05 * diff(range(logFC_SR)),
              y = max(logFC_LR) - 0.05 * diff(range(logFC_LR)),
              cor_label = cor_label[1],
              .groups = "drop")
 
  p_hex <- ggplot(hex_df, aes(x = logFC_SR, y = logFC_LR)) +
    geom_hex(bins = 50) +
    scale_fill_gradientn(colors = c("white", "lightblue", "blue", "darkblue", "red"),
                         trans = "log10", name = "Density") +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
    geom_text(data = label_df,
              aes(x = x, y = y, label = cor_label),
              hjust = 0, vjust = 1, size = 4, fontface = "bold",
              inherit.aes = FALSE) +
    facet_wrap(~ cell_type, ncol = 2) +
    labs(title    = "LR Vs SR Posterior Mean logFC",
         subtitle = "Per-cell-type comparison; dashed line = identity",
         x = "Short-Read logFC", y = "Long-Read logFC") +
    theme_minimal(base_size = 13) +
    theme(strip.text = element_text(face = "bold", size = 13),
          plot.title = element_text(face = "bold", hjust = 0.5),
          plot.subtitle = element_text(color = "grey40", hjust = 0.5))
 
  ggsave(file.path(OUT_DIR_FIGURES,
                   paste0("lr_vs_sr_logfc_2x2_", DATE_STAMP, ".png")),
         p_hex, width = 10, height = 10, dpi = 300, bg = "white")
 
  print(p_hex)
}
 
comparison_df <- do.call(rbind, comparison_list)
print(comparison_df)
 
if (nrow(comparison_df) > 0) {
  bar_data <- comparison_df %>%
    dplyr::select(cell_type, both_nmd, LR_only, SR_only) %>%
    tidyr::pivot_longer(-cell_type, names_to = "category", values_to = "n_genes")
 
  bar_data$category <- factor(bar_data$category,
                              levels = c("LR_only", "both_nmd", "SR_only"),
                              labels = c("Long-Read Only", "Both", "Short-Read Only"))
  bar_data$cell_type <- factor(bar_data$cell_type, levels = c("AT", "DD", "FB", "MV"))
 
  p_bar <- ggplot(bar_data, aes(x = cell_type, y = n_genes, fill = category)) +
    geom_bar(stat = "identity", position = "dodge") +
    scale_fill_manual(values = c("steelblue", "purple", "coral")) +
    geom_text(aes(label = n_genes),
              position = position_dodge(width = 0.9), vjust = -0.5, size = 3) +
    labs(title = "NMD-Responsive Overlap (LR Vs SR)",
         x = "Cell Type", y = "Gene Count", fill = NULL) +
    theme_minimal(base_size = 14) +
    theme(legend.position = "bottom",
          plot.title      = element_text(face = "bold", hjust = 0.5))
 
  ggsave(file.path(OUT_DIR_FIGURES,
                   paste0("lr_vs_sr_nmd_overlap_", DATE_STAMP, ".png")),
         p_bar, width = 10, height = 6, dpi = 300, bg = "white")
 
  print(p_bar)
}
```
 
# Session Info
 
```{r session-info}
sessionInfo()
```