---
title: "Long-Read Gene-Level DGE Summary: NMD Inhibition Effects"
author: "Leshem, Yul"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: lumen
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
    code_folding: hide
---

Code location: `/udd/reyle/nmd_lungcells_2026/code/final/LR_Gene-Level_DGE_Summary_mashR.Rmd`

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Analysis Overview

This report summarizes long-read gene-level differential expression (posterior
estimates) for the Smg1i (NMD inhibitor) vs DMSO contrast across four primary
human lung cell types (AT, DD, FB, MV).

Effect sizes (`logFC`) are posterior means from multivariate adaptive shrinkage,
which borrows strength across cell types. Significance is the local false sign
rate (lfsr); lfsr < 0.05 means there is < 5% probability that the sign of the
effect estimate is wrong. lfsr is carried in the `adj.P.Val` column for
column-format compatibility with topTables.

**Definitions:**

- **NMD-Responsive Gene:** posterior mean logFC > 0 AND lfsr < 0.05 (canonical Smg1i-up signature).
- **NMD-Suppressed Gene:** posterior mean logFC < 0 AND lfsr < 0.05.
- **logFC:** posterior mean log2 fold change.
- **adj.P.Val:** local false sign rate (lfsr).

```{r config, include=FALSE}
INPUT_DGE_DIR    <- "/udd/reyle/nmd_lungcells_2026/long_read/mashr_gene"
DATE_STAMP       <- "2026.3.10"
DGE_FILE_PATTERN <- "^nmd_mashr_lr_dge_.*_2026\\.3\\.10\\.csv$"

OUTPUT_DIR  <- file.path("/udd/reyle/nmd_lungcells_2026/tmp/mashr_lr_gene")
OUT_DIR_FIG <- "/udd/reyle/nmd_lungcells_2026/final_figures/lr_gene"

OUTPUT_SUMMARY            <- file.path(OUTPUT_DIR, paste0("mashr_lr_gene_summary_statistics_",     DATE_STAMP, ".tsv"))
OUTPUT_TOP_HITS           <- file.path(OUTPUT_DIR, paste0("mashr_lr_gene_top_nmd_responsive_",     DATE_STAMP, ".tsv"))
OUTPUT_CELLTYPE_OVERLAP   <- file.path(OUTPUT_DIR, paste0("mashr_lr_gene_celltype_overlap_",       DATE_STAMP, ".tsv"))
OUTPUT_CORE_TARGETS       <- file.path(OUTPUT_DIR, paste0("mashr_lr_gene_core_nmd_targets_",       DATE_STAMP, ".tsv"))
OUTPUT_CELLTYPE_SPECIFIC  <- file.path(OUTPUT_DIR, paste0("mashr_lr_gene_celltype_specific_",      DATE_STAMP, ".tsv"))
OUTPUT_CHARACTERISTICS    <- file.path(OUTPUT_DIR, paste0("mashr_lr_gene_characteristics_",        DATE_STAMP, ".tsv"))
OUTPUT_GSEA_ALL           <- file.path(OUTPUT_DIR, paste0("mashr_lr_gene_gsea_results_",           DATE_STAMP, ".tsv"))
OUTPUT_GSEA_SIG           <- file.path(OUTPUT_DIR, paste0("mashr_lr_gene_gsea_significant_",       DATE_STAMP, ".tsv"))

SIGNIFICANCE_THRESHOLD <- 0.05
TOP_HITS_N             <- 500

CELLTYPE_MAP     <- c("at" = "AT", "dd" = "DD", "fb" = "FB", "mv" = "MV")
CT_DISPLAY_ORDER <- c("AT", "DD", "FB", "MV")

if (!dir.exists(INPUT_DGE_DIR)) stop("DGE directory not found: ", INPUT_DGE_DIR)
dir.create(OUTPUT_DIR,  recursive = TRUE, showWarnings = FALSE)
dir.create(OUT_DIR_FIG, recursive = TRUE, showWarnings = FALSE)
```

<style>
body { font-size: 16px; }
p, li { font-size: 16px; line-height: 1.6; }
</style>

# Libraries

```{r load-libs}
library(tidyverse)
library(data.table)
library(DT)
library(edgeR)
library(tibble)
library(rlang)

# Bioconductor namespace masking guard
select <- dplyr::select
filter <- dplyr::filter
rename <- dplyr::rename
count  <- dplyr::count
slice  <- dplyr::slice
first  <- dplyr::first
```

# Load DGE Data

```{r load-dge}
dge_files <- list.files(INPUT_DGE_DIR, pattern = DGE_FILE_PATTERN, full.names = TRUE)
cat("Loading", length(dge_files), "DGE files\n")

# File pattern: nmd_mashr_lr_dge_<ct>_2026.3.10.csv
# Splitting by underscore: parts[1]="nmd", [2]="mashr", [3]="lr", [4]="dge",
# [5]=<ct>, [6]="2026.3.10.csv"  -> cell type lives at parts[5]
all_gene_dge <- map_df(dge_files, ~{
  df       <- read_csv(.x, show_col_types = FALSE)
  parts    <- str_split(basename(.x), "_")[[1]]
  df$cell_type_file <- parts[5]
  df
})

all_gene_dge$cell_type <- CELLTYPE_MAP[all_gene_dge$cell_type_file]
all_gene_dge <- all_gene_dge %>% filter(!is.na(cell_type))

all_gene_dge <- all_gene_dge %>%
  mutate(
    valid_symbol = !is.na(hgnc_symbol) & hgnc_symbol != "" & !str_detect(hgnc_symbol, "^ENSG"),
    gene_display = ifelse(valid_symbol, hgnc_symbol, ensembl_gene_id_version)
  )

cat("Loaded:", nrow(all_gene_dge), "rows\n")
cat("Cell types:", paste(unique(all_gene_dge$cell_type), collapse = ", "), "\n")
```

# Summary Statistics

```{r summary-stats}
summary_stats <- all_gene_dge %>%
  group_by(cell_type) %>%
  summarize(
    n_genes_tested            = n(),
    n_valid_symbols           = sum(valid_symbol),
    n_sig                     = sum(adj.P.Val < SIGNIFICANCE_THRESHOLD),
    pct_sig                   = 100 * n_sig / n(),
    n_nmd_responsive          = sum(logFC > 0 & adj.P.Val < SIGNIFICANCE_THRESHOLD),
    pct_nmd_responsive        = ifelse(n_sig > 0, 100 * n_nmd_responsive / n_sig, 0),
    n_nmd_suppressed          = sum(logFC < 0 & adj.P.Val < SIGNIFICANCE_THRESHOLD),
    pct_nmd_suppressed        = ifelse(n_sig > 0, 100 * n_nmd_suppressed / n_sig, 0),
    mean_logFC_responsive     = mean(logFC[logFC > 0 & adj.P.Val < SIGNIFICANCE_THRESHOLD]),
    median_logFC_responsive   = median(logFC[logFC > 0 & adj.P.Val < SIGNIFICANCE_THRESHOLD]),
    .groups = "drop"
  )

cat("**Table 1.** Long-Read Gene-Level Summary Statistics By Cell Type\n\n")

datatable(summary_stats,
          options = list(pageLength = 10, scrollX = TRUE),
          caption = "Table 1. Long-Read Gene-Level Summary Statistics By Cell Type") %>%
  formatRound(columns = c("pct_sig", "pct_nmd_responsive", "pct_nmd_suppressed",
                          "mean_logFC_responsive", "median_logFC_responsive"),
              digits = 2)

write_tsv(summary_stats, OUTPUT_SUMMARY)
knitr::kable(summary_stats, format = "pipe", digits = 2)
```

## Identify NMD-Responsive Genes

```{r identify-nmd-responsive}
nmd_responsive_genes <- all_gene_dge %>%
  filter(logFC > 0, adj.P.Val < SIGNIFICANCE_THRESHOLD) %>%
  select(cell_type, ensembl_gene_id_version, hgnc_symbol, gene_display,
         logFC, adj.P.Val)

cat("Total NMD-responsive rows:",   nrow(nmd_responsive_genes), "\n")
cat("Unique NMD-responsive genes:", n_distinct(nmd_responsive_genes$ensembl_gene_id_version), "\n\n")

print(
  nmd_responsive_genes %>%
    group_by(cell_type) %>%
    summarize(n_nmd_genes  = n(),
              mean_logFC   = mean(logFC),
              median_logFC = median(logFC),
              .groups = "drop")
)
```

# Top NMD-Responsive Genes

```{r top-hits}
top_hits_dt_list <- list()
table_num <- 2

for (ct in CT_DISPLAY_ORDER) {
  if (!ct %in% all_gene_dge$cell_type) next

  top_hits <- nmd_responsive_genes %>%
    filter(cell_type == ct) %>%
    arrange(adj.P.Val) %>%
    head(TOP_HITS_N) %>%
    select(hgnc_symbol, gene_display, logFC, adj.P.Val)

  top_hits_dt_list[[ct]] <- htmltools::tagList(
    htmltools::h3(paste0(ct, " — Top ", TOP_HITS_N, " NMD-Responsive Genes")),
    datatable(
      top_hits,
      options   = list(pageLength = 10, scrollX = TRUE),
      caption   = paste0("Table ", table_num, ". Top ", TOP_HITS_N, " NMD-Responsive Genes In ", ct),
      elementId = paste0("top_hits_lr_dt_", ct)
    ) %>%
      formatRound(columns = c("logFC", "adj.P.Val"), digits = 4)
  )
  table_num <- table_num + 1
}

write_tsv(nmd_responsive_genes, OUTPUT_TOP_HITS)

htmltools::tagList(top_hits_dt_list)
```

# Visualizations

## Volcano Plots

```{r volcano-plots, fig.width=10, fig.height=10}
volcano_data <- all_gene_dge %>%
  mutate(
    neg_log10_lfsr = -log10(adj.P.Val + 1e-300),
    is_nmd_responsive = logFC > 0 & adj.P.Val < SIGNIFICANCE_THRESHOLD,
    significance = case_when(
      is_nmd_responsive                  ~ "NMD-Responsive",
      adj.P.Val < SIGNIFICANCE_THRESHOLD ~ "Significant (Other)",
      TRUE                               ~ "Not Significant"
    )
  )
volcano_data$cell_type <- factor(volcano_data$cell_type, levels = CT_DISPLAY_ORDER)

p_volcano <- ggplot(volcano_data,
                    aes(x = logFC, y = neg_log10_lfsr, color = significance)) +
  geom_point(alpha = 0.5, size = 1.2) +
  facet_wrap(~ cell_type, ncol = 2) +
  scale_color_manual(values = c(
    "NMD-Responsive"      = "steelblue",
    "Significant (Other)" = "coral",
    "Not Significant"     = "gray70"
  )) +
  coord_cartesian(ylim = c(0, 50)) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "black") +
  labs(
    title    = "Volcano Plots",
    subtitle = "Posterior mean logFC vs −log10(lfsr); y-axis capped at 50",
    x        = "Posterior Mean log2 FC (Smg1i Vs DMSO)",
    y        = expression(-log[10]("lfsr")),
    color    = "Status"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    legend.position = "top",
    legend.title    = element_text(face = "bold"),
    strip.text      = element_text(size = 13, face = "bold"),
    plot.title      = element_text(size = 14, face = "bold"),
    plot.subtitle   = element_text(color = "grey40")
  )

ggsave(file.path(OUT_DIR_FIG, paste0("volcano_2x2_lr_gene_", DATE_STAMP, ".png")),
       p_volcano, width = 10, height = 10, dpi = 300, bg = "white")

print(p_volcano)
```

## logFC Distribution

```{r logfc-distribution, fig.width=10, fig.height=6}
median_logfc <- nmd_responsive_genes %>%
  group_by(cell_type) %>%
  summarize(median_logFC = median(logFC), .groups = "drop")

p_dens <- ggplot(nmd_responsive_genes, aes(x = logFC, fill = cell_type, color = cell_type)) +
  geom_density(alpha = 0.3, linewidth = 1) +
  geom_vline(data = median_logfc, aes(xintercept = median_logFC, color = cell_type),
             linetype = "dashed", linewidth = 0.8) +
  labs(
    title    = "logFC Distribution For NMD-Responsive Genes",
    subtitle = "Posterior mean logFC; dashed lines = per-cell-type median",
    x = "Posterior Mean log2 FC (Smg1i Vs DMSO)",
    y = "Density",
    fill = "Cell Type", color = "Cell Type"
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "top",
        plot.title = element_text(face = "bold"))

ggsave(file.path(OUT_DIR_FIG, paste0("logfc_density_lr_gene_", DATE_STAMP, ".png")),
       p_dens, width = 10, height = 6, dpi = 300, bg = "white")

print(p_dens)
```

# Cell-Type Comparison

## Overlap Counts

```{r celltype-overlap}
gene_celltype_counts <- nmd_responsive_genes %>%
  group_by(ensembl_gene_id_version, hgnc_symbol, gene_display) %>%
  summarize(
    n_celltypes = n_distinct(cell_type),
    celltypes   = paste(sort(unique(cell_type)), collapse = ", "),
    mean_logFC  = mean(logFC),
    .groups = "drop"
  ) %>%
  arrange(desc(n_celltypes), desc(mean_logFC))

overlap_summary <- gene_celltype_counts %>%
  count(n_celltypes, name = "n_genes") %>%
  mutate(pct_genes = 100 * n_genes / sum(n_genes))

cat("**Table 8.** Genes By Number Of Cell Types With NMD Response\n\n")
print(overlap_summary)

p_overlap <- ggplot(overlap_summary,
                    aes(x = factor(n_celltypes), y = n_genes, fill = factor(n_celltypes))) +
  geom_col() +
  geom_text(aes(label = n_genes), vjust = -0.5, size = 4) +
  scale_fill_brewer(palette = "Set3") +
  labs(
    title = "NMD-Responsive Genes By Cell-Type Consistency",
    x = "Number Of Cell Types With NMD Response",
    y = "Number Of Genes"
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "none",
        plot.title = element_text(face = "bold"))

ggsave(file.path(OUT_DIR_FIG, paste0("celltype_overlap_lr_gene_", DATE_STAMP, ".png")),
       p_overlap, width = 8, height = 5, dpi = 300, bg = "white")

print(p_overlap)
write_tsv(gene_celltype_counts, OUTPUT_CELLTYPE_OVERLAP)
```

```{r upset-summary-table}
upset_summary <- nmd_responsive_genes %>%
  group_by(ensembl_gene_id_version) %>%
  summarise(
    n_cts      = n_distinct(cell_type),
    cts        = paste(sort(unique(cell_type)), collapse = " & "),
    mean_logFC = round(mean(logFC), 3),
    .groups = "drop"
  ) %>%
  count(n_cts, cts, name = "n_genes") %>%
  arrange(desc(n_cts), desc(n_genes))

knitr::kable(upset_summary,
             caption = "NMD-Responsive Gene Intersection Sizes By Cell-Type Combination")
```

## NMD-Responsive In 2+ Cell Types

```{r multi-celltype-table}
multi_celltype_genes <- gene_celltype_counts %>%
  filter(n_celltypes >= 2) %>%
  arrange(desc(n_celltypes), desc(mean_logFC))

cat("Genes responsive in 2+ cell types:", nrow(multi_celltype_genes), "\n\n")

if (nrow(multi_celltype_genes) > 0) {
  multi_celltype_wide <- nmd_responsive_genes %>%
    filter(cell_type %in% CT_DISPLAY_ORDER) %>%
    semi_join(multi_celltype_genes, by = "ensembl_gene_id_version") %>%
    select(ensembl_gene_id_version, gene_display, cell_type, logFC) %>%
    pivot_wider(
      id_cols     = c(ensembl_gene_id_version, gene_display),
      names_from  = cell_type,
      values_from = logFC,
      values_fill = NA,
      names_prefix = "logFC_"
    ) %>%
    left_join(
      gene_celltype_counts %>% select(ensembl_gene_id_version, n_celltypes),
      by = "ensembl_gene_id_version"
    ) %>%
    select(gene_display, n_celltypes, starts_with("logFC_")) %>%
    filter(if_any(starts_with("logFC_"), ~ !is.na(.))) %>%
    arrange(desc(n_celltypes))

  output_multi_celltype <- file.path(OUTPUT_DIR,
                                     paste0("mashr_lr_gene_multi_celltype_nmd_", DATE_STAMP, ".tsv"))
  write_tsv(multi_celltype_wide, output_multi_celltype)

  logfc_cols <- colnames(multi_celltype_wide)[grepl("^logFC_", colnames(multi_celltype_wide))]

  datatable(
    multi_celltype_wide,
    options = list(pageLength = 25, scrollX = TRUE),
    caption = "Table 8A. NMD-Responsive Genes Present In 2 Or More Cell Types"
  ) %>%
    formatRound(columns = logfc_cols, digits = 3)
}
```

## Core NMD Targets (≥ 4 Cell Types)

```{r core-targets}
core_targets <- gene_celltype_counts %>%
  filter(n_celltypes >= 4) %>%
  arrange(desc(n_celltypes), desc(mean_logFC))

cat("Core NMD targets (>=4 cell types):", nrow(core_targets), "\n\n")

if (nrow(core_targets) > 0) {
  core_targets_wide <- nmd_responsive_genes %>%
    semi_join(core_targets, by = "ensembl_gene_id_version") %>%
    select(ensembl_gene_id_version, gene_display, cell_type, logFC) %>%
    group_by(ensembl_gene_id_version, gene_display, cell_type) %>%
    summarise(logFC = mean(logFC), .groups = "drop") %>%
    pivot_wider(
      id_cols     = c(ensembl_gene_id_version, gene_display),
      names_from  = cell_type,
      values_from = logFC
    ) %>%
    select(-ensembl_gene_id_version)

  cat("**Table 9.** Top 50 Core NMD Targets\n\n")
  datatable(
    core_targets_wide %>% head(50),
    options = list(pageLength = 10, scrollX = TRUE),
    caption = "Table 9. Top 50 Core NMD Targets"
  ) %>%
    formatRound(columns = 2:ncol(core_targets_wide), digits = 3)

  if (nrow(core_targets) >= 30) {
    top_core <- core_targets %>% head(30) %>% pull(ensembl_gene_id_version)

    heatmap_df <- nmd_responsive_genes %>%
      filter(ensembl_gene_id_version %in% top_core) %>%
      mutate(display_name = ifelse(!is.na(hgnc_symbol) & hgnc_symbol != "",
                                   hgnc_symbol, gene_display)) %>%
      group_by(ensembl_gene_id_version, display_name, cell_type) %>%
      summarise(logFC = mean(logFC), .groups = "drop")
    heatmap_df$cell_type <- factor(heatmap_df$cell_type, levels = CT_DISPLAY_ORDER)

    p_core <- ggplot(heatmap_df, aes(x = cell_type, y = display_name, fill = logFC)) +
      geom_tile(color = "white") +
      scale_fill_gradient(low = "white", high = "steelblue", name = "logFC") +
      labs(title = "Top 30 Core NMD Targets",
           x = "Cell Type", y = NULL) +
      theme_minimal(base_size = 12) +
      theme(axis.text.y = element_text(size = 8),
            plot.title  = element_text(hjust = 0.5, face = "bold"))

    ggsave(file.path(OUT_DIR_FIG, paste0("core_nmd_targets_heatmap_lr_gene_", DATE_STAMP, ".png")),
           p_core, width = 8, height = 9, dpi = 300, bg = "white")
    print(p_core)
  }

  write_tsv(core_targets_wide, OUTPUT_CORE_TARGETS)
}
```

## Cell-Type-Specific Genes (1 Cell Type Only)

```{r celltype-specific}
celltype_specific <- gene_celltype_counts %>%
  filter(n_celltypes == 1) %>%
  left_join(
    nmd_responsive_genes %>% select(ensembl_gene_id_version, cell_type, logFC, adj.P.Val),
    by = "ensembl_gene_id_version"
  ) %>%
  arrange(cell_type, desc(logFC))

cat("Cell-type-specific genes:", nrow(celltype_specific), "\n\n")

specific_summary <- celltype_specific %>%
  group_by(cell_type) %>%
  summarize(
    n_genes               = n(),
    mean_logFC_specific   = mean(logFC),
    median_logFC_specific = median(logFC),
    .groups = "drop"
  )

cat("**Table 10.** Cell-Type-Specific NMD-Responsive Genes\n\n")
print(specific_summary)

if (nrow(celltype_specific) > 0) {
  cat("\n**Top 10 Cell-Type-Specific Genes Per Cell Type:**\n\n")
  for (ct in CT_DISPLAY_ORDER) {
    ct_specific <- celltype_specific %>%
      filter(cell_type == ct) %>%
      select(gene_display, logFC, adj.P.Val) %>%
      head(10)
    if (nrow(ct_specific) > 0) {
      cat("**", ct, "**\n")
      print(knitr::kable(ct_specific, digits = 4))
      cat("\n")
    }
  }
  write_tsv(celltype_specific, OUTPUT_CELLTYPE_SPECIFIC)
}
```

# Gene Characteristics

## Per-Sample Expression (CPM)

```{r load-dgelist-for-expression}
DGELIST_PATH <- "/udd/reyle/nmd_lungcells_2026/long_read/dge_gene_longread_filtered_2026.3.10.rds"

if (file.exists(DGELIST_PATH)) {
  dge_gene       <- readRDS(DGELIST_PATH)
  cpm_normalized <- cpm(dge_gene, normalized = TRUE, log = FALSE)

  sample_metadata <- dge_gene$samples %>%
    rownames_to_column("sample_name")

  if ("treatment" %in% colnames(sample_metadata) && "ct" %in% colnames(sample_metadata)) {
    sample_metadata <- sample_metadata %>%
      select(sample_name, treatment, ct) %>%
      mutate(cell_type = ifelse(ct == "AT2", "AT", ct)) %>%
      select(-ct)
    has_condition_data <- TRUE
  } else {
    has_condition_data <- FALSE
  }

  if (exists("sample_metadata") && has_condition_data) {
    cpm_long <- as.data.frame(cpm_normalized) %>%
      mutate(dgelist_gene_id = rownames(.)) %>%
      pivot_longer(-dgelist_gene_id, names_to = "sample_name", values_to = "cpm") %>%
      left_join(sample_metadata, by = "sample_name")

    gene_info <- dge_gene$genes %>% rownames_to_column("dgelist_gene_id")
    ensembl_col <- if ("ensembl_gene_id_version" %in% colnames(gene_info)) {
      "ensembl_gene_id_version"
    } else if ("gene_id" %in% colnames(gene_info)) {
      "gene_id"
    } else NULL

    gene_info <- if (!is.null(ensembl_col)) {
      gene_info %>% select(dgelist_gene_id, !!sym(ensembl_col), hgnc_symbol)
    } else {
      gene_info %>% select(dgelist_gene_id, hgnc_symbol)
    }

    cpm_long <- cpm_long %>% left_join(gene_info, by = "dgelist_gene_id")
    gene_id_column <- ensembl_col
  }
} else {
  has_condition_data <- FALSE
  gene_id_column <- NULL
}
```

## Expression Distribution Of NMD-Responsive Genes (DMSO)

```{r expression-violin-dmso, fig.width=10, fig.height=10}
if (has_condition_data && !is.null(gene_id_column)) {
  responsive_ids <- unique(nmd_responsive_genes$ensembl_gene_id_version)

  cpm_responsive <- cpm_long %>%
    filter(treatment == "DMSO",
           !!sym(gene_id_column) %in% responsive_ids,
           cell_type %in% CT_DISPLAY_ORDER)

  if (nrow(cpm_responsive) > 0) {
    cpm_responsive$cell_type <- factor(cpm_responsive$cell_type, levels = CT_DISPLAY_ORDER)

    p_dmso <- ggplot(cpm_responsive, aes(x = cell_type, y = log2(cpm + 1), fill = cell_type)) +
      geom_violin(alpha = 0.7, scale = "width") +
      geom_boxplot(width = 0.15, outlier.alpha = 0.2, fill = "white") +
      facet_wrap(~ cell_type, ncol = 2, scales = "free_x") +
      labs(
        title    = "Baseline Expression Of NMD-Responsive Genes (DMSO)",
        subtitle = "log2(CPM + 1) per sample",
        x = "Cell Type", y = "log2(CPM + 1)"
      ) +
      theme_minimal(base_size = 13) +
      theme(legend.position = "none",
            strip.text      = element_text(face = "bold", size = 13),
            plot.title      = element_text(face = "bold", hjust = 0.5),
            plot.subtitle   = element_text(color = "grey40", hjust = 0.5))

    ggsave(file.path(OUT_DIR_FIG,
                     paste0("expression_dmso_2x2_lr_gene_", DATE_STAMP, ".png")),
           p_dmso, width = 10, height = 10, dpi = 300, bg = "white")
    print(p_dmso)
  }
}
```

## Expression Distribution Of NMD-Responsive Genes (Smg1i)

```{r expression-violin-smg1i, fig.width=10, fig.height=10}
if (has_condition_data && !is.null(gene_id_column)) {
  cpm_responsive_smg1i <- cpm_long %>%
    filter(treatment == "Smg1i",
           !!sym(gene_id_column) %in% responsive_ids,
           cell_type %in% CT_DISPLAY_ORDER)

  if (nrow(cpm_responsive_smg1i) > 0) {
    cpm_responsive_smg1i$cell_type <- factor(cpm_responsive_smg1i$cell_type,
                                             levels = CT_DISPLAY_ORDER)

    p_smg1i <- ggplot(cpm_responsive_smg1i, aes(x = cell_type, y = log2(cpm + 1), fill = cell_type)) +
      geom_violin(alpha = 0.7, scale = "width") +
      geom_boxplot(width = 0.15, outlier.alpha = 0.2, fill = "white") +
      facet_wrap(~ cell_type, ncol = 2, scales = "free_x") +
      labs(
        title    = "Smg1i Expression Of NMD-Responsive Genes",
        subtitle = "log2(CPM + 1) per sample",
        x = "Cell Type", y = "log2(CPM + 1)"
      ) +
      theme_minimal(base_size = 13) +
      theme(legend.position = "none",
            strip.text      = element_text(face = "bold", size = 13),
            plot.title      = element_text(face = "bold", hjust = 0.5),
            plot.subtitle   = element_text(color = "grey40", hjust = 0.5))

    ggsave(file.path(OUT_DIR_FIG,
                     paste0("expression_smg1i_2x2_lr_gene_", DATE_STAMP, ".png")),
           p_smg1i, width = 10, height = 10, dpi = 300, bg = "white")
    print(p_smg1i)
  }
}
```

## Effect Size Distribution

```{r effect-size, fig.width=10, fig.height=6}
nmd_responsive_genes$cell_type <- factor(nmd_responsive_genes$cell_type, levels = CT_DISPLAY_ORDER)

p_effect <- ggplot(nmd_responsive_genes, aes(x = cell_type, y = logFC, fill = cell_type)) +
  geom_boxplot(alpha = 0.7, outlier.alpha = 0.3) +
  geom_hline(yintercept = 2, linetype = "dashed", color = "red", linewidth = 0.8) +
  labs(
    title    = "logFC Distribution For NMD-Responsive Genes",
    subtitle = "Red line: logFC = 2 (4-fold). Values are posterior means.",
    x = "Cell Type", y = "Posterior Mean log2 FC"
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "none",
        plot.title      = element_text(face = "bold"))

ggsave(file.path(OUT_DIR_FIG, paste0("logfc_boxplot_lr_gene_", DATE_STAMP, ".png")),
       p_effect, width = 10, height = 6, dpi = 300, bg = "white")

print(p_effect)

extreme_fc_genes <- nmd_responsive_genes %>%
  filter(abs(logFC) > 2) %>%
  arrange(desc(logFC))

cat("\n**Table 12.** Genes With |logFC| > 2:", nrow(extreme_fc_genes), "\n\n")
print(extreme_fc_genes %>% group_by(cell_type) %>% summarize(n_extreme = n(), .groups = "drop"))

characteristics_data <- nmd_responsive_genes %>%
  select(cell_type, ensembl_gene_id_version, hgnc_symbol, gene_display, logFC, adj.P.Val) %>%
  mutate(extreme_fc = abs(logFC) > 2)

write_tsv(characteristics_data, OUTPUT_CHARACTERISTICS)
```

# GSEA: Pathway Enrichment

## Load Gene Sets

```{r load-genesets}
options(repos = c(CRAN = "https://cloud.r-project.org"))

if (!require("fgsea",   quietly = TRUE)) BiocManager::install("fgsea")
if (!require("msigdbr", quietly = TRUE)) install.packages("msigdbr")

library(fgsea)
library(msigdbr)

GSEA_MIN_SIZE      <- 15
GSEA_MAX_SIZE      <- 500
GSEA_FDR_THRESHOLD <- 0.05

msigdb_h        <- msigdbr(species = "Homo sapiens", collection = "H")
msigdb_bp       <- msigdbr(species = "Homo sapiens", collection = "C5", subcollection = "GO:BP")
msigdb_kegg     <- msigdbr(species = "Homo sapiens", collection = "C2", subcollection = "CP:KEGG_LEGACY")
msigdb_reactome <- msigdbr(species = "Homo sapiens", collection = "C2", subcollection = "CP:REACTOME")

all_genesets <- bind_rows(
  msigdb_h        %>% mutate(collection = "Hallmark"),
  msigdb_bp       %>% mutate(collection = "GO_BP"),
  msigdb_kegg     %>% mutate(collection = "KEGG"),
  msigdb_reactome %>% mutate(collection = "Reactome")
)

pathways <- all_genesets %>% split(x = .$gene_symbol, f = .$gs_name)
pathways_filtered <- pathways[
  sapply(pathways, length) >= GSEA_MIN_SIZE &
  sapply(pathways, length) <= GSEA_MAX_SIZE
]

cat("Pathways after size filter:", length(pathways_filtered), "\n")
```

## Build Per-Cell-Type Rankings

```{r prepare-rankings}
rankings_per_celltype <- list()

for (ct in CT_DISPLAY_ORDER) {
  if (!ct %in% all_gene_dge$cell_type) next

  ranked_data <- all_gene_dge %>%
    filter(cell_type == ct, valid_symbol) %>%
    group_by(hgnc_symbol) %>%
    slice_max(order_by = abs(logFC), n = 1, with_ties = FALSE) %>%
    ungroup() %>%
    arrange(desc(logFC))

  ranked_vector <- setNames(ranked_data$logFC, ranked_data$hgnc_symbol)
  rankings_per_celltype[[ct]] <- ranked_vector
}
```

## Run GSEA

```{r run-gsea}
gsea_results_all <- list()
gsea_dt_list     <- list()

for (ct in names(rankings_per_celltype)) {
  ranked_list <- rankings_per_celltype[[ct]]

  gsea_res <- fgsea(
    pathways = pathways_filtered,
    stats    = ranked_list,
    minSize  = GSEA_MIN_SIZE,
    maxSize  = GSEA_MAX_SIZE
  )
  gsea_res$cell_type <- ct
  gsea_results_all[[ct]] <- gsea_res

  top_100 <- gsea_res %>%
    filter(NES > 0, padj < 0.05) %>%
    arrange(pval) %>%
    head(100) %>%
    select(pathway, pval, padj, ES, NES, size, leadingEdge) %>%
    mutate(
      pathway_short = str_trunc(pathway, 60),
      leadingEdge   = sapply(leadingEdge, function(x) paste(head(x, 5), collapse = ", "))
    ) %>%
    select(pathway_short, pval, padj, ES, NES, size, leadingEdge)

  gsea_dt_list[[ct]] <- htmltools::tagList(
    htmltools::h3(paste0(ct, " — Top GSEA Results")),
    datatable(
      top_100,
      options   = list(pageLength = 10, scrollX = TRUE),
      caption   = paste0("Top GSEA Results — ", ct),
      elementId = paste0("gsea_lr_dt_", ct)
    ) %>%
      formatRound(columns = c("pval", "padj", "ES", "NES"), digits = 4)
  )
}

gsea_combined     <- bind_rows(gsea_results_all)
gsea_combined_pos <- gsea_combined %>% filter(NES > 0)

gsea_significant <- gsea_combined_pos %>%
  filter(padj < GSEA_FDR_THRESHOLD) %>%
  arrange(padj)

gsea_combined$leadingEdge <- sapply(gsea_combined$leadingEdge, paste, collapse = ", ")
write_tsv(gsea_combined, OUTPUT_GSEA_ALL)

if (nrow(gsea_significant) > 0) {
  gsea_significant$leadingEdge <- sapply(gsea_significant$leadingEdge, paste, collapse = ", ")
  write_tsv(gsea_significant, OUTPUT_GSEA_SIG)
}

htmltools::tagList(gsea_dt_list)
```

## Cross-Cell-Type Pathway Consistency

```{r pathway-consistency}
pathway_counts_strict  <- gsea_combined_pos %>%
  filter(padj < GSEA_FDR_THRESHOLD) %>% count(pathway) %>% arrange(desc(n)) %>% filter(n > 1)
pathway_counts_relaxed <- gsea_combined_pos %>%
  filter(padj < 0.1) %>%                count(pathway) %>% arrange(desc(n)) %>% filter(n > 1)

cat("Pathways significant in >1 cell type (FDR <", GSEA_FDR_THRESHOLD, "):",
    nrow(pathway_counts_strict), "\n")

top_n <- 5

if (nrow(pathway_counts_strict) > 0) {
  cat("### Top Consistent Pathways (FDR <", GSEA_FDR_THRESHOLD, ")\n\n")
  for (pw in head(pathway_counts_strict$pathway, top_n)) {
    cat("\n**", pw, "**\n")
    print(knitr::kable(
      gsea_combined_pos %>%
        filter(pathway == pw, padj < GSEA_FDR_THRESHOLD) %>%
        select(cell_type, NES, padj) %>%
        arrange(cell_type),
      digits = 4
    ))
  }
}
```

# Key Findings

```{r key-findings}
cat("**Long-Read Gene-Level NMD Response:**\n")
cat("- Average per-cell-type NMD-responsive genes:",
    round(mean(summary_stats$n_nmd_responsive)),
    " (", round(mean(summary_stats$pct_nmd_responsive), 1), "%)\n")
cat("- Mean posterior logFC:",
    round(mean(summary_stats$mean_logFC_responsive),   2), "\n")
cat("- Median posterior logFC:",
    round(mean(summary_stats$median_logFC_responsive), 2), "\n\n")

cat("**Cell-Type Consistency:**\n")
cat("- Unique NMD-responsive genes:",
    n_distinct(nmd_responsive_genes$ensembl_gene_id_version), "\n")
if (exists("core_targets")      && nrow(core_targets)      > 0)
  cat("- Core NMD targets (>=4 cell types):",     nrow(core_targets),      "\n")
if (exists("celltype_specific") && nrow(celltype_specific) > 0)
  cat("- Cell-type-specific:", nrow(celltype_specific), "\n")
```

# Session Info

```{r session-info}
sessionInfo()
```