---
title: "NMD-target genes are enriched for RNA-binding proteins"
subtitle: "Long-read DIE isoform results (mashr NMD calls) intersected with the Gerstberger 2014 RBP census"
author: "Yul Leshem"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output:
  html_document: { toc: true, toc_float: true, df_print: paged, code_folding: hide, number_sections: true }
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE, warning=FALSE, message=FALSE, fig.width=10, fig.height=5, dpi=110)
options(bitmapType="cairo"); .libPaths(c("/udd/reyle/Rlibs", .libPaths()))
```

```{r libs}
suppressPackageStartupMessages({
  library(data.table); library(dplyr); library(tidyr); library(tibble)
  library(ggplot2); library(knitr); library(readxl)
})
select <- dplyr::select; filter <- dplyr::filter; rename <- dplyr::rename
count  <- dplyr::count;  slice  <- dplyr::slice
```

# Overview

Self-contained analysis: load the **long-read differential-isoform-expression (DIE) results**
(mashr per-cell-type isoform posterior means + lfsr), define **NMD-target genes** (any gene with
a confident NMD isoform), and test whether they are enriched for **RNA-binding proteins** using
the **Gerstberger, Hafner & Tuschl 2014** census of 1,542 human RBPs.

- "NMD isoform" = `lfsr < 0.05 & posterior mean > 0` in a cell type (NMD up-call).
- "NMD-target gene" = a gene with ≥1 NMD isoform in ≥1 cell type.
- Background = all genes with an isoform tested in the DIE/mashr set.

```{r params}
proj    <- "/udd/reyle/nmd_lungcells_2026"
OUT_DIR <- file.path(proj, "long_read/nmd_rbp_enrichment")
dir.create(OUT_DIR, recursive=TRUE, showWarnings=FALSE)
DATE_STAMP   <- format(Sys.Date(), "%Y.%m.%d")
CELL_TYPES   <- c("AT","DD","FB","MV")
LFSR_NMD_THR <- 0.05

# long-read isoform DGEList (for the isoform -> gene/symbol mapping)
DGE_ISO_PATH  <- file.path(proj, "short_read/dge_isoform_longread_filtered_2026.3.3.rds")
# long-read DIE isoform mashr results
PM_ISO_PATH   <- file.path(proj, "long_read/mashr/mashr_isoform_posterior_means_2026.3.10.csv")
LFSR_ISO_PATH <- file.path(proj, "long_read/mashr/mashr_isoform_lfsr_2026.3.10.csv")
# Gerstberger 2014 RBP census (Supplementary Table S3). Either the xls or the exported csv.
RBP_XLS <- file.path(proj, "41576_2014_BFnrg3813_MOESM25_ESM.xls")   # <- transfer this here
RBP_CSV <- file.path(proj, "gerstberger_2014_rbp_census.csv")        # <- or this (if you exported it)
```

# Load long-read DIE isoform results

```{r load}
dge_iso  <- readRDS(DGE_ISO_PATH)
pm_iso   <- read.csv(PM_ISO_PATH,   check.names=FALSE, stringsAsFactors=FALSE)
lfsr_iso <- read.csv(LFSR_ISO_PATH, check.names=FALSE, stringsAsFactors=FALSE)
colnames(pm_iso)[1] <- "isoform_id"; colnames(lfsr_iso)[1] <- "isoform_id"

# isoform -> gene / symbol map from the DGEList
iso_col  <- intersect(c("isoform_id","isoform","txid","transcript_id"), colnames(dge_iso$genes))[1]
gene_col <- intersect(c("gene_id","associated_gene","ensembl_gene_id","gene_id_ens115_sqanti"), colnames(dge_iso$genes))[1]
sym_col  <- intersect(c("hgnc_symbol","gene_name","external_gene_name","symbol"), colnames(dge_iso$genes))[1]
iso2gene <- data.frame(isoform_id=dge_iso$genes[[iso_col]], gene_id=dge_iso$genes[[gene_col]],
                       hgnc_symbol=dge_iso$genes[[sym_col]], stringsAsFactors=FALSE)

resolve_mash_col <- function(df, ct){ cn<-colnames(df); if (ct %in% cn) return(ct)
  h<-cn[grepl(paste0("_in_",ct,"$"),cn)]; if(length(h)==1) return(h)
  h<-cn[grepl(paste0("^",ct,"[._:-]"),cn)]; if(length(h)==1) return(h); stop("no mash col ",ct) }

cat(sprintf("DIE isoforms loaded: %d; mapped to %d genes\n",
            nrow(lfsr_iso), length(unique(iso2gene$gene_id))))
```

# Load Gerstberger 2014 RBP census

```{r load-rbp}
sv <- function(x) sub("\\..*$","",x)   # strip Ensembl version
if (file.exists(RBP_CSV)) {
  rbp <- fread(RBP_CSV)                                   # gene_symbol, ensembl_gene_id, consensus_rna_target, domains
} else {
  rr  <- read_excel(RBP_XLS, sheet="RBP table")
  rbp <- data.frame(gene_symbol          = rr[["gene name"]],
                    ensembl_gene_id      = rr[["gene id"]],
                    consensus_rna_target = rr[["consensus RNA target"]],
                    stringsAsFactors=FALSE)
}
rbp_ens   <- unique(sv(rbp$ensembl_gene_id))
rbp_sym   <- unique(rbp$gene_symbol)
sr_family <- paste0("SRSF", 1:12)   # canonical 12 human SR proteins (Manley & Krainer 2010)

cat(sprintf("RBP census: %d RBPs (%d with Ensembl IDs)\n", length(rbp_sym), length(rbp_ens)))
```

# NMD-target genes from the DIE results

```{r nmd-genes}
nmd_gene_ct <- bind_rows(lapply(CELL_TYPES, function(ct){
  l <- lfsr_iso[[resolve_mash_col(lfsr_iso, ct)]]; p <- pm_iso[[resolve_mash_col(pm_iso, ct)]]
  iso <- lfsr_iso$isoform_id[!is.na(l) & !is.na(p) & l < LFSR_NMD_THR & p > 0]
  data.frame(ct=ct, gene_id=unique(iso2gene$gene_id[iso2gene$isoform_id %in% iso]))
}))
nmd_genes <- unique(nmd_gene_ct$gene_id)
bg_genes  <- unique(iso2gene$gene_id[iso2gene$isoform_id %in% lfsr_iso$isoform_id])  # all DIE-tested genes

cat("NMD-target genes per cell type:\n")
print(nmd_gene_ct %>% count(ct, name="n_nmd_genes"))
cat(sprintf("\nUnion NMD-target genes (>=1 NMD isoform, any CT): %d of %d tested\n",
            length(nmd_genes), length(bg_genes)))
```

# RBP enrichment among NMD-target genes

```{r rbp-enrichment}
g2sym  <- iso2gene %>% distinct(gene_id, hgnc_symbol)
sym_of <- function(gid) g2sym$hgnc_symbol[match(gid, g2sym$gene_id)]
is_rbp <- function(gid) (sv(gid) %in% rbp_ens) | (sym_of(gid) %in% rbp_sym)

cat(sprintf("NMD-target genes that are RBPs (of 1,542): %d (%.1f%%)\n",
            sum(is_rbp(nmd_genes)), 100*mean(is_rbp(nmd_genes))))

# enrichment vs all DIE-tested genes (one-sided Fisher)
tab <- table(NMD_target = bg_genes %in% nmd_genes, RBP = is_rbp(bg_genes))
print(tab)
ft <- fisher.test(tab, alternative="greater")
cat(sprintf("\nRBP enrichment among NMD targets: OR=%.2f (95%% CI %.2f-%.2f), p=%.2e\n",
            ft$estimate, ft$conf.int[1], ft$conf.int[2], ft$p.value))

kable(data.frame(
  set            = c("NMD-target genes","background (DIE-tested)"),
  n_genes        = c(length(nmd_genes), length(bg_genes)),
  n_RBP          = c(sum(is_rbp(nmd_genes)), sum(is_rbp(bg_genes))),
  pct_RBP        = round(100*c(mean(is_rbp(nmd_genes)), mean(is_rbp(bg_genes))),1)),
  caption="RBP fraction among NMD-target genes vs all DIE-tested genes (Gerstberger 2014 census).")
```

# Per-cell-type RBP enrichment

```{r rbp-enrichment-by-ct}
kable(bind_rows(lapply(CELL_TYPES, function(ct){
  gset <- unique(nmd_gene_ct$gene_id[nmd_gene_ct$ct==ct])
  tb   <- table(NMD = bg_genes %in% gset, RBP = is_rbp(bg_genes))
  f    <- fisher.test(tb, alternative="greater")
  data.frame(ct=ct, n_nmd=length(gset), n_rbp=sum(is_rbp(gset)),
             pct_rbp=round(100*mean(is_rbp(gset)),1),
             OR=round(f$estimate,2), p=signif(f$p.value,3))
})), caption="RBP enrichment among NMD-target genes, per cell type.")
```

# SR proteins (SRSF1-12): which are NMD targets, by cell type

```{r sr-distribution}
sr_tab <- nmd_gene_ct %>% mutate(sym=sym_of(gene_id)) %>% filter(sym %in% sr_family) %>%
  distinct(sym, ct) %>% mutate(present=1L) %>%
  pivot_wider(names_from=ct, values_from=present, values_fill=0) %>%
  mutate(n_celltypes = rowSums(across(all_of(intersect(CELL_TYPES, colnames(.)))))) %>%
  arrange(desc(n_celltypes))
kable(sr_tab, caption="Canonical SR proteins (SRSF1-12) that are NMD targets, by cell type (1 = NMD isoform present).")
```

# Full list of NMD-target RBPs (supplement)

```{r nmd-rbp-list}
nmd_rbp <- data.frame(gene_id = nmd_genes[is_rbp(nmd_genes)]) %>%
  mutate(gene_symbol = sym_of(gene_id),
         consensus_rna_target = rbp$consensus_rna_target[match(gene_symbol, rbp$gene_symbol)]) %>%
  arrange(gene_symbol)
cat(sprintf("NMD-target RBPs: %d\n", nrow(nmd_rbp)))
print(nmd_rbp %>% count(consensus_rna_target, name="n") %>% arrange(desc(n)))
fwrite(nmd_rbp, file.path(OUT_DIR, paste0("nmd_target_rbps_", DATE_STAMP, ".csv")))
fwrite(nmd_gene_ct, file.path(OUT_DIR, paste0("nmd_target_genes_by_ct_", DATE_STAMP, ".csv")))
cat("Saved to:", OUT_DIR, "\n")
```

```{r session}
sessionInfo()
```
