---
title: "Productive output response to NMD inhibition"
author: "Yul Leshem"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output:
  html_document: { toc: true, toc_float: true, df_print: paged, code_folding: hide, number_sections: true }
---

Code location: `/udd/reyle/nmd_lungcells_2026/code/final/productive_response.Rmd`

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE, warning=FALSE, message=FALSE, fig.width=11, fig.height=5, dpi=110)
options(bitmapType="cairo"); .libPaths(c("/udd/reyle/Rlibs", .libPaths()))
```

```{r libs}
suppressPackageStartupMessages({
  library(data.table); library(dplyr); library(tidyr); library(tibble)
  library(ggplot2); library(scales); library(knitr); library(limma); library(edgeR)
})
select <- dplyr::select; filter <- dplyr::filter; rename <- dplyr::rename; count <- dplyr::count
```

# Overview and conventions

This document builds the **custom NMD-anchored normalization** for the productive-output analysis,
then walks the per-gene quantities through their distributions, modeling (limma/mashr), a
reproducibility battery, the up/down model, pathways, and the two-program model, across the four
primary lung cell types (AT2, LAE, FB, MV). The DGEList stores the original ct codes `AT` (= AT2)
and `DD` (= LAE); we relabel to display names right after loading and use AT2 / LAE throughout.

**Variable glossary (no cryptic abbreviations).** Every per-gene quantity below is a mean across
that cell type's donors:

| name | meaning |
|---|---|
| `prodCPM_DMSO`     | productive CPM, DMSO |
| `unprodCPM_DMSO`   | unproductive CPM, DMSO |
| `prodCPM_SMG_raw`  | productive CPM, SMG1i (raw, whole-library) |
| `unprodCPM_SMG`    | unproductive CPM, SMG1i |
| `prodCPM_SMG_adj`  | productive CPM, SMG1i (**custom-adjusted**: unproductive surge removed) |
| `totalCPM_DMSO`    | `prodCPM_DMSO + unprodCPM_DMSO` (gene total, DMSO) |
| `totalCPM_SMG`     | `prodCPM_SMG_raw + unprodCPM_SMG` (gene total, SMG1i) |
| `prodfrac_DMSO`    | `prodCPM_DMSO / totalCPM_DMSO` (productive share, DMSO) |
| `prodfrac_SMG`     | `prodCPM_SMG_raw / totalCPM_SMG` (productive share, SMG1i) |
| `custom`           | adjusted productive log2FC = `log2((prodCPM_SMG_adj+1)/(prodCPM_DMSO+1))` — **surge removed** |
| `Kitagawa`         | raw productive log2FC = `log2((prodCPM_SMG_raw+1)/(prodCPM_DMSO+1))` — surge included |
| `prod_shift_adj`   | **productive-only output shift** = `(prodCPM_SMG_adj − prodCPM_DMSO)/totalCPM_DMSO` — fixed DMSO denominator, surge removed |
| `prodfrac_change`  | `prodfrac_SMG − prodfrac_DMSO` — productive *share* change (**includes** the unproductive surge) |
| `dprod_adj`        | `prodCPM_SMG_adj − prodCPM_DMSO` (adjusted productive CPM change) |
| `dprod_raw`        | `prodCPM_SMG_raw − prodCPM_DMSO` (raw productive CPM change) |
| `unprod_gain`      | `unprodCPM_SMG − unprodCPM_DMSO` (the unproductive surge) |
| `n_iso`            | isoforms per gene |

**Other conventions:**

- "NMD gene" = a gene with **≥1 mashr-confident unproductive isoform** (`lfsr<0.05 & posterior mean>0`).
- "**change in productive**" always means the **adjusted** measure (`custom` in log space, or
  `prod_shift_adj` / `dprod_adj` in CPM space) — i.e. the unproductive surge is removed.
- Main figures show **NMD vs non-NMD**. The **uncertain ("unsure")** group is set aside to the supplement.
- FB is treated as **provisional** (smaller effect sizes; fails part of the reproducibility battery) and
  is excluded from the mechanism / Kitagawa summaries, which use AT2 / LAE / MV.
- Every data-shaping step prints a `cat()` summary so the numbers are auditable inline.

```{r params}
proj    <- "/udd/reyle/nmd_lungcells_2026"
OUT_DIR <- file.path(proj, "long_read/productive_response/basics_ordered")
dir.create(OUT_DIR, recursive=TRUE, showWarnings=FALSE)
DATE_STAMP   <- format(Sys.Date(), "%Y.%m.%d")
CELL_TYPES   <- c("AT2","LAE","FB","MV")
LFSR_NMD_THR    <- 0.05
LFSR_UNSURE_THR <- 0.20
FLOOR_BOTH <- 1; FLOOR_DMSO <- 2
pass_floor <- function(td, ts) (td>=FLOOR_BOTH & ts>=FLOOR_BOTH) | (td>=FLOOR_DMSO)
DROP_THR <- 0.3
ct_colors  <- c(AT2="#e74c3c", LAE="#3498db", FB="#16a085", MV="#9b59b6")
cat_colors <- c("NMD"="#c0392b","non-NMD"="#7f8c8d","unsure"="#e67e22")
MAIN_CATS  <- c("NMD","non-NMD")

DGE_ISO_PATH  <- file.path(proj, "short_read/dge_isoform_longread_filtered_2026.3.3.rds")
PM_ISO_PATH   <- file.path(proj, "long_read/mashr/mashr_isoform_posterior_means_2026.3.10.csv")
LFSR_ISO_PATH <- file.path(proj, "long_read/mashr/mashr_isoform_lfsr_2026.3.10.csv")
SQANTI_PATH   <- "/proj/regeps/regep00/studies/ExternalCellLines/analyses/repjc/Randell_Lung_Cells_2025/code/sqanti3_by_chromosome/projects/nmd_lungcells/results/nmd_lungcells_classification.txt"
SR_DIR        <- "/udd/reyle/nmd_lungcells_2026/shortread_dge/mashr"
SR_PM_PATH    <- file.path(SR_DIR, "mashr_posterior_means_2026.3.10.csv")
SR_LFSR_PATH  <- file.path(SR_DIR, "mashr_lfsr_2026.3.10.csv")
```

# Step 1 — Load, classify, annotate

```{r load}
dge_iso  <- readRDS(DGE_ISO_PATH)
pm_iso   <- read.csv(PM_ISO_PATH,   check.names=FALSE, stringsAsFactors=FALSE)
lfsr_iso <- read.csv(LFSR_ISO_PATH, check.names=FALSE, stringsAsFactors=FALSE)
colnames(pm_iso)[1] <- "isoform_id"; colnames(lfsr_iso)[1] <- "isoform_id"

iso_col  <- intersect(c("isoform_id","isoform","txid","transcript_id"), colnames(dge_iso$genes))[1]
gene_col <- intersect(c("gene_id","associated_gene","ensembl_gene_id","gene_id_ens115_sqanti"), colnames(dge_iso$genes))[1]
sym_col  <- intersect(c("hgnc_symbol","gene_name","external_gene_name","symbol"), colnames(dge_iso$genes))[1]
iso2gene <- data.frame(isoform_id=dge_iso$genes[[iso_col]], gene_id=dge_iso$genes[[gene_col]],
                       hgnc_symbol=dge_iso$genes[[sym_col]], stringsAsFactors=FALSE)
n_iso_tab <- iso2gene %>% count(gene_id, name="n_iso")

samples <- as.data.frame(dge_iso$samples)
ct_col    <- intersect(c("ct","cell_type"), colnames(samples))[1]
trt_col   <- intersect(c("treatment","trt"), colnames(samples))[1]
donor_col <- intersect(c("id","sample_id","donor","subject_id","donor_id","subject","patient"), colnames(samples))[1]
samples$cell_type <- as.character(samples[[ct_col]]); samples$treatment <- as.character(samples[[trt_col]])
samples$sample_id <- as.character(samples[[donor_col]])
# relabel internal codes to display names AFTER loading (rows aligned to counts by rowname)
samples$cell_type[samples$cell_type=="AT"] <- "AT2"
samples$cell_type[samples$cell_type=="DD"] <- "LAE"
samples <- samples[samples$cell_type %in% CELL_TYPES, ]
counts_iso <- dge_iso$counts[, rownames(samples)]

cat(sprintf("Loaded: %d isoforms x %d samples; %d genes (%d multi-isoform).\n",
            nrow(counts_iso), ncol(counts_iso), nrow(n_iso_tab), sum(n_iso_tab$n_iso>1)))
cat("Samples per CT x treatment:\n"); print(table(samples$cell_type, samples$treatment))

resolve_mash_col <- function(df, ct){ cn<-colnames(df); if (ct %in% cn) return(ct)
  h<-cn[grepl(paste0("_in_",ct,"$"),cn)]; if(length(h)==1) return(h)
  h<-cn[grepl(paste0("^",ct,"[._:-]"),cn)]; if(length(h)==1) return(h); stop("no mash col ",ct) }
```

```{r pc-annot}
sq <- fread(SQANTI_PATH, select=c("isoform","associated_gene","coding"))
gcs <- sq %>% inner_join(iso2gene %>% select(isoform_id, gene_id) %>% distinct(),
                         by=c("isoform"="isoform_id")) %>%
  group_by(gene_id) %>% summarise(is_pc = sum(coding=="coding")>0, .groups="drop")
PC_GENES <- gcs$gene_id[gcs$is_pc]
iso2gene <- iso2gene %>% left_join(gcs, by="gene_id")
cat(sprintf("Protein-coding genes (>=1 coding isoform in SQANTI): %d\n", length(PC_GENES)))
```

```{r helpers}
# ---- shared helpers (defined ONCE here; referenced throughout) ----
suppressPackageStartupMessages({ library(org.Hs.eg.db); library(AnnotationDbi) })

svf <- function(x) sub("\\..*$","", x)          # strip Ensembl version suffix

# GO term(s) -> SYMBOL (entrez route)  and  GO term(s) -> ENSEMBL
go_to_symbols <- function(go_ids){
  e <- unique(unlist(AnnotationDbi::mget(go_ids, org.Hs.egGO2ALLEGS, ifnotfound=NA)))
  unique(na.omit(AnnotationDbi::mapIds(org.Hs.eg.db, keys=e, column="SYMBOL",
                                       keytype="ENTREZID", multiVals="first")))
}
go_genes <- function(go_ids){
  e <- tryCatch(unique(unlist(AnnotationDbi::mget(go_ids, org.Hs.egGO2ALLEGS))),
                error=function(...) character(0))
  unique(na.omit(AnnotationDbi::mapIds(org.Hs.eg.db, keys=e, column="ENSEMBL",
                                       keytype="ENTREZID", multiVals="first")))
}

# genes (gene_id) with >=1 isoform significantly UP (NMD up-call) / DOWN in a cell type
nmd_iso_genes <- function(ct){
  l <- lfsr_iso[[resolve_mash_col(lfsr_iso,ct)]]; p <- pm_iso[[resolve_mash_col(pm_iso,ct)]]
  unique(iso2gene$gene_id[iso2gene$isoform_id %in%
           lfsr_iso$isoform_id[!is.na(l)&!is.na(p)&l<LFSR_NMD_THR&p>0]])
}
down_iso_genes <- function(ct){
  l <- lfsr_iso[[resolve_mash_col(lfsr_iso,ct)]]; p <- pm_iso[[resolve_mash_col(pm_iso,ct)]]
  unique(iso2gene$gene_id[iso2gene$isoform_id %in%
           lfsr_iso$isoform_id[!is.na(l)&!is.na(p)&l<LFSR_NMD_THR&p<0]])
}

# one-sided Fisher with safe fallback; and a set-enrichment wrapper given a background
safe_fisher <- function(hit, inset){
  tab <- table(factor(hit, c(FALSE,TRUE)), factor(inset, c(FALSE,TRUE)))
  tryCatch(fisher.test(tab, alternative="greater"),
           error=function(e) list(estimate=NA_real_, p.value=NA_real_))
}
test_set <- function(hit_syms, set_syms, bg_syms){
  if (length(hit_syms) < 3) return(data.frame(n_in=sum(hit_syms %in% set_syms), OR=NA, p=NA))
  ft <- safe_fisher(bg_syms %in% hit_syms, bg_syms %in% set_syms)
  data.frame(n_in=sum(hit_syms %in% set_syms), OR=round(as.numeric(ft$estimate),2), p=signif(ft$p.value,3))
}
sym_of <- function(g) iso2gene$hgnc_symbol[match(g, iso2gene$gene_id)]

# Kitagawa level/composition decomposition for a base-like df
kitagawa_decomp <- function(df){
  df %>% dplyr::mutate(
    T_d = prodCPM_DMSO + unprodCPM_DMSO,
    T_s = prodCPM_SMG_raw + unprodCPM_SMG,
    f_d = prodCPM_DMSO / T_d,
    f_s = prodCPM_SMG_raw / T_s,
    dP  = prodCPM_SMG_raw - prodCPM_DMSO,
    level       = ((f_s + f_d)/2) * (T_s - T_d),
    composition = ((T_s + T_d)/2) * (f_s - f_d),
    driver = ifelse(abs(composition) > abs(level), "composition-dominated", "level-dominated"))
}

# Kitagawa decomposition on significant NMD genes, built once (FB excluded: provisional CT).
# References prod_dir_mashr and base, which are created in later steps (resolved at call time).
kit_nmd_table <- function(exclude_fb = TRUE){
  d <- prod_dir_mashr %>% dplyr::filter(dir %in% c("up","down"), gene_category=="NMD")
  if (exclude_fb) d <- d %>% dplyr::filter(ct != "FB")
  d %>% dplyr::left_join(base %>% dplyr::select(ct, gene_id, prodCPM_DMSO, prodCPM_SMG_raw,
                          unprodCPM_DMSO, unprodCPM_SMG), by=c("ct","gene_id")) %>%
    kitagawa_decomp()
}

# mechanism split for productive-direction genes (needs sr_logFC, prodfrac_DMSO/SMG joined)
classify_mech <- function(df, dir=c("down","up")){
  dir <- match.arg(dir)
  if (dir=="down")
    df %>% dplyr::mutate(prodfrac_change = prodfrac_SMG - prodfrac_DMSO,
      mechanism = dplyr::case_when(
        !is.na(sr_logFC) & sr_logFC < -0.2                            ~ "transcriptional",
        prodfrac_change < -0.10 & (is.na(sr_logFC) | sr_logFC > -0.1) ~ "composition",
        TRUE                                                          ~ "mixed"))
  else
    df %>% dplyr::mutate(prodfrac_change = prodfrac_SMG - prodfrac_DMSO,
      mechanism = dplyr::case_when(
        !is.na(sr_logFC) & sr_logFC > 0.2                            ~ "transcriptional",
        prodfrac_change > 0.10 & (is.na(sr_logFC) | sr_logFC < 0.1)  ~ "composition",
        TRUE                                                         ~ "mixed"))
}

# ---- theme gene-sets (built once) + reusable enrichment (used by S6 / S7) ----
THEME_SETS <- list(
  "Mitotic cell cycle" = go_to_symbols("GO:0000278"),
  "DNA replication"    = go_to_symbols("GO:0006260"),
  "DNA repair"         = go_to_symbols("GO:0006281"),
  "RNA splicing"       = go_to_symbols("GO:0008380"),
  "SR/hnRNP factors"   = c(paste0("SRSF",1:12),
                           paste0("HNRNP",c("A1","A2B1","A3","C","D","F","H1","K","L","M","U")),
                           "TRA2A","TRA2B","RBM39","PTBP1","PTBP2","U2AF1","U2AF2"),
  "ISR/UPR"            = go_to_symbols(c("GO:0006986","GO:0140467")))
theme_enrichment <- function(trans_g, comp_g, universe_g, sets = THEME_SETS){
  bg <- unique(na.omit(sym_of(universe_g)))
  dplyr::bind_rows(lapply(names(sets), function(s){
    t <- test_set(unique(na.omit(sym_of(trans_g))), sets[[s]], bg)
    c <- test_set(unique(na.omit(sym_of(comp_g))),  sets[[s]], bg)
    data.frame(theme=s, trans_OR=t$OR, trans_p=t$p, trans_n=t$n_in,
               comp_OR=c$OR, comp_p=c$p, comp_n=c$n_in) }))
}
```

# Step 2 — Custom NMD-anchored normalization (equations)

The whole-library CPM denominator is inflated by the unproductive surge under SMG1i, which makes
every other isoform look artificially smaller. The custom normalization removes exactly that
artifact by **pinning each unproductive isoform to the DMSO proportion of its gene's productive
compartment**, then rescaling the library.

**Raw CPM** for isoform $i$ in sample $s$:
$$\text{CPM}_{i,s} = 10^6\cdot\frac{c_{i,s}}{\sum_j c_{j,s}}$$

**Productive compartment of gene $g$ at DMSO** (sum over productive isoforms, mean across DMSO replicates) — this is `prodCPM_DMSO`:
$$\texttt{prodCPM\_DMSO}_g = \sum_{i\,\in\,\mathrm{prod}(g)} \overline{\text{CPM}}^{DMSO}_i$$

**DMSO share** of each unproductive isoform within its gene's productive compartment:
$$\text{share}_i = \frac{\overline{\text{CPM}}^{DMSO}_i}{\texttt{prodCPM\_DMSO}_g}$$

**Adjusted unproductive CPM** under SMG1i sample $s$ (anchorable isoforms re-pinned to the DMSO share
of the *current* productive compartment; non-anchorable isoforms held at their DMSO mean):
$$\widetilde{u}_{i,s} = \text{share}_i \cdot \Big(\textstyle\sum_{j\in\mathrm{prod}(g)}\text{CPM}^{SMG1i}_{j,s}\Big)$$

**Library rescale** so each adjusted SMG1i library sums to $10^6$:
$$\lambda_s = \frac{10^6}{\sum_i \widetilde{\text{CPM}}_{i,s}},\qquad
\text{CPM}^{adj}_{i,s} = \widetilde{\text{CPM}}_{i,s}\cdot \lambda_s$$

**Per-gene productive quantities** (summed over a gene's productive isoforms, mean across donors):
`prodCPM_DMSO`, `prodCPM_SMG_raw` (from raw SMG1i CPM), and `prodCPM_SMG_adj` (from the adjusted CPM above).

```{r custom-build}
build_adj <- function(ct) {
  l <- lfsr_iso[[resolve_mash_col(lfsr_iso, ct)]]; p <- pm_iso[[resolve_mash_col(pm_iso, ct)]]
  is_unprod <- !is.na(l)&!is.na(p)&l<LFSR_NMD_THR&p>0
  is_unsure <- !is.na(l)&!is.na(p)&l>=LFSR_NMD_THR&l<LFSR_UNSURE_THR&p>0
  is_down   <- !is.na(l)&!is.na(p)&l<LFSR_NMD_THR&p<0
  cls <- setNames(ifelse(is_unprod,"unprod","prod"), lfsr_iso$isoform_id)
  uns <- setNames(is_unsure, lfsr_iso$isoform_id)
  dwn <- setNames(is_down,   lfsr_iso$isoform_id)
  cs <- rownames(samples)[samples$cell_type==ct]; cnt <- counts_iso[, cs, drop=FALSE]
  trt <- setNames(samples[cs,"treatment"], cs); donor <- setNames(samples[cs,"sample_id"], cs)
  rawcpm <- sweep(cnt, 2, colSums(cnt), "/") * 1e6
  ids <- rownames(rawcpm); gene <- iso2gene$gene_id[match(ids, iso2gene$isoform_id)]
  class_v <- cls[ids]
  unsure_v <- uns[ids]; unsure_v[is.na(unsure_v)] <- FALSE
  down_v   <- dwn[ids]; down_v[is.na(down_v)] <- FALSE
  anchor <- !is.na(gene) & !grepl("::", gene, fixed=TRUE) & !is.na(class_v)
  dmso <- cs[trt[cs]=="DMSO"]; smg <- cs[trt[cs]=="Smg1i"]
  dmso_mean <- rowMeans(rawcpm[, dmso, drop=FALSE])
  prod_idx  <- anchor & class_v=="prod"
  Pdmso     <- tapply(dmso_mean[prod_idx], gene[prod_idx], sum)
  iu <- which(anchor & class_v=="unprod"); pg <- Pdmso[gene[iu]]
  hp <- !is.na(pg) & pg > 0; share <- ifelse(hp, dmso_mean[iu]/pg, NA_real_)
  ia <- iu[hp]; ina <- iu[!hp]
  adj <- rawcpm; rescale <- setNames(numeric(length(smg)), smg)
  for (s in smg) {
    col <- rawcpm[, s]; Pgs <- tapply(col[prod_idx], gene[prod_idx], sum); nc <- col
    nc[ia] <- Pgs[gene[ia]] * share[hp]; nc[ina] <- dmso_mean[ina]; nc[!is.finite(nc)] <- 0
    f <- 1e6/sum(nc); rescale[s] <- f; adj[, s] <- nc * f
  }
  list(rawcpm=rawcpm, adj=adj, gene=gene, class_v=class_v, unsure_v=unsure_v, down_v=down_v,
       anchor=anchor, trt=trt, donor=donor, rescale=rescale)
}

process_ct <- function(ct) {
  o <- build_adj(ct); k <- o$anchor; fk <- paste(o$gene[k], o$class_v[k], sep="|||")
  rawf <- rowsum(o$rawcpm[k,,drop=FALSE], fk); adjf <- rowsum(o$adj[k,,drop=FALSE], fk)
  fm <- data.frame(feature=rownames(rawf)); fm$gene <- sub("\\|\\|\\|.*$","",fm$feature); fm$class <- sub("^.*\\|\\|\\|","",fm$feature)
  sm <- data.frame(sample=colnames(rawf), donor=o$donor[colnames(rawf)], trt=o$trt[colnames(rawf)])
  melt <- function(m,v){d<-as.data.frame(m); d$feature<-rownames(m); pivot_longer(d,-feature,names_to="sample",values_to=v)}
  long <- melt(adjf,"adj") %>% left_join(melt(rawf,"raw"), by=c("feature","sample")) %>%
    left_join(fm, by="feature") %>% left_join(sm, by="sample")
  # per-donor gene-level columns (clear names)
  dm <- long %>% filter(trt=="DMSO") %>% select(gene,class,donor,cpm=adj) %>%
    pivot_wider(names_from=class, values_from=cpm, values_fill=0) %>%
    rename(prodCPM_DMSO=prod, unprodCPM_DMSO=unprod)
  sg <- long %>% filter(trt=="Smg1i") %>% select(gene,class,donor,adj,raw) %>%
    pivot_wider(names_from=class, values_from=c(adj,raw), values_fill=0) %>%
    rename(prodCPM_SMG_adj=adj_prod, unprodCPM_SMG_adj=adj_unprod,
           prodCPM_SMG_raw=raw_prod, unprodCPM_SMG=raw_unprod)
  nmd_g <- unique(o$gene[o$anchor & o$class_v=="unprod"])
  unsure_g <- setdiff(unique(o$gene[o$anchor & (o$unsure_v | o$down_v)]), nmd_g)
  full_join(dm, sg, by=c("gene","donor")) %>%
    mutate(ct=ct, gene_category = case_when(gene %in% nmd_g ~ "NMD", gene %in% unsure_g ~ "unsure", TRUE ~ "non-NMD"))
}

gene_sample <- bind_rows(lapply(CELL_TYPES, process_ct)) %>%
  left_join(iso2gene %>% select(gene_id, hgnc_symbol, is_pc) %>% distinct(), by=c("gene"="gene_id")) %>%
  rename(gene_id = gene)
cat(sprintf("Per-gene-per-donor rows assembled: %d\n", nrow(gene_sample)))
```

**Derived per-gene quantities** (means across donors). Totals and fractions:
$$\texttt{totalCPM\_DMSO}=\texttt{prodCPM\_DMSO}+\texttt{unprodCPM\_DMSO},\quad
\texttt{totalCPM\_SMG}=\texttt{prodCPM\_SMG\_raw}+\texttt{unprodCPM\_SMG}$$
$$\texttt{prodfrac\_DMSO}=\frac{\texttt{prodCPM\_DMSO}}{\texttt{totalCPM\_DMSO}},\qquad
\texttt{prodfrac\_SMG}=\frac{\texttt{prodCPM\_SMG\_raw}}{\texttt{totalCPM\_SMG}}$$

Productive-**output** measures (unproductive surge removed via the adjustment):
$$\underbrace{\texttt{custom}=\log_2\frac{\texttt{prodCPM\_SMG\_adj}+1}{\texttt{prodCPM\_DMSO}+1}}_{\text{adjusted productive log2FC}}\qquad
\underbrace{\texttt{prod\_shift\_adj}=\frac{\texttt{prodCPM\_SMG\_adj}-\texttt{prodCPM\_DMSO}}{\texttt{totalCPM\_DMSO}}}_{\text{productive-only shift, fixed DMSO denominator}}$$

Productive-**share** / raw comparators (unproductive surge included):
$$\underbrace{\texttt{Kitagawa}=\log_2\frac{\texttt{prodCPM\_SMG\_raw}+1}{\texttt{prodCPM\_DMSO}+1}}_{\text{raw whole-library log2FC}}\qquad
\underbrace{\texttt{prodfrac\_change}=\texttt{prodfrac\_SMG}-\texttt{prodfrac\_DMSO}}_{\text{productive share change (incl. surge)}}$$

```{r base-build}
base <- gene_sample %>%
  filter(!is.na(prodCPM_DMSO), !is.na(prodCPM_SMG_raw)) %>%
  mutate(across(c(prodCPM_DMSO,unprodCPM_DMSO,prodCPM_SMG_raw,unprodCPM_SMG,prodCPM_SMG_adj),
                ~dplyr::coalesce(.x,0))) %>%
  group_by(ct, gene_id, hgnc_symbol, is_pc, gene_category) %>%
  summarise(prodCPM_DMSO    = mean(prodCPM_DMSO),
            unprodCPM_DMSO  = mean(unprodCPM_DMSO),
            prodCPM_SMG_raw = mean(prodCPM_SMG_raw),
            unprodCPM_SMG   = mean(unprodCPM_SMG),
            prodCPM_SMG_adj = mean(prodCPM_SMG_adj), .groups="drop") %>%
  mutate(totalCPM_DMSO = prodCPM_DMSO + unprodCPM_DMSO,
         totalCPM_SMG  = prodCPM_SMG_raw + unprodCPM_SMG) %>%
  filter(pass_floor(totalCPM_DMSO, totalCPM_SMG)) %>%
  mutate(
    # productive OUTPUT (unproductive surge removed)
    custom         = log2((prodCPM_SMG_adj + 1)/(prodCPM_DMSO + 1)),
    prod_shift_adj = (prodCPM_SMG_adj - prodCPM_DMSO)/totalCPM_DMSO,
    dprod_adj      = prodCPM_SMG_adj - prodCPM_DMSO,
    # raw / share comparators (surge included)
    Kitagawa        = log2((prodCPM_SMG_raw + 1)/(prodCPM_DMSO + 1)),
    dprod_raw       = prodCPM_SMG_raw - prodCPM_DMSO,
    prodfrac_DMSO   = prodCPM_DMSO   / totalCPM_DMSO,
    prodfrac_SMG    = prodCPM_SMG_raw / totalCPM_SMG,
    prodfrac_change = prodfrac_SMG - prodfrac_DMSO,   # NEGATIVE = productive SHARE lower in SMG1i
    # unproductive surge
    unprod_gain     = unprodCPM_SMG - unprodCPM_DMSO) %>%
  left_join(n_iso_tab, by="gene_id")

cat("Genes per CT by category (after expression floor):\n")
print(base %>% count(ct, gene_category) %>% pivot_wider(names_from=gene_category, values_from=n, values_fill=0))
cat(sprintf("\nMedian custom (adjusted productive log2FC, NMD) per CT: %s\n",
    paste(sprintf("%s=%.3f", CELL_TYPES,
      sapply(CELL_TYPES, function(c) median(base$custom[base$ct==c & base$gene_category=="NMD"]))), collapse="  ")))
cat(sprintf("Median prod_shift_adj (productive-only shift, NMD) per CT: %s\n",
    paste(sprintf("%s=%.4f", CELL_TYPES,
      sapply(CELL_TYPES, function(c) median(base$prod_shift_adj[base$ct==c & base$gene_category=="NMD"]))), collapse="  ")))
fwrite(base, file.path(OUT_DIR, paste0("base_custom_", DATE_STAMP, ".csv")))
```

```{r cat-defs}
cat_defs <- tibble::tribble(
  ~gene_category, ~isoform_level_filter, ~used_in,
  "NMD",      "≥1 isoform with mashr lfsr<0.05 & pm>0 (confident unproductive / NMD up-call)", "main + supplement",
  "non-NMD",  "no isoform with any significant or borderline change (clean control)",          "main + supplement",
  "unsure",   "NOT NMD, but ≥1 isoform borderline-up (0.05≤lfsr<0.20 & pm>0) OR significantly DOWN (lfsr<0.05 & pm<0)", "SUPPLEMENT ONLY")
kable(cat_defs, caption="Gene-category filters.")
```

# Step 3 — Distributions of every computed variable

CPM-type variables are strictly non-negative and heavily right-skewed, so they are shown on
`log10(x+1)`; change/difference variables on their natural (signed) scale. Each panel is faceted
by cell type and colored by gene class (**NMD vs non-NMD** in main; **unsure** added in the supplement).

```{r dist-helper}
within_sd <- function(x, k=3){ m <- mean(x, na.rm=TRUE); s <- sd(x, na.rm=TRUE); is.finite(x) & abs(x - m) <= k*s }

plot_dist <- function(df, var, title, xlab, cats=MAIN_CATS, logx=FALSE, xlim=NULL, vline=NULL, trim_sd=NULL){
  d <- df %>% filter(gene_category %in% cats) %>% mutate(.v = .data[[var]]) %>% filter(is.finite(.v))
  if (logx){ d$.v <- log10(d$.v + 1); xlab <- paste0("log10(", xlab, " + 1)") }
  sub <- NULL
  if (!is.null(trim_sd)){
    n0 <- nrow(d)
    d  <- d %>% group_by(ct) %>% filter(within_sd(.v, trim_sd)) %>% ungroup()
    sub <- sprintf("display trimmed to ±%g SD of the per-cell-type mean (%d of %d points removed for the plot only)",
                   trim_sd, n0 - nrow(d), n0)
  }
  g <- ggplot(d, aes(.v, fill=gene_category, colour=gene_category)) +
    geom_density(alpha=0.30, linewidth=0.5) +
    facet_wrap(~ct, nrow=1, scales="free_y") +
    scale_fill_manual(values=cat_colors, name=NULL) + scale_colour_manual(values=cat_colors, name=NULL) +
    labs(x=xlab, y="density", title=title, subtitle=sub) +
    theme_bw(base_size=11) + theme(legend.position="top", strip.background=element_rect(fill="grey92"))
  if (!is.null(vline)) g <- g + geom_vline(xintercept=vline, linetype=2, colour="grey40")
  if (!is.null(xlim))  g <- g + coord_cartesian(xlim=xlim)
  g
}
med_tab <- function(df, var, cats=MAIN_CATS){
  df %>% filter(gene_category %in% cats) %>% group_by(ct, gene_category) %>%
    summarise(n=n(), median=round(median(.data[[var]], na.rm=TRUE),3),
              q25=round(quantile(.data[[var]],.25,na.rm=TRUE),3),
              q75=round(quantile(.data[[var]],.75,na.rm=TRUE),3), .groups="drop") %>%
    pivot_wider(names_from=gene_category, values_from=c(n, median, q25, q75))
}
```

## 3.1 Productive-only output shift (adjusted; the unproductive response removed)

`prod_shift_adj = (prodCPM_SMG_adj − prodCPM_DMSO) / totalCPM_DMSO`. Numerator is the **adjusted**
productive CPM change (surge removed); denominator is **fixed at the DMSO gene total**. −0.10 means
productive output fell by 10% of baseline total; ~0 means preserved.

```{r dist-prodshift, fig.height=4}
print(plot_dist(base, "prod_shift_adj",
  "Productive-only output shift (negative = productive output lower in SMG1i)",
  xlab="change in productive output (fraction of DMSO gene total)",
  xlim=c(-0.5,0.5), vline=0))
kable(med_tab(base, "prod_shift_adj"),
  caption="Productive-only output shift (adjusted numerator, fixed DMSO denominator).")
cat(sprintf("NMD genes with prod_shift_adj<0 (productive output lower in SMG1i) per CT: %s\n",
  paste(sprintf("%s=%.1f%%", CELL_TYPES,
    sapply(CELL_TYPES, function(c){x<-base$prod_shift_adj[base$ct==c&base$gene_category=="NMD"];100*mean(x<0)})), collapse="  ")))
```

## 3.1b Productive-share change — for contrast (this one INCLUDES the unproductive surge)

```{r dist-prodfrac, fig.height=4}
print(plot_dist(base, "prodfrac_change",
  "Productive-SHARE change (negative = share lower in SMG1i; INCLUDES unproductive surge)",
  xlab="change in productive fraction of the gene",
  xlim=c(-0.75,0.5), vline=0))
xo <- base$prodfrac_change[base$gene_category %in% MAIN_CATS]
cat(sprintf("prodfrac_change display window [-0.75, 0.5]: %.2f%% of genes fall outside (%.2f%% below -0.75, %.2f%% above 0.5); data clipped for the view only.\n",
            100*mean(xo < -0.75 | xo > 0.5, na.rm=TRUE),
            100*mean(xo < -0.75, na.rm=TRUE), 100*mean(xo > 0.5, na.rm=TRUE)))
kable(med_tab(base, "prodfrac_change"),
  caption="Productive-share change (includes the unproductive surge via the growing denominator).")
```

## 3.2 CPM of DMSO productive (`prodCPM_DMSO`)

```{r dist-prodDMSO, fig.height=4}
print(plot_dist(base, "prodCPM_DMSO", "DMSO productive CPM", xlab="productive CPM (DMSO)", logx=TRUE))
kable(med_tab(base, "prodCPM_DMSO"), caption="DMSO productive CPM by class.")
```

## 3.3 CPM of SMG1i productive — raw (`prodCPM_SMG_raw`)

```{r dist-prodSMGraw, fig.height=4}
print(plot_dist(base, "prodCPM_SMG_raw", "SMG1i productive CPM (raw)", xlab="productive CPM (SMG1i, raw)", logx=TRUE))
kable(med_tab(base, "prodCPM_SMG_raw"), caption="Raw SMG1i productive CPM by class.")
```

## 3.4 CPM of SMG1i productive — adjusted (`prodCPM_SMG_adj`)

```{r dist-prodSMGadj, fig.height=4}
print(plot_dist(base, "prodCPM_SMG_adj", "SMG1i productive CPM (custom-adjusted)", xlab="productive CPM (SMG1i, adjusted)", logx=TRUE))
kable(med_tab(base, "prodCPM_SMG_adj"), caption="Adjusted SMG1i productive CPM by class.")
cat("Note: prodCPM_SMG_adj > prodCPM_SMG_raw on average because rescaling redistributes the CPM the unproductive surge had stolen.\n")
```

## 3.5 Change in productive (CPM, adjusted) `dprod_adj = prodCPM_SMG_adj − prodCPM_DMSO`

```{r dist-dprod, fig.height=4}
print(ggplot(base %>% filter(gene_category %in% MAIN_CATS),
       aes(gene_category, dprod_adj, fill=gene_category)) +
  geom_hline(yintercept=0, colour="grey40") +
  geom_boxplot(outlier.size=0.3, outlier.alpha=0.2, width=0.6, colour="grey25") +
  facet_wrap(~ct, nrow=1) + coord_cartesian(ylim=c(-25, 25)) +
  scale_fill_manual(values=cat_colors, guide="none") +
  labs(x=NULL, y="change in productive CPM (SMG1i − DMSO, adjusted)",
       title="Change in productive (adjusted CPM), NMD vs non-NMD") +
  theme_bw(base_size=11) + theme(strip.background=element_rect(fill="grey92")))
kable(med_tab(base, "dprod_adj"), caption="Change in productive (adjusted CPM) by class.")
```

## 3.6 Other variables (unproductive, totals, log2FC, isoform count)

```{r dist-others, fig.height=4}
print(plot_dist(base, "unprodCPM_DMSO", "Unproductive CPM at DMSO", xlab="unproductive CPM (DMSO)", logx=TRUE))
print(plot_dist(base, "unprodCPM_SMG",  "Unproductive CPM at SMG1i", xlab="unproductive CPM (SMG1i)", logx=TRUE))
print(plot_dist(base, "unprod_gain",    "Unproductive gain (SMG1i − DMSO)", xlab="change in unproductive CPM", vline=0, xlim=c(0, 25)))
print(plot_dist(base, "custom",         "Adjusted productive log2FC", xlab="productive log2 fold-change (adjusted)", xlim=c(-2,2), vline=0))
print(plot_dist(base, "totalCPM_DMSO",  "Total gene CPM at DMSO", xlab="total gene CPM (DMSO)", logx=TRUE))
print(plot_dist(base, "n_iso",          "Isoforms per gene", xlab="isoforms per gene", xlim=c(0,40)))
kable(med_tab(base, "unprod_gain"), caption="Unproductive gain (unprodCPM_SMG − unprodCPM_DMSO) by class.")
kable(med_tab(base, "custom"),      caption="Adjusted productive log2FC by class.")
```

# Step 4 — FIGURE R1: productive change, NMD vs non-NMD

```{r R1, fig.height=4}
t_med <- base %>% filter(gene_category %in% MAIN_CATS) %>%
  group_by(ct, gene_category) %>%
  summarise(n=n(), median_custom_log2FC=round(median(custom),3),
            pct_down_0.3=round(100*mean(custom < -DROP_THR),1),
            pct_up_0.3  =round(100*mean(custom >  DROP_THR),1), .groups="drop")
kable(t_med %>% pivot_wider(names_from=gene_category,
        values_from=c(n, median_custom_log2FC, pct_down_0.3, pct_up_0.3)),
      caption="TABLE R1. Median adjusted productive log2FC (custom), NMD vs matched non-NMD.")
print(ggplot(base %>% filter(gene_category %in% MAIN_CATS),
       aes(gene_category, custom, fill=gene_category)) +
  geom_hline(yintercept=0, colour="grey40") +
  geom_boxplot(outlier.size=0.3, outlier.alpha=0.2, width=0.6, colour="grey25") +
  facet_wrap(~ct, nrow=1) + coord_cartesian(ylim=c(-1.2,1.2)) +
  scale_fill_manual(values=cat_colors, guide="none") +
  labs(x=NULL, y="productive log2FC (custom)", title="FIGURE R1. Productive output, NMD vs matched non-NMD") +
  theme_bw(base_size=11) + theme(strip.background=element_rect(fill="grey92")))
ggsave(file.path(OUT_DIR,"FIG_R1_productive_box.pdf"), width=10, height=4)
```

# Step 5 — Correlation: adjusted SMG1i vs DMSO (normalization sanity check)

```{r R5-control-corr, fig.height=4}
ctl_corr <- base %>% filter(gene_category=="non-NMD") %>% group_by(ct) %>%
  summarise(n=n(),
            pearson_log=round(cor(log2(prodCPM_DMSO+1), log2(prodCPM_SMG_adj+1)),4),
            spearman=round(cor(prodCPM_DMSO, prodCPM_SMG_adj, method="spearman"),4),
            slope_log=round(coef(lm(log2(prodCPM_SMG_adj+1) ~ log2(prodCPM_DMSO+1)))[2],3),
            median_log2fc=round(median(custom),3),
            pct_within_2fold=round(100*mean(abs(custom)<1),1), .groups="drop")
kable(ctl_corr, caption="Control (non-NMD) genes: DMSO vs adjusted-SMG1i productive CPM. slope~1, r~1, median log2FC~0 => normalization leaves controls unchanged.")
print(ggplot(base %>% filter(gene_category=="non-NMD"), aes(log2(prodCPM_DMSO+1), log2(prodCPM_SMG_adj+1))) +
  geom_abline(slope=1, intercept=0, linetype="dashed", colour="grey50") +
  geom_point(alpha=0.10, size=0.4, colour="#7f8c8d") +
  geom_smooth(method="lm", se=FALSE, colour="#c0392b", linewidth=0.6) +
  facet_wrap(~ct, nrow=1) +
  labs(x="log2(DMSO productive CPM + 1)", y="log2(adjusted SMG1i productive CPM + 1)",
       title="Control genes on the identity line (normalization sanity check)") +
  theme_bw(base_size=11) + theme(strip.background=element_rect(fill="grey92")))
```

# Step 6 — CPM-space relationships

## 6.1 DMSO productive CPM vs change in productive (CPM)

```{r R6-prodDMSO-vs-dprod, fig.height=4.5}
d6a <- base %>% filter(gene_category %in% MAIN_CATS) %>% group_by(ct) %>%
  filter(within_sd(dprod_adj, 3)) %>% ungroup()
cat(sprintf("6.1 scatter: %d points beyond ±3 SD of dprod_adj removed for display.\n",
            sum(base$gene_category %in% MAIN_CATS) - nrow(d6a)))
print(ggplot(d6a, aes(log10(prodCPM_DMSO+1), dprod_adj, colour=gene_category)) +
  geom_hline(yintercept=0, colour="grey70") +
  geom_point(alpha=0.12, size=0.5) + geom_smooth(method="loess", se=FALSE, linewidth=0.7) +
  facet_wrap(~ct, nrow=1) +
  scale_colour_manual(values=cat_colors, name=NULL) +
  labs(x="log10(DMSO productive CPM + 1)", y="change in productive (adjusted CPM)",
       title="DMSO productive level vs change in productive",
       subtitle="display trimmed to ±3 SD of change-in-productive (per cell type)") +
  theme_bw(base_size=11) + theme(legend.position="top", strip.background=element_rect(fill="grey92")))
```

## 6.2 Unproductive gain (raw SMG1i) vs change in productive (CPM)

```{r R6-unprodgain-vs-dprod, fig.height=4.5}
d6b <- base %>% filter(gene_category=="NMD") %>% group_by(ct) %>%
  filter(within_sd(unprod_gain, 3) & within_sd(dprod_adj, 3)) %>% ungroup()
cat(sprintf("6.2 scatter: %d points beyond ±3 SD (of unprod_gain or dprod_adj) removed for display.\n",
            sum(base$gene_category=="NMD") - nrow(d6b)))
print(ggplot(d6b, aes(unprod_gain, dprod_adj)) +
  geom_hline(yintercept=0, colour="grey70") + geom_vline(xintercept=0, colour="grey85") +
  geom_point(alpha=0.15, size=0.5, colour="#c0392b") +
  geom_smooth(method="lm", se=TRUE, colour="black", linewidth=0.7) +
  facet_wrap(~ct, nrow=1, scales="free") +
  labs(x="unproductive gain (raw CPM, SMG1i − DMSO)", y="change in productive (adjusted CPM)",
       title="Unproductive gain vs productive change (NMD genes): flat/positive = no sibling-drop",
       subtitle="display trimmed to ±3 SD (per cell type); slope table below uses ALL genes") +
  theme_bw(base_size=11) + theme(strip.background=element_rect(fill="grey92")))
slope_cpm <- base %>% filter(gene_category=="NMD") %>% group_by(ct) %>%
  summarise(n=n(),
    slope=round(coef(lm(dprod_adj ~ unprod_gain + log10(prodCPM_DMSO+1)))["unprod_gain"],4),
    p=signif(summary(lm(dprod_adj ~ unprod_gain + log10(prodCPM_DMSO+1)))$coefficients["unprod_gain","Pr(>|t|)"],3),
    rho=round(cor(unprod_gain, dprod_adj, method="spearman", use="complete.obs"),3), .groups="drop")
kable(slope_cpm, caption="CPM-space slope of productive change on unproductive gain (NMD genes; controls for DMSO level). ~0/positive = decoupled.")
```

# Step 7 — Short-read induction loading (independent axis)

```{r sr-load}
sr_pm   <- fread(SR_PM_PATH); sr_lfsr <- fread(SR_LFSR_PATH)
sr_id   <- intersect(c("gene_id","ensembl_gene_id_version"), colnames(sr_pm))[1]
resolve_sr <- function(df, ct){ cn<-colnames(df); h<-cn[grepl(paste0("_in_",ct,"$"),cn)]
  if(length(h)==1) return(h); if (ct %in% cn) return(ct); stop("no SR col ",ct) }
sr_long <- bind_rows(lapply(CELL_TYPES, function(ct) data.frame(
  gene_id=sr_pm[[sr_id]], ct=ct,
  sr_logFC=sr_pm[[resolve_sr(sr_pm,ct)]], sr_lfsr=sr_lfsr[[resolve_sr(sr_lfsr,ct)]],
  stringsAsFactors=FALSE))) %>%
  mutate(sr_responsive = !is.na(sr_lfsr)&!is.na(sr_logFC)&sr_lfsr<0.05&sr_logFC>0)
cat(sprintf("SR-induced (lfsr<0.05 & logFC>0) genes per CT: %s\n",
    paste(sprintf("%s=%d", CELL_TYPES,
      sapply(CELL_TYPES, function(c) sum(sr_long$sr_responsive[sr_long$ct==c], na.rm=TRUE))), collapse="  ")))
```

# Step 8 — FIGURE R5d: NMD genes split by SR induction, with non-NMD overlaid

```{r R5d, fig.height=4.5}
nmd_sr <- base %>% filter(gene_category=="NMD") %>%
  inner_join(sr_long, by=c("gene_id","ct")) %>% filter(!is.na(sr_logFC)) %>%
  mutate(sr = ifelse(sr_responsive,"SR NMD+","SR NMD-"))
ctrl <- base %>% filter(gene_category=="non-NMD") %>% group_by(ct) %>%
  summarise(ctrl=median(custom, na.rm=TRUE), .groups="drop")

box_dat <- bind_rows(
  nmd_sr %>% transmute(ct, custom, grp=sr),
  base %>% filter(gene_category=="non-NMD") %>% transmute(ct, custom, grp="non-NMD")
) %>% mutate(grp=factor(grp, levels=c("SR NMD+","SR NMD-","non-NMD")))
print(ggplot(box_dat, aes(grp, custom, fill=grp)) +
  geom_hline(yintercept=0, colour="grey85") +
  geom_hline(data=ctrl, aes(yintercept=ctrl), colour="#2980b9", linetype="dashed") +
  geom_boxplot(outlier.size=0.3, outlier.alpha=0.2, width=0.6, colour="grey25") +
  facet_wrap(~ct, nrow=1) + coord_cartesian(ylim=c(-1.2,1.2)) +
  scale_fill_manual(values=c("SR NMD+"="#c0392b","SR NMD-"="#e67e22","non-NMD"="#7f8c8d"), guide="none") +
  labs(x=NULL, y="productive log2FC (custom)",
       title="FIGURE R5d. NMD genes by SR induction, with non-NMD overlaid (blue dashed = non-NMD median)") +
  theme_bw(base_size=11) + theme(strip.background=element_rect(fill="grey92")))

dens_dat <- bind_rows(
  nmd_sr %>% transmute(ct, custom, grp=sr),
  base %>% filter(gene_category=="non-NMD") %>% transmute(ct, custom, grp="non-NMD")
)
print(ggplot(dens_dat, aes(custom, fill=grp, colour=grp)) +
  geom_vline(xintercept=0, colour="grey70") + geom_density(alpha=0.25, linewidth=0.5) +
  facet_wrap(~ct, nrow=1) + coord_cartesian(xlim=c(-1.5,1.5)) +
  scale_fill_manual(values=c("SR NMD+"="#c0392b","SR NMD-"="#e67e22","non-NMD"="#7f8c8d"), name=NULL) +
  scale_colour_manual(values=c("SR NMD+"="#c0392b","SR NMD-"="#e67e22","non-NMD"="#7f8c8d"), guide="none") +
  labs(x="productive log2FC (custom)", title="Distribution: SR+ vs SR- NMD genes vs non-NMD") +
  theme_bw(base_size=11) + theme(legend.position="top", strip.background=element_rect(fill="grey92")))

kable(nmd_sr %>% group_by(ct, sr) %>% summarise(n=n(), median_custom=round(median(custom),3), .groups="drop") %>%
        pivot_wider(names_from=sr, values_from=c(n, median_custom)) %>%
        left_join(ctrl %>% rename(nonNMD_ctrl=ctrl), by="ct"),
      caption="TABLE R5d. SR-called vs not within NMD genes, vs non-NMD baseline. SR- ~ non-NMD = induction-gating.")
```

# Step 9 — FIGURE R5a: productive output vs short-read induction quintile

```{r R5a, fig.height=4.5}
dose_nmd <- nmd_sr %>% group_by(ct) %>% mutate(q=ntile(sr_logFC,5)) %>%
  group_by(ct,q) %>% summarise(median_sr_logFC=round(median(sr_logFC),3),
                               custom=round(median(custom),3), class="NMD", .groups="drop")
nonnmd_sr <- base %>% filter(gene_category=="non-NMD") %>%
  inner_join(sr_long, by=c("gene_id","ct")) %>% filter(!is.na(sr_logFC))
dose_non <- nonnmd_sr %>% group_by(ct) %>% mutate(q=ntile(sr_logFC,5)) %>%
  group_by(ct,q) %>% summarise(median_sr_logFC=round(median(sr_logFC),3),
                               custom=round(median(custom),3), class="non-NMD", .groups="drop")
dose <- bind_rows(dose_nmd, dose_non)
print(ggplot(dose, aes(factor(q), custom, group=class, colour=class)) +
  geom_hline(yintercept=0, colour="grey85") +
  geom_line(linewidth=1) + geom_point(size=2) + facet_wrap(~ct, nrow=1) +
  scale_colour_manual(values=c("NMD"="#c0392b","non-NMD"="#7f8c8d"), name=NULL) +
  labs(x="short-read induction quintile", y="median productive log2FC (custom)",
       title="FIGURE R5a. Productive output rises monotonically with independent induction (NMD vs non-NMD)") +
  theme_bw(base_size=11) + theme(legend.position="top", strip.background=element_rect(fill="grey92")))

r5stats <- bind_rows(lapply(CELL_TYPES, function(ct_i){
  d <- nmd_sr %>% filter(ct==ct_i); m <- lm(custom ~ sr_logFC, d); s <- summary(m)
  data.frame(ct=ct_i, class="NMD", n=nrow(d), slope=round(coef(m)[2],3), R2=round(s$r.squared,3),
             p=signif(s$coefficients[2,4],3), spearman=round(cor(d$custom,d$sr_logFC,method="spearman"),3))
}))
kable(r5stats, caption="TABLE R5a. Productive change (custom) ~ short-read induction, per CT (NMD genes).")

sl <- bind_rows(lapply(CELL_TYPES, function(ct_i){
  d <- base %>% filter(ct==ct_i, gene_category %in% MAIN_CATS) %>%
    inner_join(sr_long, by=c("gene_id","ct")) %>% filter(!is.na(sr_logFC))
  s <- summary(lm(custom ~ sr_logFC*gene_category, data=d))$coefficients
  data.frame(ct=ct_i, slope_nonNMD=round(s["sr_logFC","Estimate"],3),
             slope_diff_NMDvsNon=round(s[grep("sr_logFC:",rownames(s)),"Estimate"],3),
             p_diff=signif(s[grep("sr_logFC:",rownames(s)),"Pr(>|t|)"],3))
}))
kable(sl, caption="TABLE R5b. Productive ~ induction slope by class. slope_diff >= 0 => NMD tracks induction at least as well as controls.")
```

# Step 10 — Kitagawa (raw) metric and decomposition (verified)

```{r kitagawa-check}
chk <- base %>% filter(gene_category %in% MAIN_CATS) %>%
  kitagawa_decomp() %>%
  mutate(resid = dP - (level + composition))
cat(sprintf("Decomposition identity check: max |dP - (level+composition)| = %.3e (should be ~0)\n",
            max(abs(chk$resid), na.rm=TRUE)))
```

```{r kitagawa-dist, fig.height=4}
print(plot_dist(base, "Kitagawa", "Raw (Kitagawa) productive log2FC", xlab="productive log2 fold-change (raw)", xlim=c(-2,2), vline=0))
kdiff <- base %>% filter(gene_category %in% MAIN_CATS) %>% group_by(ct, gene_category) %>%
  summarise(n=n(), median=round(median(Kitagawa),3), mean=round(mean(Kitagawa),3),
            sd=round(sd(Kitagawa),3), pct_pos=round(100*mean(Kitagawa>0),1), .groups="drop") %>%
  pivot_wider(names_from=gene_category, values_from=c(n,median,mean,sd,pct_pos))
kable(kdiff, caption="Kitagawa distribution stats. The peak at 0 reflects the buffered majority; medians/means/SD differ across CT and class.")
```

```{r kitagawa-decomp, fig.height=4}
dec <- chk
kable(dec %>% mutate(driver_cd=ifelse(abs(composition)>abs(level),"composition-driven","level-driven")) %>%
        group_by(ct, gene_category) %>%
        summarise(median_level=round(median(level),2), median_composition=round(median(composition),2),
                  median_dP=round(median(dP),2),
                  pct_composition_driven=round(100*mean(driver_cd=="composition-driven"),1), .groups="drop") %>%
        pivot_wider(names_from=gene_category, values_from=c(median_level, median_composition, median_dP, pct_composition_driven)),
      caption="TABLE. Kitagawa decomposition. Non-NMD composition ~0 confirms composition is the NMD-specific term.")
decl <- dec %>% filter(gene_category=="NMD") %>% select(ct, level, composition) %>%
  pivot_longer(c(level, composition), names_to="term", values_to="contribution")
print(ggplot(decl, aes(contribution, fill=term)) + geom_vline(xintercept=0, colour="grey50") +
  geom_density(alpha=0.5) + facet_wrap(~ct, nrow=1, scales="free_y") + coord_cartesian(xlim=c(-30,30)) +
  scale_fill_manual(values=c("level"="#16a085","composition"="#8e44ad"), name=NULL) +
  labs(x="contribution to productive change (CPM)", title="Level (output) vs composition (isoform shift), NMD genes") +
  theme_bw(base_size=11) + theme(legend.position="top", strip.background=element_rect(fill="grey92")))
fwrite(dec, file.path(OUT_DIR, paste0("kitagawa_decomposition_", DATE_STAMP, ".csv")))
```

# Step 11 — Per-gene significance: limma (custom CPM)

```{r limma}
fit_custom_limma <- function(ct){
  o <- build_adj(ct)
  prodk <- o$anchor & o$class_v=="prod"
  pe <- rowsum(o$adj[prodk,,drop=FALSE], o$gene[prodk])
  unprod_genes <- unique(o$gene[o$anchor & o$class_v=="unprod"])
  unsure_genes <- setdiff(unique(o$gene[o$anchor & (o$unsure_v | o$down_v)]), unprod_genes)
  cs <- colnames(pe)
  samp <- data.frame(row.names=cs, treatment=relevel(factor(o$trt[cs]), ref="DMSO"),
                     sample_id=factor(o$donor[cs]))
  keep <- rowSums(pe >= 1) >= ceiling(ncol(pe)/2)
  E <- log2(pe[keep,,drop=FALSE] + 1)
  design <- model.matrix(~treatment, data=samp)
  cf  <- limma::duplicateCorrelation(E, design, block=samp$sample_id)
  fit <- limma::eBayes(limma::lmFit(E, design, block=samp$sample_id,
                                    correlation=cf$consensus.correlation), trend=TRUE)
  ci  <- grep("Smg1i", colnames(design)); tt <- limma::topTable(fit, coef=ci, number=Inf, sort.by="none")
  bhat <- fit$coefficients[,ci]; se <- bhat/fit$t[,ci]
  gc <- ifelse(rownames(tt) %in% unprod_genes, "NMD",
        ifelse(rownames(tt) %in% unsure_genes, "unsure", "non-NMD"))
  data.frame(gene_id=rownames(tt), ct=ct, logFC=tt$logFC, se=se[rownames(tt)],
             P=tt$P.Value, adjP=tt$adj.P.Val,
             gene_category=gc, stringsAsFactors=FALSE)
}
prod_dir <- bind_rows(lapply(CELL_TYPES, fit_custom_limma)) %>%
  mutate(dir = case_when(adjP<0.05 & logFC>0 ~ "up", adjP<0.05 & logFC<0 ~ "down", TRUE ~ "ns")) %>%
  left_join(iso2gene %>% select(gene_id, hgnc_symbol) %>% distinct(), by="gene_id")
kable(prod_dir %>% filter(gene_category %in% MAIN_CATS) %>% count(ct, gene_category, dir) %>%
        pivot_wider(names_from=dir, values_from=n, values_fill=0) %>% arrange(ct, gene_category),
      caption="TABLE. Productive direction (limma-trend on custom CPM, donor-blocked). Per-CT power is LAE-driven (n=4 vs 3); see mashr for the powered cross-CT view.")
fwrite(prod_dir, file.path(OUT_DIR, paste0("productive_direction_custom_", DATE_STAMP, ".csv")))
```

# Step 12 — Cross-cell-type significance: mashr

```{r mashr}
suppressPackageStartupMessages({ library(mashr); library(ashr) })
all_genes <- unique(prod_dir$gene_id)
bhat <- matrix(NA_real_, length(all_genes), length(CELL_TYPES), dimnames=list(all_genes, CELL_TYPES)); shat <- bhat
for (ct in CELL_TYPES){ d <- prod_dir %>% filter(ct==!!ct)
  bhat[d$gene_id, ct] <- d$logFC; shat[d$gene_id, ct] <- d$se }
fin <- apply(shat,1,function(x) all(is.finite(x)&x>0)) & apply(bhat,1,function(x) all(is.finite(x)))
bhat<-bhat[fin,,drop=FALSE]; shat<-shat[fin,,drop=FALSE]
mi<-mash_set_data(bhat,shat); m1<-mash_1by1(mi); st<-get_significant_results(m1,0.05)
set.seed(42); ri<-sample(nrow(bhat),min(20000,nrow(bhat)))
Vh<-estimate_null_correlation_simple(mash_set_data(bhat[ri,,drop=FALSE],shat[ri,,drop=FALSE]))
md<-mash_set_data(bhat,shat,V=Vh); Up<-if(length(st)>=5) cov_pca(md,npc=min(5,ncol(bhat)-1),subset=st) else list()
m<-mash(md, Ulist=c(Up,cov_canonical(md)), outputlevel=2)
pm<-get_pm(m); lf<-get_lfsr(m)
prod_dir_mashr <- bind_rows(lapply(CELL_TYPES, function(ct)
  data.frame(gene_id=rownames(pm), ct=ct, pm=pm[,ct], lfsr=lf[,ct], stringsAsFactors=FALSE))) %>%
  mutate(dir=case_when(lfsr<0.05&pm>0~"up", lfsr<0.05&pm<0~"down", TRUE~"ns")) %>%
  left_join(prod_dir %>% select(gene_id, ct, gene_category, hgnc_symbol), by=c("gene_id","ct"))
kable(prod_dir_mashr %>% filter(gene_category %in% MAIN_CATS) %>% count(ct, gene_category, dir) %>%
        pivot_wider(names_from=dir, values_from=n, values_fill=0) %>% arrange(ct, gene_category),
      caption="TABLE. Productive direction after cross-CT mashr on custom effects (lfsr<0.05).")
fwrite(prod_dir_mashr, file.path(OUT_DIR, paste0("productive_direction_mashr_custom_", DATE_STAMP, ".csv")))
saveRDS(m, file.path(OUT_DIR, paste0("mashr_custom_", DATE_STAMP, ".rds")))
```

# Step 12b — Is the productive direction real, or just noise?

Null: a gene's productive up/down direction is noise (its within-donor change has no real
treatment effect; sign is random). We attack it several independent ways, all on the **adjusted**
productive change (surge removed) and respecting donor pairing. The reproducibility battery figure
collects the **four retained tests** (sign-flip permutation, cross-donor concordance, effect-size
variance ratio, Storey π₀); the additional cross-platform, leave-one-donor-out, and Gaussian-mixture
tests below are **exploratory and not reported** in the manuscript.

```{r noise-helpers}
if (!"P" %in% names(prod_dir)) prod_dir$P <- 2 * pnorm(-abs(prod_dir$logFC / prod_dir$se))

if (requireNamespace("matrixStats", quietly=TRUE)) {
  rowSds <- matrixStats::rowSds
} else {
  rowSds <- function(X){ n<-ncol(X); sqrt((rowMeans(X^2) - rowMeans(X)^2) * n/(n-1)) }
}
prod_logfc_mat <- function(ct_i, cats=MAIN_CATS){
  M <- gene_sample %>% filter(ct==ct_i, gene_category %in% cats,
        !is.na(prodCPM_DMSO), !is.na(prodCPM_SMG_adj)) %>%
       mutate(D = log2((prodCPM_SMG_adj+1)/(prodCPM_DMSO+1))) %>%
       select(gene_id, donor, D) %>%
       tidyr::pivot_wider(names_from=donor, values_from=D) %>%
       tibble::column_to_rownames("gene_id") %>% as.matrix()
  M[rowSums(is.finite(M))==ncol(M), , drop=FALSE]
}
storey_pi0 <- function(p, lambda = 0.5){ p <- p[is.finite(p)]; min(1, mean(p > lambda)/(1 - lambda)) }
```

```{r leading-edge-check}
upr_ens <- unique(na.omit(
  AnnotationDbi::select(org.Hs.eg.db, keys="GO:0006986",
                        keytype="GOALL", columns="ENSEMBL")$ENSEMBL))

for (ct_i in CELL_TYPES) {
  up_g <- prod_dir_mashr %>% filter(ct==ct_i, gene_category=="NMD", dir=="up") %>% pull(gene_id)
  le   <- intersect(svf(up_g), upr_ens)
  syms <- AnnotationDbi::mapIds(org.Hs.eg.db, keys=le, column="SYMBOL",
                                keytype="ENSEMBL", multiVals="first")
  cat(sprintf("%s — UPR productive-UP genes (%d): %s\n",
              ct_i, length(le), paste(sort(na.omit(syms)), collapse=", ")))
}
le_all <- unique(unlist(lapply(CELL_TYPES, function(ct_i){
  up_g <- prod_dir_mashr %>% filter(ct==ct_i, gene_category=="NMD", dir=="up") %>% pull(gene_id)
  intersect(svf(up_g), upr_ens)
})))
cat("\nUnion UPR productive-UP genes:\n",
    paste(sort(na.omit(AnnotationDbi::mapIds(org.Hs.eg.db, keys=le_all,
          column="SYMBOL", keytype="ENSEMBL", multiVals="first"))), collapse=", "), "\n")
```

## TEST 1 — Sign-flip permutation test (per-gene) [retained]

```{r noise-signflip}
signflip_test <- function(ct_i, B = 2000, seed = 1){
  M <- prod_logfc_mat(ct_i); n <- ncol(M); G <- nrow(M)
  tstat  <- function(X) rowMeans(X) / (rowSds(X) / sqrt(n))
  cutoff <- qt(0.975, n - 1)
  obs    <- sum(abs(tstat(M)) >= cutoff, na.rm = TRUE)
  set.seed(seed); null <- numeric(B)
  for (b in 1:B){ S <- matrix(sample(c(-1,1), G*n, TRUE), G, n); null[b] <- sum(abs(tstat(M*S)) >= cutoff, na.rm=TRUE) }
  data.frame(ct=ct_i, n_donor=n, n_genes=G, observed_sig=obs,
             null_mean_sig=round(mean(null),1),
             empirical_FDR=round(mean(null)/obs,4),
             perm_p=round((sum(null>=obs)+1)/(B+1),4))
}
kable(bind_rows(lapply(CELL_TYPES, signflip_test)),
  caption="TEST 1. Within-donor sign-flip permutation. FDR<<1 & perm_p~0 reject 'noise'.")
```

## TEST 2 — Cross-donor sign concordance [retained]

```{r noise-concordance}
kable(bind_rows(lapply(CELL_TYPES, function(ct_i){
  M <- prod_logfc_mat(ct_i); n <- ncol(M); G <- nrow(M)
  all_pos <- sum(rowSums(M>0)==n); all_neg <- sum(rowSums(M<0)==n)
  bt <- binom.test(all_pos+all_neg, G, p=2^(1-n), alternative="greater")
  data.frame(ct=ct_i, n_donor=n, n_genes=G, all_up=all_pos, all_down=all_neg,
             observed_concordant=all_pos+all_neg, expected_under_noise=round(G*2^(1-n)),
             p=signif(bt$p.value,3))
})), caption="TEST 2. Cross-donor sign concordance vs binomial chance 2*(1/2)^n.")
```

## TEST 3 — Cross-platform concordance (exploratory)

```{r noise-3-crossplatform}
xp <- base %>% filter(gene_category=="non-NMD") %>%
  inner_join(sr_long, by=c("gene_id","ct")) %>% filter(is.finite(custom), is.finite(sr_logFC))
kable(xp %>% group_by(ct) %>% summarise(
  n=n(), spearman=round(cor(custom, sr_logFC, method="spearman"),3),
  sign_concordance=round(mean(sign(custom)==sign(sr_logFC)),3),
  p=signif(cor.test(custom, sr_logFC, method="spearman")$p.value,3), .groups="drop"),
  caption="TEST 3 (exploratory). Cross-platform concordance (non-NMD): long-read productive logFC vs short-read gene logFC.")
```

## TEST 4 — Leave-one-donor-out reproducibility (exploratory)

```{r noise-4-loo}
kable(bind_rows(lapply(CELL_TYPES, function(ct_i){
  M <- prod_logfc_mat(ct_i)
  d1 <- M[,1]; drest <- rowMeans(M[,-1,drop=FALSE])
  data.frame(ct=ct_i, n_genes=nrow(M),
             rho=round(cor(d1,drest,method="spearman"),3),
             sign_concordance=round(mean(sign(d1)==sign(drest)),3),
             p=signif(cor.test(d1,drest,method="spearman")$p.value,3))
})), caption="TEST 4 (exploratory). Leave-one-donor-out reproducibility: donor 1 vs mean of the rest.")
```

## TEST 5 — Storey pi0 / calibrated FDR (limma moderated p-values) [retained]

```{r noise-5-pi0}
kable(bind_rows(lapply(CELL_TYPES, function(ct_i){
  p <- prod_dir$P[prod_dir$ct==ct_i & prod_dir$gene_category %in% MAIN_CATS]; p <- p[is.finite(p)]
  pi0 <- storey_pi0(p)
  data.frame(ct=ct_i, n=length(p), pi0=round(pi0,3),
             est_nonnull_fraction=round(1 - pi0, 3),
             n_BH_FDR05=sum(p.adjust(p, "BH") < 0.05))
})), caption="TEST 5. Storey pi0 (lambda=0.5): 1-pi0 = estimated fraction of genes with a real productive effect.")
```

## TEST 7 — Effect-size variance ratio (change distribution wider than noise) [retained]

```{r noise-7-distribution-width}
distrib_test <- function(ct_i, B = 200){
  M <- prod_logfc_mat(ct_i); n <- ncol(M); G <- nrow(M)
  eff <- rowMeans(M)
  set.seed(1)
  eff_null <- unlist(lapply(1:B, function(b){ S <- matrix(sample(c(-1,1), G*n, TRUE), G, n); rowMeans(M*S) }))
  data.frame(ct=ct_i, var_obs=round(var(eff),4), var_null=round(var(eff_null),4),
             var_ratio=round(var(eff)/var(eff_null),2),
             ks_p=signif(ks.test(eff, sample(eff_null, min(length(eff_null),5e4)))$p.value,3))
}
kable(bind_rows(lapply(CELL_TYPES, distrib_test)),
  caption="TEST 7. Per-gene mean productive log2FC vs the sign-flip null. var_ratio>1 + KS p<<0.05 = wider than noise.")
```

## TEST 12 — Single Gaussian vs mixture (exploratory)

```{r noise-12-mixture, eval = requireNamespace("mclust", quietly=TRUE)}
suppressPackageStartupMessages(library(mclust))
res12 <- bind_rows(lapply(CELL_TYPES, function(ct_i){
  x <- base$custom[base$ct==ct_i & base$gene_category %in% MAIN_CATS]; x <- x[is.finite(x)]
  m1 <- mclust::densityMclust(x, G=1,   plot=FALSE, verbose=FALSE)
  mk <- mclust::densityMclust(x, G=1:4, plot=FALSE, verbose=FALSE)
  data.frame(ct=ct_i, n=length(x), best_G=mk$G,
             BIC_1comp = round(max(m1$BIC), 1),
             BIC_best  = round(max(mk$BIC), 1),
             dBIC_best_minus_1 = round(max(mk$BIC) - max(m1$BIC), 1))
}))
kable(res12, caption="TEST 12 (exploratory). Single Gaussian vs mixture (mclust BIC; higher = better).")
```

## Reproducibility battery figure (four retained tests)

```{r reproducibility-battery, fig.height=4.5}
# Values computed from the retained test functions above (NOT hardcoded), so the figure
# cannot drift from the tables. Saved to OUT_DIR alongside the other outputs.
battery <- bind_rows(lapply(CELL_TYPES, function(ct_i){
  M    <- prod_logfc_mat(ct_i); n <- ncol(M); G <- nrow(M)
  sf   <- signflip_test(ct_i)$empirical_FDR                            # TEST 1
  conc <- (sum(rowSums(M>0)==n) + sum(rowSums(M<0)==n)) / (G*2^(1-n))  # TEST 2 fold over expected
  vr   <- distrib_test(ct_i)$var_ratio                                 # TEST 7 variance ratio
  p    <- prod_dir$P[prod_dir$ct==ct_i & prod_dir$gene_category %in% MAIN_CATS]
  nn   <- 1 - storey_pi0(p[is.finite(p)])                              # TEST 5 non-null fraction
  data.frame(ct=ct_i,
    `Sign-flip permutation\n(empirical FDR; pass <1)`  = round(sf,3),
    `Cross-donor concordance\n(x expected; pass >1)`   = round(conc,2),
    `Effect-size variance ratio\n(obs/null; pass >1)`  = round(vr,2),
    `Storey non-null fraction\n(1-pi0; pass >0.1)`     = round(nn,3),
    check.names=FALSE)
}))
long <- battery %>% pivot_longer(-ct, names_to="test", values_to="stat") %>%
  mutate(pass = case_when(grepl("FDR", test) ~ stat < 1,
                          grepl("Storey", test) ~ stat > 0.1,
                          TRUE ~ stat > 1),
         label = sprintf("%.2f", stat),
         ct = factor(ct, levels=c("AT2","LAE","FB","MV")),
         test = factor(test, levels=rev(unique(test))))
p_batt <- ggplot(long, aes(ct, test, fill=pass)) +
  geom_tile(color="white", linewidth=1.1) +
  geom_text(aes(label=label), fontface="bold", size=3.6, color=ifelse(long$pass,"grey15","white")) +
  scale_fill_manual(values=c(`TRUE`="#4393c3",`FALSE`="#d6604d"),
                    labels=c(`TRUE`="meets criterion", `FALSE`="fails criterion"), name=NULL) +
  scale_x_discrete(position="top") +
  labs(title="Reproducibility of the productive (non-NMD) output response",
       subtitle="Four retained tests; pass criterion in each row label. FB is the provisional cell type.",
       x=NULL, y=NULL) +
  theme_minimal(base_size=12) +
  theme(panel.grid=element_blank(), axis.text.x.top=element_text(face="bold",size=12),
        axis.text.y=element_text(size=9, lineheight=0.95), plot.title=element_text(face="bold"),
        plot.subtitle=element_text(color="grey40", size=9.5), legend.position="top")
ggsave(file.path(OUT_DIR, "reproducibility_battery.png"), p_batt, width=9, height=4.5, dpi=300, bg="white")
print(p_batt)
```

# Step 13 — What makes a gene's productive output go up vs down?

An **inferential (associational)** logistic regression — we read the **odds ratios / p-values**,
estimated on **all** the data, so there is deliberately **no train/test split** (a split is the
right tool only for a prediction claim, provided as a cross-validated AUC in §13.1).

```{r updown-model}
mdl <- prod_dir_mashr %>% filter(dir %in% c("up","down")) %>%
  left_join(base %>% select(ct, gene_id, prodCPM_DMSO, unprodCPM_DMSO, unprodCPM_SMG,
                            unprod_gain, totalCPM_DMSO, n_iso), by=c("ct","gene_id")) %>%
  left_join(sr_long %>% select(ct, gene_id, sr_logFC), by=c("ct","gene_id")) %>%
  mutate(up = as.integer(dir=="up"),
         is_NMD = as.integer(gene_category=="NMD"),
         log_expr       = log10(totalCPM_DMSO + 1),
         log_unprodDMSO = log10(unprodCPM_DMSO + 1),
         log_unprodSMG  = log10(unprodCPM_SMG + 1),
         log_niso       = log10(n_iso + 1)) %>%
  filter(is.finite(log_expr))
cat(sprintf("Model rows: %d (up=%d, down=%d)\n", nrow(mdl), sum(mdl$up), sum(1-mdl$up)))

fit_glm <- glm(up ~ is_NMD + log_expr + log_unprodDMSO + log_unprodSMG + unprod_gain + log_niso + ct,
               data=mdl, family=binomial)
co <- summary(fit_glm)$coefficients
kable(data.frame(term=rownames(co), beta=round(co[,1],3),
                 odds_ratio=round(exp(co[,1]),3), p=signif(co[,4],3)),
      caption="TABLE. Logistic model: P(productive UP) across all genes (mashr-significant only).")

mdl_sr <- mdl %>% filter(is.finite(sr_logFC))
fit_glm_sr <- glm(up ~ is_NMD + log_expr + log_unprodDMSO + log_unprodSMG + unprod_gain + log_niso + sr_logFC + ct,
                  data=mdl_sr, family=binomial)
co2 <- summary(fit_glm_sr)$coefficients
kable(data.frame(term=rownames(co2), beta=round(co2[,1],3),
                 odds_ratio=round(exp(co2[,1]),3), p=signif(co2[,4],3)),
      caption="TABLE. Same model + short-read induction. If sr_logFC dominates and is_NMD shrinks toward OR~1, productive direction is induction-gated.")
```

## 13.1 — Predictive check: cross-validated AUC (this one DOES use train/test)

```{r updown-cv-auc}
auc <- function(y, p){
  r <- rank(p); n1 <- sum(y==1); n0 <- sum(y==0)
  if (n1==0 || n0==0) return(NA_real_)
  (sum(r[y==1]) - n1*(n1+1)/2) / (n1*n0)
}
cv_auc <- function(df, form, K=5, seed=1){
  set.seed(seed); df <- df[is.finite(rowSums(model.matrix(form, df)[,-1,drop=FALSE])), , drop=FALSE]
  fold <- sample(rep(1:K, length.out=nrow(df))); pred <- rep(NA_real_, nrow(df))
  for (k in 1:K){
    tr <- df[fold!=k,,drop=FALSE]; te <- df[fold==k,,drop=FALSE]
    fit <- suppressWarnings(glm(form, data=tr, family=binomial))
    pred[fold==k] <- predict(fit, newdata=te, type="response")
  }
  auc(df$up, pred)
}
auc_full  <- cv_auc(mdl_sr, up ~ is_NMD + log_expr + log_unprodDMSO + log_unprodSMG + unprod_gain + log_niso + sr_logFC + ct)
auc_ind   <- cv_auc(mdl_sr, up ~ sr_logFC)
auc_noind <- cv_auc(mdl_sr, up ~ is_NMD + log_expr + log_unprodDMSO + log_unprodSMG + unprod_gain + log_niso + ct)
auc_nmd   <- cv_auc(mdl_sr, up ~ is_NMD)
kable(data.frame(
  model = c("full (all features + induction)","induction only (sr_logFC)",
            "all features, NO induction","is_NMD only"),
  cv_AUC = round(c(auc_full, auc_ind, auc_noind, auc_nmd), 3)),
  caption="TABLE 13.1. 5-fold cross-validated AUC for predicting productive direction.")
```

## 13.2 — Separate "goes up" and "goes down" models (one-vs-rest)

```{r updown-separate}
mdl_all <- prod_dir_mashr %>%
  filter(gene_category %in% MAIN_CATS) %>%
  left_join(base %>% select(ct, gene_id, prodCPM_DMSO, unprodCPM_DMSO, unprodCPM_SMG,
                            unprod_gain, totalCPM_DMSO, n_iso), by=c("ct","gene_id")) %>%
  left_join(sr_long %>% select(ct, gene_id, sr_logFC), by=c("ct","gene_id")) %>%
  mutate(is_NMD = as.integer(gene_category=="NMD"),
         log_expr       = log10(totalCPM_DMSO + 1),
         log_unprodDMSO = log10(unprodCPM_DMSO + 1),
         log_unprodSMG  = log10(unprodCPM_SMG + 1),
         log_niso       = log10(n_iso + 1),
         goes_up   = as.integer(dir=="up"),
         goes_down = as.integer(dir=="down")) %>%
  filter(is.finite(log_expr))
cat(sprintf("All-gene one-vs-rest set: n=%d (up=%d, down=%d, ns=%d)\n",
    nrow(mdl_all), sum(mdl_all$goes_up), sum(mdl_all$goes_down), sum(mdl_all$dir=="ns")))

run_or <- function(df, outcome){
  f <- as.formula(paste(outcome,
        "~ is_NMD + log_expr + log_unprodDMSO + log_unprodSMG + unprod_gain + log_niso + ct"))
  co <- summary(glm(f, data=df, family=binomial))$coefficients
  data.frame(term=rownames(co), beta=round(co[,1],3), odds_ratio=round(exp(co[,1]),3), p=signif(co[,4],3))
}
kable(run_or(mdl_all, "goes_up"),   caption="TABLE 13.2a. P(productive UP vs NOT up), one-vs-rest.")
kable(run_or(mdl_all, "goes_down"), caption="TABLE 13.2b. P(productive DOWN vs NOT down), one-vs-rest.")
cmp <- bind_rows(
  run_or(mdl_all, "goes_up")   %>% filter(term=="is_NMD") %>% mutate(model="goes_up (vs rest)"),
  run_or(mdl_all, "goes_down") %>% filter(term=="is_NMD") %>% mutate(model="goes_down (vs rest)")
) %>% select(model, is_NMD_OR=odds_ratio, p)
kable(cmp, caption="TABLE 13.2c. is_NMD odds ratio in each one-vs-rest model.")
```

# Step 14 — Is the dominant trend "NMD up -> productive up", and what confounds it?

```{r nmd-up-confounders, eval = requireNamespace("org.Hs.eg.db", quietly=TRUE) && requireNamespace("AnnotationDbi", quietly=TRUE)}
sf_ens <- go_genes("GO:0008380")                                  # splicing factors
dn_ens <- go_genes(c("GO:0006281","GO:0006260","GO:0000278"))     # proliferation/DNA-repair (down program)

nmd_feat <- prod_dir_mashr %>% filter(gene_category=="NMD", dir %in% c("up","down")) %>%
  left_join(base %>% select(ct, gene_id, totalCPM_DMSO, unprod_gain), by=c("ct","gene_id")) %>%
  left_join(sr_long %>% select(ct, gene_id, sr_logFC), by=c("ct","gene_id")) %>%
  mutate(up=as.integer(dir=="up"), g0=svf(gene_id),
         is_SF=as.integer(g0 %in% sf_ens), is_down_prog=as.integer(g0 %in% dn_ens),
         log_expr=log10(totalCPM_DMSO+1)) %>% filter(is.finite(log_expr))
cat(sprintf("NMD genes with a significant productive call: up=%d, down=%d\n",
            sum(nmd_feat$up), sum(1-nmd_feat$up)))

fit_nmd <- glm(up ~ sr_logFC + is_SF + is_down_prog + log_expr + unprod_gain,
               data=nmd_feat %>% filter(is.finite(sr_logFC)), family=binomial)
cn <- summary(fit_nmd)$coefficients
kable(data.frame(term=rownames(cn), beta=round(cn[,1],3),
                 odds_ratio=round(exp(cn[,1]),3), p=signif(cn[,4],3)),
      caption="TABLE. Within NMD genes: predictors of productive UP. OR<1 (is_SF, is_down_prog) = confounders that PREVENT coupling; sr_logFC OR>1 = induction drives it.")

enr <- nmd_feat %>% group_by(dir) %>%
  summarise(n=n(), pct_SF=round(100*mean(is_SF),1), pct_down_prog=round(100*mean(is_down_prog),1),
            median_sr_logFC=round(median(sr_logFC, na.rm=TRUE),3), .groups="drop")
kable(enr, caption="TABLE. Feature rates among productive-UP vs productive-DOWN NMD genes.")
```

# Step 15 — Pathways (topGO, two GO-defined programs)

```{r topgo, eval = requireNamespace("topGO", quietly=TRUE) && requireNamespace("org.Hs.eg.db", quietly=TRUE)}
suppressPackageStartupMessages({ library(topGO); library(org.Hs.eg.db) })
run_bp <- function(universe_ens, interesting_ens, label, topn=12){
  m <- suppressMessages(AnnotationDbi::select(org.Hs.eg.db, keys=svf(unique(universe_ens)),
        columns="ENTREZID", keytype="ENSEMBL"))
  m <- m[!is.na(m$ENTREZID)&!duplicated(m$ENSEMBL),]; e <- setNames(m$ENTREZID,m$ENSEMBL)
  uni<-unique(na.omit(e[svf(universe_ens)])); intr<-unique(na.omit(e[svf(interesting_ens)]))
  if(length(intr)<5){cat(sprintf("\n[%s] %d mapped -- skipped\n",label,length(intr)));return(invisible(NULL))}
  gl<-factor(as.integer(uni%in%intr),levels=c(0,1)); names(gl)<-uni
  GOd<-suppressMessages(new("topGOdata",ontology="BP",allGenes=gl,annot=annFUN.org,
        mapping="org.Hs.eg.db",ID="entrez",nodeSize=10))
  tb<-GenTable(GOd,elimFisher=suppressMessages(runTest(GOd,algorithm="elim",statistic="fisher")),
        orderBy="elimFisher",topNodes=topn)
  cat(sprintf("\n=== %s (interesting=%d of universe=%d) ===\n",label,length(intr),length(uni)));print(tb)
  invisible(tb)
}
go_collect <- list()
for (ct_i in CELL_TYPES) for (cat_i in MAIN_CATS) for (dir_i in c("up","down")){
  d <- prod_dir_mashr %>% filter(ct==ct_i, gene_category==cat_i)
  tb <- run_bp(d$gene_id, d$gene_id[d$dir==dir_i], sprintf("[mashr] %s | %s | productive-%s", ct_i, cat_i, toupper(dir_i)))
  if (!is.null(tb)) { tb$ct<-ct_i; tb$category<-cat_i; tb$direction<-dir_i; go_collect[[length(go_collect)+1]] <- tb }
}
go_all <- bind_rows(go_collect)
go_all$elimFisher <- as.numeric(gsub("<","",go_all$elimFisher))
fwrite(go_all, file.path(OUT_DIR, paste0("topgo_all_mashr_", DATE_STAMP, ".csv")))
```

```{r topgo-consensus, eval = exists("go_all") && nrow(go_all) > 0}
consensus <- go_all %>% filter(elimFisher < 0.05) %>%
  group_by(category, direction, GO.ID, Term) %>%
  summarise(n_ct=n_distinct(ct), cts=paste(sort(unique(ct)), collapse=","),
            min_p=signif(min(elimFisher),2), .groups="drop") %>%
  filter(n_ct >= 2) %>% arrange(category, direction, desc(n_ct), min_p)
kable(consensus, caption="TABLE. GO BP terms recurring in >=2 cell types (elim Fisher<0.05). NMD-UP = ISR/UPR program; *-DOWN = proliferation/DNA-repair program.")
fwrite(consensus, file.path(OUT_DIR, paste0("topgo_consensus_", DATE_STAMP, ".csv")))
```

```{r two-program-GO, eval = requireNamespace("org.Hs.eg.db", quietly=TRUE) && requireNamespace("AnnotationDbi", quietly=TRUE), fig.height=4.5}
up_go <- go_genes(c("GO:0006986","GO:0140467"))                 # ISR/UPR
dn_go <- go_genes(c("GO:0006281","GO:0006260","GO:0000278"))    # DNA repair + replication + mitotic cell cycle
both  <- intersect(up_go, dn_go); up_go <- setdiff(up_go, both); dn_go <- setdiff(dn_go, both)
cat(sprintf("GO sets: up(ISR/UPR)=%d, down(prolif/DNA-repair)=%d, dropped overlap=%d\n",
            length(up_go), length(dn_go), length(both)))
pg <- base %>% mutate(g0=svf(gene_id),
        program=case_when(g0 %in% up_go ~ "ISR/UPR (up)",
                          g0 %in% dn_go ~ "Proliferation/DNA-repair (down)", TRUE ~ NA_character_)) %>%
  filter(!is.na(program))
kable(pg %>% group_by(ct, program) %>%
        summarise(n=n(), median_custom=round(median(custom),3),
                  p_vs_zero=signif(suppressWarnings(wilcox.test(custom)$p.value),3), .groups="drop") %>%
        pivot_wider(names_from=program, values_from=c(n, median_custom, p_vs_zero)),
      caption="TABLE. Productive log2FC for GO-defined ISR/UPR (up) vs proliferation/DNA-repair (down) gene sets, per CT.")
print(ggplot(pg, aes(program, custom, fill=program)) +
  geom_hline(yintercept=0, colour="grey40") +
  geom_boxplot(outlier.size=0.3, outlier.alpha=0.3, width=0.6, colour="grey25") +
  facet_wrap(~ct, nrow=1) + coord_cartesian(ylim=c(-1.2,1.2)) +
  scale_fill_manual(values=c("ISR/UPR (up)"="#16a085","Proliferation/DNA-repair (down)"="#c0392b"), guide="none") +
  labs(x=NULL, y="productive log2FC (custom)", title="Two opposing productive programs (GO-defined)") +
  theme_bw(base_size=11) + theme(strip.background=element_rect(fill="grey92"), axis.text.x=element_text(angle=15, hjust=1)))
```

# Step 16 — Two transcriptional programs and their sharing

```{r step16-setup}
if (!requireNamespace("msigdbr", quietly=TRUE)) stop("Package 'msigdbr' required — install before knitting.")
library(msigdbr)
hm   <- tryCatch(msigdbr(species="Homo sapiens", category="H"),
                 error=function(e) msigdbr(species="Homo sapiens", collection="H"))
hset    <- function(nm) unique(hm$gene_symbol[hm$gs_name==nm])
hm_list <- split(hm$gene_symbol, hm$gs_name)
down_sets <- list(E2F=hset("HALLMARK_E2F_TARGETS"), G2M=hset("HALLMARK_G2M_CHECKPOINT"), DNArep=hset("HALLMARK_DNA_REPAIR"))
nfkb_set  <- hset("HALLMARK_TNFA_SIGNALING_VIA_NFKB")
fmt_p <- function(p) ifelse(is.na(p), NA, ifelse(p>=1e-3, formatC(p,format="f",digits=3), formatC(p,format="e",digits=1)))
```

## 16.1 DOWN program = E2F-driven cell-cycle/proliferation arrest

```{r step16-1-down-thresholded}
enrich_down_ct <- function(ct_i, class, set){
  d <- prod_dir_mashr %>% filter(ct==ct_i, gene_category==class)
  hit <- unique(d$hgnc_symbol[d$dir=="down"]); bg <- unique(d$hgnc_symbol)
  ft <- safe_fisher(bg %in% hit, bg %in% set)
  data.frame(n_down=length(hit), n_in_set=sum(hit %in% set), OR=round(as.numeric(ft$estimate),2), p=fmt_p(ft$p.value))
}
res_dn <- bind_rows(lapply(names(down_sets), function(prog)
  bind_rows(lapply(CELL_TYPES, function(c) bind_rows(
    cbind(ct=c, class="NMD",     prog=prog, enrich_down_ct(c,"NMD",     down_sets[[prog]])),
    cbind(ct=c, class="non-NMD", prog=prog, enrich_down_ct(c,"non-NMD", down_sets[[prog]])))))))
kable(res_dn, caption="TABLE 16.1a. Productive-DOWN genes vs E2F/G2M/DNA-repair (thresholded Fisher), by cell type and class.")
```

```{r step16-1-down-pooled}
nn <- prod_dir_mashr %>% filter(gene_category=="non-NMD")
nn_down <- unique(nn$hgnc_symbol[nn$dir=="down"]); nn_bg <- unique(nn$hgnc_symbol)
for (prog in names(down_sets)) { ft <- safe_fisher(nn_bg %in% nn_down, nn_bg %in% down_sets[[prog]])
  cat(sprintf("%-7s pooled non-NMD down: %d/%d in set, OR=%.2f, p=%s\n",
              prog, sum(nn_down %in% down_sets[[prog]]), length(nn_down), ft$estimate, fmt_p(ft$p.value))) }
```

```{r step16-1-down-continuous}
gsea_scan <- function(ct_i, class="non-NMD"){
  d <- prod_dir_mashr %>% filter(ct==ct_i, gene_category==class, !is.na(pm)); d <- d[!duplicated(d$hgnc_symbol),]; med <- median(d$pm)
  bind_rows(lapply(names(hm_list), function(s){ inset <- d$hgnc_symbol %in% hm_list[[s]]
    if (sum(inset) < 10) return(NULL); w <- wilcox.test(d$pm[inset], d$pm[!inset])
    data.frame(pathway=sub("HALLMARK_","",s), n_in=sum(inset), median_pm_in=round(median(d$pm[inset]),3),
               dir=ifelse(median(d$pm[inset])<med,"DOWN","up"), p=w$p.value) })) %>%
    arrange(p) %>% mutate(padj=signif(p.adjust(p,"BH"),3), p=fmt_p(p))
}
for (c in CELL_TYPES){ cat("\n==", c, "non-NMD — threshold-free Hallmark scan (top 8) ==\n"); print(head(gsea_scan(c), 8)) }
```

## 16.2 UP program = NF-κB/TNFα inflammatory (secondary)

```{r step16-2-up-NFkB}
gsea_one <- function(ct_i, class, set){
  d <- prod_dir_mashr %>% filter(ct==ct_i, gene_category==class, !is.na(pm)); d <- d[!duplicated(d$hgnc_symbol),]
  inset <- d$hgnc_symbol %in% set; w <- wilcox.test(d$pm[inset], d$pm[!inset])
  data.frame(ct=ct_i, class=class, n_in=sum(inset), median_pm_in=round(median(d$pm[inset]),3),
             dir=ifelse(median(d$pm[inset])>median(d$pm[!inset]),"UP","down"), p=fmt_p(w$p.value))
}
kable(bind_rows(lapply(CELL_TYPES, function(c) bind_rows(gsea_one(c,"non-NMD",nfkb_set), gsea_one(c,"NMD",nfkb_set)))),
  caption="TABLE 16.2. HALLMARK_TNFA_SIGNALING_VIA_NFKB (continuous test on productive pm), by cell type and class.")
```

## 16.3 Cross-cell-type sharing of the up and down programs

```{r step16-3-sharing}
emat <- prod_dir %>% filter(gene_category %in% MAIN_CATS) %>%
  select(ct, gene_id, logFC) %>%
  tidyr::pivot_wider(names_from = ct, values_from = logFC) %>%
  tibble::column_to_rownames("gene_id") %>% as.matrix()
emat <- emat[rowSums(is.finite(emat)) == ncol(emat), , drop = FALSE]

resp <- unique(prod_dir_mashr$gene_id[prod_dir_mashr$dir != "ns"])
emat_resp <- emat[rownames(emat) %in% resp, , drop = FALSE]

C   <- cor(emat_resp, method = "spearman", use = "pairwise.complete.obs")
cdf <- as.data.frame(as.table(C)); names(cdf) <- c("CT1", "CT2", "rho")
print(ggplot(cdf, aes(CT1, CT2, fill = rho)) +
  geom_tile(colour = "white", linewidth = 0.8) +
  geom_text(aes(label = sprintf("%.2f", rho)), fontface = "bold",
            colour = ifelse(cdf$rho > 0.6, "white", "grey20")) +
  scale_fill_gradient2(low = "#D73027", mid = "white", high = "#2166AC",
                       midpoint = min(C), limits = c(min(C), 1), name = "Spearman") +
  labs(title = "Pairwise correlation of the productive response across cell types",
       subtitle = sprintf("unshrunken limma logFC; %d genes significant in >=1 cell type", nrow(emat_resp)),
       x = NULL, y = NULL) +
  theme_minimal(base_size = 12) +
  theme(panel.grid = element_blank(), axis.text = element_text(face = "bold")))
```

```{r picking-genes}
isr_set <- c("ATF4","ATF3","DDIT3","CHAC1","TRIB3","ASNS","PSAT1","PHGDH","SHMT2","MTHFD2",
             "SLC7A11","SLC7A5","SLC3A2","WARS","CEBPB","PPP1R15A","DDIT4","HERPUD1","HSPA5",
             "XBP1","EIF2AK3","ATF6","DNAJB9","SDF2L1","MANF","CALR","PDIA4","SEC61B","SEL1L")

up_fig <- base %>%
  dplyr::filter(gene_category=="NMD", custom > 0.3, hgnc_symbol %in% isr_set) %>%
  dplyr::mutate(
    nmd_frac_DMSO = round(unprodCPM_DMSO/(prodCPM_DMSO + unprodCPM_DMSO), 3),
    nmd_frac_SMG  = round(unprodCPM_SMG /(prodCPM_SMG_raw + unprodCPM_SMG), 3)) %>%
  dplyr::left_join(sr_long %>% dplyr::select(ct, gene_id, sr_logFC), by=c("ct","gene_id")) %>%
  dplyr::transmute(hgnc_symbol, ct,
    prod_DMSO=round(prodCPM_DMSO,1), prod_SMG=round(prodCPM_SMG_raw,1),
    unprod_DMSO=round(unprodCPM_DMSO,1), unprod_SMG=round(unprodCPM_SMG,1),
    nmd_frac_DMSO, nmd_frac_SMG, custom=round(custom,2), sr_logFC=round(sr_logFC,2)) %>%
  dplyr::arrange(dplyr::desc(nmd_frac_SMG))
cat("ISR/UPR NMD-up genes, ranked by NMD-isoform visibility (nmd_frac_SMG):\n")
print(head(as.data.frame(up_fig), 20), row.names=FALSE)

sf_syms <- c(paste0("SRSF",1:12), paste0("HNRNP",c("A1","A2B1","C","D","K","L","M","U")),
             "TRA2A","TRA2B","RBM39","PTBP1","PTBP2","U2AF1","U2AF2")
ex_down <- prod_dir_mashr %>%
  filter(gene_category=="NMD", dir=="down", hgnc_symbol %in% sf_syms) %>%
  left_join(base %>% select(ct, gene_id, prodCPM_DMSO, prodCPM_SMG_adj, unprod_gain,
                            custom, prodfrac_DMSO, prodfrac_SMG), by=c("ct","gene_id")) %>%
  arrange(custom)
cat("\nDOWN candidates (splicing factors, NMD susceptible, productive DOWN):\n")
print(ex_down %>% transmute(hgnc_symbol, ct,
  prod_DMSO=round(prodCPM_DMSO,1), prod_SMG_adj=round(prodCPM_SMG_adj,1),
  unprod_gain=round(unprod_gain,1), custom_log2FC=round(custom,2),
  prodfrac_DMSO=round(prodfrac_DMSO,2), prodfrac_SMG=round(prodfrac_SMG,2)) %>% head(12))
```

```{r dna-down-examples}
dna_set <- unique(c(
  go_to_symbols(c("GO:0006281","GO:0006260","GO:0006310")),       # DNA repair, replication, recombination
  if (exists("down_sets")) down_sets$DNArep else character(0)))    # Hallmark DNA_REPAIR if Step 16 ran

dna_down <- prod_dir_mashr %>%
  dplyr::filter(gene_category=="NMD", dir=="down", hgnc_symbol %in% dna_set) %>%
  dplyr::left_join(base %>% dplyr::select(ct, gene_id, prodCPM_DMSO, prodCPM_SMG_adj,
        unprod_gain, custom, prodfrac_DMSO, prodfrac_SMG), by=c("ct","gene_id")) %>%
  dplyr::left_join(sr_long %>% dplyr::select(ct, gene_id, sr_logFC, sr_lfsr), by=c("ct","gene_id")) %>%
  classify_mech("down") %>%
  dplyr::arrange(custom)

cat(sprintf("DNA-repair/replication NMD genes with productive DOWN: %d gene-CT rows\n", nrow(dna_down)))
print(head(dna_down %>% dplyr::transmute(hgnc_symbol, ct,
  prod_DMSO=round(prodCPM_DMSO,1), prod_SMG=round(prodCPM_SMG_adj,1),
  unprod_gain=round(unprod_gain,1), custom=round(custom,2),
  prodfrac_DMSO=round(prodfrac_DMSO,2), prodfrac_SMG=round(prodfrac_SMG,2),
  sr_logFC=round(sr_logFC,2), mechanism), 20), row.names=FALSE)

cat("\nMechanism tally among DNA-repair productive-down genes:\n")
print(dna_down %>% dplyr::count(mechanism, name="n"))
```

```{r isopair-example-genes}
suppressPackageStartupMessages({
  library(ggplot2); library(dplyr); library(tidyr); library(tibble)
  library(patchwork); library(Isopair); library(scales)
})
select <- dplyr::select; filter <- dplyr::filter   # re-assert dplyr masking (Isopair loads its own select/filter)
MIN_FRAC <- 0.05; MIN_CPM  <- 1; MAX_ISO  <- 8
## ---- the three examples: gene, CT (display code), and its story ----
EXAMPLES <- tibble::tribble(
  ~gene,    ~ct,   ~subtitle,
  "SHMT2",  "MV",  "ISR/ATF4 induction overrides NMD: unproductive isoform surges to the majority, yet productive output still rises",
  "SRSF2",  "MV",  "Splicing-factor autoregulation: unproductive isoform surges, productive output falls",
  "PCNA",   "LAE", "Transcriptional repression: whole-gene output falls, isoform composition preserved")
ct_disp <- c(AT2="AT2", LAE="LAE", FB="FB", MV="MV")

COMBINED_GTF <- paste0(
  "/proj/regeps/regep00/studies/ExternalCellLines/analyses/repjc/",
  "Randell_Lung_Cells_2025/code/isocall_v1/projects/nmd_lungcells/",
  "results/call/nmd_isocall.isoforms.gtf.gz")
OUT_DIR_EX <- file.path(proj, "long_read/productive_response/examples")
dir.create(OUT_DIR_EX, recursive = TRUE, showWarnings = FALSE)

EXON_HALFH <- 0.22
BODY <- 10; HEADER <- 12; TAG_SIZE <- 14
PAL_NMD   <- c("Unproductive" = "#c0392b", "Productive" = "#16a085")
TREAT_COL <- c("DMSO" = "grey25", "Smg1i" = "#C0392B")

make_example_fig <- function(EXAMPLE_GENE, EXAMPLE_CT, FIG_SUB) {
  ctd <- ct_disp[[EXAMPLE_CT]]
  gid <- iso2gene %>% filter(hgnc_symbol == EXAMPLE_GENE) %>% pull(gene_id) %>% unique()
  if (length(gid) == 0) { cat(sprintf("** %s not found\n", EXAMPLE_GENE)); return(NULL) }
  gid <- gid[1]
  TARGETS <- iso2gene %>% filter(gene_id == gid) %>% pull(isoform_id) %>% unique()

  iso_status <- data.frame(
    isoform_id = lfsr_iso$isoform_id,
    lfsr = lfsr_iso[[resolve_mash_col(lfsr_iso, EXAMPLE_CT)]],
    pm   = pm_iso[[resolve_mash_col(pm_iso,   EXAMPLE_CT)]]) %>%
    filter(isoform_id %in% TARGETS) %>%
    mutate(nmd_class = ifelse(!is.na(lfsr) & !is.na(pm) & lfsr < LFSR_NMD_THR & pm > 0,
                              "Unproductive", "Productive")) %>%
    arrange(desc(nmd_class), isoform_id)

  cs <- rownames(samples)[samples$cell_type == EXAMPLE_CT]
  cpm_ct <- edgeR::cpm(counts_iso[, cs, drop = FALSE], log = FALSE)
  keep <- rownames(cpm_ct)[rowMeans(cpm_ct[, , drop=FALSE]) >= 1]
  iso_status <- iso_status %>% filter(isoform_id %in% keep)
  if (nrow(iso_status) < 2) { cat(sprintf("** %s: <2 expressed isoforms\n", EXAMPLE_GENE)); return(NULL) }

  structures <- parseIsoformStructures(COMBINED_GTF, isoform_ids = iso_status$isoform_id, verbose = FALSE)
  iso_status <- iso_status %>% filter(isoform_id %in% structures$isoform_id)
  ISOFORMS <- iso_status$isoform_id
  strand <- structures$strand[1]
  y_centers <- setNames(seq(length(ISOFORMS), 1L), ISOFORMS)

  make_introns <- function(iso){ s<-structures[structures$isoform_id==iso,]
    es<-s$exon_starts[[1]]; ee<-s$exon_ends[[1]]; o<-order(es); es<-es[o]; ee<-ee[o]
    if(length(es)<=1L) return(NULL)
    data.frame(isoform_id=iso, x=ee[-length(ee)], xend=es[-1L]) }
  introns <- bind_rows(lapply(ISOFORMS, make_introns))
  if (nrow(introns)) { introns$y <- y_centers[introns$isoform_id]; introns$yend <- introns$y }
  make_chevrons <- function(intr, strand, arm_size=0.07, min_intron_bp=30){
    if(is.null(intr)||nrow(intr)==0) return(NULL)
    rng <- max(intr$xend)-min(intr$x); spacing <- max(rng*0.03, 200); arm_w_default <- spacing*0.3
    out<-list()
    for(i in seq_len(nrow(intr))){ a<-intr$x[i]; b<-intr$xend[i]; yv<-intr$y[i]; len<-b-a
      if(len<min_intron_bp) next; arm_w<-min(arm_w_default,len*0.4)
      nch<-max(1L,floor(len/spacing)); step<-len/(nch+1); cxs<-a+step*seq_len(nch)
      for(cx in cxs){ if(strand=="+"){xs<-c(cx-arm_w,cx-arm_w)}else{xs<-c(cx+arm_w,cx+arm_w)}
        out[[length(out)+1L]]<-data.frame(x=xs,xend=c(cx,cx),y=c(yv+arm_size,yv-arm_size),yend=c(yv,yv)) } }
    if(length(out)) do.call(rbind,out) else NULL }
  chevrons <- bind_rows(lapply(ISOFORMS, function(iso) make_chevrons(introns[introns$isoform_id==iso,], strand)))

  regs <- bind_rows(lapply(ISOFORMS, function(iso){ s<-structures[structures$isoform_id==iso,]
    data.frame(isoform_id=iso, xmin=s$exon_starts[[1]], xmax=s$exon_ends[[1]]) }))
  regs$y_center <- y_centers[regs$isoform_id]
  regs$nmd_class <- iso_status$nmd_class[match(regs$isoform_id, iso_status$isoform_id)]
  regs$ymin <- regs$y_center - EXON_HALFH; regs$ymax <- regs$y_center + EXON_HALFH

  x_min<-min(regs$xmin); x_max<-max(regs$xmax); x_buf<-max((x_max-x_min)*0.04,50)
  label_df <- tibble(x=x_min-x_buf*1.5, y=y_centers[ISOFORMS], isoform_id=ISOFORMS) %>%
    left_join(iso_status %>% select(isoform_id,nmd_class), by="isoform_id") %>% mutate(label=isoform_id)

  p_A <- ggplot() +
    { if (nrow(introns)) geom_segment(data=introns, aes(x,xend=xend,y=y,yend=yend), color="gray55", linewidth=0.4) } +
    { if (!is.null(chevrons)&&nrow(chevrons)) geom_segment(data=chevrons, aes(x,xend=xend,y=y,yend=yend), color="gray55", linewidth=0.3) } +
    geom_rect(data=regs, aes(xmin=xmin,xmax=xmax,ymin=ymin,ymax=ymax,fill=nmd_class), color="black", linewidth=0.3) +
    scale_fill_manual(values=PAL_NMD, breaks=c("Unproductive","Productive"), name=NULL) +
    geom_text(data=label_df, aes(x,y,label=label,colour=nmd_class), hjust=1, size=3.0, fontface="bold") +
    scale_colour_manual(values=PAL_NMD, guide="none") +
    scale_y_continuous(breaks=NULL, limits=c(min(y_centers)-0.5, max(y_centers)+0.5)) +
    coord_cartesian(xlim=c(x_min-x_buf*4, x_max+x_buf), clip="off") +
    labs(x=sprintf("Genomic position (%s strand)", strand), y=NULL,
         title=sprintf("%s isoforms — structure and NMD status in %s", EXAMPLE_GENE, ctd)) +
    theme_minimal() +
    theme(panel.grid.major.y=element_blank(), panel.grid.minor=element_blank(),
          legend.position="bottom", plot.title=element_text(size=HEADER,face="bold"),
          plot.margin=margin(5,10,5,110))

  cpmL <- edgeR::cpm(counts_iso[ISOFORMS, cs, drop=FALSE], log=TRUE, prior.count=1)
  dcols <- cs[samples[cs,"treatment"]=="DMSO"]; scols <- cs[samples[cs,"treatment"]=="Smg1i"]
  expr <- data.frame(isoform_id=rownames(cpmL),
                     DMSO=rowMeans(cpmL[,dcols,drop=FALSE]),
                     Smg1i=rowMeans(cpmL[,scols,drop=FALSE])) %>%
    mutate(isoform_id=factor(isoform_id, levels=rev(ISOFORMS)))
  expr_long <- expr %>% pivot_longer(c(DMSO,Smg1i), names_to="treatment", values_to="log2cpm")
  yax_col <- PAL_NMD[iso_status$nmd_class[match(rev(ISOFORMS), iso_status$isoform_id)]]
  p_B <- ggplot() +
    geom_segment(data=expr, aes(x=DMSO,xend=Smg1i,y=isoform_id,yend=isoform_id), color="grey70", linewidth=0.8) +
    geom_point(data=expr_long, aes(log2cpm,isoform_id,color=treatment), size=3) +
    scale_color_manual(values=TREAT_COL, name=NULL) +
    labs(x="log2(CPM)", y=NULL, title=paste0(EXAMPLE_GENE," expression in ",ctd),
         subtitle="TMM-normalised; mean DMSO vs Smg1i") +
    theme_minimal(base_size=BODY) +
    theme(panel.grid.minor=element_blank(), panel.grid.major.y=element_blank(),
          legend.position="top", plot.title=element_text(face="bold",size=HEADER),
          plot.subtitle=element_text(color="gray30",size=BODY), axis.text.y=element_text(color=yax_col))

  donors <- unique(samples[cs,"sample_id"])
  cpm_lin <- edgeR::cpm(counts_iso[, cs, drop=FALSE], log=FALSE)
  fc <- bind_rows(lapply(ISOFORMS, function(iso){
    v <- vapply(donors, function(d){
      dc <- rownames(samples)[samples$cell_type==EXAMPLE_CT & samples$sample_id==d & samples$treatment=="DMSO"]
      sc <- rownames(samples)[samples$cell_type==EXAMPLE_CT & samples$sample_id==d & samples$treatment=="Smg1i"]
      if (length(dc)!=1 || length(sc)!=1) return(NA_real_)
      log2((cpm_lin[iso,sc]+1)/(cpm_lin[iso,dc]+1)) }, numeric(1))
    v <- v[is.finite(v)]
    data.frame(isoform_id=iso, logFC=mean(v), se=ifelse(length(v)>1, sd(v)/sqrt(length(v)), NA), n=length(v)) })) %>%
    left_join(iso_status %>% select(isoform_id, nmd_class, lfsr), by="isoform_id") %>%
    mutate(CI.L=logFC-1.96*se, CI.R=logFC+1.96*se,
           isoform_id=factor(isoform_id, levels=rev(ISOFORMS)), sig=lfsr<0.05)
  p_C <- ggplot(fc, aes(logFC, isoform_id)) +
    geom_vline(xintercept=0, color="grey60", linetype="dashed") +
    geom_pointrange(aes(xmin=CI.L, xmax=CI.R, color=nmd_class, shape=sig), size=0.5, linewidth=0.7) +
    scale_color_manual(values=PAL_NMD, name=NULL) +
    scale_shape_manual(values=c(`TRUE`=16,`FALSE`=1), breaks=c("TRUE","FALSE"),
                       labels=c("lfsr < 0.05","n.s."), name=NULL) +
    labs(x="Smg1i log2 fold-change (donor-paired, 95% CI)", y=NULL,
         title=paste0("Smg1i effect on ",EXAMPLE_GENE," (",ctd,")"),
         subtitle="per-isoform, donor-paired; colored by NMD class") +
    theme_minimal(base_size=BODY) +
    theme(panel.grid.minor=element_blank(), panel.grid.major.y=element_blank(),
          legend.position="top", plot.title=element_text(face="bold",size=HEADER),
          plot.subtitle=element_text(color="gray30",size=BODY))

  bottom <- (p_B + labs(title=NULL,subtitle=NULL)) +
            (p_C + labs(title=NULL,subtitle=NULL) +
               theme(axis.text.y=element_blank(), axis.ticks.y=element_blank())) +
            plot_layout(widths=c(1.25,1))
  combined <- (p_A + labs(title=NULL,subtitle=NULL)) / bottom +
    plot_layout(heights=c(1.55,1)) +
    plot_annotation(tag_levels="A",
      title=sprintf("%s in %s", EXAMPLE_GENE, ctd), subtitle=FIG_SUB) &
    theme(plot.tag=element_text(face="bold",size=TAG_SIZE),
          plot.title=element_text(face="bold",size=HEADER),
          plot.subtitle=element_text(color="gray30",size=BODY),
          plot.background=element_rect(color="grey25",fill=NA,linewidth=0.5),
          plot.margin=margin(6,6,6,6))

  ggsave(file.path(OUT_DIR_EX, sprintf("FIG_%s_%s.pdf", EXAMPLE_GENE, EXAMPLE_CT)),
         combined, width=13, height=9, units="in")
  ggsave(file.path(OUT_DIR_EX, sprintf("FIG_%s_%s.png", EXAMPLE_GENE, EXAMPLE_CT)),
         combined, width=13, height=9, units="in", dpi=300)
  print(combined)
  cat(sprintf("Saved %s (%s): %d isoforms (%d unproductive)\n",
              EXAMPLE_GENE, ctd, nrow(iso_status), sum(iso_status$nmd_class=="Unproductive")))
  invisible(combined)
}

for (i in seq_len(nrow(EXAMPLES)))
  make_example_fig(EXAMPLES$gene[i], EXAMPLES$ct[i], EXAMPLES$subtitle[i])
```

```{r specific-splicing}
splice_set <- go_to_symbols(c("GO:0008380","GO:0000380","GO:0043484"))   # RNA splicing + reg. of AS
sf_curated <- c(paste0("SRSF",1:12),
                paste0("HNRNP",c("A1","A2B1","A3","C","D","F","H1","K","L","M","U")),
                "TRA2A","TRA2B","RBM39","PTBP1","PTBP2","RBFOX2","CELF1","MBNL1")

enrich_down_byset <- function(set,label) dplyr::bind_rows(lapply(CELL_TYPES,function(ct_i)
  dplyr::bind_rows(lapply(MAIN_CATS,function(cl){
    d   <- dplyr::filter(prod_dir_mashr, ct==ct_i, gene_category==cl)
    hit <- unique(d$hgnc_symbol[d$dir=="down"]); bg <- unique(d$hgnc_symbol)
    ft  <- safe_fisher(bg %in% hit, bg %in% set)
    data.frame(set=label, ct=ct_i, class=cl, n_down=length(hit),
               n_down_in_set=sum(hit %in% set),
               OR=round(as.numeric(ft$estimate),2), p=signif(ft$p.value,3)) }))))
kable(enrich_down_byset(splice_set,"RNA splicing (GO)"),
      caption="Splicing-regulation genes enriched among productive-DOWN genes (one-sided Fisher).")
kable(enrich_down_byset(sf_curated,"SR/hnRNP curated"),
      caption="Curated SR/hnRNP factors enriched among productive-DOWN genes.")

kable(prod_dir_mashr %>% dplyr::filter(hgnc_symbol %in% paste0("SRSF",1:12)) %>%
        dplyr::group_by(ct) %>%
        dplyr::summarise(n_SR_tested=dplyr::n_distinct(hgnc_symbol),
                  n_SR_down=dplyr::n_distinct(hgnc_symbol[dir=="down"]),
                  n_SR_up=dplyr::n_distinct(hgnc_symbol[dir=="up"]), .groups="drop"),
      caption="Canonical SR proteins (SRSF1-12): productive direction per cell type.")

kable(base %>% dplyr::filter(hgnc_symbol %in% sf_curated) %>%
        dplyr::group_by(ct) %>%
        dplyr::summarise(n=dplyr::n(), median_custom=round(median(custom),3),
                  pct_down_0.3=round(100*mean(custom < -DROP_THR),1),
                  wilcox_p=signif(suppressWarnings(wilcox.test(custom)$p.value),3), .groups="drop"),
      caption="Productive log2FC of curated splicing factors per CT (median; Wilcoxon vs 0).")
```

# SUPPLEMENT

## S0 — The uncertain group on every distribution

```{r S0-unsure-dists, fig.height=4}
ALL3 <- c("NMD","non-NMD","unsure")
print(plot_dist(base, "custom",          "Adjusted productive log2FC — incl. unsure", xlab="productive log2 fold-change (adjusted)", cats=ALL3, xlim=c(-2,2), vline=0))
print(plot_dist(base, "prod_shift_adj",  "Productive-only output shift (neg = lower in SMG1i) — incl. unsure", xlab="change in productive output (fraction of DMSO gene total)", cats=ALL3, xlim=c(-0.5,0.5), vline=0))
print(plot_dist(base, "prodfrac_change", "Productive-SHARE change (incl. surge) — incl. unsure", xlab="change in productive fraction of the gene", cats=ALL3, xlim=c(-0.75,0.5), vline=0))
print(plot_dist(base, "prodCPM_DMSO",    "DMSO productive CPM — incl. unsure",         xlab="productive CPM (DMSO)", cats=ALL3, logx=TRUE))
print(plot_dist(base, "prodCPM_SMG_adj", "Adjusted SMG1i productive CPM — incl. unsure", xlab="productive CPM (SMG1i, adjusted)", cats=ALL3, logx=TRUE))
kable(med_tab(base, "custom", cats=ALL3), caption="TABLE S0. Adjusted productive log2FC including the unsure group.")
```

```{r S0-unsure-box, fig.height=4}
g3 <- base %>% mutate(gene_category=factor(gene_category, levels=c("NMD","unsure","non-NMD")))
print(ggplot(g3, aes(gene_category, custom, fill=gene_category)) +
  geom_hline(yintercept=0, colour="grey40") +
  geom_boxplot(outlier.size=0.3, outlier.alpha=0.2, width=0.65, colour="grey25") +
  facet_wrap(~ct, nrow=1) + coord_cartesian(ylim=c(-1.2,1.2)) +
  scale_fill_manual(values=cat_colors, guide="none") +
  labs(x=NULL, y="productive log2FC (custom)", title="FIGURE S0. Including the uncertain group") +
  theme_bw(base_size=11) + theme(strip.background=element_rect(fill="grey92")))
```

## S1 — Concordance: custom vs raw (Kitagawa)

```{r S1-concordance, fig.height=4}
agree <- base %>% filter(gene_category %in% MAIN_CATS) %>% group_by(ct) %>%
  summarise(n=n(), pearson=round(cor(custom, Kitagawa),3),
            spearman=round(cor(custom, Kitagawa, method="spearman"),3),
            median_custom_minus_Kitagawa=round(median(custom-Kitagawa),3), .groups="drop")
kable(agree, caption="TABLE S1. Per-gene agreement between custom and raw (Kitagawa).")
print(ggplot(base %>% filter(gene_category %in% MAIN_CATS), aes(Kitagawa, custom, colour=gene_category)) +
  geom_abline(slope=1, intercept=0, linetype="dashed", colour="grey50") +
  geom_point(alpha=0.1, size=0.4) + facet_wrap(~ct, nrow=1) +
  coord_cartesian(xlim=c(-1.5,1.5), ylim=c(-1.5,1.5)) +
  scale_colour_manual(values=cat_colors, name=NULL) +
  labs(title="FIGURE S1. custom vs Kitagawa per gene (dashed = identity)") +
  theme_bw(base_size=11) + theme(strip.background=element_rect(fill="grey92"), legend.position="top"))
```

## S2 — Down-isoform genes

```{r S2-down, fig.height=4.5}
down_tab <- bind_rows(lapply(CELL_TYPES, function(ct) data.frame(ct=ct, gene_id=down_iso_genes(ct), stringsAsFactors=FALSE)))
grp_dn <- bind_rows(
  base %>% transmute(ct, custom, group=gene_category),
  base %>% inner_join(down_tab, by=c("ct","gene_id")) %>% transmute(ct, custom, group="down-isoform")
) %>% mutate(group=factor(group, levels=c("NMD","unsure","non-NMD","down-isoform")))
kable(grp_dn %>% group_by(ct, group) %>%
        summarise(n=n(), median_custom=round(median(custom),3),
                  pct_down_0.3=round(100*mean(custom < -DROP_THR),1), .groups="drop") %>%
        pivot_wider(names_from=group, values_from=c(n, median_custom, pct_down_0.3)),
      caption="TABLE S2. Productive change for categories + the down-isoform group (overlapping membership).")
ov_tab <- bind_rows(lapply(CELL_TYPES, function(ct){
  dn <- down_iso_genes(ct); nm <- nmd_iso_genes(ct); k <- length(intersect(dn, nm))
  data.frame(ct=ct, n_down_gene=length(dn), n_down_with_NMD_iso=k,
             pct_with_NMD_iso=round(100*k/length(dn),1), n_down_only=length(setdiff(dn,nm)))
}))
kable(ov_tab, caption="TABLE S2b. Of down-isoform genes, fraction also carrying an NMD (up) isoform vs down-only.")
```

```{r S2-down-topgo, eval = requireNamespace("topGO", quietly=TRUE) && requireNamespace("org.Hs.eg.db", quietly=TRUE)}
if (exists("run_bp")) for (ct_i in CELL_TYPES){
  uni <- base$gene_id[base$ct==ct_i]
  run_bp(uni, intersect(down_iso_genes(ct_i), uni), sprintf("down-isoform | %s (all)", ct_i))
  run_bp(uni, intersect(setdiff(down_iso_genes(ct_i), nmd_iso_genes(ct_i)), uni),
         sprintf("down-isoform | %s (down-only)", ct_i))
}
```

## S3 — Where do short-read DOWN-regulated genes land? (split by NMD-isoform status)

```{r S3-sr-down-split, fig.height=4.5}
sr_down <- sr_long %>% filter(!is.na(sr_lfsr), !is.na(sr_logFC), sr_lfsr<0.05, sr_logFC<0) %>%
  select(ct, gene_id) %>% distinct()
nmd_iso_tab <- bind_rows(lapply(CELL_TYPES, function(ct) data.frame(ct=ct, gene_id=nmd_iso_genes(ct), stringsAsFactors=FALSE)))
sr_down_split <- sr_down %>%
  mutate(has_nmd_iso = paste(ct,gene_id) %in% paste(nmd_iso_tab$ct, nmd_iso_tab$gene_id)) %>%
  mutate(group = ifelse(has_nmd_iso, "SR-down (NMD-iso)", "SR-down (no NMD-iso)"))
grp_sr <- bind_rows(
  base %>% filter(gene_category %in% MAIN_CATS) %>% transmute(ct, custom, group=gene_category),
  base %>% inner_join(sr_down_split %>% select(ct, gene_id, group), by=c("ct","gene_id")) %>%
    transmute(ct, custom, group)
) %>% mutate(group=factor(group, levels=c("NMD","non-NMD","SR-down (NMD-iso)","SR-down (no NMD-iso)")))
kable(grp_sr %>% group_by(ct, group) %>%
        summarise(n=n(), median_custom=round(median(custom),3),
                  pct_down_0.3=round(100*mean(custom < -DROP_THR),1), .groups="drop") %>%
        pivot_wider(names_from=group, values_from=c(n, median_custom, pct_down_0.3)),
      caption="TABLE S3. SR-down genes split by NMD-isoform status vs NMD and non-NMD baselines.")
print(ggplot(grp_sr, aes(group, custom, fill=group)) +
  geom_hline(yintercept=0, colour="grey40") +
  geom_boxplot(outlier.size=0.3, outlier.alpha=0.2, width=0.6, colour="grey25") +
  facet_wrap(~ct, nrow=1) + coord_cartesian(ylim=c(-1.5,1.5)) +
  scale_fill_manual(values=c("NMD"="#c0392b","non-NMD"="#7f8c8d",
                             "SR-down (NMD-iso)"="#8e44ad","SR-down (no NMD-iso)"="#2980b9"), guide="none") +
  labs(x=NULL, y="productive log2FC (custom)", title="FIGURE S3. SR down-regulated genes, split by NMD-isoform status") +
  theme_bw(base_size=11) + theme(strip.background=element_rect(fill="grey92"), axis.text.x=element_text(angle=20, hjust=1)))
```

```{r S3-sr-down-topgo, eval = requireNamespace("topGO", quietly=TRUE) && requireNamespace("org.Hs.eg.db", quietly=TRUE)}
if (exists("run_bp")) for (ct_i in CELL_TYPES){
  uni <- base$gene_id[base$ct==ct_i]
  run_bp(uni, intersect(sr_down_split$gene_id[sr_down_split$ct==ct_i & sr_down_split$has_nmd_iso],  uni), sprintf("SR-down NMD-iso | %s", ct_i))
  run_bp(uni, intersect(sr_down_split$gene_id[sr_down_split$ct==ct_i & !sr_down_split$has_nmd_iso], uni), sprintf("SR-down no-NMD-iso | %s", ct_i))
}
```

## S4 — Per-donor productive change with error bars

```{r S4-perdonor-candidate, fig.height=4}
perdonor <- gene_sample %>%
  filter(gene_category %in% MAIN_CATS, !is.na(prodCPM_DMSO), !is.na(prodCPM_SMG_adj)) %>%
  mutate(custom_d = log2((prodCPM_SMG_adj+1)/(prodCPM_DMSO+1)))
pd_sum <- perdonor %>% group_by(ct, donor, gene_category) %>%
  summarise(median_custom=median(custom_d, na.rm=TRUE), .groups="drop")
print(ggplot(pd_sum, aes(gene_category, median_custom, colour=gene_category)) +
  geom_hline(yintercept=0, colour="grey70") +
  geom_point(position=position_jitter(width=0.12), size=2, alpha=0.8) +
  stat_summary(fun=median, geom="crossbar", width=0.5, colour="black", linewidth=0.3) +
  facet_wrap(~ct, nrow=1) +
  scale_colour_manual(values=cat_colors, guide="none") +
  labs(x=NULL, y="per-donor median productive log2FC (custom)",
       title="S4. Per-donor productive change (each point = one donor)") +
  theme_bw(base_size=11) + theme(strip.background=element_rect(fill="grey92")))
```

## S5 — Kitagawa on all significant NMD genes (FB excluded)

```{r kitagawa-on-all}
# General Kitagawa result: all significant NMD genes (up or down), mashr-called (AT2/LAE/MV)
kit_nmd_table() %>% dplyr::count(driver) %>% dplyr::mutate(pct = round(100*n/sum(n),1))
```

## S6 — Productive-DOWN: does it co-occur with a down-isoform, and which mechanism?

```{r check-die-down}
prod_down <- prod_dir_mashr %>% dplyr::filter(dir=="down") %>%
  dplyr::select(ct, gene_id) %>% dplyr::distinct() %>% dplyr::mutate(prod_down=TRUE)
down_iso <- dplyr::bind_rows(lapply(CELL_TYPES, function(ct)
  data.frame(ct=ct, gene_id=down_iso_genes(ct), down_iso=TRUE)))

cmp_di <- prod_dir_mashr %>% dplyr::select(ct, gene_id) %>% dplyr::distinct() %>%
  dplyr::left_join(prod_down, by=c("ct","gene_id")) %>%
  dplyr::left_join(down_iso,  by=c("ct","gene_id")) %>%
  dplyr::mutate(prod_down=!is.na(prod_down), down_iso=!is.na(down_iso))

xtab <- dplyr::bind_rows(lapply(CELL_TYPES, function(ct_i){
  d <- cmp_di %>% dplyr::filter(ct==ct_i)
  ft <- fisher.test(table(factor(d$prod_down,c(F,T)), factor(d$down_iso,c(F,T))), alternative="greater")
  data.frame(ct=ct_i, n=nrow(d),
             prod_down=sum(d$prod_down), down_iso=sum(d$down_iso),
             both=sum(d$prod_down & d$down_iso),
             prod_down_only=sum(d$prod_down & !d$down_iso),
             down_iso_only=sum(!d$prod_down & d$down_iso),
             pct_proddown_with_downiso=round(100*mean(d$down_iso[d$prod_down]),1),
             OR=round(as.numeric(ft$estimate),2), p=signif(ft$p.value,3))
}))
print(xtab, row.names=FALSE)

mech <- cmp_di %>% dplyr::filter(prod_down) %>%
  dplyr::left_join(base %>% dplyr::select(ct,gene_id,unprod_gain,custom,prodfrac_DMSO,prodfrac_SMG),
                   by=c("ct","gene_id")) %>%
  dplyr::left_join(sr_long %>% dplyr::select(ct,gene_id,sr_logFC), by=c("ct","gene_id")) %>%
  dplyr::mutate(group=ifelse(down_iso,"prod-down WITH down-isoform","prod-down WITHOUT down-isoform")) %>%
  dplyr::group_by(group) %>%
  dplyr::summarise(n=dplyr::n(),
            median_sr_logFC=round(median(sr_logFC,na.rm=TRUE),2),
            median_unprod_gain=round(median(unprod_gain,na.rm=TRUE),2),
            median_prodfrac_drop=round(median(prodfrac_SMG-prodfrac_DMSO,na.rm=TRUE),3),
            .groups="drop")
print(mech, row.names=FALSE)
```

## S6b — Down-arm mechanism split (transcriptional vs composition; AT2/LAE/MV)

```{r down-arm-mechanism}
# classify productive-down genes once (FB excluded: provisional CT)
pd_mech <- prod_dir_mashr %>% dplyr::filter(dir=="down", ct != "FB") %>%
  dplyr::left_join(base %>% dplyr::select(ct,gene_id,prodfrac_DMSO,prodfrac_SMG), by=c("ct","gene_id")) %>%
  dplyr::left_join(sr_long %>% dplyr::select(ct,gene_id,sr_logFC), by=c("ct","gene_id")) %>%
  classify_mech("down")
cat("Down-arm mechanism split:\n")
print(pd_mech %>% dplyr::count(mechanism) %>% dplyr::mutate(pct = round(100*n/sum(n),1)))

trans_g    <- pd_mech %>% dplyr::filter(mechanism=="transcriptional") %>% dplyr::pull(gene_id) %>% unique()
comp_g     <- pd_mech %>% dplyr::filter(mechanism=="composition")     %>% dplyr::pull(gene_id) %>% unique()
universe_g <- unique(prod_dir_mashr$gene_id[prod_dir_mashr$ct != "FB"])
cat(sprintf("transcriptional=%d | composition=%d | universe=%d\n", length(trans_g), length(comp_g), length(universe_g)))
cat("\nTheme enrichment: transcriptional vs composition productive-down genes\n")
print(theme_enrichment(trans_g, comp_g, universe_g), row.names=FALSE)
if (exists("run_bp")) { run_bp(universe_g, trans_g, "TRANSCRIPTIONAL productive-down (GO BP)")
                        run_bp(universe_g, comp_g,  "COMPOSITION productive-down (GO BP)") }
```

```{r kitagawa-proof}
# Kitagawa driver split by direction (AT2/LAE/MV)
kit_nmd_table() %>% dplyr::count(dir, driver) %>%
  dplyr::group_by(dir) %>% dplyr::mutate(pct=round(100*n/sum(n),1)) %>% dplyr::ungroup()
```

## S7 — Up-arm mechanism split + decomposition + theme enrichment

```{r up-arm-stuff}
## ---- 1) mechanism classification of productive-UP genes (AT2/LAE/MV) ----
pu <- prod_dir_mashr %>% dplyr::filter(dir=="up", ct != "FB") %>%
  dplyr::left_join(base %>% dplyr::select(ct,gene_id,prodfrac_DMSO,prodfrac_SMG), by=c("ct","gene_id")) %>%
  dplyr::left_join(sr_long %>% dplyr::select(ct,gene_id,sr_logFC), by=c("ct","gene_id")) %>%
  classify_mech("up")
cat("Up-arm mechanism split:\n")
print(pu %>% dplyr::count(mechanism) %>% dplyr::mutate(pct = round(100*n/sum(n),1)))

## ---- 2) Kitagawa decomposition on productive-UP genes ----
cat("\nKitagawa decomposition (productive-UP):\n")
print(kit_nmd_table() %>% dplyr::filter(dir=="up") %>%
  dplyr::count(driver) %>% dplyr::mutate(pct=round(100*n/sum(n),1)))

## ---- 3) theme enrichment: transcriptional-up vs composition-up ----
trans_g    <- pu %>% dplyr::filter(mechanism=="transcriptional") %>% dplyr::pull(gene_id) %>% unique()
comp_g     <- pu %>% dplyr::filter(mechanism=="composition")     %>% dplyr::pull(gene_id) %>% unique()
universe_g <- unique(prod_dir_mashr$gene_id[prod_dir_mashr$ct != "FB"])
cat(sprintf("\ntranscriptional-up=%d | composition-up=%d | universe=%d\n", length(trans_g), length(comp_g), length(universe_g)))
# up arm adds the NF-kB set to the shared theme list
sets_up <- c(THEME_SETS, list("NF-kB/TNFA (GO)" = go_to_symbols(c("GO:0051092","GO:0034612"))))
cat("\nTheme enrichment: transcriptional-up vs composition-up\n")
print(theme_enrichment(trans_g, comp_g, universe_g, sets_up), row.names=FALSE)
if (exists("run_bp")) { run_bp(universe_g, trans_g, "TRANSCRIPTIONAL productive-UP (GO BP)")
  if (length(comp_g) >= 5) run_bp(universe_g, comp_g, "COMPOSITION productive-UP (GO BP)")
  else cat(sprintf("\nComposition-up group too small for topGO (n=%d)\n", length(comp_g))) }
```

```{r session}
sessionInfo()
```
