---
title: "NMD Transcriptional Output: Cross-Pipeline Analysis"
author: "Leshem, Yul"
date: "2026-03-18"
output:
  html_document:
    theme: lumen
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
    code_folding: hide
---

Code location: `/udd/reyle/nmd_lungcells_2026/code/final/transcriptional_output.Rmd`

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, dev = "png")
options(bitmapType = "cairo")

```

# Overview

This report quantifies how much of the long-read transcriptome's output is
suppressed by NMD under DMSO conditions, by measuring how much of each
isoform's (and gene's) Smg1i CPM appears only after NMD inhibition.

For each cell-type x donor pair we compute the per-isoform delta
(Smg1i − DMSO), keep only positive deltas (NMD-rescued), and divide by Smg1i
CPM to get the fraction of output that NMD was degrading. Results are
aggregated per cell type at both isoform and gene level, and compared with
short-read gene-level mashr DGE counts.

Naming: the DGEList stores the original ct codes `AT` (= AT2) and `DD` (= LAE);
we relabel to the display names right after loading and use AT2 / LAE
throughout.

```{r libraries}
suppressPackageStartupMessages({
  library(data.table)
  library(dplyr)
  library(ggplot2)
  library(tidyr)
  library(edgeR)
  library(DT)
  library(tibble)
})

# Bioconductor namespace masking guard
select <- dplyr::select
filter <- dplyr::filter
rename <- dplyr::rename
count  <- dplyr::count
slice  <- dplyr::slice
first  <- dplyr::first

# -----------------------------------------------------------------------------
# Paths and constants
# -----------------------------------------------------------------------------
BASE              <- "/udd/reyle/nmd_lungcells_2026"
DGELIST_PATH      <- file.path(BASE, "long_read/dge_isoform_longread_2026.3.3.rds")
DGELIST_FILT_PATH <- file.path(BASE, "long_read/dge_isoform_longread_filtered_2026.3.3.rds")
SR_DGE_DIR        <- file.path(BASE, "shortread_dge/mashr")
OUTPUT_DIR        <- file.path(BASE, "long_read/transcriptional_output")

dir.create(OUTPUT_DIR, recursive = TRUE, showWarnings = FALSE)

# 4 cell types only (DD_ALI / DO_ALI dropped upstream), display naming.
# AT2 = old AT, LAE = old DD.
CT_KEEP    <- c("LAE", "AT2", "FB", "MV")
SR_PATTERN <- "^nmd_mashr_dge_.*_2026\\.3\\.10\\.csv$"

# SR mashr file codes -> display names. Pipeline writes nmd_mashr_dge_at2/lae/fb/mv.
CELLTYPE_MAP <- c("at2" = "AT2", "lae" = "LAE", "fb" = "FB", "mv" = "MV")

SIGNIFICANCE_THRESHOLD <- 0.05
```

# Load DGEList And Parse Metadata

```{r load}
dge <- readRDS(DGELIST_PATH)
cat("DGEList dimensions:", dim(dge), "\n")

counts_raw <- dge$counts

samp <- dge$samples %>%
  rownames_to_column("bam_id") %>%
  mutate(ct        = as.character(ct),
         treatment = as.character(treatment),
         donor_id  = id)

# Relabel ct to display names AFTER loading (AT -> AT2, DD -> LAE).
# counts are keyed by bam_id, so renaming ct does not disturb alignment.
samp$ct[samp$ct == "AT"] <- "AT2"
samp$ct[samp$ct == "DD"] <- "LAE"

# Drop ALI cell types
n_before <- nrow(samp)
samp     <- samp %>% filter(ct %in% CT_KEEP)
cat(sprintf("FILTER: keep CT_KEEP {%s}. Samples before: %d / after: %d (dropped %d outside CT_KEEP).\n",
            paste(CT_KEEP, collapse = ", "), n_before, nrow(samp), n_before - nrow(samp)))

counts_raw <- counts_raw[, samp$bam_id]

# Isoform annotation (one row per txid)
iso_info <- dge$genes %>%
  select(any_of(c("txid", "gene_id", "hgnc_symbol")))

cat("\nCell types:", paste(sort(unique(samp$ct)), collapse = ", "), "\n")
cat("Treatments:", paste(unique(samp$treatment), collapse = ", "), "\n")
cat("Donors:",     paste(sort(unique(samp$donor_id)), collapse = ", "), "\n")

```

# Identify Matched Donor Pairs

For each cell type, we keep only donors with both Smg1i and DMSO samples. The
delta calculation is paired within donor.

```{r matched-pairs}
paired_donors <- samp %>%
  group_by(ct, donor_id) %>%
  summarise(has_smg1i = any(treatment == "Smg1i"),
            has_dmso  = any(treatment == "DMSO"),
            .groups = "drop") %>%
  filter(has_smg1i & has_dmso)

cat("Matched donor pairs per cell type:\n")
paired_donors %>%
  group_by(ct) %>%
  summarise(n_pairs = n(),
            donors  = paste(donor_id, collapse = ", "),
            .groups = "drop") %>%
  print(n = 20)
```

# Normalize Counts (Plain CPM, No TMM)

TMM is not appropriate here because Smg1i genuinely shifts the transcriptome
composition by rescuing NMD targets — TMM would partially scrub that signal.
Instead we use plain CPM (library-size correction only).

```{r normalize}
# normalized.lib.sizes = FALSE tells edgeR to ignore norm.factors and use raw lib sizes
cpm_mat <- cpm(dge[, samp$bam_id], normalized.lib.sizes = FALSE, log = FALSE)

cat("CPM matrix:", nrow(cpm_mat), "isoforms x", ncol(cpm_mat), "samples\n")
cat("CPM range:", round(min(cpm_mat), 2), "–", round(max(cpm_mat), 1), "\n")

```

# Per-Isoform Output Lost To NMD

For each isoform $i$ in cell type $c$ and donor $d$:

$$\Delta_{i,d} = \text{CPM}_{i,d}^{\text{Smg1i}} - \text{CPM}_{i,d}^{\text{DMSO}}$$

$$\Delta_{i,d}^{+} = \max(\Delta_{i,d},\; 0)$$

$$f_{i,d} = \begin{cases} \dfrac{\Delta_{i,d}^{+}}{\text{CPM}_{i,d}^{\text{Smg1i}}} & \text{if } \text{CPM}_{i,d}^{\text{Smg1i}} > 0 \\[6pt] 0 & \text{otherwise} \end{cases}$$

$f_{i,d}$ is the fraction of isoform $i$'s transcriptional output lost to NMD
in donor $d$.

```{r compute-frac-lost}
results_list <- list()

for (ct_i in sort(unique(paired_donors$ct))) {
  donors_i <- paired_donors %>% filter(ct == ct_i) %>% pull(donor_id)

  for (d in donors_i) {
    smg1i_id <- samp %>%
      filter(ct == ct_i, donor_id == d, treatment == "Smg1i") %>% pull(bam_id)
    dmso_id  <- samp %>%
      filter(ct == ct_i, donor_id == d, treatment == "DMSO")  %>% pull(bam_id)

    stopifnot(length(smg1i_id) == 1, length(dmso_id) == 1)

    val_smg1i <- cpm_mat[, smg1i_id]
    val_dmso  <- cpm_mat[, dmso_id]
    delta     <- val_smg1i - val_dmso
    delta_pos <- pmax(delta, 0)
    frac_lost <- ifelse(val_smg1i > 0, delta_pos / val_smg1i, 0)

    results_list[[paste(ct_i, d, sep = "_")]] <- data.frame(
      txid      = rownames(cpm_mat),
      ct        = ct_i,
      donor_id  = d,
      smg1i_val = val_smg1i,
      dmso_val  = val_dmso,
      delta     = delta,
      delta_pos = delta_pos,
      frac_lost = frac_lost,
      stringsAsFactors = FALSE
    )
  }
}

results_all <- bind_rows(results_list)

stopifnot("txid must be unique in iso_info" = !anyDuplicated(iso_info$txid))
results_all <- results_all %>% left_join(iso_info, by = "txid")

cat("Total rows (isoform x ct x donor):", nrow(results_all), "\n")
```

## Average Fraction Lost Across Donors

$$\bar{f}_{i,c} = \frac{1}{|D_c|} \sum_{d \in D_c} f_{i,d}$$

where $D_c$ is the set of matched donors for cell type $c$.

```{r avg-frac-lost}
avg_lost <- results_all %>%
  group_by(ct, txid) %>%
  summarise(mean_frac_lost = mean(frac_lost, na.rm = TRUE),
            mean_delta_pos = mean(delta_pos, na.rm = TRUE),
            mean_smg1i_val = mean(smg1i_val, na.rm = TRUE),
            mean_dmso_val  = mean(dmso_val,  na.rm = TRUE),
            n_donors       = n(),
            gene_id        = first(gene_id),
            hgnc_symbol    = first(hgnc_symbol),
            .groups = "drop") %>%
  filter(mean_frac_lost > 0)

cat("Isoforms with positive mean loss to NMD per cell type:\n")
avg_lost %>%
  group_by(ct) %>%
  summarise(n_isoforms = n(), .groups = "drop") %>%
  print()

```

## Isoforms With Total Loss (100%)

Isoforms with `mean_frac_lost == 1.0` are present only in Smg1i (DMSO == 0 in
every matched donor) — entirely destroyed by NMD under normal conditions.

```{r pct-total-loss}
total_loss <- avg_lost %>%
  mutate(pct_lost = 100 * mean_frac_lost) %>%
  group_by(ct) %>%
  summarise(n_total_loss   = sum(pct_lost == 100),
            n_positive     = n(),
            pct_total_loss = round(100 * n_total_loss / n_positive, 2),
            .groups = "drop")

print(total_loss)
```

## Global % Output Lost Per Cell Type

$$\% \text{ Output Lost}_c = 100 \times \frac{\displaystyle\sum_{i=1}^{N}\sum_{d \in D_c} \Delta_{i,d}^{+}}{\displaystyle\sum_{i=1}^{N}\sum_{d \in D_c} \text{CPM}_{i,d}^{\text{Smg1i}}}$$

```{r global-metric}
global_metric <- results_all %>%
  group_by(ct) %>%
  summarise(sum_delta_pos   = sum(delta_pos),
            total_smg1i     = sum(smg1i_val),
            pct_output_lost = 100 * sum_delta_pos / total_smg1i,
            n_isoform_donor = n(),
            .groups = "drop") %>%
  arrange(desc(pct_output_lost))

print(global_metric)

```

```{r barplot-global, fig.width=10, fig.height=6}
p_global <- ggplot(global_metric,
                   aes(x = reorder(ct, pct_output_lost),
                       y = pct_output_lost, fill = ct)) +
  geom_col(alpha = 0.8) +
  geom_text(aes(label = paste0(round(pct_output_lost, 2), "%")),
            hjust = -0.1, size = 4) +
  coord_flip() +
  labs(
    title    = "% Transcriptional Output Lost To NMD By Cell Type",
    subtitle = "Isoform-level",
    x        = "Cell Type",
    y        = "% Output Lost To NMD"
  ) +
  theme_minimal(base_size = 14) +
  theme(legend.position = "none",
        plot.title      = element_text(face = "bold"),
        plot.subtitle   = element_text(color = "grey40")) +
  expand_limits(y = max(global_metric$pct_output_lost) * 1.15)

print(p_global)

ggsave(file.path(OUTPUT_DIR, "pct_output_lost_barplot.png"),
       p_global, width = 10, height = 6, dpi = 300, bg = "white")
```

## % Output Lost Vs Number Of NMD Susceptible Isoforms

```{r scatter-output-vs-nisoforms, fig.width=8, fig.height=6}
n_responsive <- avg_lost %>%
  group_by(ct) %>%
  summarise(n_nmd_isoforms = n(), .groups = "drop")

scatter_df <- global_metric %>% left_join(n_responsive, by = "ct")

fit       <- lm(pct_output_lost ~ n_nmd_isoforms, data = scatter_df)
r2        <- summary(fit)$r.squared
slope     <- coef(fit)[2]
fit_label <- sprintf("slope = %.2e, R² = %.3f", slope, r2)

p_scatter <- ggplot(scatter_df, aes(x = n_nmd_isoforms, y = pct_output_lost)) +
  geom_smooth(method = "lm", se = FALSE, linetype = "dashed", color = "grey40") +
  geom_point(aes(color = ct), size = 4) +
  geom_text(aes(label = ct), vjust = -1, size = 4.5) +
  annotate("text",
           x     = min(scatter_df$n_nmd_isoforms),
           y     = max(scatter_df$pct_output_lost) * 1.05,
           label = fit_label, hjust = 0, size = 4, color = "grey30") +
  labs(title = "% Output Lost Vs NMD Susceptible Isoforms",
       x = "Number Of NMD Susceptible Isoforms",
       y = "% Output Lost To NMD") +
  theme_minimal(base_size = 14) +
  theme(legend.position = "none",
        plot.title      = element_text(face = "bold")) +
  expand_limits(y = max(scatter_df$pct_output_lost) * 1.15,
                x = max(scatter_df$n_nmd_isoforms) * 1.05)

print(p_scatter)

ggsave(file.path(OUTPUT_DIR, "output_lost_vs_n_isoforms_scatter.png"),
       p_scatter, width = 8, height = 6, dpi = 300, bg = "white")

```

## % Output Lost Vs Significant Short-Read DGE Genes

Cross-pipeline check: does long-read transcriptional output loss track with
short-read gene-level NMD susceptible counts?

```{r scatter-output-vs-sig-genes, fig.width=8, fig.height=6}
sr_files <- list.files(SR_DGE_DIR, pattern = SR_PATTERN, full.names = TRUE)
cat("SR files matching pattern:", length(sr_files), "\n"); print(basename(sr_files))

# SR file pattern: nmd_mashr_dge_<ct>_<date>.csv -> sub() pulls <ct>
n_sig_genes <- lapply(sr_files, function(f) {
  ct_code <- sub("^nmd_mashr_dge_(.+)_2026.*", "\\1", basename(f))
  ct_name <- CELLTYPE_MAP[ct_code]
  if (is.na(ct_name) || !ct_name %in% CT_KEEP) return(NULL)

  df    <- fread(f)
  n_sig <- sum(df$adj.P.Val < SIGNIFICANCE_THRESHOLD & df$logFC > 0, na.rm = TRUE)
  data.frame(ct = ct_name, n_sig_genes = n_sig)
}) %>% bind_rows()

stopifnot("No SR files mapped — check CELLTYPE_MAP vs filenames" = nrow(n_sig_genes) > 0)

scatter_sr <- global_metric %>% left_join(n_sig_genes, by = "ct")

fit_sr       <- lm(pct_output_lost ~ n_sig_genes, data = scatter_sr)
r2_sr        <- summary(fit_sr)$r.squared
slope_sr     <- coef(fit_sr)[2]
fit_label_sr <- sprintf("slope = %.2e, R² = %.3f", slope_sr, r2_sr)

p_scatter_sr <- ggplot(scatter_sr, aes(x = n_sig_genes, y = pct_output_lost)) +
  geom_smooth(method = "lm", se = FALSE, linetype = "dashed", color = "grey40") +
  geom_point(aes(color = ct), size = 4) +
  geom_text(aes(label = ct), vjust = -1, size = 4.5) +
  annotate("text",
           x     = min(scatter_sr$n_sig_genes),
           y     = max(scatter_sr$pct_output_lost) * 1.05,
           label = fit_label_sr, hjust = 0, size = 4, color = "grey30") +
  labs(title    = "% Output Lost Vs Significant Short-Read DGE Genes",
       subtitle = "NMD susceptible genes: lfsr < 0.05 & logFC > 0",
       x = "Number Of NMD Susceptible Genes (Short-Read)",
       y = "% Output Lost To NMD") +
  theme_minimal(base_size = 14) +
  theme(legend.position = "none",
        plot.title      = element_text(face = "bold"),
        plot.subtitle   = element_text(color = "grey40")) +
  expand_limits(y = max(scatter_sr$pct_output_lost) * 1.15,
                x = max(scatter_sr$n_sig_genes) * 1.05)

print(p_scatter_sr)

ggsave(file.path(OUTPUT_DIR, "output_lost_vs_sig_sr_genes_scatter.png"),
       p_scatter_sr, width = 8, height = 6, dpi = 300, bg = "white")
```

## Per-Isoform Fraction Lost (Filtered Isoforms)

The filtered DGEList drops lowly-expressed isoforms. We use it here to keep
the per-isoform distribution comparable to what limma-voom actually tested.

```{r boxplot-isoform, fig.width=10, fig.height=6}
dge_filt      <- readRDS(DGELIST_FILT_PATH)
filt_isoforms <- rownames(dge_filt$counts)
cat("Filtered isoform set:", length(filt_isoforms), "\n")

plot_data <- results_all %>%
  filter(frac_lost > 0, txid %in% filt_isoforms)

cat("Isoform-donor rows after filtering:", nrow(plot_data), "\n")

p_isoform_dist <- ggplot(plot_data, aes(x = ct, y = frac_lost, fill = ct)) +
  geom_boxplot(alpha = 0.7, outlier.size = 0.3, outlier.alpha = 0.3) +
  labs(title    = "Per-Isoform Fraction Lost To NMD",
       subtitle = "(Smg1i − DMSO) / Smg1i CPM, positive deltas, filtered isoforms",
       x = "Cell Type", y = "Fraction Lost To NMD") +
  theme_minimal(base_size = 14) +
  theme(legend.position = "none",
        plot.title      = element_text(face = "bold"),
        plot.subtitle   = element_text(color = "grey40"))

print(p_isoform_dist)

ggsave(file.path(OUTPUT_DIR, "isoform_frac_lost_boxplot.png"),
       p_isoform_dist, width = 10, height = 6, dpi = 300, bg = "white")

boxplot_stats <- plot_data %>%
  group_by(ct) %>%
  summarise(n      = n(),
            min    = round(min(frac_lost), 4),
            Q1     = round(quantile(frac_lost, 0.25), 4),
            median = round(median(frac_lost),         4),
            mean   = round(mean(frac_lost),           4),
            Q3     = round(quantile(frac_lost, 0.75), 4),
            max    = round(max(frac_lost), 4),
            sd     = round(sd(frac_lost),  4),
            IQR    = round(IQR(frac_lost), 4),
            .groups = "drop")

print(boxplot_stats)

rm(dge_filt)

```

## Top 10 Isoforms By Mean % Lost

```{r top10}
top10_per_ct <- avg_lost %>%
  group_by(ct) %>%
  arrange(desc(mean_frac_lost), desc(mean_delta_pos)) %>%
  slice_head(n = 10) %>%
  mutate(pct_lost       = round(100 * mean_frac_lost, 2),
         mean_smg1i_val = round(mean_smg1i_val, 1),
         mean_dmso_val  = round(mean_dmso_val,  1),
         mean_delta_pos = round(mean_delta_pos, 1)) %>%
  select(ct, txid, hgnc_symbol, gene_id, pct_lost, mean_delta_pos,
         mean_smg1i_val, mean_dmso_val, n_donors)

for (ct_i in sort(unique(top10_per_ct$ct))) {
  cat("\n--- ", ct_i, " ---\n")
  print(top10_per_ct %>% filter(ct == ct_i), n = 10)
}
```

## Bottom 10 Isoforms By Mean % Lost

Smallest positive fractional loss — barely affected by NMD.

```{r bottom10}
bottom10_per_ct <- avg_lost %>%
  group_by(ct) %>%
  arrange(mean_frac_lost, mean_delta_pos) %>%
  slice_head(n = 10) %>%
  mutate(pct_lost       = round(100 * mean_frac_lost, 2),
         mean_smg1i_val = round(mean_smg1i_val, 1),
         mean_dmso_val  = round(mean_dmso_val,  1),
         mean_delta_pos = round(mean_delta_pos, 1)) %>%
  select(ct, txid, hgnc_symbol, gene_id, pct_lost, mean_delta_pos,
         mean_smg1i_val, mean_dmso_val, n_donors)

for (ct_i in sort(unique(bottom10_per_ct$ct))) {
  cat("\n--- ", ct_i, " ---\n")
  print(bottom10_per_ct %>% filter(ct == ct_i), n = 10)
}

```

# Gene-Level Output Lost To NMD

Isoform CPMs are summed within gene before the same delta calculation:

$$\Delta_{g,d}^{+} = \max\!\left(\text{CPM}_{g,d}^{\text{Smg1i}} - \text{CPM}_{g,d}^{\text{DMSO}},\; 0\right)$$

$$\% \text{ Output Lost}_c = 100 \times \frac{\displaystyle\sum_{g=1}^{G}\sum_{d \in D_c} \Delta_{g,d}^{+}}{\displaystyle\sum_{g=1}^{G}\sum_{d \in D_c} \text{CPM}_{g,d}^{\text{Smg1i}}}$$

```{r gene-level-output}
iso_to_gene <- iso_info %>%
  select(txid, gene_id, hgnc_symbol) %>%
  filter(!is.na(gene_id) & gene_id != "")

# Subset CPM matrix to isoforms with gene annotation
keep_iso <- intersect(rownames(cpm_mat), iso_to_gene$txid)
cpm_gene <- cpm_mat[keep_iso, samp$bam_id]

# Aggregate isoforms -> genes (rowsum groups by the second arg; alignment is critical)
gene_ids <- iso_to_gene$gene_id[match(keep_iso, iso_to_gene$txid)]
stopifnot(length(gene_ids) == nrow(cpm_gene))
stopifnot(!any(is.na(gene_ids)))

cpm_gene_agg <- rowsum(cpm_gene, group = gene_ids)
cat("Gene-level CPM matrix:", nrow(cpm_gene_agg), "genes x", ncol(cpm_gene_agg), "samples\n")

# One row per gene
gene_anno <- iso_to_gene %>%
  group_by(gene_id) %>%
  summarise(hgnc_symbol = first(hgnc_symbol),
            n_isoforms  = n(),
            .groups = "drop")

gene_results_list <- list()

for (ct_i in sort(unique(paired_donors$ct))) {
  donors_i <- paired_donors %>% filter(ct == ct_i) %>% pull(donor_id)

  for (d in donors_i) {
    smg1i_id <- samp %>%
      filter(ct == ct_i, donor_id == d, treatment == "Smg1i") %>% pull(bam_id)
    dmso_id  <- samp %>%
      filter(ct == ct_i, donor_id == d, treatment == "DMSO")  %>% pull(bam_id)

    stopifnot(length(smg1i_id) == 1, length(dmso_id) == 1)

    val_smg1i <- cpm_gene_agg[, smg1i_id]
    val_dmso  <- cpm_gene_agg[, dmso_id]
    delta     <- val_smg1i - val_dmso
    delta_pos <- pmax(delta, 0)
    frac_lost <- ifelse(val_smg1i > 0, delta_pos / val_smg1i, 0)

    gene_results_list[[paste(ct_i, d, sep = "_")]] <- data.frame(
      gene_id   = rownames(cpm_gene_agg),
      ct        = ct_i,
      donor_id  = d,
      smg1i_val = val_smg1i,
      dmso_val  = val_dmso,
      delta     = delta,
      delta_pos = delta_pos,
      frac_lost = frac_lost,
      stringsAsFactors = FALSE
    )
  }
}

gene_results_all <- bind_rows(gene_results_list) %>%
  left_join(gene_anno, by = "gene_id")

cat("Total rows (gene x ct x donor):", nrow(gene_results_all), "\n")
```

## Global Gene-Level Metric

```{r gene-level-global-metric}
gene_global_metric <- gene_results_all %>%
  group_by(ct) %>%
  summarise(sum_delta_pos   = sum(delta_pos),
            total_smg1i     = sum(smg1i_val),
            pct_output_lost = 100 * sum_delta_pos / total_smg1i,
            n_gene_donor    = n(),
            .groups = "drop") %>%
  arrange(desc(pct_output_lost))

print(gene_global_metric)

# Isoform-level vs gene-level side-by-side
compare_level <- global_metric %>%
  select(ct, pct_isoform = pct_output_lost) %>%
  left_join(gene_global_metric %>% select(ct, pct_gene = pct_output_lost),
            by = "ct") %>%
  mutate(diff_pct = pct_isoform - pct_gene,
         ratio    = pct_isoform / pct_gene)

cat("\nIsoform-level vs Gene-level:\n")
print(compare_level)

```

```{r gene-level-barplot, fig.width=10, fig.height=6}
p_gene_global <- ggplot(gene_global_metric,
                        aes(x = reorder(ct, pct_output_lost),
                            y = pct_output_lost, fill = ct)) +
  geom_col(alpha = 0.8) +
  geom_text(aes(label = paste0(round(pct_output_lost, 2), "%")),
            hjust = -0.1, size = 4) +
  coord_flip() +
  labs(title    = "% Transcriptional Output Lost To NMD By Cell Type",
       subtitle = "Gene-level",
       x = "Cell Type",
       y = "% Output Lost To NMD") +
  theme_minimal(base_size = 14) +
  theme(legend.position = "none",
        plot.title      = element_text(face = "bold"),
        plot.subtitle   = element_text(color = "grey40")) +
  expand_limits(y = max(gene_global_metric$pct_output_lost) * 1.15)

print(p_gene_global)

ggsave(file.path(OUTPUT_DIR, "pct_output_lost_gene_level_barplot.png"),
       p_gene_global, width = 10, height = 6, dpi = 300, bg = "white")
```

## Genes With Positive Mean Loss

```{r gene-level-avg-lost}
gene_avg_lost <- gene_results_all %>%
  group_by(ct, gene_id) %>%
  summarise(mean_frac_lost = mean(frac_lost, na.rm = TRUE),
            mean_delta_pos = mean(delta_pos, na.rm = TRUE),
            mean_smg1i_val = mean(smg1i_val, na.rm = TRUE),
            mean_dmso_val  = mean(dmso_val,  na.rm = TRUE),
            n_donors       = n(),
            hgnc_symbol    = first(hgnc_symbol),
            .groups = "drop") %>%
  filter(mean_frac_lost > 0)

gene_avg_lost %>%
  group_by(ct) %>%
  summarise(n_genes = n(), .groups = "drop") %>%
  print()

```

## Top 100 Genes By Output Lost

```{r top100-gene-table}
top100_genes <- gene_avg_lost %>%
  group_by(ct) %>%
  arrange(desc(mean_delta_pos)) %>%
  slice_head(n = 100) %>%
  mutate(pct_lost       = round(100 * mean_frac_lost, 2),
         mean_smg1i_val = round(mean_smg1i_val, 2),
         mean_dmso_val  = round(mean_dmso_val,  2),
         mean_delta_pos = round(mean_delta_pos, 2)) %>%
  select(ct, gene_id, hgnc_symbol, pct_lost, mean_delta_pos,
         mean_smg1i_val, mean_dmso_val, n_donors) %>%
  ungroup()

datatable(
  top100_genes,
  filter  = "top",
  options = list(pageLength = 20, scrollX = TRUE, scrollY = "500px"),
  caption = "Top 100 Genes By Mean CPM Lost To NMD Per Cell Type"
) %>%
  formatRound(columns = c("pct_lost", "mean_delta_pos",
                          "mean_smg1i_val", "mean_dmso_val"),
              digits = 2)
```

# Shared Vs Cell-Type-Specific Genes

How many genes lose output in only one cell type vs across multiple?

```{r shared-vs-unique-all-genes, fig.width=10, fig.height=6}
all_gene_cts_full <- gene_avg_lost %>%
  select(gene_id, ct) %>%
  distinct()

gene_n_cts_full <- all_gene_cts_full %>%
  group_by(gene_id) %>%
  summarise(n_cell_types = n(), .groups = "drop")

all_gene_cts_full <- all_gene_cts_full %>%
  left_join(gene_n_cts_full, by = "gene_id") %>%
  mutate(category = ifelse(n_cell_types == 1,
                           "Cell-Type-Specific",
                           "Shared (2+ cell types)"))

n_ct <- length(CT_KEEP)

sharing_summary_full <- all_gene_cts_full %>%
  group_by(ct, category) %>%
  summarise(n_genes = n(), .groups = "drop")

sharing_detail_full <- all_gene_cts_full %>%
  mutate(sharing_bin = case_when(
    n_cell_types == 1     ~ "1 (unique)",
    n_cell_types == n_ct  ~ paste0(n_ct, " (all)"),
    TRUE                  ~ as.character(n_cell_types)
  )) %>%
  group_by(ct, sharing_bin) %>%
  summarise(n_genes = n(), .groups = "drop")

cat("Total unique genes across all cell types:",
    n_distinct(all_gene_cts_full$gene_id), "\n\n")

gene_n_cts_full %>%
  group_by(n_cell_types) %>%
  summarise(n_genes = n(), .groups = "drop") %>%
  mutate(label = case_when(
    n_cell_types == 1    ~ "unique to 1 CT",
    n_cell_types == n_ct ~ paste0("shared across all ", n_ct, " CTs"),
    TRUE                 ~ paste0("shared across ", n_cell_types, " CTs")
  )) %>%
  print()

p_sharing_full <- ggplot(sharing_summary_full,
                         aes(x = ct, y = n_genes, fill = category)) +
  geom_col(alpha = 0.8) +
  geom_text(aes(label = n_genes), position = position_stack(vjust = 0.5), size = 4) +
  scale_fill_manual(values = c("Cell-Type-Specific"     = "#E69F00",
                               "Shared (2+ cell types)" = "#56B4E9")) +
  labs(title = "Genes With Output Lost: Shared Vs Cell-Type-Specific",
       x = "Cell Type", y = "Number Of Genes", fill = NULL) +
  theme_minimal(base_size = 14) +
  theme(plot.title      = element_text(face = "bold"),
        legend.position = "top")

print(p_sharing_full)

ggsave(file.path(OUTPUT_DIR, "all_genes_output_lost_shared_vs_unique.png"),
       p_sharing_full, width = 10, height = 6, dpi = 300, bg = "white")

p_detail_full <- ggplot(sharing_detail_full,
                        aes(x = ct, y = n_genes, fill = sharing_bin)) +
  geom_col(alpha = 0.8) +
  geom_text(aes(label = n_genes), position = position_stack(vjust = 0.5), size = 3.5) +
  scale_fill_brewer(palette = "YlGnBu") +
  labs(title = "Genes With Output Lost: Sharing Across Cell Types",
       x = "Cell Type", y = "Number Of Genes", fill = "# Cell Types") +
  theme_minimal(base_size = 14) +
  theme(plot.title      = element_text(face = "bold"),
        legend.position = "top")

print(p_detail_full)

ggsave(file.path(OUTPUT_DIR, "all_genes_output_lost_sharing_detail.png"),
       p_detail_full, width = 10, height = 6, dpi = 300, bg = "white")

```

# Save Results

```{r save-results}
# Isoform-level
fwrite(results_all,     file.path(OUTPUT_DIR, "transcriptional_output_all_donors_cpm.csv"))
fwrite(avg_lost,        file.path(OUTPUT_DIR, "transcriptional_output_avg_by_ct_cpm.csv"))
fwrite(global_metric,   file.path(OUTPUT_DIR, "pct_output_lost_by_ct_cpm.csv"))
fwrite(top10_per_ct,    file.path(OUTPUT_DIR, "top10_isoforms_lost_per_ct.csv"))
fwrite(bottom10_per_ct, file.path(OUTPUT_DIR, "bottom10_isoforms_lost_per_ct.csv"))
fwrite(boxplot_stats,   file.path(OUTPUT_DIR, "boxplot_summary_stats.csv"))
fwrite(total_loss,      file.path(OUTPUT_DIR, "pct_isoforms_total_loss.csv"))

# Gene-level
fwrite(gene_results_all,    file.path(OUTPUT_DIR, "transcriptional_output_gene_level_all_donors.csv"))
fwrite(gene_avg_lost,       file.path(OUTPUT_DIR, "transcriptional_output_gene_level_avg_by_ct.csv"))
fwrite(gene_global_metric,  file.path(OUTPUT_DIR, "pct_output_lost_gene_level_by_ct.csv"))
fwrite(compare_level,       file.path(OUTPUT_DIR, "isoform_vs_gene_level_comparison.csv"))
fwrite(top100_genes,        file.path(OUTPUT_DIR, "top100_genes_output_lost_per_ct.csv"))

# Top 500 per cell type — saved as a separate "all genes" reference table
all_genes_table <- gene_avg_lost %>%
  group_by(ct) %>%
  arrange(desc(mean_delta_pos)) %>%
  slice_head(n = 500) %>%
  ungroup() %>%
  mutate(pct_lost       = round(100 * mean_frac_lost, 2),
         mean_smg1i_val = round(mean_smg1i_val, 2),
         mean_dmso_val  = round(mean_dmso_val,  2),
         mean_delta_pos = round(mean_delta_pos, 2)) %>%
  select(ct, gene_id, hgnc_symbol, pct_lost, mean_delta_pos,
         mean_smg1i_val, mean_dmso_val, n_donors)

fwrite(all_genes_table, file.path(OUTPUT_DIR, "all_genes_output_lost_per_ct.csv"))

cat("All outputs saved to:", OUTPUT_DIR, "\n")
```

# Session Info

```{r session-info}
sessionInfo()
```
