---
title: "Supplemental figures — NMD long-read manuscript"
output:
  html_document:
    toc: true
    toc_depth: 2
    self_contained: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
knitr::opts_knit$set(root.dir = normalizePath("."))
```

This document regenerates the supplemental figures produced from the
`nmd_fig_data/` bundle. Each section runs its render script and displays the
result.

**Scope note.** The following supplements are rendered in the *Isopair* /
collaborator repository (`figures/multipanel/figureS1_*`, `figureS2_*`), **not**
here, and are therefore not built by this document: SR↔LR correlation-by-cell-type,
isoform-length distribution, pairwise expression similarity, SQANTI3 structural
categories, mashr pairwise sharing, and proportion-vs-expression. The supplements
below are the ones whose render code lives in this repository.

**Path note.** `make_norm_sanity.R`, `make_reproducibility_battery.R`, and
`make_sr_isopair.R` currently write their outputs to a hardcoded `HERE` directory
(a `figures/multipanel/figureS*` path). Before publishing, repoint each script's
`HERE` constant to a local output directory. The display chunks below search for
each output by filename so they render regardless of that path.

```{r helpers}
DATA_DIR <- "nmd_fig_data"
RSCRIPT  <- file.path(R.home("bin"), "Rscript")
PYTHON   <- Sys.which("python"); if (!nzchar(PYTHON)) PYTHON <- Sys.which("python3")

run_r <- function(script, args = character()) {
  stopifnot(file.exists(script))
  out <- system2(RSCRIPT, c(shQuote(script), args), stdout = TRUE, stderr = TRUE)
  cat(paste(utils::tail(out, 12), collapse = "\n"), "\n")
  st <- attr(out, "status"); if (!is.null(st) && st != 0)
    stop(sprintf("%s failed (exit %s)", script, st))
  invisible(out)
}
run_py <- function(script) {
  stopifnot(file.exists(script), nzchar(PYTHON))
  out <- system2(PYTHON, shQuote(script), stdout = TRUE, stderr = TRUE)
  cat(paste(utils::tail(out, 6), collapse = "\n"), "\n")
  invisible(out)
}
## Find a produced figure by basename, searching likely output locations.
show_fig <- function(basename_png) {
  hits <- list.files(".", pattern = basename_png, recursive = TRUE, full.names = TRUE)
  if (!length(hits)) { cat("Not found:", basename_png, "\n"); return(invisible()) }
  knitr::include_graphics(hits[[which.min(nchar(hits))]])
}
```

# S — KEGG ER-stress pathway across all four cell types

`make_pathway_allct.R` renders hsa04141 colored by short-read gene logFC per cell
type; `figure_pathway_allct.py` composes the 2×2. Inputs:
`nmd_mashr_dge_{at2,dd,fb,mv}_2026.3.10.csv`.

```{r s-pathway, results='hide'}
run_r("make_pathway_allct.R")
run_py(file.path("fig_panels", "figure_pathway_allct.py"))
```

```{r s-pathway-show, echo=FALSE, out.width='100%'}
show_fig("figure_pathway_allct\\.png$")
```

# S — Normalization sanity check

`make_norm_sanity.R`: control (non-NMD) genes are essentially unchanged by the
custom NMD-anchored normalization (per-gene DMSO vs adjusted-SMG1i). Inputs:
`dge_isoform_longread_filtered_2026.3.3.rds`,
`mashr_isoform_{posterior_means,lfsr}_2026.3.10.csv`.

```{r s-normsanity, results='hide'}
run_r("make_norm_sanity.R")
```

```{r s-normsanity-show, echo=FALSE, out.width='100%'}
show_fig("figureS3_norm_sanity\\.png$")
```

# S — Reproducibility battery

`make_reproducibility_battery.R`: the four retained robustness tests for the
productive-direction calls. Same data inputs as above.

```{r s-repro, results='hide'}
run_r("make_reproducibility_battery.R")
```

```{r s-repro-show, echo=FALSE, out.width='100%'}
show_fig("figureS3_reproducibility\\.png$")
```

# S — SRSF isoform structures (short-read isopair)

`make_sr_isopair.R`: per-gene SRSF isoform structure + donor-paired logFC.
Inputs add `nmd_fig_data/srsf.gtf`, `srsf.cds.gff3`, `srsf.fasta` (Isopair).
Produces `SR_SRSF*.png` (one per gene).

```{r s-srsf, results='hide'}
run_r("make_sr_isopair.R")
```

```{r s-srsf-show, echo=FALSE, out.width='100%'}
srsf <- list.files(".", pattern = "^SR_SRSF.*\\.png$", recursive = TRUE, full.names = TRUE)
if (length(srsf)) knitr::include_graphics(srsf) else cat("No SR_SRSF*.png found\n")
```

# S — Cell-type-specific NMD analysis

`ct_specific_nmd_analysis.R`: expression-vs-regulation basis of cell-type-specific
NMD, isoform and short-read gene level. Writes a 4-panel composite and an HTML
report into `ct_specific_nmd_report/`. Inputs: `dge_isoform_longread_2026.3.3.rds`,
`dge_gene_unfiltered_2026.1.2.rds`, `nmd_mashr_die_*`, `nmd_mashr_dge_*`.

```{r s-ctspecific, results='hide'}
run_r("ct_specific_nmd_analysis.R")
```

```{r s-ctspecific-show, echo=FALSE, out.width='100%'}
show_fig("figS_ct_specific_nmd_composite\\.png$")
```

# S — NMD machinery expression and functional activity

`_nmd_machinery_report.R`: core-factor expression score, canonical-isoform CPM for
the four well-captured factors, and the endogenous-substrate functional panel.
Writes PNGs + an HTML report into `nmd_machinery_report/`. Inputs:
`dge_gene_unfiltered_2026.1.2.rds`, `dge_isoform_longread_2026.3.3.rds`,
`nmd_mashr_dge_*`, `nmd_mashr_die_*`.

```{r s-machinery, results='hide'}
run_r("_nmd_machinery_report.R")
```

```{r s-machinery-show, echo=FALSE, out.width='100%'}
show_fig("substrate_panel_heatmap\\.png$")
show_fig("machinery_heatmap\\.png$")
show_fig("canonical_isoform_cpm\\.png$")
```

# S — RBP enrichment

The RBP-category enrichment supplement is produced by `nmd_rbp_enrichment.Rmd`
(rosters: `encode_rbp_roster_vannostrand2020.csv`,
`gerstberger_2014_rbp_census.csv`). Knit that document separately, or add it as a
child here once its output paths are parameterized.

# Session info

```{r session, echo=FALSE}
sessionInfo()
```
